### 2022-01-14 ms/manuscript.Rmd
Add manuscript Rmd

---
title: "An open and continuously updated fern tree of life (FTOL)"
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
---

---
bibliography: 
  - `r here::here("ms/references.yaml")`
  - `r here::here("ms/references_other.yaml")`
  - `r here::here("ms/references_other.bib")`
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Set left justification -->
\raggedright

<!-- Don't indent -->
\setlength{\parindent}{0ex}

**Joel H. Nitta^1^, Eric Schuettpelz^2^, and Wataru Iwasaki^3^**

^1^Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan

^2^Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., USA

^3^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

**\*Correspondence:**    
Joel H. Nitta  
joelnitta@gmail.com

Keywords: Fern~1~, Phylogeney~2~, Plastome~3~, PPGI~4~, Pteridophyte~5~, *rbcL*~6~.

# Abstract

Ferns (ca. 12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for ca. 1/3 of all fern species deposited in GenBank.
Global trees including all ferns have been published periodically, but as molecular systematics research continues at a rapid pace, such trees quickly become outdated.
While some large projects exist to generate trees for all plants or all life, such "one-size-fits-all" approaches are not expected to generate optimal results for specific groups because they must work across a very wide range of taxa.
Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from sequence data on GenBank.
Our tailored sampling strategy combines whole plastomes (few taxa, many genes) with Sanger-sequenced genes (many taxa, few genes) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
We provide multiple versions of the phylogeny including a time-calibrated tree and a synthetic tree with species lacking molecular data added by taxonomy.
We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
The current FTOL (v0.0.1) includes 4,492 species, an increase of ca. 20% relative to the most recent global fern phylogeny while accounting for synonyms.
FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have easy access to the most recent, comprehensively sampled fern phylogeny.
FTOL will be useful for anybody studying this important group of plants at scales from smaller (e.g., genus or lower) clades to the entire tree; we anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale.
Finally, we envision FTOL will be integrated into the next PPG, PPGII, to produce a "living" taxonomic system with direct linkage between a community-driven fern taxonomy and the most recent hypothesis of phylogeny.

# Introduction

# Materials and Methods

## Locus sampling

The plastid genome has been the most widely sequenced genomic compartment in ferns by far, so we used plastid loci to build our tree.
Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome genes").
*trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*.
*rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication.

## Sequence extraction

All sequences were downloaded from GenBank.
GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< 7000 bp) or plastome sequences (> 7000 bp) with the "[SLEN]" term.
Furthermore, there is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.

To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of known sequences [i.e., a "baited search"; @Smith2009b; @Smith2019]
Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
We then filtered the reference sequences to a single representative longest sequence per genus (Sanger loci only).
Next, we aligned the sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" on the sequences downloaded from GenBank, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences to use as references that lacked annotations in the first round.
The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences.

## Taxonomic name resolution

One goal of this project is generate a phylogeny that is consistent with the community-driven Pteridophyte Phylogeny Group I taxonomic system [@PteridophytePhylogenyGroupI2016; hereafter "PPGI"].
Species names in GenBank do not necessarily conform to PPGI.
Therefore, we standardized all species names in the GenBank sequences against a reference taxonomy.
We designed the reference taxonomy for this project based on the World Ferns database [@Hassler2022], which conforms to PPGI and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
We used the taxastand R package [@Nitta2021] to resolve species names, which can account for differences in formatting of taxonomic authors (e.g., parenthetical basionym author present or absent) as well as perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
We manually inspected any fuzzily matched or non-matching names.
This revealed some species names in GenBank that were completely lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (e.g., changes in synonymy).
Therefore, we updated and edited the World Ferns database using the dwctaxon R package, which is designed to work with taxonomic data in the Darwin Core format.
We refer to the resulting taxonomic database as "pteridocat", and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.

## Automated removal of contaminations

Another potential issue with GenBank sequences is the presence of misidentified sequences and contaminations (hereafter "contaminations").
We removed putative contaminations from Sanger sequences as follows.
First, we constructed a BLAST library including all extracted Sanger sequences.
Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
We compared the families (following PPGI) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contamination and excluded from further analysis.
For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
While this method cannot account for contaminations at finer taxonomic levels, it is an efficient method to remove clearly incorrect sequences.

### 2022-01-14 ms/manuscript.Rmd
Update MS
Load targets from _targets cache submodule

@@
     reference_docx: Frontiers_Template.docx
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+library(targets)
+library(assertr)
+library(ape)
+library(scales)
+library(tidyverse)
+```
+
+```{r load-targets}
+tar_load(
+	c(
+		sanger_alignment,
+		plastome_alignment,
+		plastome_tree,
+		plastome_outgroups,
+		ppgi_taxonomy),
+	store = here::here("ftol_cache")
+)
+```
+
+```{r abstract-stats}
+# Make tibble with one row per species in Sanger sampling
+sanger_sampling <- tibble(
+	species = rownames(sanger_alignment)
+) %>%
+	# Add higher-level taxonomy
+	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
+	left_join(
+		select(ppgi_taxonomy, genus, family, order), by = "genus"
+	)
+
+# Count number of fern species in Sanger data
+n_sanger_ferns <-
+	sanger_sampling %>% 
+	# outgroup taxa have no family data
+	filter(!is.na(family)) %>%
+	pull(species) %>%
+	n_distinct()
 ```
@@
 We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
-The current FTOL (v0.0.1) includes 4,492 species, an increase of ca. 20% relative to the most recent global fern phylogeny while accounting for synonyms.
+The current FTOL (v0.0.1) includes `r comma(n_sanger_ferns)` species, an increase of ca. 20% relative to the most recent global fern phylogeny while accounting for synonyms.
 FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have easy access to the most recent, comprehensively sampled fern phylogeny.
@@
 To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
-Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of known sequences [i.e., a "baited search"; @Smith2009b; @Smith2019]
+Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of known sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019]
 Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
@@
 
-One goal of this project is generate a phylogeny that is consistent with the community-driven Pteridophyte Phylogeny Group I taxonomic system [@PteridophytePhylogenyGroupI2016; hereafter "PPGI"].
+One goal of this project is generate a phylogeny that is consistent with the community-driven Pteridophyte Phylogeny Group I taxonomic system [@PteridophytePhylogenyGroupI2016\; hereafter "PPGI"].
 Species names in GenBank do not necessarily conform to PPGI.

### 2022-01-17 R/functions.R
Fix extract_citations to handle cases with escaped ;
e.g., [@ref\;]

@@ -9,7 +9,7 @@ extract_citations <- function(rmd_file) {
 		stringr::str_split(" |;") %>% 
 		unlist %>% 
 		magrittr::extract(., stringr::str_detect(., "@")) %>% 
-		stringr::str_remove_all("\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}") %>% 
+		stringr::str_remove_all("\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}|\\\\") %>% 
 		magrittr::extract(., stringr::str_detect(., "^@|^-@")) %>% 
 		stringr::str_remove_all("^@|^-@") %>% 
 		unique %>% 

### 2022-01-17 ms/manuscript.Rmd
Update MS Rmd

@@
 		plastome_tree,
-		plastome_outgroups,
+		plastome_metadata_renamed,
 		ppgi_taxonomy),
@@
 ```{r abstract-stats}
+# Make tibble of outgroup species
+og_species <- 
+	plastome_metadata_renamed %>%
+	select(species, outgroup) %>%
+	filter(outgroup == TRUE)
+
 # Make tibble with one row per species in Sanger sampling
@@
 		select(ppgi_taxonomy, genus, family, order), by = "genus"
-	)
+	) %>% 
+	# Add outgroup status
+	left_join(og_species, by = "species") %>%
+	mutate(outgroup = replace_na(outgroup, FALSE)) %>%
+	verify(nrow(.) == nrow(sanger_alignment)) %>%
+	verify(sum(outgroup) == nrow(og_species))
 
@@
 	sanger_sampling %>% 
-	# outgroup taxa have no family data
-	filter(!is.na(family)) %>%
-	pull(species) %>%
-	n_distinct()
+	filter(outgroup == FALSE) %>%
+	n_distinct(.$species)
 ```
@@
 *rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
-The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication.
+The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes).
 
@@
 To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
-Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of known sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019]
+Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
 Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
+
 We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
-We then filtered the reference sequences to a single representative longest sequence per genus (Sanger loci only).
-Next, we aligned the sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
-To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" on the sequences downloaded from GenBank, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences to use as references that lacked annotations in the first round.
-The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences.
+We then filtered the sequences to a single representative longest sequence per genus.
+Next, we aligned the filtered sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
+To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
+The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from sequences downloaded from GenBank.
 
@@
 While this method cannot account for contaminations at finer taxonomic levels, it is an efficient method to remove clearly incorrect sequences.
+
+## Sequence selection and concatenation
+
+A typical step in multilocus workflows is to concatenate loci across samples.
+For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
+This is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
+Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions sometimes have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
+First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
+We then concatenated loci for monophyletic species by selecting the longest accession per locus within each species.
+Next, for the remaining species, we concatenated loci across accessions if all accessions originated from the same voucher specimen; this could not be done in all cases as voucher information is not always annotated on GenBank.
+Next, for the remaining species, we concatenated loci across accessions if all accessions for a given species originated from only one publication, on the assumption that the authors of the study intended to do so.
+For the remaining species that could not be concatenated using any of the above methods, we selected one longest accession per locus and discarded the others.
+
+We then selected the final set of concatenated loci for each species. 
+We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and we expect that maximal sampling of one locus should improve phylogenetic analysis.
+Our selection is based on the following criteria, in order of priority:
+
+1. The sample (set of accessions joined by the methods described above) with the longest concatenated sequence length; sample includes *rbcL* and additional loci.
+2. The sample with the longest *rbcL* sequence; sample includes only *rbcL* and no additional loci.
+3. The sample with the longest concatenated sequence length; sample does not include *rbcL*.
+
+The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
+For these, we selected one specimen per species with the longest combined sequence length across all plastome genes.
+
+## Sequence alignment
+
+For non-spacer regions, we aligned each locus separately with MAFFT.
+
+Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (e.g., within genera) where slower-evolving genes like *rbcL* may not provide enough resolution.
+Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "--merge" option.
+This retains the indels within each subalignment while aligning across subalignments.
+For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".
+
+Next, we removed poorly aligned regions from the alignments using trimAl.
+We removed regions with >1% or >5% of sequences having gaps from coding regions or spacers, respectively.
+
+## Phylogenetic analysis

### 2022-01-25 ms/manuscript.Rmd
Update MS

@@
 
-**Joel H. Nitta^1^, Eric Schuettpelz^2^, and Wataru Iwasaki^3^**
+**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^4^**
 
@@
 
-^3^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+^3^ Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico
+
+^4^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
 
@@
 # Introduction
+
+Ferns (ca.
+12,000 spp.) are the second most diverse clade of vascular plants after seed plants (ca.
+260,000 spp.), and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
+As Dobzhansky famously said, "Nothing in biology makes sense except in the light of evolution"; likewise, a well-sampled molecular phylogeny is key to any investigation of fern evolutionary biology.
+Fortunately, ferns have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for ca.
+1/3 of all fern species deposited in GenBank.
+Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.
+
+Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
+Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a].
+We believe there is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
+This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
+
+We recognize that multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
+While these may be appropriate in some cases, we contend that they cannot produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth.
+By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.
+
+Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be possible with a "tree of all life" or "tree of all plants".
+Recently, a genus-level taxonomy of ferns and lycophytes was established following a open, community-driven model [@PteridophytePhylogenyGroupI2016].
+Thanks to this community-first approach, the taxonomy (PPG I) has been widely accepted and used (cited 475 times on Google Scholar as of writing).
+However, there were clearly problematic (non-monophyletic) genera at the time PPG I was published, and many new taxonomic changes have been (and will continue to be) made since [e.g., @Almeida2017; @Shang2018; @Zhang2020].
+It is anticipated that "PPG II" will be a online, open resource that can be updated as necessary (Schuettpelz, pers. comm.).
+We envision an open, continuously updated global fern phylogeny that is directly integrated with PPG II so that taxonomic decisions reflecting phylogeny can be made based on community consensus using the most recently available data.
+
+Here, we leverage our specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
+We plan to run the pipeline on a semi-annual basis and make the results freely available online through a web portal and R package.
+This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
+We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) It will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) It will allow for continuous assessment of taxonomy, and indicate what parts of the tree are in need of taxonomic revision; 3) It will be an important source of data for any phylogenetic comparative study of ferns.
 
@@
 For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".
+Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data.
+We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
 

### 2022-01-25 ms/manuscript.Rmd
Update MS

@@
 
-^3^ Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico
+^3^Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico
 
@@
 Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from sequence data on GenBank.
-Our tailored sampling strategy combines whole plastomes (few taxa, many genes) with Sanger-sequenced genes (many taxa, few genes) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
+Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with Sanger-sequenced loci (many taxa, few loci) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
 We provide multiple versions of the phylogeny including a time-calibrated tree and a synthetic tree with species lacking molecular data added by taxonomy.
 We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
-The current FTOL (v0.0.1) includes `r comma(n_sanger_ferns)` species, an increase of ca. 20% relative to the most recent global fern phylogeny while accounting for synonyms.
+The current FTOL (v0.0.1) includes `r comma(n_sanger_ferns)` species, an increase of > 20% relative to the most recent global fern phylogeny.
 FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have easy access to the most recent, comprehensively sampled fern phylogeny.
@@
 The plastid genome has been the most widely sequenced genomic compartment in ferns by far, so we used plastid loci to build our tree.
-Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome genes").
+Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome loci").
 *trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*.
 *rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
-The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes).
+The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes), then combined with the two spacers (79 loci total).
+The plastome loci include all seven Sanger loci.
 
@@
 The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
-For these, we selected one specimen per species with the longest combined sequence length across all plastome genes.
+For these, we selected one specimen per species with the longest combined sequence length across all plastome loci.
 
@@
 ## Phylogenetic analysis
+
+Phylogenetic analysis was conducted in two steps.
+First, we generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome loci in IQ TREE [@Nguyen2015] with 1000 ultrafast rapid bootstrap replicates [@Minh2013].
+We applied the 'GTR+I+G' to a concatenated dataset with no partitioning.
+
+Next, we used the backbone phylogeny as a constraint tree to conduct phylogenetic analysis of the Sanger loci in IQ TREE.
+Settings for the ML analysis of the Sanger loci were otherwise identical to the backbone phylogenetic analysis.
+
+## Reproducibility
+
+The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
+Code used to generate FTOL is available at https://github.com/joelnitta/ftol.
+A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.
+
+# References
+<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
+<div id="refs"></div>

### 2022-01-30 ms/manuscript.Rmd
Update MS

@@
 library(tidyverse)
+library(rgnparser)
+library(glue)
+library(phytools)
+
+# Specify percentage format
+percent <- function(...) {scales::percent(...)}
+number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Specify FTOL version
+ftol_ver <- "v0.0.2"
 ```
@@
 		plastome_metadata_renamed,
-		ppgi_taxonomy),
+		ppgi_taxonomy,
+		plastid_calibration_dates,
+		pteridocat),
 	store = here::here("ftol_cache")
@@
 GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
 There is no formal definition of genomic vs.
@@
 
+## Molecular dating
+
+We dated the tree using treePL [@Smith2012], which is one of the only programs capable of dating very large alignments.
+We applied a set of `r nrow (plastid_calibration_dates)` calibration points, mostly based on @Testo2016a and references within.
+
 ## Reproducibility
@@
 
+# Results
+
+## Sampling
+
+```{r sampling-stats, message=FALSE}
+# Make list of FTOL sampling
+ftol_sampling_by_rank <-
+	sanger_sampling %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup) %>%
+	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
+	group_by(rank) %>%
+	count(name) %>%
+	ungroup
+
+# Make list of all accepted species
+accepted_species <-
+	pteridocat %>% 
+	# Only keep accepted names
+	filter(str_detect(taxonomicStatus, "accepted")) %>%
+	# Only keep species
+	# FIXME: need to add rank for newly added names in pteridocat_maker
+	# then don't need is.na()
+	filter(taxonRank == "species" | is.na(taxonRank)) %>%
+	select(scientificName) %>%
+	mutate(
+		rgnparser::gn_parse_tidy(scientificName) %>% select(species = canonicalsimple),
+		species = str_replace_all(species, " ", "_")
+	) %>%
+	# Map on higher level taxonomy
+	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
+	left_join(select(ppgi_taxonomy, genus, family, order, class), by = "genus") %>%
+	# FIXME: update class for nothogenera in PPG
+	mutate(class = case_when(
+		str_detect(scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
+		TRUE ~ class
+	)) %>%
+	# Drop lycophytes
+	assert(not_na, class) %>%
+	filter(class == "Polypodiopsida") %>%
+	select(-class)
+
+accepted_sampling_by_rank <-
+	accepted_species %>%
+	select(-scientificName) %>%
+	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
+	group_by(rank) %>%
+	count(name) %>%
+	ungroup %>%
+	# Note that nothogenera have no family or order
+	filter(!is.na(name))
+
+coverage <- left_join(
+	ftol_sampling_by_rank,
+	rename(accepted_sampling_by_rank, n_accepted = n),
+	by = c("rank", "name")
+) %>%
+	mutate(coverage = n/n_accepted)
+
+# Format overall sampling for printing in MS
+total_sampling <-
+	# Start with FTOL sampling by genus, family, order
+	ftol_sampling_by_rank %>% 
+	count(rank) %>%
+	# Add n species
+	bind_rows(
+		sanger_sampling %>%
+			filter(outgroup == FALSE) %>%
+			select(-outgroup) %>%
+			summarize(n = n(), rank = "species")
+	) %>%
+	# Join to accepted number of species, genus, family, order
+	left_join(
+		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
+			bind_rows(
+				summarize(accepted_species, n_accepted = n(), rank = "species")
+			),
+		by = "rank"
+	) %>%
+	# Calculate coverage
+	mutate(
+		coverage_p = percent(n / n_accepted),
+		coverage = glue("{number(n)}/{number(n_accepted)}")) %>%
+	# Split for printing
+	split(.$rank)
+```
+
+Taxon sampling for the current version of FTOL (`r ftol_ver`) includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
+
+# Figures
+
+```{r backbone-tree}
+og_taxa <-
+	sanger_sampling %>%
+	filter(outgroup == TRUE) %>%
+	pull(species)
+
+backbone_tree_full <- 
+	plastome_tree %>%
+	ape::root.phylo("Anthoceros_angustus") %>%
+	ape::ladderize() %>%
+	ape::drop.tip(og_taxa)
+
+family_exemp <-
+tibble(
+	species = backbone_tree_full$tip.label
+) %>%
+	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
+	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
+	group_by(family) %>%
+	slice(1) %>%
+	ungroup() %>%
+	pull(species)
+
+backbone_tree_exemp <- ape::keep.tip(backbone_tree_full, family_exemp)
+
+bb_trans_table <-
+	tibble(species = backbone_tree_exemp$tip.label) %>%
+	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
+	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
+	left_join(accepted_sampling_by_rank, by = c(family = "name")) %>%
+	assert(not_na, species, family, n) %>%
+	select(tip.label = species, clade.label = family, N = n) %>%
+	## set crown node depth to 1/2 the maximum depth
+	mutate(
+		depth = sapply(backbone_tree_exemp$tip.label, function(x,y) 
+	0.5*y$edge.length[which(backbone_tree_exemp$edge[,2] ==
+	which(y$tip.label == x))], y = backbone_tree_exemp)
+	)
+
+## convert
+tt <- phylo.toBackbone(backbone_tree_exemp, bb_trans_table)
+## plot
+plot(tt)
+```
+
 # References

### 2022-01-30 ms/manuscript.Rmd
Update MS

@@
 knitr::opts_chunk$set(echo = FALSE)
+library(rmarkdown)
+library(bookdown)
 library(targets)
@@
 library(glue)
-library(phytools)
 
@@
 
-backbone_tree_full <- 
-	plastome_tree %>%
-	ape::root.phylo("Anthoceros_angustus") %>%
-	ape::ladderize() %>%
-	ape::drop.tip(og_taxa)
-
 family_exemp <-
-tibble(
-	species = backbone_tree_full$tip.label
-) %>%
-	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
+tibble(species = plastome_tree$tip.label) %>%
+	mutate(
+		genus = str_split(species, "_") %>% map_chr(1),
+		phy_order = seq_len(nrow(.))) %>%
 	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
+	mutate(family = replace_na(family, "og")) %>%
 	group_by(family) %>%
@@
 	ungroup() %>%
-	pull(species)
+	arrange(phy_order)
 
-backbone_tree_exemp <- ape::keep.tip(backbone_tree_full, family_exemp)
+backbone_tree <- ape::keep.tip(plastome_tree, family_exemp$species)
+backbone_tree$tip.label <- family_exemp$family
 
-bb_trans_table <-
-	tibble(species = backbone_tree_exemp$tip.label) %>%
-	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
-	left_join(select(ppgi_taxonomy, family, genus), by = "genus") %>%
-	left_join(accepted_sampling_by_rank, by = c(family = "name")) %>%
-	assert(not_na, species, family, n) %>%
-	select(tip.label = species, clade.label = family, N = n) %>%
-	## set crown node depth to 1/2 the maximum depth
-	mutate(
-		depth = sapply(backbone_tree_exemp$tip.label, function(x,y) 
-	0.5*y$edge.length[which(backbone_tree_exemp$edge[,2] ==
-	which(y$tip.label == x))], y = backbone_tree_exemp)
-	)
+backbone_tree$node.label <- str_replace(backbone_tree$node.label, "100", "")
 
-## convert
-tt <- phylo.toBackbone(backbone_tree_exemp, bb_trans_table)
-## plot
-plot(tt)
+backbone_tree <-
+	backbone_tree %>%
+	root.phylo("og") %>%
+	ladderize() %>%
+	drop.tip("og")
+
+plot(backbone_tree, font = 1)
+nodelabels(backbone_tree$node.label, frame = "none", col = "grey", cex = 0.8)
 ```

### 2022-01-31 ms/manuscript.Rmd
Update MS
Add phylogeny figures

--- 2022-01-25 ms/manuscript.Rmd
@@
   chunk_output_type: console
+params:
+  doc_type: doc
 ---
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+# Load packages
+library(conflicted)
+library(rmarkdown)
+library(bookdown)
 library(targets)
@@
 library(tidyverse)
+library(rgnparser)
+library(ggtree)
+library(glue)
+library(Polychrome)
+library(patchwork)
+library(magrittr)
+
+# Resolve conflicts
+conflict_prefer("filter", "dplyr")
+conflict_prefer("extract", "magrittr")
+
+# Load functions
+source(here::here("R/functions.R"))
+
+# Specify percentage format
+percent <- function(...) {scales::percent(...)}
+number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Specify FTOL version
+ftol_ver <- "v0.0.2"
 ```
@@
 		plastome_metadata_renamed,
-		ppgi_taxonomy),
+		ppgi_taxonomy,
+		plastome_alignment,
+		plastid_calibration_dates,
+		pteridocat,
+		sanger_alignment_tbl),
 	store = here::here("ftol_cache")
 )
+
+# FIXME: read in real ultrametric tree when ready
+sanger_tree <- ape::read.tree(here::here("working/sanger_alignment.phy.contree.tre"))
 ```
@@
 ```{r abstract-stats}
+# Helper function to add plot_group:
+# a "rank" that may be at order or suborder level for plotting
+add_plot_group <- function(data) {
+	data %>%
+		mutate(
+			plot_group = coalesce(suborder, order),
+			plot_group = case_when(
+				plot_group %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales",
+				TRUE ~ plot_group
+			)
+		)
+}
+
 # Make tibble of outgroup species
@@
 	left_join(
-		select(ppgi_taxonomy, genus, family, order), by = "genus"
+		select(ppgi_taxonomy, genus, family, suborder, order), by = "genus"
 	) %>% 
+	# Add plot_group
+	add_plot_group() %>%
 	# Add outgroup status
@@
 GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
 There is no formal definition of genomic vs.
 Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< 7000 bp) or plastome sequences (> 7000 bp) with the "[SLEN]" term.
-Furthermore, there is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.
 
+There is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.
 To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
@@
 We manually inspected any fuzzily matched or non-matching names.
-This revealed some species names in GenBank that were completely lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (e.g., changes in synonymy).
+This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (e.g., changes in synonymy).
 Therefore, we updated and edited the World Ferns database using the dwctaxon R package, which is designed to work with taxonomic data in the Darwin Core format.
@@
 For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
-While this method cannot account for contaminations at finer taxonomic levels, it is an efficient method to remove clearly incorrect sequences.
+Species belonging to monotypic families were not considered for this filtering.
+While this method cannot account for contaminations at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
 
@@
 For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
-This is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
-Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions sometimes have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
+However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
+Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
 First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
@@
 We then selected the final set of concatenated loci for each species. 
-We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and we expect that maximal sampling of one locus should improve phylogenetic analysis.
+We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus should improve phylogenetic analysis [@Talavera2021].
 Our selection is based on the following criteria, in order of priority:
@@
 Phylogenetic analysis was conducted in two steps.
-First, we generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome loci in IQ TREE [@Nguyen2015] with 1000 ultrafast rapid bootstrap replicates [@Minh2013].
-We applied the 'GTR+I+G' to a concatenated dataset with no partitioning.
+First, we generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome loci in IQ TREE [@Nguyen2015] with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
+We applied the 'GTR+I+G' model to a concatenated dataset with no partitioning.
 
@@
 Settings for the ML analysis of the Sanger loci were otherwise identical to the backbone phylogenetic analysis.
+
+## Molecular dating
+
+We dated the tree using treePL [@Smith2012], which is one of the only programs capable of dating very large alignments.
+We applied a set of `r nrow (plastid_calibration_dates)` calibration points, mostly based on @Testo2016a and references within.
 
@@
 
+# Results
+
+## Taxon sampling
+
+```{r sampling-stats, message=FALSE}
+# FTOL sampling ----
+
+# Helper function to add plot_group:
+# a "rank" that may be at order or suborder level for plotting
+add_plot_group <- function(data) {
+	data %>%
+		mutate(
+			plot_group = coalesce(suborder, order),
+			plot_group = case_when(
+				plot_group %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales",
+				TRUE ~ plot_group
+			)
+		)
+}
+
+# Tally FTOL sampling by rank
+ftol_sampling_by_rank <-
+	sanger_sampling %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup) %>%
+	# add plot_group
+	add_plot_group() %>%
+	# drop suborder
+	select(-suborder) %>%
+	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
+	group_by(rank) %>%
+	count(name) %>%
+	ungroup
+
+# Make list of all accepted species
+accepted_species <-
+	pteridocat %>% 
+	# Only keep accepted names
+	filter(str_detect(taxonomicStatus, "accepted")) %>%
+	# Only keep species
+	# FIXME: need to add rank for newly added names in pteridocat_maker
+	# then don't need is.na()
+	filter(taxonRank == "species" | is.na(taxonRank)) %>%
+	select(scientificName) %>%
+	mutate(
+		rgnparser::gn_parse_tidy(scientificName) %>% select(species = canonicalsimple),
+		species = str_replace_all(species, " ", "_")
+	) %>%
+	# Map on higher level taxonomy
+	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
+	left_join(select(ppgi_taxonomy, genus, family, suborder, order, class), by = "genus") %>%
+	# FIXME: update class for nothogenera in PPG
+	mutate(class = case_when(
+		str_detect(scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
+		TRUE ~ class
+	)) %>%
+	# Drop lycophytes
+	assert(not_na, class) %>%
+	filter(class == "Polypodiopsida") %>%
+	select(-class) %>%
+	# Add plot_group
+	add_plot_group() %>%
+	# drop suborder
+	select(-suborder)
+
+# Tally accepted sampling by rank
+accepted_sampling_by_rank <-
+	accepted_species %>%
+	select(-scientificName) %>%
+	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
+	group_by(rank) %>%
+	count(name) %>%
+	ungroup %>%
+	# Note that nothogenera have no family or order
+	filter(!is.na(name))
+
+# Format coverage at order-ish level for plotting
+coverage_data <- left_join(
+	ftol_sampling_by_rank,
+	rename(accepted_sampling_by_rank, n_accepted = n),
+	by = c("rank", "name")
+) %>%
+	mutate(coverage = n/n_accepted)
+
+# Format for text in MS
+tax_coverage <-
+coverage_data %>%
+	filter(rank == "plot_group") %>%
+	arrange(coverage) %>%
+	transmute(clade = name, coverage = percent(coverage)) %>%
+	split(.$clade)
+
+# Format overall sampling for printing in MS
+total_sampling <-
+	# Start with FTOL sampling by genus, family, order
+	ftol_sampling_by_rank %>% 
+	count(rank) %>%
+	# Add n species
+	bind_rows(
+		sanger_sampling %>%
+			filter(outgroup == FALSE) %>%
+			select(-outgroup) %>%
+			summarize(n = n(), rank = "species")
+	) %>%
+	# Join to accepted number of species, genus, family, order
+	left_join(
+		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
+			bind_rows(
+				summarize(accepted_species, n_accepted = n(), rank = "species")
+			),
+		by = "rank"
+	) %>%
+	# Calculate coverage
+	mutate(
+		coverage_p = percent(n / n_accepted),
+		coverage = glue("{number(n)}/{number(n_accepted)}")) %>%
+	# Split for printing
+	split(.$rank)
+
+# Plastome (backbone) sampling ----
+bb_sampling <-
+	tibble(species = plastome_tree$tip.label) %>%
+	# Add higher taxonomy
+	left_join(sanger_sampling, by = "species")
+
+bb_sampling_by_rank <-
+	bb_sampling %>%
+	# Drop OG
+	filter(outgroup == FALSE) %>%
+	select(-outgroup) %>%
+	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
+	group_by(rank) %>%
+	count(name) %>%
+	ungroup
+
+# Format sampling for printing in MS
+bb_total_sampling <-
+	# Start with backbone sampling by genus, family, order
+	bb_sampling_by_rank %>% 
+	count(rank) %>%
+	# Add n species
+	bind_rows(
+		bb_sampling %>%
+			filter(outgroup == FALSE) %>%
+			select(-outgroup) %>%
+			summarize(n = n(), rank = "species")
+	) %>%
+	# Join to accepted number of species, genus, family, order
+	left_join(
+		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
+			bind_rows(
+				summarize(accepted_species, n_accepted = n(), rank = "species")
+			),
+		by = "rank"
+	) %>%
+	# Calculate coverage
+	mutate(
+		coverage_p = percent(n / n_accepted),
+		coverage = glue("{number(n)}/{number(n_accepted)}")) %>%
+	# Split for printing
+	split(.$rank)
+```
+
+Taxon sampling for the current version of FTOL (`r ftol_ver`) includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
+Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
+
+Taxon sampling for the backbone phylogeny (plastome genes) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.
+
+## DNA alignments
+
+```{r dna-stats}
+# Define missing base codes in ape::base.freq
+missing_base_codes <- c("n", "-", "?")
+
+# Calculate frequence of missing bases in each dataset
+plastome_base_freq <- base.freq(plastome_alignment, all = TRUE)
+plastome_missing_freq <- sum(plastome_base_freq[missing_base_codes])
+sanger_base_freq <- base.freq(sanger_alignment, all = TRUE)
+sanger_missing_freq <- sum(sanger_base_freq[missing_base_codes])
+
+sanger_missing_by_gene <-
+	sanger_alignment_tbl %>%
+	rename(gene = target) %>%
+	mutate(
+		# Format gene names for printing
+		gene = str_replace_all(gene, "-", "--"),
+		gene = str_replace_all(gene, "([^-]+)", "*\\1*"),
+		# Calculate percent missing data per gene
+		missing = map_dbl(align_trimmed, ~sum(base.freq(., all = TRUE)[missing_base_codes])),
+		missing = percent(missing)) %>%
+	select(-align_trimmed) %>%
+	arrange(missing) %>%
+	split(.$gene)
+```
+
+The Sanger DNA alignment was `r number(ncol(sanger_alignment))` bp with `r percent(sanger_missing_freq, accuracy = 0.1)` missing data (missing bases or gaps) overall; rates of missing data by locus ranged from `r sanger_missing_by_gene %>% first() %>% extract2("missing")` (`r sanger_missing_by_gene %>% first() %>% extract2("gene")`) to `r sanger_missing_by_gene %>% last() %>% extract2("missing")` (`r sanger_missing_by_gene %>% last() %>% extract2("gene")`).
+The plastome DNA alignment was `r number(ncol(plastome_alignment))` bp with `r percent(plastome_missing_freq, accuracy = 0.1)` missing data.
+
+## Topology
+
+Ferns are strongly supported as monophyletic ("strongly supported" indicates BS 100% unless otherwise mentioned; `r figure("backbone")`).
+The first split within ferns separates a clade including Equisetales and Ophioglossales from all other species.
+The next split separates Marattiales from leptosporangiates (Polypodiidae).
+
+Within leptosporangiate ferns, Osmundales is sister to all other species.
+The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales, followed by Schizaeales.
+The backbone phylogeny recovered Salviniales as sister to Cyatheales but without strong support (), which are in turn sister to Polypodiales.
+
+The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods.
+There are two major clades within eupolypods, eupolypods I and II.
+
+Within eupolypods I, Hypodematiaceae is sister to all remaining species, which diverge successfully in the order Didymochlaenaceae, Dryopteridaceae, Lomariopsidaceae, Nephrolepidaceae, Tectariaceae, Oleandraceae, then Polypodiaceae sister to Davalliaceae.
+
+Within eupolypods II, we recover two large clades.
+One clade consists of Cyctopteridaceae, Rhacidosoraceae, Diplaziopsidaceae, and a clade comprised of Aspleniaceae and Hemidictyaceae.
+The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and a clade comprised of Athyriaceae and Desmophlebiaceae.
+
+## Divergence times
+
+# Discussion
+
+## FTOL phylogeny
+
+The position of Equisetales as sister to Ophioglossales agrees with some plastome studies [ML analysis of @Lehtonen2019] and many plastid studies including fewer genes, but contradicts nuclear studies, which have recovered Equisetales as sister to all other ferns [@Rothfels2015a; @Qi2018; @Shen2018].
+@Rothfels2015a argued that resolution of Equisetales sister to Ophioglossales may have been due to insufficient sampling of plastid genes available at that time, but this is not supported by our study or other recent plastid phylogenomic analyses [@Lehtonen2019].
+
+## Accesibility and usage of FTOL
+
+We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
+
+## Future directions
+
+Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions.
+Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
+However, the plastid tree does not necessarily agree with trees inferred from other data sources.
+Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing, and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
+Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
+Therefore, we have no expectation that there is a single fern tree of life; rather, multiple trees exist that each capture an element of biological reality, and FTOL is merely one such possible tree.
+
+Recent studies using transcriptomics are clarifying the backbone of the fern phylogeny using many (25--2400) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
+The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFLAG) project, which seeks to generate genomic data (ca.
+300 single-to-low copy nuclear gene regions) for all flagellate plants (bryophytes, lycophytes, ferns, and gymnosperms) [@Breinholt2021].
+GoFLAG data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFLAG markers are to be expected in the near future.
+The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data.
+Furthermore, many subclades of ferns are under active investigation using plastid markers using both Sanger and next-gen sequencing, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
+We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for several more years.
+Future efforts should focus on integrating across plastid and nuclear DNA sequences to gain better insight into the complex evolutionary history of ferns.
+
 # References
@@
 <div id="refs"></div>
+
+# Figures
+
+```{r plot-prep}
+# Make vector of outgroup taxa
+og_taxa <-
+	sanger_sampling %>%
+	filter(outgroup == TRUE) %>%
+	pull(species)
+
+# Make tibble of backbone tips with higher-level taxonomy
+backbone_taxonomy <-
+	tibble(species = plastome_tree$tip.label) %>%
+	# Add higher taxonomy
+	left_join(sanger_sampling, by = "species") %>%
+	# Add column for phylogenetic order
+	mutate(phy_order = seq_len(nrow(.))) %>%
+	# Outgroup lacks family data, so call "og"
+	mutate(
+		family = case_when(
+			outgroup == TRUE ~ "og",
+			TRUE ~ family),
+		plot_group = case_when(
+			outgroup == TRUE ~ "og",
+			TRUE ~ plot_group)
+	)
+```
+
+```{r ftol-make}
+# Coverage bar plot ----
+
+# Make exemplar set of species in backbone representing plot groups
+plot_group_exemp_taxa <-
+	backbone_taxonomy %>%
+	# Slice to one exemplar species per family
+	group_by(plot_group) %>%
+	slice(1) %>%
+	ungroup() %>%
+	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
+	arrange(phy_order)
+
+# Get names of exemplar tips in ladderized phylo order
+# Trim tree to exemplar species, rename by family
+phy <- ape::keep.tip(plastome_tree, plot_group_exemp_taxa$species)
+phy$tip.label <- plot_group_exemp_taxa$plot_group
+
+# Make tibble of plot groups in phylogenetic order
+plot_group_phy_order <-
+	phy %>%
+	# Reroot, ladderize, drop outgroup
+	root.phylo("og") %>%
+	ladderize() %>%
+	drop.tip("og") %>%
+	jntools::get_tips_in_ape_plot_order() %>%
+	tibble(
+		name = .,
+		phy_order = seq_along(.)
+	)
+
+rm(phy)
+
+# Make a color palette for major clades
+# Start with list of green.armytage.colors names,
+# exclude those that are too close or stand out too much
+# https://www.jstatsoft.org/article/view/v090c01
+green_army_use_cols <-
+c(
+	"amethyst",
+	"blue",
+	"caramel",
+	"damson",
+	# "ebony",
+	# "forest",
+	"green",
+	"honeydew",
+	# "iron",
+	# "jade",
+	# "khaki",
+	# "lime",
+	"mallow",
+	"navy",
+	# "orpiment",
+	"pink",
+	# "quagmire",
+	"red",
+	"sky",
+	"turquoise",
+	"uranium",
+	"violet",
+	"wine",
+	# "xanthin",
+	# "yellow",
+	"zinnia"
+)
+
+clade_cols <-
+Polychrome::green.armytage.colors() %>%
+  extract(green_army_use_cols) %>%
+	extract(seq_len(nrow(plot_group_phy_order))) %>%
+	set_names(plot_group_phy_order$name)
+
+# Format coverage at order-ish level for plotting
+b <-
+	coverage_data %>%
+	filter(rank == "plot_group") %>%
+	left_join(plot_group_phy_order, by = "name") %>%
+	arrange(phy_order) %>%
+	rename(clade = name) %>%
+	mutate(
+		clade = fct_reorder(clade, phy_order) %>%
+			fct_rev(),
+		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
+	) %>%
+	ggplot(aes(x = coverage, y = clade, fill = clade)) +
+	geom_col(color = "grey10", size = 0.2) +
+	geom_text(
+		aes(label = coverage_text),
+		hjust = 0,
+		nudge_x = 0.01,
+		size = 6*.36 # 6 pt font
+	) +
+	scale_x_continuous(
+		labels = percent,
+		limits = c(0,1),
+		expand = expansion(mult = c(0, 0))
+		) +
+	scale_fill_manual(values = clade_cols) +
+	labs(x = "Coverage") +
+	theme_bw() +
+	theme(
+		axis.text.y = element_text(size = 6),
+		axis.title.y = element_blank(),
+		legend.position = "none",
+		plot.background = element_rect(fill = "transparent", colour = NA),
+		panel.grid.major.y = element_blank(),
+		panel.grid.minor.x = element_blank()
+		)
+
+# Tree plot ----
+
+phy <- sanger_tree %>%
+	root("Anthoceros_angustus") %>%
+	ladderize()
+
+# Make tibble for annotating major clades
+# FIXME: need to do this better when full tree is available
+clade_tibble <-
+sanger_sampling %>%
+	inner_join(
+		tibble(species = phy$tip.label),
+		by = "species"
+	) %>%
+	filter(outgroup == FALSE) %>%
+	select(species, genus, plot_group) %>%
+	group_by(genus, plot_group) %>%
+	slice(1) %>%
+	ungroup() %>%
+	select(-genus) %>%
+	group_by(plot_group) %>%
+	slice(1:2) %>%
+	add_count() %>%
+	filter(n > 1) %>%
+	select(-n) %>%
+	nest() %>%
+	ungroup() %>%
+	mutate(mrca = map_dbl(data, ~getMRCA(phy, .$species))) %>%
+	select(-data) %>%
+	mutate(dummy = "")
+
+a <- ggtree(
+	phy, 
+	layout = "circular", right = FALSE, size = 0.2,
+	# FIXME: change to phylogram when ultrametric tree is done
+	branch.length = "none"
+	) +
+	geom_cladelab(
+		data = clade_tibble,
+		mapping = aes(
+			node = mrca,
+			color = plot_group,
+			label = dummy
+		),
+	offset = 0.1,
+	barsize = 2) +
+	scale_color_manual(values = clade_cols) +
+	theme(legend.position = "none")
+
+ftol_plot <-
+a + inset_element(b, left = 0.85, bottom = 0.14, right = 1.3, top = .8) +
+	plot_annotation(
+		theme = theme(
+			plot.margin = unit(c(-2,5,-2,-1.6), "cm"))
+		)
+```
+
+```{r ftol-save, eval = params$doc_type == "doc"}
+# Save plot if rendering to doc
+ggsave(
+	result_file(figure("ftol"), "pdf"),
+	ftol_plot, width = 7, height = 7, units = "in")
+```
+
+```{r ftol-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
+# Print plot if rendering to PDF
+ftol_plot
+```
+
+**`r figure("ftol")`**. Fern Tree of Life (FTOL) `r ftol_ver`. Inset plot shows coverage by major clade. Numbers next to each bar show sampled species out of total number of species.
+
+```{r backbone-make}
+# Make exemplar set of species in backbone representing families
+family_exemp <-
+	backbone_taxonomy %>%
+	# Slice to one exemplar species per family
+	group_by(family) %>%
+	slice(1) %>%
+	ungroup() %>%
+	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
+	arrange(phy_order)
+
+# Trim tree to exemplar species, rename by family
+phy <- keep.tip(plastome_tree, family_exemp$species)
+phy$tip.label <- family_exemp$family
+# Convert BS support values to numeric
+phy$node.label <- parse_number(phy$node.label)
+# Reroot, ladderize, drop outgroup
+phy <-
+	phy %>%
+	root.phylo("og") %>%
+	ladderize() %>%
+	drop.tip("og")
+
+# Make tibble for printing BS values
+bs_tibble <- tibble(
+	# hard-code node ID: internal nodes start after tip nodes,
+	# and phy$node.label is in the same order as internal nodes
+	node = 1:Nnode(phy) + Ntip(phy), 
+	# Don't print BS = 100%
+	bootstrap = ifelse(phy$node.label == 100, "", phy$node.label))
+
+# Use the ggtree::`%<+%` operator to map the 
+# bootstrap values onto the ggtree object
+backbone_plot <- ggtree(phy) %<+% bs_tibble +
+	# now we can show BS values using geom_text()
+	geom_text(aes(label = bootstrap), hjust = -.25, size = 2.5) +
+	# add tip labels
+	geom_tiplab(aes(label = label)) +
+	xlim(0,1.1)
+```
+
+```{r backbone-tree-save, eval = params$doc_type == "doc"}
+# Save plot if rendering to doc
+ggsave(
+	result_file(figure("backbone"), "pdf"),
+	backbone_plot, width = 7, height = 9, units = "in")
+```
+
+```{r backbone-tree-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
+# Print plot if rendering to PDF
+backbone_plot
+```
+
+**`r figure("backbone")`**. Backbone phylogeny. One exemplar tip is shown per family. Unlabeled bootstrap support values are all 100%. Outgroup (seed plants, lycophytes, and bryophytes) not shown.

### 2022-01-31 R/functions.R

@@ -1,3 +1,17 @@
+# Formatting ----
+
+# Specify percentage format
+percent <- function(...) {scales::percent(...)}
+
+# Specify number format
+number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Abbreviations
+ie <- "*i.e.*"
+eg <- "*e.g.*"
+
+# References ----
+
 #' Extract citations (formatted like `[@key]`) from an Rmd file
 #'
 #' @param rmd_file Character vector; path to Rmd file
@@ -61,3 +75,63 @@ filter_refs_yaml <- function(rmd_file, yaml_in = "ms/main_library.yaml", yaml_ou
 	# Write out the YAML file
 	yaml::write_yaml(ref_yaml_filtered, file = yaml_out)
 }
+
+# Figure output ----
+
+#' Generate a path to save a results file
+#' 
+#' Only works for figures or tables cited in text, and outputs
+#' files to the "results" folder
+#'
+#' @param result_num Number of result, e.g. "Fig. 2"
+#' @param extension Extension to use for file
+#'
+#' @return String.
+#' 
+result_file <- function (result_num, extension) {
+	
+	fs::path(
+		here::here("results"),
+		result_num %>%
+			str_remove_all("\\.") %>% 
+			str_replace_all(" ", "_")
+	) %>%
+		fs::path_ext_set(extension)
+}
+
+# Captions ----
+
+# Follow Frontiers in Plant Sciences style
+# Captions should be preceded by the appropriate label,
+# for example "Figure 1." 
+# Figure panels are referred to by bold capital letters in brackets: 
+# (A), (B), (C), (D), etc.
+
+# - First define the "full" version, which would include a caption
+# (except I never use the caption in the function, and instead replace with 'blank')
+figure_full <- captioner::captioner(prefix = "Figure ", auto_space = FALSE)
+table_full <- captioner::captioner(prefix = "Table ", auto_space = FALSE)
+s_figure_full <- captioner::captioner(prefix = "Figure S ", auto_space = FALSE)
+s_table_full <- captioner::captioner(prefix = "Table S ", auto_space = FALSE)
+
+# - Make a short function that prints only the object type and number, e.g., "Fig. 1"
+figure <- pryr::partial(figure_full, display = "cite", caption = "blank")
+table <- pryr::partial(table_full, display = "cite", caption = "blank")
+s_figure <- pryr::partial(s_figure_full, display = "cite", caption = "blank")
[…]

### 2022-02-02 ms/manuscript.Rmd
Update MS

@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(
+	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
 
@@
 			plot_group = case_when(
-				plot_group %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales",
+				plot_group %in% c("Ophioglossales", "Psilotales") ~ 
+					"Ophioglossales + Psilotales",
 				TRUE ~ plot_group
@@
 	mutate(
-		rgnparser::gn_parse_tidy(scientificName) %>% select(species = canonicalsimple),
+		rgnparser::gn_parse_tidy(scientificName) %>% 
+			select(species = canonicalsimple),
 		species = str_replace_all(species, " ", "_")
@@
 	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
-	left_join(select(ppgi_taxonomy, genus, family, suborder, order, class), by = "genus") %>%
+	left_join(
+		select(ppgi_taxonomy, genus, family, 
+					 suborder, order, class), by = "genus"
+		) %>%
 	# FIXME: update class for nothogenera in PPG
 	mutate(class = case_when(
-		str_detect(scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
+		str_detect(
+			scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
 		TRUE ~ class
@@
 		coverage_p = percent(n / n_accepted),
-		coverage = glue("{number(n)}/{number(n_accepted)}")) %>%
+		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
 	# Split for printing
@@
 
-Ferns are strongly supported as monophyletic ("strongly supported" indicates BS 100% unless otherwise mentioned; `r figure("backbone")`).
-The first split within ferns separates a clade including Equisetales and Ophioglossales from all other species.
-The next split separates Marattiales from leptosporangiates (Polypodiidae).
+```{r topology-stats}
+#' Function to get support value in BS%
+#'
+#' @param phy an object of class "phylo".
+#' @param tips a vector of mode numeric or character specifying the tips
+#'
+#' @return Bootstrap support as % of that node
+get_bs <- function(phy, tips) {
+	node_select <- ape::getMRCA(phy, tips)
+	as_tibble(phy) %>%
+	filter(node == node_select) %>%
+		pull(label) %>%
+		parse_number() %>%
+		magrittr::multiply_by(0.01) %>%
+		percent()
+}
+
+ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
+hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
+cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
+denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
+```
+
+Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
+The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (`r figure("backbone")`).
+The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
 
 Within leptosporangiate ferns, Osmundales is sister to all other species.
-The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales, followed by Schizaeales.
-The backbone phylogeny recovered Salviniales as sister to Cyatheales but without strong support (), which are in turn sister to Polypodiales.
+The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
+Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales.
 
-The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods.
-There are two major clades within eupolypods, eupolypods I and II.
+The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
+There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).
 
-Within eupolypods I, Hypodematiaceae is sister to all remaining species, which diverge successfully in the order Didymochlaenaceae, Dryopteridaceae, Lomariopsidaceae, Nephrolepidaceae, Tectariaceae, Oleandraceae, then Polypodiaceae sister to Davalliaceae.
+Within Polypodiineae, Hypodematiaceae is sister to all remaining species, which diverge successfully in the order Didymochlaenaceae, Dryopteridaceae, Lomariopsidaceae, Nephrolepidaceae, Tectariaceae, Oleandraceae, then finally Polypodiaceae sister to Davalliaceae.
 
-Within eupolypods II, we recover two large clades.
+Within Aspleniineae, we recover two large clades.
 One clade consists of Cyctopteridaceae, Rhacidosoraceae, Diplaziopsidaceae, and a clade comprised of Aspleniaceae and Hemidictyaceae.
@@
 
-The position of Equisetales as sister to Ophioglossales agrees with some plastome studies [ML analysis of @Lehtonen2019] and many plastid studies including fewer genes, but contradicts nuclear studies, which have recovered Equisetales as sister to all other ferns [@Rothfels2015a; @Qi2018; @Shen2018].
-@Rothfels2015a argued that resolution of Equisetales sister to Ophioglossales may have been due to insufficient sampling of plastid genes available at that time, but this is not supported by our study or other recent plastid phylogenomic analyses [@Lehtonen2019].
+The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
+We recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
+This contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have recovered Equisetidae as sister to all other ferns; some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae with low support [@Pryer2001; @Qiu2006; @Qiu2007].
+Aside from alternative resolutions of Equisetidae dependent on nuclear envelope or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
+The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is lacking.
+
+Another enigmatic relationship in ferns is the placement of Hymenophyllales.
+The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates.
+However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
+Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a non-monophyletic Gleicheniales [@Qi2018; @Shen2018].
+We recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
+However, current plastome sampling does not include all genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
 
@@
 We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
+FTOL is available via the "ftol" R package, as well as a web portal (FIXME: add URL when ready).
+
+A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data from joining [@Page2008].
+During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/joelnitta/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand) .
+The "pteridocat" package includes the pteridcat taxonomic database as a data frame (tibble) in Darwin Core format.
+The "taxastand" package includes functions to resolve taxonomic names while taking account variation in taxonomic author format and orthographic variation.
+By using these two packages in combination, it should be straightforward for other researchers to map their own data onto FTOL, thus greatly enabling and enhancing studies including, but not limited to, comparative phylogenetics, biogeography, and community ecology in ferns.
 
@@
 # Tree plot ----
-
 phy <- sanger_tree %>%
@@
 # Make tibble for annotating major clades
-# FIXME: need to do this better when full tree is available
 clade_tibble <-
-sanger_sampling %>%
-	inner_join(
-		tibble(species = phy$tip.label),
-		by = "species"
+tibble(
+	Polypodiineae = c("Leucostegia_truncata", "Didymochlaena_fijiensis"),
+	Aspleniineae = c("Acystopteris_taiwaniana", "Rhachidosorus_pulcher"),
+	Dennstaedtiineae = c("Monachosorum_flagellare", "Dennstaedtia_auriculata"),
+	Pteridineae = c("Coniogramme_affinis", "Pteris_dispar"),
+	Lindsaeineae = c("Cystodium_sorbifolium", "Odontosoria_retusa"),
+	Saccolomatineae = c("Saccoloma_inaequale", "Orthiopteris_campylura"),
+	Cyatheales = c("Cyathea_bipinnatifida", "Culcita_macrocarpa"),
+	Salviniales = c("Marsilea_nashii", "Salvinia_oblongifolia"),
+	Schizaeales = c("Schizaea_elegans", "Lygodium_palmatum"),
+	Gleicheniales = c("Matonia_foxworthyi", "Sticherus_bifidus"),
+	Hymenophyllales = c("Cephalomanes_javanicum", "Hymenophyllum_tortuosum"),
+	Marattiales = c("Angiopteris_evecta", "Danaea_erecta"),
+	`Ophioglossales + Psilotales` = c("Psilotum_nudum", "Sceptridium_nipponicum"),
+	Osmundales = c("Osmundastrum_cinnamomeum", "Todea_barbara"),
+	Equisetales = c("Equisetum_bogotense", "Equisetum_hyemale")
 ) %>%
-	filter(outgroup == FALSE) %>%
-	select(species, genus, plot_group) %>%
-	group_by(genus, plot_group) %>%
-	slice(1) %>%
-	ungroup() %>%
-	select(-genus) %>%
+	pivot_longer(names_to = "plot_group", values_to = "taxa", cols = everything()) %>%
+	arrange(plot_group) %>%
 	group_by(plot_group) %>%
-	slice(1:2) %>%
-	add_count() %>%
-	filter(n > 1) %>%
-	select(-n) %>%
-	nest() %>%
-	ungroup() %>%
-	mutate(mrca = map_dbl(data, ~getMRCA(phy, .$species))) %>%
-	select(-data) %>%
-	mutate(dummy = "")
+	summarize(taxa = list(taxa)) %>%
+	mutate(
+		mrca = map_chr(taxa, ~getMRCA(phy, .))
+	) %>%
+	transmute(
+		plot_group,
+		mrca,
+		dummy = ""
+	) %>%
+	mutate(mrca = parse_number(mrca))
 
@@
 	phy, 
-	layout = "circular", right = FALSE, size = 0.2,
+	layout = "circular", right = FALSE, size = 0.1,
 	# FIXME: change to phylogram when ultrametric tree is done
@@
 
+tibble(
+	Polypodiineae = c("Hypodematiaceae", "Davalliaceae"),
+	Aspleniineae = c("Cystopteridaceae", "Athyriaceae"),
+	Lindsaeineae = c("Lindsaeaceae", "Cystodiaceae"),
+	Cyatheales = c("Cyatheaceae", "Plagiogyriaceae"),
+	Salviniales = c("Marsileaceae", "Salviniaceae"),
+	Schizaeales = c("")
+) %>%
+	pivot_longer(names_to = "plot_group", values_to = "taxa", cols = everything()) %>%
+	arrange(plot_group) %>%
+	group_by(plot_group)
+
+# Make tibble for labeling clades
+phy_tip <- as_tibble(phy)
+
+suborder_labels_singletons <-
+ family_exemp %>% 
+	filter(outgroup == FALSE) %>%
+	group_by(plot_group) %>%
+	select(plot_group, family) %>% 
+	filter(n_distinct(family) == 1) %>%
+ 	ungroup() %>%
+ 	left_join(phy_tip, by = c(family = "label")) %>%
+ 	select(plot_group, mrca = node)
+
+suborder_labels <-
+  family_exemp %>% 
+	filter(outgroup == FALSE) %>%
+	group_by(plot_group) %>%
+	select(plot_group, family) %>% 
+	filter(n_distinct(family) > 1) %>%
+	summarize(taxa = list(family)) %>%
+	mutate(
+		mrca = map_chr(taxa, ~getMRCA(phy, .))
+	) %>%
+	transmute(
+		plot_group,
+		mrca
+	) %>%
+	mutate(mrca = parse_number(mrca)) %>%
+	bind_rows(suborder_labels_singletons)
+
 # Use the ggtree::`%<+%` operator to map the 
 # bootstrap values onto the ggtree object
-backbone_plot <- ggtree(phy) %<+% bs_tibble +
+backbone_plot <- 
+	ggtree(phy) %<+% bs_tibble +
 	# now we can show BS values using geom_text()
@@
 	geom_tiplab(aes(label = label)) +
-	xlim(0,1.1)
+	# label clades
+	geom_cladelab(
+		data = suborder_labels,
+		mapping = aes(
+			node = mrca,
+			label = plot_group
+		),
+		offset = 0.26,
+		align = TRUE,
+		extend = 0.3,
+		textcolour = "black",
+		barsize = 2) +
+	# highlight clades
+	geom_hilight(
+		data = suborder_labels,
+		mapping = aes(
+			node = mrca,
+			fill = plot_group
+		),
+		alpha = 0.2,
+		align = "right",
+		extend = 0.28,
+		gradient.direction = "rt",
+		gradient.length.out = 2) +
+	scale_fill_manual(values = clade_cols) +
+	xlim(0, 1.7) +
+	theme(legend.position = "none")
 ```

### 2022-02-02 R/functions.R

@@ -9,6 +9,7 @@ number <- function(...) {scales::number(big.mark = ",", ...)}
 # Abbreviations
 ie <- "*i.e.*"
 eg <- "*e.g.*"
+ca <- "*ca.*"
 
 # References ----
 

### 2022-02-03 ms/manuscript.Rmd
Update MS

@@
 library(magrittr)
+library(ggrepel)
 
@@
 # FIXME: read in real ultrametric tree when ready
-sanger_tree <- ape::read.tree(here::here("working/sanger_alignment.phy.contree.tre"))
+sanger_tree <- ape::read.tree(here::here("working/sanger_alignment_2022-02-02.tre"))
 ```
@@
 ```{r abstract-stats}
-# Helper function to add plot_group:
-# a "rank" that may be at order or suborder level for plotting
-add_plot_group <- function(data) {
-	data %>%
-		mutate(
-			plot_group = coalesce(suborder, order),
-			plot_group = case_when(
-				plot_group %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales",
-				TRUE ~ plot_group
-			)
-		)
-}
-
 # Make tibble of outgroup species
@@
 
-## Automated removal of contaminations
+## Automated removal of rogue sequences
 
-Another potential issue with GenBank sequences is the presence of misidentified sequences and contaminations (hereafter "contaminations").
-We removed putative contaminations from Sanger sequences as follows.
+Another potential issue with GenBank sequences is the presence of misidentified sequences and contaminations (hereafter "rogues").
+We removed putative rogues from Sanger sequences as follows.
 First, we constructed a BLAST library including all extracted Sanger sequences.
@@
 Species belonging to monotypic families were not considered for this filtering.
-While this method cannot account for contaminations at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
+While this method cannot account for rogues at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
 
@@
 
+We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic methods as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable phylogenetic methods.
+
+## Manual inspection
+
+After generating the initial ML tree, we manually inspected it to identify any putative remaining rogues at finer levels that were missed by our automated detection at the family level.
+Where applicable, we updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect.
+However, this is not possible in all cases, particularly in groups that are in need of taxonomic revision (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
+While we consider our phylogeny as an important guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the Code of Botanical Taxonomy.
+In a very small number of cases we assigned a different higher level taxonomy for recently described names if no existing combinations were available that were compatible with our taxonomic system.
+These are indicated in our supplemental data files as "comb. ined.".
+We compiled a list of putative rogue GenBank accessions, re-ran all analyses to this point after excluding rogues, then continued with the manually inspected phylogeny.
+
 ## Molecular dating
 
-We dated the tree using treePL [@Smith2012], which is one of the only programs capable of dating very large alignments.
-We applied a set of `r nrow (plastid_calibration_dates)` calibration points, mostly based on @Testo2016a and references within.
+We dated the tree using treePL [@Smith2012].
+We applied a set of `r nrow (plastid_calibration_dates)` calibration points as minimum ages.
+Our calibration points are all based on fossils (not secondary calibration points) following a careful literature review.
+We used the species identification of each fossil to assign lineage only after consulting the original publication; previous studies [`r eg`, @Testo2016a] have used genus to assign lineage, but this can be problematic as genus concepts may vary between the original publication and current taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
 
@@
 A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data from joining [@Page2008].
-During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/joelnitta/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand) .
-The "pteridocat" package includes the pteridcat taxonomic database as a data frame (tibble) in Darwin Core format.
+During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/joelnitta/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand).
+We selected R because it is widely used by the ecological research community, well established, and freely available.
+The "pteridocat" package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
 The "taxastand" package includes functions to resolve taxonomic names while taking account variation in taxonomic author format and orthographic variation.
@@
 
-## Future directions
+## FTOL as a living, community-driven resource
+
+One feature of FTOL that sets it apart from the vast majority of other fern phylogenetic studies is its iterative nature.
+Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript (`r ftol_ver`) is not meant to be the last.
+Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree regularly (`ie`, semi-annually).
+
+Furthermore, FTOL will not only grow in size with time, but also become more refined.
+We are aware that our methodology cannot produce a "perfect" tree, nor is that our goal.
+Each round of analysis involves a step to manually inspect the tree and make taxonomic modifications (`ie`, change opinion regarding synonyms) and flag rogue accessions for exclusion.
+However, it is impossible to catch everything since it is possible the subsequent round of GenBank mining will recover additional rogues among the replacements for the excluded sequences.
+Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked, or proper taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
+That is why we have made our methodology (code), data, and software (R packages and docker image) completely open and available to the research community.
+It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through GitHub (https://github.com/joelnitta/ftol) or by contacting us by email.
+This way, FTOL will continually improve and keep pace with the currently available data and taxonomic hypotheses of ferns.
+
+We want to make it clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
+FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphologic, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
+
+## Plastid vs. nuclear trees
 
@@
 	ladderize() %>%
-	drop.tip("og")
+	drop.tip("og") %>%
+	# Scale to total height of 100 for easy plotting
+	rescale_tree(100)
 
@@
 	bootstrap = ifelse(phy$node.label == 100, "", phy$node.label))
-
-tibble(
-	Polypodiineae = c("Hypodematiaceae", "Davalliaceae"),
-	Aspleniineae = c("Cystopteridaceae", "Athyriaceae"),
-	Lindsaeineae = c("Lindsaeaceae", "Cystodiaceae"),
-	Cyatheales = c("Cyatheaceae", "Plagiogyriaceae"),
-	Salviniales = c("Marsileaceae", "Salviniaceae"),
-	Schizaeales = c("")
-) %>%
-	pivot_longer(names_to = "plot_group", values_to = "taxa", cols = everything()) %>%
-	arrange(plot_group) %>%
-	group_by(plot_group)
 
@@
 	mutate(mrca = parse_number(mrca)) %>%
-	bind_rows(suborder_labels_singletons)
+	bind_rows(suborder_labels_singletons) %>%
+	mutate(
+		clade_label = case_when(
+			plot_group == "Aspleniineae" ~ "Aspleniineae\n(eupolypods II)",
+			plot_group == "Polypodiineae" ~ "Polypodiineae\n(eupolypods I)",
+			TRUE ~ plot_group
+		)
+	)
 
+# Get data of x and y co-ordinates for ggtree plot
+phy_plot_data <- ggplot_build(ggtree(phy)) %>%
+	magrittr::extract2("data") %>%
+	magrittr::extract2(1) %>%
+	as_tibble()
+
+# Make a dataframe for plotting various clade labels
+other_clade_labels <-
+list(
+	"eupolypods" = c("Polypodiaceae", "Aspleniaceae"),
+	"Polypodiales" = c("Polypodiaceae", "Saccolomataceae"),
+	"Polypodiidae\n(leptosporangiates)" = c("Polypodiaceae", "Osmundaceae")
+) %>%
+	tibble(
+		clade_label = names(.),
+		x = map_dbl(., ~node_height_from_tips(phy, .)),
+		node = map_dbl(., ~ape::getMRCA(phy, .))
+	) %>%
+	left_join(
+		select(phy_plot_data, node, y), by = "node"
+	) %>%
+	select(-`.`, -node)
+
+segment_cols <-
+rep("grey20", length(other_clade_labels$clade_label)) %>%
+	set_names(other_clade_labels$clade_label)
+
+set.seed(123)
 # Use the ggtree::`%<+%` operator to map the 
@@
 	geom_tiplab(aes(label = label)) +
-	# label clades
-	geom_cladelab(
-		data = suborder_labels,
-		mapping = aes(
-			node = mrca,
-			label = plot_group
-		),
-		offset = 0.26,
-		align = TRUE,
-		extend = 0.3,
-		textcolour = "black",
-		barsize = 2) +
-	# highlight clades
+	# highlight major clades
 	geom_hilight(
@@
 		align = "right",
-		extend = 0.28,
+		extend = 30,
 		gradient.direction = "rt",
+		roundrect = TRUE,
 		gradient.length.out = 2) +
 	scale_fill_manual(values = clade_cols) +
-	xlim(0, 1.7) +
+	# label major clades
+	geom_cladelab(
+		data = suborder_labels,
+		mapping = aes(
+			node = mrca,
+			label = clade_label
+		),
+		offset = 28,
+		align = TRUE,
+		extend = 0.3,
+		hjust = 0,
+		barcolour = "grey20",
+		barsize = 2) +
+	# Label other groups
+	geom_label_repel(
+		data = other_clade_labels %>%
+			filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
+		aes(x = x, y = y, label = clade_label),
+		hjust = 0,
+		xlim = c(NA, 10),
+		ylim = c(13, NA),
+		color = "grey20",
+		segment.color = "grey20"
+	) +
+	geom_label_repel(
+		data = other_clade_labels %>%
+			filter(clade_label == "eupolypods"),
+		aes(x = x, y = y, label = clade_label),
+		hjust = 0,
+		xlim = c(NA, 64),
+		ylim = c(32, NA),
+		color = "grey20",
+		segment.color = "grey20"
+	) +
+	geom_label_repel(
+		data = other_clade_labels %>%
+			filter(clade_label == "Polypodiales"),
+		aes(x = x, y = y, label = clade_label),
+		hjust = 0,
+		nudge_x = -25,
+		nudge_y = 4,
+		color = "grey20",
+		segment.color = "grey20"
+	) +
+	xlim(0, 180) +
+	coord_cartesian(clip = "off") +
 	theme(legend.position = "none")
@@
 
-**`r figure("backbone")`**. Backbone phylogeny. One exemplar tip is shown per family. Unlabeled bootstrap support values are all 100%. Outgroup (seed plants, lycophytes, and bryophytes) not shown.
+**`r figure("backbone")`**.
+Fern tree of life (FTOL) backbone phylogeny.
+One exemplar tip is shown per family.
+Ultrafast bootstrap support values (%) shown at nodes; unlabeled nodes are 100%.
+Outgroup (seed plants, lycophytes, and bryophytes) not shown.
+Colors of major clades correspond to those used in `r figure("ftol")`.
+Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.

### 2022-02-03 R/functions.R

@@ -136,3 +136,49 @@ s_table_num <- function(name) {s_table(name) %>% str_remove("Table S ")}
 #'
 #' Breaks a page depending if the output format is PDF
 pagebreak_pdf <- function(rmd_params = params) {ifelse(rmd_params$doc_type == "pdf", return("\\clearpage"), return(""))}
+
+# Phylogeny ----
+
+#' Rescale tree height
+#' 
+#' from:
+#' http://blog.phytools.org/2012/02/quicker-way-to-rescale-total-length-of.html
+#'
+#' @param tree Input phylogeny (list of class "phylo")
+#' @param scale Numeric: total height to scale to
+#'
+#' @return Phylogeny (list of class "phylo")
+#' 
+rescale_tree <- function(tree, scale){
+	tree$edge.length <-
+		tree$edge.length/max(phytools::nodeHeights(tree)[,2])*scale
+	return(tree)
+}
+
+#' Get node height corresponding to the most recent common
+#' ancestor of two or more taxa in a phylogeny
+#'
+#' @param phy List of class "phylo"
+#' @param tips Character vector; tips of the phylogeny
+#'
+#' @return Height of the MRCA node of the tips
+#' 
+node_height_from_tips <- function(phy, tips) {
+	ape::getMRCA(phy, tips) %>%
+		phytools::nodeheight(phy, .)
+}
+
+# Etc -----
+# Helper function to add plot_group:
+# a "rank" that may be at order or suborder level for plotting
+add_plot_group <- function(data) {
+	data %>%
+		mutate(
+			plot_group = coalesce(suborder, order),
+			plot_group = case_when(
+				plot_group %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales",
+				TRUE ~ plot_group
+			)
+		)
+}
+

### 2022-02-03 render.R
Add script for rendering MS to results folder

@@ -1,0 +1,1 @@
+rmarkdown::render("ms/manuscript.Rmd", output_dir = "results")

### 2022-02-03 ms/manuscript.Rmd
Update MS

@@
 We dated the tree using treePL [@Smith2012].
-We applied a set of `r nrow (plastid_calibration_dates)` calibration points as minimum ages.
+We applied a set of FIXME calibration points as minimum ages.
 Our calibration points are all based on fossils (not secondary calibration points) following a careful literature review.
@@
 	# Label other groups
-	geom_label_repel(
+	geom_text_repel(
 		data = other_clade_labels %>%
@@
 		xlim = c(NA, 10),
-		ylim = c(13, NA),
+		ylim = c(15, NA),
 		color = "grey20",
@@
 	) +
-	geom_label_repel(
+	geom_text_repel(
 		data = other_clade_labels %>%
@@
 		hjust = 0,
-		nudge_x = -25,
-		nudge_y = 4,
+		xlim = c(NA, 48),
+		ylim = c(24, NA),
 		color = "grey20",

### 2022-02-03 ms/manuscript.Rmd
Move functions to ms_functions

@@
 # Load functions
-source(here::here("R/functions.R"))
+source(here::here("R/ms_functions.R"))
 
@@
 The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates.
-However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
+However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
 Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a non-monophyletic Gleicheniales [@Qi2018; @Shen2018].
@@
 	)
+```
+
+```{r gb-make}
+
+gb_acc_by_dna_type %>%
+	filter(year == 2020)
+
+ggplot(gb_acc_by_dna_type, aes(x = year, y = hits, color = type)) +
+	geom_line() +
+	geom_point()
+
 ```

### 2022-02-04 R/functions.R
Use functions.R for targets functions

@@ -1,0 +1,212 @@
+
+# GenBank ----
+
+#' Fetch metadata from GenBank
+#'
+#' @param query String to use for querying GenBank
+#' @param col_select Character vector; columns of metadata to retain in output
+#'
+#' @return Tibble
+#' 
+fetch_metadata <- function(
+	query = NULL,
+	col_select = c("gi", "caption", "taxid", "title", "slen", "subtype", "subname")) {
+	
+	assertthat::assert_that(assertthat::is.string(query))
+	
+	assertthat::assert_that(is.character(col_select))
+	
+	# Do an initial search without downloading any IDs to see how many hits
+	# we get.
+	initial_genbank_results <- rentrez::entrez_search(
+		db = "nucleotide",
+		term = query,
+		use_history = FALSE
+	)
+	
+	# If no results, return empty tibble
+	if (initial_genbank_results$count < 1) return(tibble(taxid = NA))
+
+	# Download IDs with maximum set to 1 more than the total number of hits.
+	genbank_results <- rentrez::entrez_search(
+		db = "nucleotide",
+		term = query,
+		use_history = FALSE,
+		retmax = initial_genbank_results$count + 1
+	)
+	
+	# Define internal function to download genbank data into tibble
+	entrez_summary_gb <- function(id, col_select) {
+		# Download data
+		rentrez::entrez_summary(db = "nucleotide", id = id) %>%
+			# Extract selected columns from result
+			purrr::map_dfr(magrittr::extract, col_select) %>%
+			# Make sure taxid column is character
+			mutate(taxid = as.character(taxid)) %>%
+			assert(not_na, taxid)
+	}
+	
+	# Extract list of IDs from search results
+	genbank_ids <- genbank_results$ids
+	
+	# Fetch metadata for each ID and extract selected columns
+	if (length(genbank_ids) == 1) {
+		rentrez_results <- rentrez::entrez_summary(db = "nucleotide", id = genbank_ids) %>%
+			magrittr::extract(col_select) %>%
+			tibble::as_tibble() %>%
+			mutate(taxid = as.character(taxid)) %>%
+			assert(not_na, taxid)
+	} else {
+		# Split input vector into chunks
+		n <- length(genbank_ids)
+		chunk_size <- 200
+		r <- rep(1:ceiling(n/chunk_size), each = chunk_size)[1:n]
[…]

### 2022-02-04 ms/manuscript.Rmd
Use targets to render MS

@@
 ```{r load-targets}
+# Targets from FTOL workflow
 tar_load(
@@
 		ppgi_taxonomy,
-		plastome_alignment,
 		plastid_calibration_dates,
@@
 
+# Targets from MS workflow
+tar_load(
+	c(
+		gb_species_by_year,
+		ref_files
+	),
+	store = here::here("_targets")
+)
+
 # FIXME: read in real ultrametric tree when ready
-sanger_tree <- ape::read.tree(here::here("working/sanger_alignment_2022-02-02.tre"))
+sanger_tree <- ape::read.tree(here::here("working/sanger_alignment.phy.tre"))
 ```
@@
 Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a].
+Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a\; `r figure("genbank")`].
 We believe there is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
@@
 
-The plastid genome has been the most widely sequenced genomic compartment in ferns by far, so we used plastid loci to build our tree.
+The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
 Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome loci").
@@
 
+```{r phy-discussion-stats}
+# Calculate number of genera sampled for Gleicheniales
+ppgi_gleich_genera_n <-
+	ppgi_taxonomy %>%
+	filter(order == "Gleicheniales") %>%
+	pull(genus) %>%
+	n_distinct()
+
+plastome_gleich_genera_n <-
+	bb_sampling %>%
+	filter(plot_group == "Gleicheniales") %>%
+	pull(genus) %>%
+	n_distinct()
+```
+
 The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
@@
 Another enigmatic relationship in ferns is the placement of Hymenophyllales.
-The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates.
+The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates [@Pryer2001; @Schneider2004d; @Schuettpelz2007; @Kuo2011; @Testo2016a].
 However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
@@
 We recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
-However, current plastome sampling does not include all genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
+However, current plastome sampling only includes `r plastome_gleich_genera_n` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
+
+Within Polypodiales, the relationship between suborders Pteridineae, Dennstaedtiineae and the eupolypods (Polypodiineae and Aspleniineae) has been difficult to resolve (here designated "P", "D", and "e", respectively).
+Most previous plastid studies based on Sanger sequencing have recovered (P, (D, e)) [@Kuo2011; @Schuettpelz2006; @Schuettpelz2007; @Testo2016a; but see @Lehtonen2011].
+We recover (D, (P, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] or nuclear data [@Rothfels2015a; @Qi2018; @Shen2018] as well as a plastid supermatrix [@Lehtonen2011].
+Recently the whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others.
+
+Taken together, our results for nodes of contention in the fern phylogeny generally agree with other plastid phylogenomic analyses and are well within the realm of plausible hypotheses generated to date.
+We do not consider any of these nodes "solved" by our analysis.
+Rather, conclusive resolution apparently still awaits additional sampling and perhaps innovation in phylogenetic methods.
 
@@
 
-## Plastid vs. nuclear trees
-
-Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions.
+## Plastid vs. nuclear fern trees
+
+Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions `r figure("genbank")`.
 Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
@@
 GoFLAG data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFLAG markers are to be expected in the near future.
-The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data.
+The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data `r figure("genbank")`.
 Furthermore, many subclades of ferns are under active investigation using plastid markers using both Sanger and next-gen sequencing, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
@@
 
-```{r gb-make}
-
-gb_acc_by_dna_type %>%
-	filter(year == 2020)
-
-ggplot(gb_acc_by_dna_type, aes(x = year, y = hits, color = type)) +
-	geom_line() +
-	geom_point()
-
-```
+```{r genbank-make}
+fern_tree_sampling <-
+	tribble(
+		~source, ~year, ~n_species,
+		"Hasebe1995", 1995, 107,
+		"Schuettpelz2007", 2007, 400,
+		"Lehtonen2011", 2011, 2957,
+		"Testo2016a", 2016, 3973,
+		"this_study", 2021, n_sanger_ferns
+	) %>%
+	mutate(source = fct_reorder(source, year))
+
+genbank_plot <-
+	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
+	geom_line(aes(color = type)) +
+	geom_point(
+		data = fern_tree_sampling,
+		aes(shape = source)
+	)
+```
+
+```{r genbank-save, eval = params$doc_type == "doc" }
+# Save plot if rendering to doc
+ggsave(
+	result_file(figure("genbank"), "pdf"),
+	genbank_plot, width = 3.3, height = 5, units = "in")
+```
+
+```{r genbank-print, fig.width = 3.3, fig.height = 5, eval = params$doc_type == "pdf"}
+# Print plot if rendering to PDF
+genbank_plot
+```
+
+**`r figure("genbank")`**.
+Number of fern species in GenBank by year and genomic compartment.
+Points indicate number of species sampled in selected studies of global fern phylogeny [@Hasebe1995; @Schuettpelz2007; @Lehtonen2011; @Testo2016a\; this study].
+@Schuettpelz2007 did not attempt exhaustive sampling but rather proportional sampling according to lineage size.
+The relatively small increase in number of species in 2021 is likely due to accessions that are still embargoed at the time of writing.
+Taxonomy of GenBank species follows NCBI [@Federhen2012].
+Only accessions identified to species included; environmental samples, hybrid formulas, and names with "aff." or "cf." annotations excluded.
 
@@
 
-```{r ftol-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
+```{r ftol-print, fig.width = 7, fig.height = 7, eval = params$doc_type == "pdf"}
 # Print plot if rendering to PDF
@@
 
-**`r figure("ftol")`**. Fern Tree of Life (FTOL) `r ftol_ver`. Inset plot shows coverage by major clade. Numbers next to each bar show sampled species out of total number of species.
+**`r figure("ftol")`**.
+Fern Tree of Life (FTOL) `r ftol_ver`.
+Inset plot shows coverage by major clade (order or suborder).
+Numbers next to each bar show sampled species out of total number of species.
+Taxonomy follows [@PteridophytePhylogenyGroupI2016].
 
@@
 
-```{r backbone-tree-print, fig.width = 5, fig.height = 5, eval = params$doc_type == "pdf"}
+```{r backbone-tree-print, fig.width = 7, fig.height = 9, eval = params$doc_type == "pdf"}
 # Print plot if rendering to PDF
@@
 Outgroup (seed plants, lycophytes, and bryophytes) not shown.
-Colors of major clades correspond to those used in `r figure("ftol")`.
+Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`.
 Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.

### 2022-02-04 _targets.R

@@ -1,0 +1,39 @@
+library(targets)
+library(tarchetypes)
+library(tidyverse)
+library(assertr)
+
+source("R/functions.R")
+
+tar_plan(
+	# Analyze number of fern accessions and species in GenBank by type of DNA 
+	# (plastid, mitochondrial, or nuclear)
+	#
+	# - Make dataframe of query terms by year
+	gb_query = make_gb_query(),
+	# - Download taxids one year at a time
+	tar_target(
+		gb_taxa,
+		fetch_taxa_by_year(
+			query = gb_query$query,
+			year = gb_query$year,
+			type = gb_query$type),
+		pattern = map(gb_query)
+		),
+	# - Load NCBI taxonomy
+	tar_file(taxdump_zip_file, "working/taxdmp_2022-02-01.zip"),
+	ncbi_names = load_ncbi_names(
+		taxdump_zip_file,
+		unique(gb_taxa$taxid)
+	),
+	# - Count number of species and accessions in GenBank
+	# by genomic compartment per year
+	gb_species_by_year = count_ncbi_species_by_year(
+		gb_taxa, ncbi_names, year_range = 1990:2021),
+	# Render manuscript
+	tar_render(
+		ms,
+		"ms/manuscript.Rmd",
+		output_dir = "results"
+	)
+)

### 2022-02-04 render.R

@@ -1,1 +1,0 @@
-rmarkdown::render("ms/manuscript.Rmd", output_dir = "results")

### 2022-02-04 ms/manuscript.Rmd
Update MS

@@
 library(ggrepel)
+library(ggokabeito)
 
@@
 
-Phylogenetic analysis was conducted in two steps.
-First, we generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome loci in IQ TREE [@Nguyen2015] with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
-We applied the 'GTR+I+G' model to a concatenated dataset with no partitioning.
-
-Next, we used the backbone phylogeny as a constraint tree to conduct phylogenetic analysis of the Sanger loci in IQ TREE.
-Settings for the ML analysis of the Sanger loci were otherwise identical to the backbone phylogenetic analysis.
-
-We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic methods as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable phylogenetic methods.
-
-## Manual inspection
-
-After generating the initial ML tree, we manually inspected it to identify any putative remaining rogues at finer levels that were missed by our automated detection at the family level.
-Where applicable, we updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect.
-However, this is not possible in all cases, particularly in groups that are in need of taxonomic revision (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny as an important guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the Code of Botanical Taxonomy.
+### Backbone phylogeny
+
+We generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome dataset in IQ-TREE [@Nguyen2015] with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
+We applied the GTR+I+G model to a concatenated dataset with no partitioning.
+
+### Initial Sanger phylogeny
+
+Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE in with the "-fast" option and same model settings as the backbone phylogeny.
+We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
+
+### Manual inspection of initial Sanger phylogeny
+
+We manually inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
+We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
+We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; @Schwery2016].
+However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
+While we consider our phylogeny as an important guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
 In a very small number of cases we assigned a different higher level taxonomy for recently described names if no existing combinations were available that were compatible with our taxonomic system.
 These are indicated in our supplemental data files as "comb. ined.".
-We compiled a list of putative rogue GenBank accessions, re-ran all analyses to this point after excluding rogues, then continued with the manually inspected phylogeny.
+
+Based on the results of the manual inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses to this step.
+When re-running the analysis, it is possible that there were newly added GenBank accessions in place of the excluded rogues that themselves are rogues.
+Therefore, we repeated this process until the expected monophyly of all higher-level (`r eg` genus rank and higher) taxa that have previously been shown to be monophyletic was confirmed.
+
+### Final Sanger phylogeny
+
+After manual inspection, we generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the Sanger dataset under otherwise identical settings as those used for the backbone phylogeny.
+
+We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
 
@@
 We used the species identification of each fossil to assign lineage only after consulting the original publication; previous studies [`r eg`, @Testo2016a] have used genus to assign lineage, but this can be problematic as genus concepts may vary between the original publication and current taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
+We tested rate smoothing parameters ranging from 0.000001 to 1000 (each varies by one order of magnitude); a value of FIXME was selected with the smallest chi-squared value for final analysis [@Smith2012].
 
@@
 
+compartment_labs <- 
+	gb_species_by_year %>%
+	filter(year == max(year)) %>%
+	transmute(
+		type,
+		label = str_to_sentence(type),
+		year,
+		n_species
+	)
+
+okabe_ito_cols <-
+palette_okabe_ito(1:9) %>%
+	set_names(
+		c("orange", "light_blue", "green", 
+			"yellow", "blue", "red",
+			"magenta", "grey", "black"))
+
+compartment_cols <- c(
+	mitochondrial = okabe_ito_cols["red"] %>% set_names(NULL),
+	nuclear = okabe_ito_cols["blue"] %>% set_names(NULL),
+	plastid = okabe_ito_cols["green"] %>% set_names(NULL),
+	total = "grey40"
+)
+
 genbank_plot <-
 	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
-	geom_line(aes(color = type)) +
+	geom_line(aes(color = type), size = .7) +
+	scale_color_manual(values = compartment_cols, guide = "none") +
+	geom_label_repel(
+		data = compartment_labs %>% filter(type %in% c("nuclear", "mitochondrial")),
+		aes(color = type, label = label),
+		segment.linetype = 2, 
+		xlim = c(2020, Inf),
+    # Do not repel from top or bottom edges.
+    ylim = c(-Inf, Inf)
+	) +
+	geom_label_repel(
+		data = compartment_labs %>% filter(type %in% c("total", "plastid")),
+		aes(color = type, label = label),
+		segment.linetype = 2, 
+		xlim = c(2020, Inf),
+    # Do not repel from top or bottom edges.
+    ylim = c(-Inf, Inf),
+		min.segment.length = 0
+	) +
 	geom_point(
 		data = fern_tree_sampling,
-		aes(shape = source)
-	)
+		aes(shape = source),
+		size = 2.5,
+	) +
+	labs(shape = "Study", y = "No. species", x = "Year") +
+	theme(legend.position = "bottom") +
+	coord_cartesian(clip = "off")
 ```
@@
 	result_file(figure("genbank"), "pdf"),
-	genbank_plot, width = 3.3, height = 5, units = "in")
+	genbank_plot, width = 7, height = 5, units = "in")
 ```

### 2022-02-04 R/process_refs.R

@@ -1,6 +1,6 @@
 # Filter a list of references in YAML format to those cited in an RMD file
 library(tidyverse)
-source("R/functions.R")
+source("R/ms_functions.R")
 
 # Filter YAML references from Zotero library.
 # data_raw/main_library.yaml has been exported from Zotero

### 2022-02-04 _targets.R

@@ -30,7 +30,9 @@ tar_plan(
 	# by genomic compartment per year
 	gb_species_by_year = count_ncbi_species_by_year(
 		gb_taxa, ncbi_names, year_range = 1990:2021),
-	# Render manuscript
+	# Render manuscript ----
+	# track files used for rendering MS
+	tar_file(ref_files, list.files("ms", "references", full.names = TRUE)),
 	tar_render(
 		ms,
 		"ms/manuscript.Rmd",

### 2022-02-04 ms/manuscript.Rmd
Update MS

@@
 	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
-	geom_line(aes(color = type), size = .7) +
+	geom_line(aes(color = type)) +
 	scale_color_manual(values = compartment_cols, guide = "none") +
@@
 		data = fern_tree_sampling,
-		aes(shape = source),
-		size = 2.5,
+		aes(shape = source)
 	) +
 	labs(shape = "Study", y = "No. species", x = "Year") +
-	theme(legend.position = "bottom") +
+	theme(legend.position = "bottom",
+				plot.margin = margin(r = 1, t = 0.5, unit = "in")) +
 	coord_cartesian(clip = "off")
@@
 
-```{r genbank-print, fig.width = 3.3, fig.height = 5, eval = params$doc_type == "pdf"}
+```{r genbank-print, fig.width = 7, fig.height = 5, eval = params$doc_type == "pdf"}
 # Print plot if rendering to PDF

### 2022-02-04 ms/manuscript.Rmd
Update MS

@@
 	tribble(
-		~source, ~year, ~n_species,
-		"Hasebe1995", 1995, 107,
-		"Schuettpelz2007", 2007, 400,
-		"Lehtonen2011", 2011, 2957,
-		"Testo2016a", 2016, 3973,
-		"this_study", 2021, n_sanger_ferns
+		~source, ~year, ~n_species, ~label,
+		"Hasebe1995", 1995, 107, "Hasebe et al. 1995",
+		"Schuettpelz2007", 2007, 400, "Schuettpelz et al. 2007",
+		"Lehtonen2011", 2011, 2957, "Lehtonen 2011",
+		"Testo2016a", 2016, 3973, "Testo and Sundue 2016",
+		"this_study", 2021, n_sanger_ferns, "This study",
 	) %>%
@@
 	) +
+	scale_shape(
+    limits = fern_tree_sampling$source, 
+    labels = fern_tree_sampling$label
+	) +
 	labs(shape = "Study", y = "No. species", x = "Year") +
 	theme(legend.position = "bottom",
-				plot.margin = margin(r = 1, t = 0.5, unit = "in")) +
+				plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")) +
+	guides(shape = guide_legend(title.position = "top")) +
 	coord_cartesian(clip = "off")

### 2022-02-07 ms/manuscript.Rmd
Update MS
As per WI comments

@@
 number <- function(...) {scales::number(big.mark = ",", ...)}
-
-# Specify FTOL version
-ftol_ver <- "v0.0.2"
 ```
@@
 
-**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^4^**
+**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^1,4,5,6,7,8^**
 
@@
 ^4^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^5^Department of Computational Biology and Medical Sciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^6^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
+
+^7^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
+
+^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
 
@@
 Ferns (ca.
-12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for ca.
-1/3 of all fern species deposited in GenBank.
+12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for > 1/3 of all fern species deposited in GenBank.
 Global trees including all ferns have been published periodically, but as molecular systematics research continues at a rapid pace, such trees quickly become outdated.
-While some large projects exist to generate trees for all plants or all life, such "one-size-fits-all" approaches are not expected to generate optimal results for specific groups because they must work across a very wide range of taxa.
+
 Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from sequence data on GenBank.
 Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with Sanger-sequenced loci (many taxa, few loci) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
-We provide multiple versions of the phylogeny including a time-calibrated tree and a synthetic tree with species lacking molecular data added by taxonomy.
 We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
-The current FTOL (v0.0.1) includes `r comma(n_sanger_ferns)` species, an increase of > 20% relative to the most recent global fern phylogeny.
-FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have easy access to the most recent, comprehensively sampled fern phylogeny.
-FTOL will be useful for anybody studying this important group of plants at scales from smaller (e.g., genus or lower) clades to the entire tree; we anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale.
-Finally, we envision FTOL will be integrated into the next PPG, PPGII, to produce a "living" taxonomic system with direct linkage between a community-driven fern taxonomy and the most recent hypothesis of phylogeny.
+
+The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of > 20% relative to the previous largest fern phylogeny.
+FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have immediate access to the most recent, comprehensively sampled fern phylogeny.
+FTOL will be useful for anybody studying this important group of plants at scales from smaller clades to the entire tree.
+We anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
 
@@
 Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a\; `r figure("genbank")`].
-We believe there is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
+There is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
 This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
 
-We recognize that multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
-While these may be appropriate in some cases, we contend that they cannot produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth.
+Multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
+While these may be appropriate in some cases, they may not produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth. 
 By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.
@@
 
-## Locus sampling
+## Locus selection
 
@@
 
-## Sequence extraction
+## Dataset construction
 
@@
 We manually inspected any fuzzily matched or non-matching names.
-This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (e.g., changes in synonymy).
+This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
 Therefore, we updated and edited the World Ferns database using the dwctaxon R package, which is designed to work with taxonomic data in the Darwin Core format.
@@
 
-## Automated removal of rogue sequences
+## Removal of rogue sequences using BLAST
 
@@
 While this method cannot account for rogues at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
-
-## Sequence selection and concatenation
+We inspected all accessions flagged as rogues before excluding them.
+In the case that that mismatch of family was due to incorrect taxonomy (`r eg`, species name on GenBank matches the correct family but name used in the taxonomic database does not), we updated the taxonomic database accordingly.
+
+## Representative sequence selection and concatenation
 
@@
 
-We generated a backbone phylogeny using maximum likelihood (ML) analysis of the plastome dataset in IQ-TREE [@Nguyen2015] with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
-We applied the GTR+I+G model to a concatenated dataset with no partitioning.
-
+We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
+ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
+To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
+The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
+Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013]. 
 ### Initial Sanger phylogeny
 
-Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE in with the "-fast" option and same model settings as the backbone phylogeny.
+Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the same model as selected for the backbone phylogeny.
 We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
 
-### Manual inspection of initial Sanger phylogeny
-
-We manually inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
+### Removal of rogue sequences based on initial Sanger phylogeny
+
+We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
 We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
-We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; @Schwery2016].
-However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny as an important guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
-In a very small number of cases we assigned a different higher level taxonomy for recently described names if no existing combinations were available that were compatible with our taxonomic system.
-These are indicated in our supplemental data files as "comb. ined.".
-
-Based on the results of the manual inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses to this step.
+We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; expected monophyly follows @PteridophytePhylogenyGroupI2016].
+However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
+While we consider our phylogeny a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
+
+Based on the results of this inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses to this step.
 When re-running the analysis, it is possible that there were newly added GenBank accessions in place of the excluded rogues that themselves are rogues.
-Therefore, we repeated this process until the expected monophyly of all higher-level (`r eg` genus rank and higher) taxa that have previously been shown to be monophyletic was confirmed.
+Therefore, we repeated this process until the monophyly of all higher-level (`r eg` genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
 
@@
 
-After manual inspection, we generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the Sanger dataset under otherwise identical settings as those used for the backbone phylogeny.
-
-We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
+We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure as used for the backbone phylogeny.
+Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates
+
+We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods or models as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
 
@@
 We applied a set of FIXME calibration points as minimum ages.
-Our calibration points are all based on fossils (not secondary calibration points) following a careful literature review.
+Our calibration points are all based on fossils following a careful literature review.
 We used the species identification of each fossil to assign lineage only after consulting the original publication; previous studies [`r eg`, @Testo2016a] have used genus to assign lineage, but this can be problematic as genus concepts may vary between the original publication and current taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
@@
 
-Taxon sampling for the current version of FTOL (`r ftol_ver`) includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
+Taxon sampling for the current version of FTOL includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
@@
 
-## FTOL phylogeny
+## Nodes of contention in fern phylogeny
 
@@
 
+## Plastid vs. nuclear fern trees
+
+Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions (`r figure("genbank")`).
+Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
+However, the plastid tree does not necessarily agree with trees inferred from other data sources.
+Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing, and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
+Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
+Therefore, we have no expectation that there is a single fern tree of life; rather, multiple trees exist that each capture an element of biological reality, and FTOL is merely one such possible tree.
+
+Recent studies using transcriptomics are clarifying the backbone of the fern phylogeny using many (25--2400) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
+The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFLAG) project, which seeks to generate genomic data (ca.
+300 single-to-low copy nuclear gene regions) for all flagellate plants (bryophytes, lycophytes, ferns, and gymnosperms) [@Breinholt2021].
+GoFLAG data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFLAG markers are to be expected in the near future.
+The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data (`r figure("genbank")`).
+Furthermore, many subclades of ferns are under active investigation using plastid markers using both Sanger and next-gen sequencing, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
+We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for several more years.
+
 ## Accesibility and usage of FTOL
@@
 
+We want to make it clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
+FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphologic, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
+
 One feature of FTOL that sets it apart from the vast majority of other fern phylogenetic studies is its iterative nature.
-Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript (`r ftol_ver`) is not meant to be the last.
+Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript is not meant to be the last.
 Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree regularly (`ie`, semi-annually).
@@
 This way, FTOL will continually improve and keep pace with the currently available data and taxonomic hypotheses of ferns.
-
-We want to make it clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
-FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphologic, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
-
-## Plastid vs. nuclear fern trees
-
-Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions `r figure("genbank")`.
-Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
-However, the plastid tree does not necessarily agree with trees inferred from other data sources.
-Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing, and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
-Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
-Therefore, we have no expectation that there is a single fern tree of life; rather, multiple trees exist that each capture an element of biological reality, and FTOL is merely one such possible tree.
-
-Recent studies using transcriptomics are clarifying the backbone of the fern phylogeny using many (25--2400) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
-The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFLAG) project, which seeks to generate genomic data (ca.
-300 single-to-low copy nuclear gene regions) for all flagellate plants (bryophytes, lycophytes, ferns, and gymnosperms) [@Breinholt2021].
-GoFLAG data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFLAG markers are to be expected in the near future.
-The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data `r figure("genbank")`.
-Furthermore, many subclades of ferns are under active investigation using plastid markers using both Sanger and next-gen sequencing, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
-We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for several more years.
-Future efforts should focus on integrating across plastid and nuclear DNA sequences to gain better insight into the complex evolutionary history of ferns.
 
@@
 **`r figure("ftol")`**.
-Fern Tree of Life (FTOL) `r ftol_ver`.
+Fern Tree of Life (FTOL).
 Inset plot shows coverage by major clade (order or suborder).

### 2022-02-09 ms/manuscript.Rmd
Update MS
use initial time tree

@@
 		plastome_alignment,
-		plastome_tree,
 		plastome_metadata_renamed,
@@
 		pteridocat,
-		sanger_alignment_tbl),
-	store = here::here("ftol_cache")
+		sanger_alignment_tbl,
+		sanger_tree_fast,
+		plastome_tree,
+		plastid_tree_dated),
+	# FIXME: change to ftol_cache when ready
+	store = here::here("working/_targets")
 )
@@
 )
-
-# FIXME: read in real ultrametric tree when ready
-sanger_tree <- ape::read.tree(here::here("working/sanger_alignment.phy.tre"))
 ```
@@
 Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
+
 ### Initial Sanger phylogeny
 
-Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the same model as selected for the backbone phylogeny.
+Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
 We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
@@
 We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure as used for the backbone phylogeny.
-Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates
+Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates.
 
@@
 
+plastome_tree_rooted <- phytools::reroot(
+	plastome_tree,
+	getMRCA(
+		plastome_tree,
+		c("Physcomitrium_patens", "Marchantia_polymorpha", "Anthoceros_angustus")
+	)) %>%
+	ladderize()
+
 # Make tibble of backbone tips with higher-level taxonomy
 backbone_taxonomy <-
-	tibble(species = plastome_tree$tip.label) %>%
+	tibble(
+		species = get_tips_in_ape_plot_order(plastome_tree_rooted)
+	) %>%
 	# Add higher taxonomy
@@
 
+# Manually specify axis buffers since we will annotate the panel border manually
+year_buffer <- 
+	max(gb_species_by_year$year) %>%
+	subtract(min(gb_species_by_year$year)) %>%
+	multiply_by(0.05)
+sp_buffer <- max(gb_species_by_year$n_species) %>%
+	subtract(min(gb_species_by_year$n_species)) %>%
+	multiply_by(0.05)
+
 genbank_plot <-
@@
 	scale_color_manual(values = compartment_cols, guide = "none") +
+	# Annotate panel border manually, since the automatic border gets
+	# printed on top of label annotations with theme_bw()
+	annotate(
+		"rect", 
+		xmin = min(gb_species_by_year$year) - year_buffer,
+		xmax = max(gb_species_by_year$year) + year_buffer,
+		ymin = min(gb_species_by_year$n_species) - sp_buffer,
+		ymax = max(gb_species_by_year$n_species) + sp_buffer,
+		fill = "transparent",
+		color = "grey20") +
 	geom_label_repel(
@@
 		xlim = c(2020, Inf),
-    # Do not repel from top or bottom edges.
 		ylim = c(-Inf, Inf),
@@
 	) +
+	scale_x_continuous(expand = expansion(0, 0)) + 
+	scale_y_continuous(expand = expansion(0, 0)) + 
 	labs(shape = "Study", y = "No. species", x = "Year") +
-	theme(legend.position = "bottom",
-				plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")) +
-	guides(shape = guide_legend(title.position = "top")) +
-	coord_cartesian(clip = "off")
+	coord_cartesian(clip = "off") +
+	guides(
+		shape = guide_legend(
+			title.position = "top",
+			nrow = 2)
+	) +
+	theme_bw() +
+	theme(
+		panel.border = element_blank(),
+		legend.position = "bottom",
+		plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")
+	)
 ```
@@
 	backbone_taxonomy %>%
-	# Slice to one exemplar species per family
+	# Slice to one exemplar species per major clade
 	group_by(plot_group) %>%
@@
 	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
-	arrange(phy_order)
-
-# Get names of exemplar tips in ladderized phylo order
-# Trim tree to exemplar species, rename by family
-phy <- ape::keep.tip(plastome_tree, plot_group_exemp_taxa$species)
-phy$tip.label <- plot_group_exemp_taxa$plot_group
-
-# Make tibble of plot groups in phylogenetic order
-plot_group_phy_order <-
-	phy %>%
-	# Reroot, ladderize, drop outgroup
-	root.phylo("og") %>%
-	ladderize() %>%
-	drop.tip("og") %>%
-	jntools::get_tips_in_ape_plot_order() %>%
-	tibble(
-		name = .,
-		phy_order = seq_along(.)
-	)
-
-rm(phy)
+	arrange(phy_order) %>%
+	# Drop outgroup
+	filter(family != "og")
 
@@
 	extract(green_army_use_cols) %>%
-	extract(seq_len(nrow(plot_group_phy_order))) %>%
-	set_names(plot_group_phy_order$name)
+	extract(seq_len(nrow(plot_group_exemp_taxa))) %>%
+	set_names(plot_group_exemp_taxa$plot_group)
 
@@
 	filter(rank == "plot_group") %>%
-	left_join(plot_group_phy_order, by = "name") %>%
+	left_join(plot_group_exemp_taxa, by = c(name = "plot_group")) %>%
 	arrange(phy_order) %>%
@@
 # Tree plot ----
-phy <- sanger_tree %>%
-	root("Anthoceros_angustus") %>%
-	ladderize()
+phy <- plastid_tree_dated
 
@@
 	layout = "circular", right = FALSE, size = 0.1,
-	# FIXME: change to phylogram when ultrametric tree is done
-	branch.length = "none"
 ) +

### 2022-02-09 R/ms_functions.R

@@ -182,3 +182,19 @@ add_plot_group <- function(data) {
 		)
 }
 
+#' Get tips of a phylogenetic tree in their plotted order
+#'
+#' After re-rooting a tree, the order of tips when the tree
+#' is plotted no longer match the order of $tip.label. Use
+#' this function to get tips in the order they are plotted.
+#' @param tree List of class "phylo"
+#' @return Character vector
+get_tips_in_ape_plot_order <- function(tree) {
+	assertthat::assert_that(inherits(tree, "phylo"))
+	# First filter out internal nodes
+	# from the the second column of the edge matrix
+	is_tip <- tree$edge[,2] <= length(tree$tip.label)
+	ordered_tips <- tree$edge[is_tip, 2]
+	# Use this vector to extract the tips in the right order
+	tree$tip.label[ordered_tips]
+}

### 2022-02-10 ms/manuscript.Rmd
Update MS

@@
 # Targets from FTOL workflow
+
+# Specify path to FTOL cache
+# FIXME: change to ftol_cache when ready
+ftol_cache <- here::here("working/_targets")
+
 tar_load(
@@
 		sanger_tree_fast,
+		sanger_sampling,
+		monophy_by_clade,
 		plastome_tree,
-		plastid_tree_dated),
-	# FIXME: change to ftol_cache when ready
-	store = here::here("working/_targets")
+		plastid_tree_dated,
+		fossil_calibration_tips,
+		treepl_cv_results),
+	store = ftol_cache
 )
@@
 		gb_species_by_year,
-		ref_files
+		ref_files,
+		other_dates
 	),
@@
 
+```{r dating-methods-stats}
+# Count number of calibration points
+n_calibrations_ferns <-
+	fossil_calibration_tips %>%
+	assert(is_uniq, node_calibrated) %>%
+	filter(!affinities %in% c("Tracheophytes", "Euphyllophytes")) %>%
+	pull(node_calibrated) %>%
+	n_distinct() %>%
+	number()
+
+n_calibrations_total <-
+	fossil_calibration_tips %>%
+	assert(is_uniq, node_calibrated) %>%
+	pull(node_calibrated) %>%
+	n_distinct() %>%
+	number()
+
+# Format values used for treePL smoothing as tibble
+cv_smooth_test <-	
+	tibble(cv_result = treepl_cv_results) %>%
+	mutate(smooth = str_match(cv_result, "\\((.*)\\)") %>% 
+				 	magrittr::extract(, 2) %>%
+				 	parse_number()) %>%
+	mutate(chisq = str_match(cv_result, "\\) (.*)$") %>% 
+				 	magrittr::extract(, 2) %>%
+				 	parse_number()) %>%
+	arrange(smooth)
+
+# Extract start, end, and best smooth values
+smooth_start <- cv_smooth_test %>%
+	slice_min(n = 1, order_by = smooth) %>%
+	pull(smooth) %>%
+	number(., accuracy = .)
+
+smooth_end <- cv_smooth_test %>%
+	slice_max(n = 1, order_by = smooth) %>%
+	pull(smooth) %>%
+	number(., accuracy = .)
+
+smooth_best <-
+	cv_smooth_test %>%
+	slice_min(n = 1, order_by = chisq) %>%
+	pull(smooth) %>%
+	number(., accuracy = .)
+```
+
 We dated the tree using treePL [@Smith2012].
-We applied a set of FIXME calibration points as minimum ages.
-Our calibration points are all based on fossils following a careful literature review.
-We used the species identification of each fossil to assign lineage only after consulting the original publication; previous studies [`r eg`, @Testo2016a] have used genus to assign lineage, but this can be problematic as genus concepts may vary between the original publication and current taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
-We tested rate smoothing parameters ranging from 0.000001 to 1000 (each varies by one order of magnitude); a value of FIXME was selected with the smallest chi-squared value for final analysis [@Smith2012].
+We applied a set of `r n_calibrations_total` calibration points as minimum ages, including `r n_calibrations_ferns` calibration points within ferns.
+Our calibration points are all based on fossils.
+We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil name), since taxon concepts may vary between the original publication and currently applied taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
+We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); a value of `r smooth_best` was selected with the smallest chi-squared value for final analysis [@Smith2012].
 
@@
 	arrange(coverage) %>%
-	transmute(clade = name, coverage = percent(coverage)) %>%
+	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
 	split(.$clade)
@@
 	mutate(
-		coverage_p = percent(n / n_accepted),
+		coverage_p = percent(n / n_accepted, accuracy = 0.1),
 		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
@@
 		coverage_p = percent(n / n_accepted),
-		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
+		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 0.1)}")) %>% #nolint
 	# Split for printing
@@
 
-Within Aspleniineae, we recover two large clades.
-One clade consists of Cyctopteridaceae, Rhacidosoraceae, Diplaziopsidaceae, and a clade comprised of Aspleniaceae and Hemidictyaceae.
-The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and a clade comprised of Athyriaceae and Desmophlebiaceae.
+Within Aspleniineae, Cystopteridaceae is sister to the remaining species, which are split into two clades. 
+One clade consists of Rhacidosoraceae, Diplaziopsidaceae, Desmophlebiaceae, Hemidictyaceae and Aspleniaceae.
+The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and Athyriaceae.
 
 ## Divergence times
+
+```{r divergence-times-stats}
+# Count number of tips in clades by taxonomy for double checking
+n_tips_in_clades <-
+	sanger_sampling %>%
+	filter(outgroup == FALSE) %>%
+	select(species, family, order) %>%
+	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
+	group_by(taxon) %>%
+	summarize(number_tips_by_tax = n())
+
+tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
+
+# Define function for calculating age based on two genera
+get_age_from_genus_pair <- function(tip_pair) {
+	sanger_sampling %>%
+	filter(genus %in% tip_pair) %>%
+	group_by(genus) %>%
+	slice(1) %>%
+	pull(species) %>%
+  ape::getMRCA(plastid_tree_dated, .) %>%
+	phytools::nodeheight(plastid_tree_dated, .) %>%
+	magrittr::subtract(tree_height, .) %>%
+	number(accuracy = 0.1)
+}
+
+# Calculate age of various clades
+fern_age <- get_age_from_genus_pair(c("Equisetum", "Polypodium"))
+lepto_age <- get_age_from_genus_pair(c("Osmunda", "Polypodium"))
+polypodiales_age <- get_age_from_genus_pair(c("Lindsaea", "Polypodium"))
+eupoly_i_age <- get_age_from_genus_pair(c("Hypodematium", "Polypodium"))
+eupoly_ii_age <- get_age_from_genus_pair(c("Cystopteris", "Asplenium"))
+
+# Get ages for large clades
+clade_ages <-
+	sanger_sampling %>%
+	select(family, order) %>%
+	unique() %>%
+	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
+	filter(!is.na(taxon)) %>%
+	unique() %>%
+	left_join(
+		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
+		by = "taxon") %>%
+	mutate(across(c(mrca, number_tips), parse_number)) %>%
+	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
+	filter(monophyly == "Yes") %>%
+	mutate(
+		# Height is distance above root
+		height = map_dbl(mrca, ~phytools::nodeheight(
+			tree = plastid_tree_dated, node = .))
+	) %>%
+	mutate(
+		# Age is the total length of the tree minus height
+		age = tree_height - height
+	) %>%
+	# Double check that n_tips is as expected
+	# (so the correct node is being used)
+	left_join(n_tips_in_clades, by = "taxon") %>%
+	verify(all(number_tips == number_tips_by_tax)) %>%
+	select(-number_tips_by_tax)
+```
+
+We estimate the crown age of ferns to be `r fern_age` ma.
+Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
+Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
+Ages of fern families differed from previous studies, with some families estimated as having older crown ages while some had younger (`r figure("divtimes")`).
 
@@
 Rather, conclusive resolution apparently still awaits additional sampling and perhaps innovation in phylogenetic methods.
+
+## Revisiting fossil calibration points in ferns
+
+Although we recovered slightly older ages for ferns and other large clades relative to the most recent globally sampled fern phylogeny [@Testo2016a], our estimated ages for families do not show a clear pattern, with some older and some younger than other studies (`r figure("divtimes")`).
+This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
+Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
+However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
+Moreover, our set of calibration points more than doubles the number used by previous studies.
+Therefore, it is perhaps not surprising that we obtained different estimates for the age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).
+
+<<FIXME: add discussion of specific groups once we have final ages>>
 
@@
 backbone_taxonomy <-
-	tibble(
-		species = get_tips_in_ape_plot_order(plastome_tree_rooted)
-	) %>%
+	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
 	# Add higher taxonomy
@@
 	coord_cartesian(clip = "off") +
-	guides(
-		shape = guide_legend(
-			title.position = "top",
-			nrow = 2)
-	) +
+	guides(shape = guide_legend(nrow = 2)) +
 	theme_bw() +
@@
 		legend.position = "bottom",
+		legend.title = element_blank(),
 		plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")
@@
 	backbone_taxonomy %>%
+	# Drop outgroups %>%
+	filter(family != "og") %>%
 	# Slice to one exemplar species per family
@@
 # Trim tree to exemplar species, rename by family
-phy <- keep.tip(plastome_tree, family_exemp$species)
-phy$tip.label <- family_exemp$family
+phy <- keep.tip(plastome_tree_rooted, family_exemp$species)
+new_tip_label <-
+	tibble(species = phy$tip.label) %>%
+	left_join(family_exemp, by = "species") %>%
+	pull(family)
+phy$tip.label <- new_tip_label
 # Convert BS support values to numeric
 phy$node.label <- parse_number(phy$node.label)
-# Reroot, ladderize, drop outgroup
-phy <-
-	phy %>%
-	root.phylo("og") %>%
-	ladderize() %>%
-	drop.tip("og") %>%
 # Scale to total height of 100 for easy plotting
-	rescale_tree(100)
+phy <- rescale_tree(phy, 100)
 
@@
 Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.
+
+```{r div-times-make}
+# Convert family ages to long format
+family_ages_median <-
+clade_ages %>%
+	filter(level == "family") %>%
+	select(family = taxon, current_best = age) %>%
+	left_join(other_dates, by = "family") %>%
+	mutate(family = fct_reorder(family, current_best)) %>%
+	pivot_longer(values_to = "age", -family) %>%
+	separate(name, into = c("author", "age_type"), sep = "_") %>%
+	assert(not_na, author, age_type) %>%
+	filter(age_type %in% c("best", "median"))
+
+# Convert family errors to almost long format
+family_ages_errors <-
+	clade_ages %>%
+	filter(level == "family") %>%
+	select(family = taxon, current_best = age) %>%
+	left_join(other_dates, by = "family") %>%
+	mutate(family = fct_reorder(family, current_best)) %>%
+	select(family, matches("ts|rothfels")) %>%
+	select(-matches("median"))
+
+family_ages_errors <-
+	bind_rows(
+		transmute(family_ages_errors, family,
+					 high = ts_high, low = ts_low, author = "ts"),
+		transmute(family_ages_errors, family,
+					 high = rothfels_high, low = rothfels_low, author = "rothfels")
+	)
+
+# Define shapes and colors for points
+ages_shapes <- c(
+	current = 16, # circle
+	rothfels = 17, # triangle
+	schuettpelz = 15, # square
+	ts = 18 # diamond
+)
+
+ages_colors <- c(
+	current = palette_okabe_ito(order = "vermillion"),
+	rothfels = palette_okabe_ito(order = "bluishgreen"),
+	schuettpelz = palette_okabe_ito(order = "blue"),
+	ts = palette_okabe_ito(order = "orange")
+)
+
+# Format names of papers for printing
+ages_sampling <-
+	tribble(
+		~source, ~year, ~label,
+		"schuettpelz", 2009, "Schuettpelz & Pryer 2009",
+		"rothfels", 2015, "Rothfels et al. 2015",
+		"ts", 2016, "Testo and Sundue 2016",
+		"current", 2021, "This study",
+	) %>%
+	mutate(source = fct_reorder(source, year))
+
+div_times_plot <-
+ggplot() +
+	geom_errorbarh(
+		data = family_ages_errors,
+		aes(xmax = high, xmin = low, color = author, y = family),
+		position = position_dodge(width = 0.5)
+	) +
+	geom_point(
+		data = filter(family_ages_median, age_type %in% c("best", "median")),
+		aes(x = age, y = family, color = author, shape = author),
+		position = position_dodge(width = 0.5)) +
+	scale_shape_manual(
+		values = ages_shapes,
+		limits = ages_sampling$source, 
+		labels = ages_sampling$label) +
+	scale_color_manual(
+		values = ages_colors,
+		limits = ages_sampling$source, 
+		labels = ages_sampling$label) +
+	labs(x = "Age (millions of years)") +
+	guides(shape = guide_legend(nrow = 2)) +
+	theme_bw() +
+	theme(
+		axis.title.y = element_blank(),
+		legend.title = element_blank(),
+		legend.position = "bottom"
+	)
+```
+
+```{r div-times-save, eval = params$doc_type == "doc"}
+# Save plot if rendering to doc
+ggsave(
+	result_file(figure("divtimes"), "pdf"),
+	div_times_plot, width = 7, height = 8, units = "in")
+```
+
+```{r div-times-print, fig.width = 7, fig.height = 9, eval = params$doc_type == "pdf"}
+# Print plot if rendering to PDF
+div_times_plot
+```
+
+**`r figure("divtimes")`**. Crown age of fern families (mya) estimated by selected studies.
+For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
+For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].

### 2022-02-10 R/functions.R

@@ -210,3 +210,33 @@ count_ncbi_species_by_year <- function(gb_taxa, ncbi_names, year_range) {
 	map_df(year_range, ~sum_species(gb_species, .))
 }
 
+# Etc ----
+
+#' Parse divergene dates in the Testo and Sundue 2016 SI file
+#' 
+#' Testo WL, Sundue MA (2016) A 4000-species dataset provides new insight into 
+#' the evolution of ferns. Molecular Phylogenetics and Evolution 105:200–211. 
+#' https://doi.org/10.1016/j.ympev.2016.09.003
+#'
+#' @param testo_sundue_2016_si_path Path to Testo and Sundue 2016 SI file
+#'
+#' @return Tibble
+#' 
+parse_ts_dates <- function(testo_sundue_2016_si_path) {
+	readxl::read_excel(
+		testo_sundue_2016_si_path,
+		# Divergence dates are in the 5th sheet
+		sheet = 5,
+		skip = 1
+	) %>%
+		janitor::clean_names() %>%
+		rename(
+			ts_median = median_age_2,
+			ts_low = x95_percent_hpd_low_3,
+			ts_high = x95_percent_hpd_high_4,
+			rothfels_median = median_age_5,
+			rothfels_low = x95_percent_hpd_low_6,
+			rothfels_high = x95_percent_hpd_high_7,
+			schuettpelz_best = best_age
+		)
+}

### 2022-02-10 _targets.R

@@ -30,6 +30,12 @@ tar_plan(
 	# by genomic compartment per year
 	gb_species_by_year = count_ncbi_species_by_year(
 		gb_taxa, ncbi_names, year_range = 1990:2021),
+	# Read in estimated clade ages from Testo and Sundue 2016 SI
+	tar_files_input(
+		testo_sundue_2016_si_path,
+		"_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
+	),
+	other_dates = parse_ts_dates(testo_sundue_2016_si_path),
 	# Render manuscript ----
 	# track files used for rendering MS
 	tar_file(ref_files, list.files("ms", "references", full.names = TRUE)),

### 2022-02-13 ms/si.Rmd
Add SI

---
title: "Supplemental Information"
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
editor_options: 
  chunk_output_type: console
params:
  doc_type: doc
---

---
bibliography: 
- `r here::here("ms/references.yaml")`
- `r here::here("ms/references_other.yaml")`
- `r here::here("ms/references_other.bib")`
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')

# Load packages
library(conflicted)
library(rmarkdown)
library(bookdown)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(ggtree)
library(glue)
library(Polychrome)
library(patchwork)
library(magrittr)
library(ggrepel)
library(ggokabeito)
library(ggforce)

# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("extract", "magrittr")

# Load functions
source(here::here("R/ms_functions.R"))

# Specify percentage format
percent <- function(...) {scales::percent(...)}
number <- function(...) {scales::number(big.mark = ",", ...)}
```

```{r load-targets}
# Targets from FTOL workflow

# Specify path to FTOL cache
# FIXME: change to ftol_cache when ready
ftol_cache <- here::here("working/_targets")

tar_load(
	c(
		sanger_sampling,
		monophy_by_clade,
		plastid_tree_dated),
	store = ftol_cache
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files,
		other_dates
	),
	store = here::here("_targets")
)
```

```{r age-biplot}
# Count number of tips in clades by taxonomy for double checking
n_tips_in_clades <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(species, family, order) %>%
	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
	group_by(taxon) %>%
	summarize(number_tips_by_tax = n())

# Get ages for large clades
clade_ages <-
	sanger_sampling %>%
	select(family, order) %>%
	unique() %>%
	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
	filter(!is.na(taxon)) %>%
	unique() %>%
	left_join(
		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
		by = "taxon") %>%
	mutate(across(c(mrca, number_tips), parse_number)) %>%
	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
	filter(monophyly == "Yes") %>%
	mutate(
		# Height is distance above root
		height = map_dbl(mrca, ~phytools::nodeheight(
			tree = plastid_tree_dated, node = .))
	) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = max(phytools::nodeHeights(plastid_tree_dated)) - height
	) %>%
	# Double check that n_tips is as expected
	# (so the correct node is being used)
	left_join(n_tips_in_clades, by = "taxon") %>%
	verify(all(number_tips == number_tips_by_tax)) %>%
	select(-number_tips_by_tax)

clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	select(matches("best|median")) %>%
	ggplot(aes(x = .panel_x, y = .panel_y)) +
	geom_point() +
	geom_autodensity(na.rm = TRUE) +
	facet_matrix(vars(everything()), layer.diag = 2, grid.y.diag = FALSE)
```

Figure 1: Biplot comparing crown ages of fern families (millions of years) across different studies.

### 2022-02-13 ms/manuscript.Rmd
Calculate family stem ages

@@
 		sanger_tree_fast,
-		sanger_sampling,
 		monophy_by_clade,
@@
 # Make tibble with one row per species in Sanger sampling
-sanger_sampling <- tibble(
-	species = rownames(sanger_alignment)
-) %>%
-	# Add higher-level taxonomy
-	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
-	left_join(
-		select(ppgi_taxonomy, genus, family, suborder, order), by = "genus"
-	) %>% 
-	# Add plot_group
-	add_plot_group() %>%
-	# Add outgroup status
-	left_join(og_species, by = "species") %>%
-	mutate(outgroup = replace_na(outgroup, FALSE)) %>%
-	verify(nrow(.) == nrow(sanger_alignment)) %>%
-	verify(sum(outgroup) == nrow(og_species))
+sanger_sampling <- make_sanger_sampling_tbl(
+	plastome_metadata_renamed,
+	sanger_tree = plastid_tree_dated,
+	ppgi_taxonomy = ppgi_taxonomy)
 
@@
 ```{r divergence-times-stats}
-# Count number of tips in clades by taxonomy for double checking
-n_tips_in_clades <-
-	sanger_sampling %>%
-	filter(outgroup == FALSE) %>%
-	select(species, family, order) %>%
-	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
-	group_by(taxon) %>%
-	summarize(number_tips_by_tax = n())
-
+# Get overall tree height
 tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
@@
 
-# Get ages for large clades
-clade_ages <-
-	sanger_sampling %>%
-	select(family, order) %>%
-	unique() %>%
-	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
-	filter(!is.na(taxon)) %>%
-	unique() %>%
-	left_join(
-		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
-		by = "taxon") %>%
-	mutate(across(c(mrca, number_tips), parse_number)) %>%
-	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
+# Get ages for families
+sanger_sampling_family <- select(sanger_sampling, species, family)
+
+# Only use monophyletic families
+mono_test_family <- assess_monophy(
+	taxon_sampling = sanger_sampling_family,
+	tree = plastid_tree_dated,
+	tax_levels = "family"
+) %>%
+	get_result_monophy(1)
+
+family_ages <-
+mono_test_family %>%
 	filter(monophyly == "Yes") %>%
 	mutate(
+		mrca = parse_number(mrca),
+		stem_node = map_dbl(
+			mrca,
+			~phangorn::Ancestors(plastid_tree_dated, ., type = "parent"))
+	) %>%
+	select(family = taxon, stem_node) %>%
+mutate(
 	# Height is distance above root
-		height = map_dbl(mrca, ~phytools::nodeheight(
+	height = map_dbl(stem_node, ~phytools::nodeheight(
 		tree = plastid_tree_dated, node = .))
@@
 		age = tree_height - height
-	) %>%
-	# Double check that n_tips is as expected
-	# (so the correct node is being used)
-	left_join(n_tips_in_clades, by = "taxon") %>%
-	verify(all(number_tips == number_tips_by_tax)) %>%
-	select(-number_tips_by_tax)
+	)
 ```
@@
 family_ages_median <-
-clade_ages %>%
-	filter(level == "family") %>%
-	select(family = taxon, current_best = age) %>%
+family_ages %>%
+	select(family, current_best = age) %>%
 	left_join(other_dates, by = "family") %>%
@@
 family_ages_errors <-
-	clade_ages %>%
-	filter(level == "family") %>%
-	select(family = taxon, current_best = age) %>%
+	family_ages %>%
+	select(family, current_best = age) %>%
 	left_join(other_dates, by = "family") %>%

### 2022-02-13 R/functions.R

@@ -240,3 +240,136 @@ parse_ts_dates <- function(testo_sundue_2016_si_path) {
 			schuettpelz_best = best_age
 		)
 }
+
+# Monophyly ----
+
+#' Add "major clade" to Sanger sampling table
+#'
+#' "major clade" includes pteridophyte suborder or order, and combines
+#' Ophioglossales + Psilotales into one group
+#'
+#' @param data Tibble including columns "suborder" and "order".
+#'
+#' @return Tibble withh column "major_clade" added
+#'
+add_major_clade <- function(data) {
+	data %>%
+		mutate(
+			major_clade = coalesce(suborder, order),
+			major_clade = case_when(
+				major_clade %in% c("Ophioglossales", "Psilotales") ~ "Ophioglossales + Psilotales", #nolint
+				TRUE ~ major_clade
+			)
+		)
+}
+
+
+#' Make tibble summarizing sampling of Sanger dataset
+#'
+#' @param plastome_metadata_renamed Plastome data with final (resolved) species
+#' names.
+#' @param sanger_tree Sanger ML tree.
+#' @param ppgi_taxonomy PPGI taxonomy
+#'
+#' @return Tibble with columns "species",  "genus", "order", "suborder",
+#' "family"  "subfamily"  "major_clade" "outgroup"
+#'
+make_sanger_sampling_tbl <- function(
+	plastome_metadata_renamed,
+	sanger_tree, ppgi_taxonomy
+) {
+	
+	# check monophyly ----
+	# Make tibble of outgroup species
+	og_species <-
+		plastome_metadata_renamed %>%
+		select(species, outgroup) %>%
+		filter(outgroup == TRUE)
+	
+	# Make tibble with one row per species in Sanger sampling
+	tibble(species = sanger_tree$tip.label) %>%
+		# Add higher-level taxonomy
+		mutate(
+			genus = str_split(species, "_") %>% map_chr(1)
+		) %>%
+		left_join(
+			select(
+				ppgi_taxonomy, order, suborder, family, subfamily, genus), by = "genus"
+		) %>%
+		# Add major_clade
+		add_major_clade() %>%
+		# Add outgroup status
+		left_join(og_species, by = "species") %>%
+		mutate(outgroup = replace_na(outgroup, FALSE)) %>%
+		verify(sum(outgroup) == nrow(og_species)) %>%
+		# Check for match for tips with tree
[…]

### 2022-02-13 ms/manuscript.Rmd
Use "major_clade" instead of "plot_group"
Update text about divergence times

@@
 source(here::here("R/ms_functions.R"))
+source(here::here("R/functions.R"))
 
@@
 
-# Helper function to add plot_group:
+# Helper function to add major_clade:
 # a "rank" that may be at order or suborder level for plotting
-add_plot_group <- function(data) {
+add_major_clade <- function(data) {
 	data %>%
 		mutate(
-			plot_group = coalesce(suborder, order),
-			plot_group = case_when(
-				plot_group %in% c("Ophioglossales", "Psilotales") ~ 
+			major_clade = coalesce(suborder, order),
+			major_clade = case_when(
+				major_clade %in% c("Ophioglossales", "Psilotales") ~ 
 					"Ophioglossales + Psilotales",
-				TRUE ~ plot_group
+				TRUE ~ major_clade
 			)
@@
 	select(-outgroup) %>%
-	# add plot_group
-	add_plot_group() %>%
+	# add major_clade
+	add_major_clade() %>%
 	# drop suborder
@@
 	select(-class) %>%
-	# Add plot_group
-	add_plot_group() %>%
+	# Add major_clade
+	add_major_clade() %>%
 	# drop suborder
@@
 	coverage_data %>%
-	filter(rank == "plot_group") %>%
+	filter(rank == "major_clade") %>%
 	arrange(coverage) %>%
@@
 Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
-Ages of fern families differed from previous studies, with some families estimated as having older crown ages while some had younger (`r figure("divtimes")`).
+Estimated stem ages of fern families were generally older than previous studies. (`r figure("divtimes")`).
 
@@
 	bb_sampling %>%
-	filter(plot_group == "Gleicheniales") %>%
+	filter(major_clade == "Gleicheniales") %>%
 	pull(genus) %>%
@@
 
-Although we recovered slightly older ages for ferns and other large clades relative to the most recent globally sampled fern phylogeny [@Testo2016a], our estimated ages for families do not show a clear pattern, with some older and some younger than other studies (`r figure("divtimes")`).
+We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (`r figure("divtimes")`).
 This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
@@
 Moreover, our set of calibration points more than doubles the number used by previous studies.
-Therefore, it is perhaps not surprising that we obtained different estimates for the age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).
+It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).
 
@@
 			TRUE ~ family),
-		plot_group = case_when(
+		major_clade = case_when(
 			outgroup == TRUE ~ "og",
-			TRUE ~ plot_group)
+			TRUE ~ major_clade)
 	)
@@
 # Make exemplar set of species in backbone representing plot groups
-plot_group_exemp_taxa <-
+major_clade_exemp_taxa <-
 	backbone_taxonomy %>%
 	# Slice to one exemplar species per major clade
-	group_by(plot_group) %>%
+	group_by(major_clade) %>%
 	slice(1) %>%
@@
 	extract(green_army_use_cols) %>%
-	extract(seq_len(nrow(plot_group_exemp_taxa))) %>%
-	set_names(plot_group_exemp_taxa$plot_group)
+	extract(seq_len(nrow(major_clade_exemp_taxa))) %>%
+	set_names(major_clade_exemp_taxa$major_clade)
 
@@
 	coverage_data %>%
-	filter(rank == "plot_group") %>%
-	left_join(plot_group_exemp_taxa, by = c(name = "plot_group")) %>%
+	filter(rank == "major_clade") %>%
+	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
 	arrange(phy_order) %>%
@@
 	) %>%
-	pivot_longer(names_to = "plot_group", values_to = "taxa", cols = everything()) %>%
-	arrange(plot_group) %>%
-	group_by(plot_group) %>%
+	pivot_longer(names_to = "major_clade", values_to = "taxa", cols = everything()) %>%
+	arrange(major_clade) %>%
+	group_by(major_clade) %>%
 	summarize(taxa = list(taxa)) %>%
@@
 	transmute(
-		plot_group,
+		major_clade,
 		mrca,
@@
 			node = mrca,
-			color = plot_group,
+			color = major_clade,
 			label = dummy
@@
 	filter(outgroup == FALSE) %>%
-	group_by(plot_group) %>%
-	select(plot_group, family) %>% 
+	group_by(major_clade) %>%
+	select(major_clade, family) %>% 
 	filter(n_distinct(family) == 1) %>%
@@
 	left_join(phy_tip, by = c(family = "label")) %>%
-	select(plot_group, mrca = node)
+	select(major_clade, mrca = node)
 
@@
 	filter(outgroup == FALSE) %>%
-	group_by(plot_group) %>%
-	select(plot_group, family) %>% 
+	group_by(major_clade) %>%
+	select(major_clade, family) %>% 
 	filter(n_distinct(family) > 1) %>%
@@
 	transmute(
-		plot_group,
+		major_clade,
 		mrca
@@
 		clade_label = case_when(
-			plot_group == "Aspleniineae" ~ "Aspleniineae\n(eupolypods II)",
-			plot_group == "Polypodiineae" ~ "Polypodiineae\n(eupolypods I)",
-			TRUE ~ plot_group
+			major_clade == "Aspleniineae" ~ "Aspleniineae\n(eupolypods II)",
+			major_clade == "Polypodiineae" ~ "Polypodiineae\n(eupolypods I)",
+			TRUE ~ major_clade
 		)
@@
 			node = mrca,
-			fill = plot_group
+			fill = major_clade
 		),
@@
 
-**`r figure("divtimes")`**. Crown age of fern families (mya) estimated by selected studies.
+**`r figure("divtimes")`**. Stem age of fern families (mya) estimated by selected studies.
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].

### 2022-02-14 R/functions.R
Warn if entrez fails

@@ -36,9 +36,20 @@ fetch_metadata <- function(
 	)
 	
 	# Define internal function to download genbank data into tibble
+	# - make "safe" version of rentrez::entrez_summary to catch errors
+	safe_entrez_summary <- purrr::safely(rentrez::entrez_summary)
 	entrez_summary_gb <- function(id, col_select) {
+
 		# Download data
-		rentrez::entrez_summary(db = "nucleotide", id = id) %>%
+		res <- safe_entrez_summary(db = "nucleotide", id = id)
+		
+		# Early exit if error with entrez
+		if (!is.null(res$error)) {
+			warning("No esummary records found in file, returning empty tibble")
+			return(tibble(taxid = NA))
+		}
+		
+		res$result %>%
 			# Extract selected columns from result
 			purrr::map_dfr(magrittr::extract, col_select) %>%
 			# Make sure taxid column is character

### 2022-02-14 ms/manuscript.Rmd
Rearrange functions, calculate ages in targets plan

@@
 source(here::here("R/functions.R"))
-
-# Specify percentage format
-percent <- function(...) {scales::percent(...)}
-number <- function(...) {scales::number(big.mark = ",", ...)}
 ```
@@
 	c(
+		sanger_sampling,
 		gb_species_by_year,
 		ref_files,
-		other_dates
+		other_dates,
+		crown_ages,
+		family_stem_ages
 	),
@@
 	filter(outgroup == TRUE)
-
-# Make tibble with one row per species in Sanger sampling
-sanger_sampling <- make_sanger_sampling_tbl(
-	plastome_metadata_renamed,
-	sanger_tree = plastid_tree_dated,
-	ppgi_taxonomy = ppgi_taxonomy)
 
@@
 
-# Helper function to add major_clade:
-# a "rank" that may be at order or suborder level for plotting
-add_major_clade <- function(data) {
-	data %>%
-		mutate(
-			major_clade = coalesce(suborder, order),
-			major_clade = case_when(
-				major_clade %in% c("Ophioglossales", "Psilotales") ~ 
-					"Ophioglossales + Psilotales",
-				TRUE ~ major_clade
-			)
-		)
-}
-
 # Tally FTOL sampling by rank
@@
 	select(-outgroup) %>%
-	# add major_clade
-	add_major_clade() %>%
 	# drop suborder
@@
 ```{r topology-stats}
-#' Function to get support value in BS%
-#'
-#' @param phy an object of class "phylo".
-#' @param tips a vector of mode numeric or character specifying the tips
-#'
-#' @return Bootstrap support as % of that node
-get_bs <- function(phy, tips) {
-	node_select <- ape::getMRCA(phy, tips)
-	as_tibble(phy) %>%
-		filter(node == node_select) %>%
-		pull(label) %>%
-		parse_number() %>%
-		magrittr::multiply_by(0.01) %>%
-		percent()
-}
-
 ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
@@
 
-```{r divergence-times-stats}
-# Get overall tree height
-tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
-
-# Define function for calculating age based on two genera
-get_age_from_genus_pair <- function(tip_pair) {
-	sanger_sampling %>%
-	filter(genus %in% tip_pair) %>%
-	group_by(genus) %>%
-	slice(1) %>%
-	pull(species) %>%
-  ape::getMRCA(plastid_tree_dated, .) %>%
-	phytools::nodeheight(plastid_tree_dated, .) %>%
-	magrittr::subtract(tree_height, .) %>%
-	number(accuracy = 0.1)
-}
-
-# Calculate age of various clades
-fern_age <- get_age_from_genus_pair(c("Equisetum", "Polypodium"))
-lepto_age <- get_age_from_genus_pair(c("Osmunda", "Polypodium"))
-polypodiales_age <- get_age_from_genus_pair(c("Lindsaea", "Polypodium"))
-eupoly_i_age <- get_age_from_genus_pair(c("Hypodematium", "Polypodium"))
-eupoly_ii_age <- get_age_from_genus_pair(c("Cystopteris", "Asplenium"))
-
-# Get ages for families
-sanger_sampling_family <- select(sanger_sampling, species, family)
-
-# Only use monophyletic families
-mono_test_family <- assess_monophy(
-	taxon_sampling = sanger_sampling_family,
-	tree = plastid_tree_dated,
-	tax_levels = "family"
-) %>%
-	get_result_monophy(1)
-
-family_ages <-
-mono_test_family %>%
-	filter(monophyly == "Yes") %>%
-	mutate(
-		mrca = parse_number(mrca),
-		stem_node = map_dbl(
-			mrca,
-			~phangorn::Ancestors(plastid_tree_dated, ., type = "parent"))
-	) %>%
-	select(family = taxon, stem_node) %>%
-mutate(
-	# Height is distance above root
-	height = map_dbl(stem_node, ~phytools::nodeheight(
-		tree = plastid_tree_dated, node = .))
-) %>%
-	mutate(
-		# Age is the total length of the tree minus height
-		age = tree_height - height
-	)
-```
-
-We estimate the crown age of ferns to be `r fern_age` ma.
-Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
+We estimate the crown age of ferns to be `r crown_ages$ferns` ma.
+Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
 Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
-Estimated stem ages of fern families were generally older than previous studies. (`r figure("divtimes")`).
+Estimated stem ages of fern families were generally older than previous studies (`r figure("divtimes")`).
 
@@
 family_ages_median <-
-family_ages %>%
+family_stem_ages %>%
 	select(family, current_best = age) %>%
@@
 family_ages_errors <-
-	family_ages %>%
+	family_stem_ages %>%
 	select(family, current_best = age) %>%
@@
 div_times_plot <-
-ggplot() +
+ggplot(mapping = aes(y = family)) +
 	geom_errorbarh(
 		data = family_ages_errors,
-		aes(xmax = high, xmin = low, color = author, y = family),
+		aes(xmax = high, xmin = low, color = author),
 		position = position_dodge(width = 0.5)
@@
 		data = filter(family_ages_median, age_type %in% c("best", "median")),
-		aes(x = age, y = family, color = author, shape = author),
+		aes(x = age, color = author, shape = author),
 		position = position_dodge(width = 0.5)) +

### 2022-02-14 R/functions.R

@@ -221,7 +221,7 @@ count_ncbi_species_by_year <- function(gb_taxa, ncbi_names, year_range) {
 	map_df(year_range, ~sum_species(gb_species, .))
 }
 
-# Etc ----
+# Data loading ----
 
 #' Parse divergene dates in the Testo and Sundue 2016 SI file
 #' 
@@ -252,7 +252,25 @@ parse_ts_dates <- function(testo_sundue_2016_si_path) {
 		)
 }
 
-# Monophyly ----
+# Rmarkdown ----
+
+# Specify format for percent and number in Rmd
+percent <- function(...) {scales::percent(...)}
+number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Monophyly and ages ----
+
+#' Get parent node from a tip taxon
+#'
+#' @param tree Phylogenetic tree.
+#' @param species Single tip of the tree.
+#'
+#' @return Number of the parent node of the species
+#' 
+get_parent <- function(tree, species) {
+	node <- which(tree$tip.label == species)
+	phangorn::Ancestors(tree, node, type = "parent")
+}
 
 #' Add "major clade" to Sanger sampling table
 #'
@@ -273,7 +291,6 @@ add_major_clade <- function(data) {
 			)
 		)
 }
-
 
 #' Make tibble summarizing sampling of Sanger dataset
 #'
@@ -384,3 +401,99 @@ assess_monophy <- function(
 		as.data.frame() %>%
 		MonoPhy::AssessMonophyly(tree, .)
 }
+
+
+#' Calculate crown age based on two genera
+#'
+#' @param tip_pair Character vector of length two: the two genera to be used
+#' to define the clade
+#' @param sanger_sampling Tibble with one row per species in the tree, including
+#' columns "genus", "family", etc.
+#' @param plastid_tree_dated Dated phylogeny
+#'
+#' @return Character vector: age of the clade defined by the two genera
+#' 
+get_crown_age_from_genus_pair <- function(tip_pair, sanger_sampling, plastid_tree_dated) {
+	tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
+	sanger_sampling %>%
+		filter(genus %in% tip_pair) %>%
+		group_by(genus) %>%
+		slice(1) %>%
+		pull(species) %>%
+		ape::getMRCA(plastid_tree_dated, .) %>%
+		phytools::nodeheight(plastid_tree_dated, .) %>%
+		magrittr::subtract(tree_height, .) %>%
[…]

### 2022-02-14 R/ms_functions.R

@@ -169,6 +169,23 @@ node_height_from_tips <- function(phy, tips) {
 }
 
 # Etc -----
+
+#' Function to get support value in BS%
+#'
+#' @param phy an object of class "phylo".
+#' @param tips a vector of mode numeric or character specifying the tips
+#'
+#' @return Bootstrap support as % of that node
+get_bs <- function(phy, tips) {
+	node_select <- ape::getMRCA(phy, tips)
+	as_tibble(phy) %>%
+		filter(node == node_select) %>%
+		pull(label) %>%
+		parse_number() %>%
+		magrittr::multiply_by(0.01) %>%
+		percent()
+}
+
 # Helper function to add plot_group:
 # a "rank" that may be at order or suborder level for plotting
 add_plot_group <- function(data) {

### 2022-02-14 _targets.R

@@ -5,7 +5,35 @@ library(assertr)
 
 source("R/functions.R")
 
+# Specify path to FTOL cache
+# FIXME: change to ftol_cache when ready
+ftol_cache <- here::here("working/_targets")
+
 tar_plan(
+	
+	# Load data ----
+	# - Dates from Testo and Sundue 2016 SI
+	# (includes Rothfels 2015, Schuettpelz & Pryer 2009)
+	tar_files_input(
+		testo_sundue_2016_si_path,
+		"_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
+	),
+	other_dates = parse_ts_dates(testo_sundue_2016_si_path),
+	# - Targets from FTOL workflow
+	plastome_metadata_renamed = tar_read(
+		plastome_metadata_renamed,
+		store = ftol_cache
+	),
+	plastid_tree_dated = tar_read(
+		plastid_tree_dated,
+		store = ftol_cache
+	),
+	ppgi_taxonomy = tar_read(
+		ppgi_taxonomy,
+		store = ftol_cache
+	),
+	
+	# GenBank ----
 	# Analyze number of fern accessions and species in GenBank by type of DNA 
 	# (plastid, mitochondrial, or nuclear)
 	#
@@ -19,7 +47,7 @@ tar_plan(
 			year = gb_query$year,
 			type = gb_query$type),
 		pattern = map(gb_query)
-		),
+	),
 	# - Load NCBI taxonomy
 	tar_file(taxdump_zip_file, "working/taxdmp_2022-02-01.zip"),
 	ncbi_names = load_ncbi_names(
@@ -30,12 +58,26 @@ tar_plan(
 	# by genomic compartment per year
 	gb_species_by_year = count_ncbi_species_by_year(
 		gb_taxa, ncbi_names, year_range = 1990:2021),
-	# Read in estimated clade ages from Testo and Sundue 2016 SI
-	tar_files_input(
-		testo_sundue_2016_si_path,
-		"_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
-	),
-	other_dates = parse_ts_dates(testo_sundue_2016_si_path),
+	
+	# Calculate ages ----
+	# - Make Sanger sampling table
+	sanger_sampling = make_sanger_sampling_tbl(
+		plastome_metadata_renamed,
+		sanger_tree = plastid_tree_dated,
+		ppgi_taxonomy = ppgi_taxonomy),
[…]

### 2022-02-15 ms/manuscript.Rmd
Update plots
Add clade codes to FTOL plot

@@
 library(magrittr)
+library(deeptime)
 library(ggrepel)
 library(ggokabeito)
+library(ggtext)
 
@@
 	)
+
+# Make exemplar set of species in backbone representing plot groups
+major_clade_exemp_taxa <-
+	backbone_taxonomy %>%
+	# Slice to one exemplar species per major clade
+	group_by(major_clade) %>%
+	slice(1) %>%
+	ungroup() %>%
+	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
+	arrange(phy_order) %>%
+	# Drop outgroup
+	filter(family != "og")
+
+# Make a color palette for major clades
+# Start with list of green.armytage.colors names,
+# exclude those that are too close or stand out too much
+# https://www.jstatsoft.org/article/view/v090c01
+green_army_use_cols <-
+	c(
+		"amethyst",
+		"blue",
+		"caramel",
+		"damson",
+		# "ebony",
+		# "forest",
+		"green",
+		"honeydew",
+		# "iron",
+		# "jade",
+		# "khaki",
+		# "lime",
+		"mallow",
+		"navy",
+		# "orpiment",
+		"pink",
+		# "quagmire",
+		"red",
+		"sky",
+		"turquoise",
+		"uranium",
+		"violet",
+		"wine",
+		# "xanthin",
+		# "yellow",
+		"zinnia"
+	)
+
+clade_cols <-
+	Polychrome::green.armytage.colors() %>%
+	extract(green_army_use_cols) %>%
+	extract(seq_len(nrow(major_clade_exemp_taxa))) %>%
+	set_names(major_clade_exemp_taxa$major_clade)
+
+clade_cols_tib <-
+	tibble(
+		color = clade_cols,
+		major_clade = names(clade_cols)
+	) %>%
+	mutate(
+		code = substr(major_clade, 1, 2),
+		code = case_when(
+			major_clade == "Saccolomatineae" ~ "Sac",
+			major_clade == "Salviniales" ~ "Sal",
+			TRUE ~ code
+		)) %>%
+	assert(is_uniq, code)
+
 ```
@@
 
-# Make exemplar set of species in backbone representing plot groups
-major_clade_exemp_taxa <-
-	backbone_taxonomy %>%
-	# Slice to one exemplar species per major clade
-	group_by(major_clade) %>%
-	slice(1) %>%
-	ungroup() %>%
-	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
-	arrange(phy_order) %>%
-	# Drop outgroup
-	filter(family != "og")
-
-# Make a color palette for major clades
-# Start with list of green.armytage.colors names,
-# exclude those that are too close or stand out too much
-# https://www.jstatsoft.org/article/view/v090c01
-green_army_use_cols <-
-	c(
-		"amethyst",
-		"blue",
-		"caramel",
-		"damson",
-		# "ebony",
-		# "forest",
-		"green",
-		"honeydew",
-		# "iron",
-		# "jade",
-		# "khaki",
-		# "lime",
-		"mallow",
-		"navy",
-		# "orpiment",
-		"pink",
-		# "quagmire",
-		"red",
-		"sky",
-		"turquoise",
-		"uranium",
-		"violet",
-		"wine",
-		# "xanthin",
-		# "yellow",
-		"zinnia"
-	)
-
-clade_cols <-
-	Polychrome::green.armytage.colors() %>%
-	extract(green_army_use_cols) %>%
-	extract(seq_len(nrow(major_clade_exemp_taxa))) %>%
-	set_names(major_clade_exemp_taxa$major_clade)
+# Format y-axis text: show major clade abbreviations in bold
+clade_labs_tib <-
+	clade_cols_tib %>%
+	select(major_clade, code) %>%
+	mutate(
+		clade_short = map2_chr(major_clade, code, ~str_remove(.x, .y)),
+		label = glue("<b>{code}</b>{clade_short}"))
+
+clade_labs <- clade_labs_tib$label %>%
+	set_names(., clade_labs_tib$major_clade)
 
@@
 	scale_fill_manual(values = clade_cols) +
+	scale_y_discrete(labels = clade_labs) +
 	labs(x = "Coverage") +
@@
 	theme(
-		axis.text.y = element_text(size = 6),
+		axis.text.y = element_markdown(size = 6),
 		axis.title.y = element_blank(),
@@
 # Tree plot ----
-phy <- plastid_tree_dated
-
 # Make tibble for annotating major clades
 clade_tibble <- 
-	tibble(
-		Polypodiineae = c("Leucostegia_truncata", "Didymochlaena_fijiensis"),
-		Aspleniineae = c("Acystopteris_taiwaniana", "Rhachidosorus_pulcher"),
-		Dennstaedtiineae = c("Monachosorum_flagellare", "Dennstaedtia_auriculata"),
-		Pteridineae = c("Coniogramme_affinis", "Pteris_dispar"),
-		Lindsaeineae = c("Cystodium_sorbifolium", "Odontosoria_retusa"),
-		Saccolomatineae = c("Saccoloma_inaequale", "Orthiopteris_campylura"),
-		Cyatheales = c("Cyathea_bipinnatifida", "Culcita_macrocarpa"),
-		Salviniales = c("Marsilea_nashii", "Salvinia_oblongifolia"),
-		Schizaeales = c("Schizaea_elegans", "Lygodium_palmatum"),
-		Gleicheniales = c("Matonia_foxworthyi", "Sticherus_bifidus"),
-		Hymenophyllales = c("Cephalomanes_javanicum", "Hymenophyllum_tortuosum"),
-		Marattiales = c("Angiopteris_evecta", "Danaea_erecta"),
-		`Ophioglossales + Psilotales` = c("Psilotum_nudum", "Sceptridium_nipponicum"),
-		Osmundales = c("Osmundastrum_cinnamomeum", "Todea_barbara"),
-		Equisetales = c("Equisetum_bogotense", "Equisetum_hyemale")
-	) %>%
-	pivot_longer(names_to = "major_clade", values_to = "taxa", cols = everything()) %>%
-	arrange(major_clade) %>%
-	group_by(major_clade) %>%
-	summarize(taxa = list(taxa)) %>%
-	mutate(
-		mrca = map_chr(taxa, ~getMRCA(phy, .))
-	) %>%
-	transmute(
-		major_clade,
-		mrca,
-		dummy = ""
-	) %>%
-	mutate(mrca = parse_number(mrca))
-
-a <- ggtree(
-	phy, 
-	layout = "circular", right = FALSE, size = 0.1,
-) +
+	# Use assess_monophy to get MRCA of each major clade
+	# They should all by monophyletic anyways
+	select(sanger_sampling, species, major_clade) %>%
+	assess_monophy(
+		taxon_sampling = .,
+		tree = plastid_tree_dated,
+		tax_levels = "major_clade"
+	) %>%
+	get_result_monophy(1) %>%
+	filter(!taxon %in% c(
+		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
+	verify(all(monophyly == "Yes")) %>%
+	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
+	left_join(clade_cols_tib, by = "major_clade")
+
+# Helper function to label clades with different offset
+# For some reason, using the same offset.text around the plot
+# results in different positions of text relative to bars. 
+# Account for this by manual adjusting offset.text
+ftol_geom_cladelab <- function(code_select, offset_text) {
 	geom_cladelab(
-		data = clade_tibble,
+		data = filter(clade_tibble, code %in% code_select),
 		mapping = aes(
@@
 			color = major_clade,
-			label = dummy
+			label = code
 		),
 		offset = 0.1,
-		barsize = 2) +
+		offset.text = offset_text,
+		fontsize = 2.5,
+		textcolour = "black",
+		barsize = 2)
+}
+
+a <- ggtree(
+	plastid_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
+	ftol_geom_cladelab(
+		clade_tibble$code[clade_tibble$code %in% c("Eq", "Op", "Ma", "Os", "Gl")], 10) +
+	ftol_geom_cladelab(
+		clade_tibble$code[clade_tibble$code %in% c("Hy", "Sc", "Sal")], 14) +
+	ftol_geom_cladelab(
+		clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li", "De")], 17) +
+	ftol_geom_cladelab("Pt", 26) +
+	ftol_geom_cladelab("As", 34) +
+	ftol_geom_cladelab("Po", 17) +
 	scale_color_manual(values = clade_cols) +
@@
 
+# Assemble plots ----
 ftol_plot <-
@@
 		theme = theme(
-			plot.margin = unit(c(-2,5,-2,-1.6), "cm"))
+			plot.margin = unit(c(-2,5,-2,-1.4), "cm"))
 	)
@@
 Fern Tree of Life (FTOL).
+Tree rooted on bryophytes.
 Inset plot shows coverage by major clade (order or suborder).
+Bold part of each clade name is its code, which is also indicated on the tree.
 Numbers next to each bar show sampled species out of total number of species.
@@
 Outgroup (seed plants, lycophytes, and bryophytes) not shown.
-Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`.
-Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.
+Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`. Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.
 
@@
 	left_join(other_dates, by = "family") %>%
+		left_join(
+		unique(select(sanger_sampling, family, major_clade)),
+		by = "family"
+	) %>%
+	left_join(clade_cols_tib, by = "major_clade"
+	) %>%
 	mutate(family = fct_reorder(family, current_best)) %>%
-	pivot_longer(values_to = "age", -family) %>%
+	pivot_longer(values_to = "age", -c(family, color, code, major_clade)) %>%
 	separate(name, into = c("author", "age_type"), sep = "_") %>%
 	assert(not_na, author, age_type) %>%
-	filter(age_type %in% c("best", "median"))
+	filter(age_type %in% c("best", "median")) %>%
+	assert(not_na, major_clade, color)
 
@@
 	mutate(source = fct_reorder(source, year))
+
+# Format y-axis text: show major clade in parentheses
+family_labs_tbl <-
+family_ages_median %>%
+	select(family, code) %>%
+	unique() %>%
+	mutate(label = glue("{family} ({code})"))
+family_labs <- family_labs_tbl$label %>%
+	set_names(., family_labs_tbl$family)
+
+# Provide names of periods in caption
+# Make sure to read and check that they are all there!
+period_names <-
+	periods %>%
+	# Filter to range including Ordovician to Neogene
+	filter(min_age < 475) %>%
+	filter(min_age > 1) %>%
+	arrange(desc(min_age)) %>%
+	mutate(label = glue("{abbr} ({name})")) %>%
+	pull(label) %>%
+	paste(collapse = ", ")
 
@@
 		labels = ages_sampling$label) +
-	labs(x = "Age (millions of years)") +
+	scale_y_discrete(labels = family_labs) +
+	scale_x_reverse("Age (millions of years)") +
+	coord_geo(
+		dat = list("periods", "eras"),
+		pos = list("b", "b"),
+		abbrv = list(TRUE, FALSE),
+		xlim = c(500, 0),
+		fill = "transparent",
+		color = "grey20",
+		lab_color = "grey30",
+		height = unit(1, "lines"),
+		size = 3) +
 	guides(shape = guide_legend(nrow = 2)) +
@@
 
-**`r figure("divtimes")`**. Stem age of fern families (mya) estimated by selected studies.
+**`r figure("divtimes")`**.
+Stem age of fern families (mya) estimated by selected studies.
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
 For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
+Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
+Period name abbreviations as follows: `r period_names`.

### 2022-02-15 _targets.R

@@ -59,7 +59,7 @@ tar_plan(
 	gb_species_by_year = count_ncbi_species_by_year(
 		gb_taxa, ncbi_names, year_range = 1990:2021),
 	
-	# Calculate ages ----
+	# Analyze monophyly and ages ----
 	# - Make Sanger sampling table
 	sanger_sampling = make_sanger_sampling_tbl(
 		plastome_metadata_renamed,

### 2022-02-15 ms/manuscript.Rmd
Move calculation of coverage to targets pipeline

@@
 	c(
+		accepted_species,
 		sanger_sampling,
+		coverage_by_rank,
+		total_coverage,
+		bb_coverage_by_rank,
+		bb_total_coverage,
 		gb_species_by_year,
@@
 
-```{r sampling-stats, message=FALSE}
-# FTOL sampling ----
-
-# Tally FTOL sampling by rank
-ftol_sampling_by_rank <-
-	sanger_sampling %>%
-	filter(outgroup == FALSE) %>%
-	select(-outgroup) %>%
-	# drop suborder
-	select(-suborder) %>%
-	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
-	group_by(rank) %>%
-	count(name) %>%
-	ungroup
-
-# Make list of all accepted species
-accepted_species <-
-	pteridocat %>% 
-	# Only keep accepted names
-	filter(str_detect(taxonomicStatus, "accepted")) %>%
-	# Only keep species
-	# FIXME: need to add rank for newly added names in pteridocat_maker
-	# then don't need is.na()
-	filter(taxonRank == "species" | is.na(taxonRank)) %>%
-	select(scientificName) %>%
-	mutate(
-		rgnparser::gn_parse_tidy(scientificName) %>% 
-			select(species = canonicalsimple),
-		species = str_replace_all(species, " ", "_")
-	) %>%
-	# Map on higher level taxonomy
-	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
-	left_join(
-		select(ppgi_taxonomy, genus, family, 
-					 suborder, order, class), by = "genus"
-	) %>%
-	# FIXME: update class for nothogenera in PPG
-	mutate(class = case_when(
-		str_detect(
-			scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
-		TRUE ~ class
-	)) %>%
-	# Drop lycophytes
-	assert(not_na, class) %>%
-	filter(class == "Polypodiopsida") %>%
-	select(-class) %>%
-	# Add major_clade
-	add_major_clade() %>%
-	# drop suborder
-	select(-suborder)
-
-# Tally accepted sampling by rank
-accepted_sampling_by_rank <-
-	accepted_species %>%
-	select(-scientificName) %>%
-	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
-	group_by(rank) %>%
-	count(name) %>%
-	ungroup %>%
-	# Note that nothogenera have no family or order
-	filter(!is.na(name))
-
-# Format coverage at order-ish level for plotting
-coverage_data <- left_join(
-	ftol_sampling_by_rank,
-	rename(accepted_sampling_by_rank, n_accepted = n),
-	by = c("rank", "name")
-) %>%
-	mutate(coverage = n/n_accepted)
-
-# Format for text in MS
+```{r sampling-stats, message = FALSE}
+# Format FTOL sampling as list
 tax_coverage <-
-	coverage_data %>%
+	coverage_by_rank %>%
 	filter(rank == "major_clade") %>%
-	arrange(coverage) %>%
-	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
+	transmute(
+		# Be sure to change to factor so list is correct order
+		# when splitting
+		clade = fct_reorder(name, coverage),
+		coverage = percent(coverage, accuracy = 0.1)) %>%
 	split(.$clade)
 
-# Format overall sampling for printing in MS
-total_sampling <-
-	# Start with FTOL sampling by genus, family, order
-	ftol_sampling_by_rank %>% 
-	count(rank) %>%
-	# Add n species
-	bind_rows(
-		sanger_sampling %>%
-			filter(outgroup == FALSE) %>%
-			select(-outgroup) %>%
-			summarize(n = n(), rank = "species")
-	) %>%
-	# Join to accepted number of species, genus, family, order
-	left_join(
-		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
-			bind_rows(
-				summarize(accepted_species, n_accepted = n(), rank = "species")
-			),
-		by = "rank"
-	) %>%
-	# Calculate coverage
-	mutate(
-		coverage_p = percent(n / n_accepted, accuracy = 0.1),
-		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
-	# Split for printing
-	split(.$rank)
-
-# Plastome (backbone) sampling ----
-bb_sampling <-
-	tibble(species = plastome_tree$tip.label) %>%
-	# Add higher taxonomy
-	left_join(sanger_sampling, by = "species")
-
-bb_sampling_by_rank <-
-	bb_sampling %>%
-	# Drop OG
-	filter(outgroup == FALSE) %>%
-	select(-outgroup) %>%
-	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
-	group_by(rank) %>%
-	count(name) %>%
-	ungroup
-
-# Format sampling for printing in MS
-bb_total_sampling <-
-	# Start with backbone sampling by genus, family, order
-	bb_sampling_by_rank %>% 
-	count(rank) %>%
-	# Add n species
-	bind_rows(
-		bb_sampling %>%
-			filter(outgroup == FALSE) %>%
-			select(-outgroup) %>%
-			summarize(n = n(), rank = "species")
-	) %>%
-	# Join to accepted number of species, genus, family, order
-	left_join(
-		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
-			bind_rows(
-				summarize(accepted_species, n_accepted = n(), rank = "species")
-			),
-		by = "rank"
-	) %>%
-	# Calculate coverage
-	mutate(
-		coverage_p = percent(n / n_accepted),
-		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 0.1)}")) %>% #nolint
-	# Split for printing
-	split(.$rank)
+# Format overall FTOL sampling as list
+total_sampling <- split(total_coverage, total_coverage$rank)
+
+# Format overall plastome (backbone) sampling as list
+bb_total_sampling <- split(bb_total_coverage, bb_total_coverage$rank)
 ```
@@
 plastome_gleich_genera_n <-
-	bb_sampling %>%
-	filter(major_clade == "Gleicheniales") %>%
-	pull(genus) %>%
+	bb_coverage_by_rank %>%
+	left_join(select(ppgi_taxonomy, name = genus, order), by = "name") %>%
+	filter(order == "Gleicheniales") %>%
+	pull(name) %>%
 	n_distinct()
@@
 	assert(is_uniq, code)
-
 ```
@@
 b <-
-	coverage_data %>%
+	coverage_by_rank %>%
 	filter(rank == "major_clade") %>%

### 2022-02-15 R/functions.R

@@ -258,6 +258,220 @@ parse_ts_dates <- function(testo_sundue_2016_si_path) {
 percent <- function(...) {scales::percent(...)}
 number <- function(...) {scales::number(big.mark = ",", ...)}
 
+# Coverage ----
+
+#' Make a tibble of accepted species in pteridocat taxonomic database
+#'
+#' @param pteridocat Taxonomic database of pteridophytes at species level
+#' @param ppgi_taxonomy Pteridophyte phylogeny group I taxonomy (genus level
+#' and above)
+#'
+#' @return Tibble of accepted species with higher level taxonomy
+#' 
+get_accepted_species <- function(pteridocat, ppgi_taxonomy) {
+	pteridocat %>% 
+		# Only keep accepted names
+		filter(str_detect(taxonomicStatus, "accepted")) %>%
+		# Only keep species
+		# FIXME: need to add rank for newly added names in pteridocat_maker
+		# then don't need is.na()
+		filter(taxonRank == "species" | is.na(taxonRank)) %>%
+		select(scientificName) %>%
+		mutate(
+			rgnparser::gn_parse_tidy(scientificName) %>% 
+				select(species = canonicalsimple),
+			species = str_replace_all(species, " ", "_")
+		) %>%
+		# Map on higher level taxonomy
+		mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
+		left_join(
+			select(ppgi_taxonomy, genus, family, 
+						 suborder, order, class), by = "genus"
+		) %>%
+		# FIXME: update class for nothogenera in PPG
+		mutate(class = case_when(
+			str_detect(
+				scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
+			TRUE ~ class
+		)) %>%
+		# Drop lycophytes
+		assert(not_na, class) %>%
+		filter(class == "Polypodiopsida") %>%
+		select(-class) %>%
+		# Add major_clade
+		add_major_clade() %>%
+		# drop suborder
+		select(-suborder) 
+}
+
+#' Calculate coverage of species and other taxonomic ranks in FTOL
+#'
+#' @param sanger_sampling Tibble including sampling statistics in Sanger 
+#' dataset (one row per species, with columns for higher-level taxonomy
+#' and outgroup status)
+#' @param accepted_species Tibble of accepted species in pteridocat
[…]

### 2022-02-15 _targets.R

@@ -2,6 +2,7 @@ library(targets)
 library(tarchetypes)
 library(tidyverse)
 library(assertr)
+library(glue)
 
 source("R/functions.R")
 
@@ -24,12 +25,20 @@ tar_plan(
 		plastome_metadata_renamed,
 		store = ftol_cache
 	),
+	plastome_tree = tar_read(
+		plastome_tree,
+		store = ftol_cache
+	),
 	plastid_tree_dated = tar_read(
 		plastid_tree_dated,
 		store = ftol_cache
 	),
 	ppgi_taxonomy = tar_read(
 		ppgi_taxonomy,
+		store = ftol_cache
+	),
+	pteridocat = tar_read(
+		pteridocat,
 		store = ftol_cache
 	),
 	
@@ -58,6 +67,20 @@ tar_plan(
 	# by genomic compartment per year
 	gb_species_by_year = count_ncbi_species_by_year(
 		gb_taxa, ncbi_names, year_range = 1990:2021),
+	
+	# Coverage ----
+	# Make tibble of accepted species in pteridocat
+	accepted_species = get_accepted_species(pteridocat, ppgi_taxonomy),
+	# Calculate coverage by rank and overall coverage for FTOL
+	coverage_by_rank = calculate_ftol_coverage(
+		sanger_sampling, accepted_species, "cov_by_rank"),
+	total_coverage = calculate_ftol_coverage(
+		sanger_sampling, accepted_species, "total_sampling"),
+	# Calculate coverage by rank and overall coverage for backbone tree
+	bb_coverage_by_rank = calculate_backbone_coverage(
+		plastome_tree, sanger_sampling, accepted_species, "cov_by_rank"),
+	bb_total_coverage = calculate_backbone_coverage(
+		plastome_tree, sanger_sampling, accepted_species, "total_sampling"),
 	
 	# Analyze monophyly and ages ----
 	# - Make Sanger sampling table

### 2022-02-16 R/functions.R
Move all ms_functions to functions.R

@@ -252,12 +252,6 @@ parse_ts_dates <- function(testo_sundue_2016_si_path) {
 		)
 }
 
-# Rmarkdown ----
-
-# Specify format for percent and number in Rmd
-percent <- function(...) {scales::percent(...)}
-number <- function(...) {scales::number(big.mark = ",", ...)}
-
 # Coverage ----
 
 #' Make a tibble of accepted species in pteridocat taxonomic database
@@ -407,7 +401,7 @@ tally_accepted_sampling_by_rank <- function(accepted_species) {
 #' @param ret_type Type of output to return:
 #' - cov_by_rank: Returns coverage by taxonmic rank
 #' - total_sampling: Returns overall sampling coverage
-#'
+#' 
 #' @return Tibble
 #' 
 calculate_backbone_coverage <- function(
@@ -461,7 +455,7 @@ calculate_backbone_coverage <- function(
 		# Calculate coverage
 		mutate(
 			coverage_p = percent(n / n_accepted),
-			coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 0.1)}"))
+			coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}"))
 	
 	switch(
 		ret_type,
@@ -711,3 +705,256 @@ get_stem_family_age <- function(
 		select(family, age)
 }
 
+# Formatting ----
+
+# Specify percentage format
+percent <- function(...) {scales::percent(...)}
+
+# Specify number format
+number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Abbreviations
+ie <- "*i.e.*"
+eg <- "*e.g.*"
+ca <- "*ca.*"
+
+#' Convert a data.frame of counts to percentages, with automatic formatting
+#'
+#' Format numbers and percentage to look nice in MS
+#'
+#' @param ... Arguments passed to janitor::tabyl() 
+#'
+#' @result Dataframe
+tabyl_fmt <- function(...) {
+	tabyl(...) %>%
+		mutate(
+			percent = percent(percent, accuracy = 0.1),
+			n = number(n, accuracy = 1)
+		)
+}
+
+# References ----
+
+#' Extract citations (formatted like `[@key]`) from an Rmd file
+#'
+#' @param rmd_file Character vector; path to Rmd file
+#'
+#' @return Data frame with one column 'key'
+#' 
+extract_citations <- function(rmd_file) {
[…]

### 2022-02-16 R/ms_functions.R

@@ -1,217 +1,0 @@
-# Formatting ----
-
-# Specify percentage format
-percent <- function(...) {scales::percent(...)}
-
-# Specify number format
-number <- function(...) {scales::number(big.mark = ",", ...)}
-
-# Abbreviations
-ie <- "*i.e.*"
-eg <- "*e.g.*"
-ca <- "*ca.*"
-
-# References ----
-
-#' Extract citations (formatted like `[@key]`) from an Rmd file
-#'
-#' @param rmd_file Character vector; path to Rmd file
-#'
-#' @return Data frame with one column 'key'
-#' 
-extract_citations <- function(rmd_file) {
-	read_lines(rmd_file) %>%
-		stringr::str_split(" |;") %>% 
-		unlist %>% 
-		magrittr::extract(., stringr::str_detect(., "@")) %>% 
-		stringr::str_remove_all("\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}|\\\\") %>% 
-		magrittr::extract(., stringr::str_detect(., "^@|^-@")) %>% 
-		stringr::str_remove_all("^@|^-@") %>% 
-		unique %>% 
-		sort %>%
-		tibble(key = .)
-}
-
-#' Filter a list of references in YAML format to those occurring in an Rmd file
-#'
-#' @param rmd_file Character vector; Path to Rmd file(s)
-#' @param yaml_in String or list; if string, the path to the YAML file to filter.
-#' If list, should be result of reading in a YAML file with yaml::read_yaml()
-#' @param yaml_out Path to write filtered YAML reference file
-#'
-#' @return NULL; externally, the filtered YAML will be written to `yaml_out`
-#' 
-filter_refs_yaml <- function(rmd_file, yaml_in = "ms/main_library.yaml", yaml_out = "ms/references.yaml", silent = FALSE) {
-	
-	# Parse RMD file and extract citation keys
-	citations <- purrr::map_df(rmd_file, extract_citations)
-	
-	# Read in YAML including all references exported from Zotero
-	if (inherits(yaml_in, "character")) {
-		ref_yaml <- yaml::read_yaml(yaml_in)
-	} else if (inherits(yaml_in, "list")) {
-		ref_yaml <- yaml_in
-	} else {
-		stop("`yaml_in` must be a path to a YAML file (string) or a list read in with yaml::read_yaml()")
-	}
-	
-	# Extract all citation keys from full YAML
[…]

### 2022-02-16 ms/manuscript.Rmd
Update MS

@@
 library(ggtext)
+library(janitor)
 
@@
 # Load functions
-source(here::here("R/ms_functions.R"))
 source(here::here("R/functions.R"))
@@
 		sanger_alignment_tbl,
-		sanger_tree_fast,
+		plastome_alignment_tbl,
 		monophy_by_clade,
@@
 		crown_ages,
+		fossil_ferns_raw,
 		family_stem_ages
@@
 	n_distinct(.$species)
-```
-
-<!-- Set left justification -->
-\raggedright
-
-<!-- Don't indent -->
-\setlength{\parindent}{0ex}
-
-**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^1,4,5,6,7,8^**
-
-^1^Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan
-
-^2^Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., USA
-
-^3^Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico
-
-^4^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
-
-^5^Department of Computational Biology and Medical Sciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
-
-^6^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
-
-^7^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
-
-^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
-
-**\*Correspondence:**    
-Joel H. Nitta  
-joelnitta@gmail.com
-
-Keywords: Fern~1~, Phylogeney~2~, Plastome~3~, PPGI~4~, Pteridophyte~5~, *rbcL*~6~.
-
-# Abstract
-
-Ferns (ca.
-12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for > 1/3 of all fern species deposited in GenBank.
-Global trees including all ferns have been published periodically, but as molecular systematics research continues at a rapid pace, such trees quickly become outdated.
-
-Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from sequence data on GenBank.
-Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with Sanger-sequenced loci (many taxa, few loci) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
-We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
-
-The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of > 20% relative to the previous largest fern phylogeny.
-FTOL will be updated on a semi-annual basis and will be available via a web portal and R package, enabling researchers to have immediate access to the most recent, comprehensively sampled fern phylogeny.
-FTOL will be useful for anybody studying this important group of plants at scales from smaller clades to the entire tree.
-We anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
-
-# Introduction
-
-Ferns (ca.
-12,000 spp.) are the second most diverse clade of vascular plants after seed plants (ca.
-260,000 spp.), and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
-As Dobzhansky famously said, "Nothing in biology makes sense except in the light of evolution"; likewise, a well-sampled molecular phylogeny is key to any investigation of fern evolutionary biology.
-Fortunately, ferns have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for ca.
-1/3 of all fern species deposited in GenBank.
-Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.
-
-Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny [@Testo2016a\; `r figure("genbank")`].
-There is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
-This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
-
-Multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
-While these may be appropriate in some cases, they may not produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth. 
-By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.
-
-Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be possible with a "tree of all life" or "tree of all plants".
-Recently, a genus-level taxonomy of ferns and lycophytes was established following a open, community-driven model [@PteridophytePhylogenyGroupI2016].
-Thanks to this community-first approach, the taxonomy (PPG I) has been widely accepted and used (cited 475 times on Google Scholar as of writing).
-However, there were clearly problematic (non-monophyletic) genera at the time PPG I was published, and many new taxonomic changes have been (and will continue to be) made since [e.g., @Almeida2017; @Shang2018; @Zhang2020].
-It is anticipated that "PPG II" will be a online, open resource that can be updated as necessary (Schuettpelz, pers. comm.).
-We envision an open, continuously updated global fern phylogeny that is directly integrated with PPG II so that taxonomic decisions reflecting phylogeny can be made based on community consensus using the most recently available data.
-
-Here, we leverage our specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
-We plan to run the pipeline on a semi-annual basis and make the results freely available online through a web portal and R package.
-This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
-We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) It will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) It will allow for continuous assessment of taxonomy, and indicate what parts of the tree are in need of taxonomic revision; 3) It will be an important source of data for any phylogenetic comparative study of ferns.
-
-# Materials and Methods
-
-## Locus selection
-
-The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
-Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome loci").
-*trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*.
-*rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
-The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes), then combined with the two spacers (79 loci total).
-The plastome loci include all seven Sanger loci.
-
-## Dataset construction
-
-All sequences were downloaded from GenBank.
-GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
-Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
-There is no formal definition of genomic vs.
-Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< 7000 bp) or plastome sequences (> 7000 bp) with the "[SLEN]" term.
-
-There is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.
-To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
-Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
-Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
-
-We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
-We then filtered the sequences to a single representative longest sequence per genus.
-Next, we aligned the filtered sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
-To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
-The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from sequences downloaded from GenBank.
-
-## Taxonomic name resolution
-
-One goal of this project is generate a phylogeny that is consistent with the community-driven Pteridophyte Phylogeny Group I taxonomic system [@PteridophytePhylogenyGroupI2016\; hereafter "PPGI"].
-Species names in GenBank do not necessarily conform to PPGI.
-Therefore, we standardized all species names in the GenBank sequences against a reference taxonomy.
-We designed the reference taxonomy for this project based on the World Ferns database [@Hassler2022], which conforms to PPGI and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
-We used the taxastand R package [@Nitta2021] to resolve species names, which can account for differences in formatting of taxonomic authors (e.g., parenthetical basionym author present or absent) as well as perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
-We manually inspected any fuzzily matched or non-matching names.
-This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
-Therefore, we updated and edited the World Ferns database using the dwctaxon R package, which is designed to work with taxonomic data in the Darwin Core format.
-We refer to the resulting taxonomic database as "pteridocat", and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
-
-## Removal of rogue sequences using BLAST
-
-Another potential issue with GenBank sequences is the presence of misidentified sequences and contaminations (hereafter "rogues").
-We removed putative rogues from Sanger sequences as follows.
-First, we constructed a BLAST library including all extracted Sanger sequences.
-Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
-We compared the families (following PPGI) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contamination and excluded from further analysis.
-For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
-Species belonging to monotypic families were not considered for this filtering.
-While this method cannot account for rogues at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
-We inspected all accessions flagged as rogues before excluding them.
-In the case that that mismatch of family was due to incorrect taxonomy (`r eg`, species name on GenBank matches the correct family but name used in the taxonomic database does not), we updated the taxonomic database accordingly.
-
-## Representative sequence selection and concatenation
-
-A typical step in multilocus workflows is to concatenate loci across samples.
-For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
-However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
-Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
-First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
-We then concatenated loci for monophyletic species by selecting the longest accession per locus within each species.
-Next, for the remaining species, we concatenated loci across accessions if all accessions originated from the same voucher specimen; this could not be done in all cases as voucher information is not always annotated on GenBank.
-Next, for the remaining species, we concatenated loci across accessions if all accessions for a given species originated from only one publication, on the assumption that the authors of the study intended to do so.
-For the remaining species that could not be concatenated using any of the above methods, we selected one longest accession per locus and discarded the others.
-
-We then selected the final set of concatenated loci for each species. 
-We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus should improve phylogenetic analysis [@Talavera2021].
-Our selection is based on the following criteria, in order of priority:
-
-1. The sample (set of accessions joined by the methods described above) with the longest concatenated sequence length; sample includes *rbcL* and additional loci.
-2. The sample with the longest *rbcL* sequence; sample includes only *rbcL* and no additional loci.
-3. The sample with the longest concatenated sequence length; sample does not include *rbcL*.
-
-The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
-For these, we selected one specimen per species with the longest combined sequence length across all plastome loci.
-
-## Sequence alignment
-
-For non-spacer regions, we aligned each locus separately with MAFFT.
-
-Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (e.g., within genera) where slower-evolving genes like *rbcL* may not provide enough resolution.
-Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "--merge" option.
-This retains the indels within each subalignment while aligning across subalignments.
-For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".
-Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data.
-We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
-
-Next, we removed poorly aligned regions from the alignments using trimAl.
-We removed regions with >1% or >5% of sequences having gaps from coding regions or spacers, respectively.
-
-## Phylogenetic analysis
-
-### Backbone phylogeny
-
-We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
-ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
-To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
-The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
-Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
-
-### Initial Sanger phylogeny
-
-Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
-We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
-
-### Removal of rogue sequences based on initial Sanger phylogeny
-
-We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
-We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
-We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; expected monophyly follows @PteridophytePhylogenyGroupI2016].
-However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
-
-Based on the results of this inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses to this step.
-When re-running the analysis, it is possible that there were newly added GenBank accessions in place of the excluded rogues that themselves are rogues.
-Therefore, we repeated this process until the monophyly of all higher-level (`r eg` genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
-
-### Final Sanger phylogeny
-
-We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure as used for the backbone phylogeny.
-Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates.
-
-We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods or models as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
-
-## Molecular dating
-
-```{r dating-methods-stats}
+
 # Count number of calibration points
@@
 	number()
+```
+
+<!-- Set left justification -->
+\raggedright
+
+<!-- Don't indent -->
+\setlength{\parindent}{0ex}
+
+**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^1,4,5,6,7,8^**
+
+^1^Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan
+
+^2^Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., USA
+
+^3^Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Mexico City, Mexico
+
+^4^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^5^Department of Computational Biology and Medical Sciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
+
+^6^Atmosphere and Ocean Research Institute, The University of Tokyo, Chiba, Japan
+
+^7^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan
+
+^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
+
+**\*Correspondence:**    
+Joel H. Nitta  
+joelnitta@gmail.com
+
+Keywords: Fern~1~, Phylogeney~2~, Plastome~3~, PPGI~4~, Pteridophyte~5~, *rbcL*~6~.
+
+# Abstract
+
+Ferns (ca.
+12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for more than one third of all fern species deposited in GenBank.
+Global trees including all ferns have been published periodically, but as molecular systematics research continues at a rapid pace, such trees quickly become outdated.
+
+Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data on GenBank.
+Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with Sanger-sequenced loci (many taxa, few loci) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
+We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
+
+The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of more than 20% relative to the previous largest fern phylogeny.
+We also used an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points to date the tree, and estimated ages for most families and deeper clades to be significantly older compared to previous studies.
+
+FTOL and its accompanying datasets including the fossil list and taxonomic database will be updated on a semi-annual basis and be made available via a web portal and R packages, enabling researchers to have immediate access to the most recent, comprehensively sampled fern phylogeny.
+FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales from smaller clades to the entire tree.
+We anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
+
+# Introduction
+
+Ferns (ca.
+12,000 species) are the second most diverse clade of vascular plants after seed plants (ca.
+300,000 species), and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
+A well-sampled molecular phylogeny is key to any investigation of fern evolutionary biology.
+Fortunately, ferns have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for more than one third all currently recognized fern species deposited in GenBank.
+Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.
+
+Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
+Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny of ferns [@Testo2016a\; `r figure("genbank")`].
+There is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
+This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
+
+Multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
+While these may be appropriate in some cases, they may not produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth. 
+By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.
+Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be possible with a "tree of all life" or "tree of all plants".
+
+Recently, a genus-level taxonomy of ferns and lycophytes was established following a open, community-driven model [@PteridophytePhylogenyGroupI2016\; hereafter "PPGI"].
+Thanks to this community-first approach, PPGI has been widely accepted and used (cited 881 times on Google Scholar as of writing).
+However, there were clearly problematic (non-monophyletic) genera at the time PPGI was published, and many new taxonomic changes have been (and will continue to be) made since [e.g., @Almeida2017; @Shang2018; @Zhang2020].
+It is anticipated that "PPGII" will be an online, open resource that can be updated as necessary (Schuettpelz, pers. comm.).
+We envision an open, continuously updated global fern phylogeny that is directly integrated with PPGII so that taxonomic decisions reflecting phylogeny can be made based on community consensus using the most recently available data.
+
+Here, we leverage our specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
+We plan to run the pipeline on a semi-annual basis and make the results freely available online through a web portal and R package.
+This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
+We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
+
+# Materials and Methods
+
+## Locus selection
+
+The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
+Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome loci").
+Here, the locus *trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*; the locus *rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
+The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes), then combined with the two spacers (79 loci total).
+The plastome loci include all seven Sanger loci.
+
+## Dataset construction
+
+All sequences were downloaded from GenBank.
+GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
+There is no formal definition of genomic vs.
+Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.
+
+There is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.
+To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
+Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
+Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
+
+We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
+We then filtered the sequences to a single representative longest sequence per genus.
+Next, we aligned the filtered sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
+To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
+The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from sequences downloaded from GenBank.
+
+## Taxonomic name resolution
+
+One goal of this project is to generate a phylogeny that is consistent with PPGI.
+Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPGI.
+Furthermore, the NCBI taxonomy is not curated specifically for ferns and includes many fern synonyms.
+Therefore, we standardized all species names in the GenBank sequences against a reference taxonomy.
+We designed the reference taxonomy for this project based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPGI (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
+
+We used the taxastand R package [@Nitta2021] to resolve species names, which can account for differences in formatting of taxonomic authors (`r eg`, parenthetical basionym author present or absent) as well as perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
+We manually inspected any fuzzily matched or non-matching names.
+This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
+Therefore, we updated and edited the World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
+We refer to the resulting taxonomic database as "pteridocat", and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
+
+There are differences between pteridocat and PPGI, mostly at the genus level.
+In the time since PPGI was published, several new genera have been published.
+Furthermore, multiple genera included in PPGI were known to be non-monophyletic and included on a tentative basis [@PteridophytePhylogenyGroupI2016].
+Pteridocat includes newly published genera and some genera that were not recognized by PPGI.
+Differences between pteridocat and PPGI are summarized in Table S1.
+
+## Removal of rogue sequences using BLAST
+
+Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminations (hereafter collectively "rogues").
+We removed putative rogues from Sanger sequences as follows.
+First, we constructed a BLAST library including all extracted Sanger sequences.
+Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
+We compared the families (following PPGI) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contamination and excluded from further analysis.
+For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
+Species belonging to monotypic families were not considered for this filtering.
+While this method cannot account for rogues at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
+We inspected all accessions flagged this way as rogues before excluding them.
+In the case that that mismatch of family was due to incorrect taxonomy (`r eg`, species name on GenBank matches the correct family but name used in the taxonomic database does not), we updated the taxonomic database accordingly.
+
+## Sequence concatenation and selection
+
+A typical step in multilocus workflows is to concatenate loci across samples.
+For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
+However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
+Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
+First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
+We then concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
+Accessions not meeting any of these conditions were not concatenated.
+All calculations of sequence length excluded missing bases.
+
+We then selected the final set of concatenated loci for each species based on presence of *rbcL* and sequence length.
+We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
+Our procedure for selecting accessions for each species is as follows.
+
+1. If accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length.
+2. Otherwise, if accessions include *rbcL*, select the accession with the longest *rbcL* sequence.
+3. Otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
+
+The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
+For these, we selected one specimen per species with the longest combined sequence length across all plastome loci.
+
+## Sequence alignment
+
+For non-spacer regions, we aligned each locus separately with MAFFT with automatic adjustment for sequence direction and other settings on default.
+Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
+Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "\-\-merge" option.
+This retains the indels within each subalignment while aligning across subalignments.
+For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "\-\-merge".
+Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data.
+We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
+
+Next, we removed poorly aligned regions from the alignments using trimAl.
+We removed regions with >1% or >5% of sequences having gaps from coding regions or spacers, respectively.
+
+## Phylogenetic analysis
+
+### Backbone phylogeny
+
+We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
+ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
+To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
+The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
+Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
+
+### Initial Sanger phylogeny
+
+Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
+We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
+
+### Removal of rogue sequences based on initial Sanger phylogeny
+
+We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
+We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
+We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; expected monophyly follows @PteridophytePhylogenyGroupI2016].
+However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
+While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
+
+Based on the results of this inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses up to this step.
+When re-running the analysis, it is possible that there were newly added GenBank accessions in place of the excluded rogues that themselves are rogues.
+Therefore, we repeated this process until the monophyly of all higher-level (`r eg` genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
+
+### Final Sanger phylogeny
+
+We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
+We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods or models as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
+
+## Molecular dating
+
+```{r dating-methods-stats}
+# Count total number of fern fossils in database
+n_fern_fossils_total <- 
+	fossil_ferns_raw %>%
+	# Exclude non-fern fossils
+	filter(!affinities %in% c("Euphyllophytes", "Tracheophytes")) %>%
+	pull(fossil_taxon) %>%
+	n_distinct() %>%
+	number()
+
+# Format fossils for printing in MS
+fossils <-
+	fossil_ferns_raw %>%
+	mutate(age = number(minimum_age, accuracy = 0.1)) %>%
+  split(.$affinities)
 
@@
 
-We dated the tree using treePL [@Smith2012].
-We applied a set of `r n_calibrations_total` calibration points as minimum ages, including `r n_calibrations_ferns` calibration points within ferns.
-Our calibration points are all based on fossils.
-We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil name), since taxon concepts may vary between the original publication and currently applied taxonomy [`r eg`, *Kuylisporites*\; @Loiseau2020].
-We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); a value of `r smooth_best` was selected with the smallest chi-squared value for final analysis [@Smith2012].
+We dated the ML tree using treePL [@Smith2012].
+We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa.
+In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
+We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil species or genus) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
+Taxon concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be calibrated for a given fossil will also change.
+We also applied two calibration points outside of ferns including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma [@Testo2016a].
+All calibration points other than the root are minimum ages.
+
+We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); a value of `r smooth_best` was selected with the smallest chi-squared value for final analysis.
+To construct confidence intervals for the estimated divergence times, we constructed 100 bootstrap trees (trees with identical topology but different branchlengths) using IQ-TREE and dated each using treePL with the same settings used for the ML tree [@Maurin2020].
 
@@
 
-## Taxon sampling
+## Taxon and locus sampling
 
 ```{r sampling-stats, message = FALSE}
+# Taxon sampling ----
 # Format FTOL sampling as list
@@
 bb_total_sampling <- split(bb_total_coverage, bb_total_coverage$rank)
+
+# Locus sampling ----
+# Make tibble with one row per locus per species
+sanger_locus_sampling <-
+sanger_alignment_tbl %>%
+	mutate(taxon = map(align_trimmed, rownames)) %>%
+	select(locus = target, species = taxon) %>%
+	unnest(cols = species) %>%
+	# Exclude outgroups
+	left_join(
+		unique(select(sanger_sampling, species, outgroup)), by = "species") %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup)
+
+# Summarize mean number of loci per species
+locus_summ <-
+sanger_locus_sampling %>%
+	group_by(species) %>%
+	summarize(
+		n_loci = n_distinct(locus)
+	) %>%
+	summarize(
+		max = max(n_loci) %>% number(accuracy = 1),
+		min = min(n_loci) %>% number(accuracy = 1),
+		mean = mean(n_loci) %>% number(accuracy = 0.1),
+		sd = sd(n_loci) %>% number(accuracy = 0.1)
+	)
+
+# Format total number of samples per locus as list for MS
+locus_total <-
+sanger_locus_sampling %>%
+	count(locus) %>%
+	mutate(
+		locus = fct_reorder(locus, n),
+		n = number(n)) %>%
+	split(.$locus)
+
+# Make tibble with number of species per locus combination
+locus_comb_tib <-
+sanger_locus_sampling %>%
+	group_by(species) %>%
+	summarize(comb = paste(sort(locus), collapse = "_")) %>%
+	count(comb) %>%
+	arrange(desc(n)) %>%
+	mutate(
+		rbcl_present = str_detect(comb, "rbcL"),
+		n_loci = str_count(comb, "_") + 1)
+
+# Format as list for MS
+locus_comb <-
+	locus_comb_tib %>%
+	mutate(
+		comb = fct_reorder(comb, n),
+		n = number(n, accuracy = 1)) %>%
+	split(.$comb)
+
+# Verify the most frequent locus combination is rbcL alone
+assertthat::assert_that(
+	locus_comb_tib %>%
+		filter(n == max(n)) %>%
+		pull(comb) == "rbcL")
+
+# Verify that the second most frequent locus combination is all seven species
+assertthat::assert_that(
+	locus_comb_tib %>%
+		arrange(desc(n)) %>%
+		slice(2) %>%
+		pull(n_loci) == 7)
+
+# Calculate number of species with > 1 locus
+sp_mult_loci <-
+sanger_locus_sampling %>%
+	group_by(species) %>%
+	summarize(mult_loci = if_else(n_distinct(locus) > 1, "true", "false")) %>%
+	tabyl_fmt(mult_loci) %>%
+	split(.$mult_loci)
+
+# Backbone locus sampling ----
+bb_locus_tbl <-
+plastome_alignment_tbl %>%
+	mutate(taxon = map(align_trimmed, rownames)) %>%
+	select(locus = target, species = taxon) %>%
+	unnest(cols = species) %>%
+	group_by(species) %>%
+	summarize(n_loci = n_distinct(locus)) %>%
+	arrange(desc(n_loci)) %>%
+	left_join(
+		unique(select(sanger_sampling, species, outgroup)), by = "species") %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup)
+
+bb_locus_summ <-
+bb_locus_tbl %>%
+	summarize(
+		max = max(n_loci) %>% number(accuracy = 1),
+		min = min(n_loci) %>% number(accuracy = 1),
+		mean = mean(n_loci) %>% number(accuracy = 0.1),
+		sd = sd(n_loci) %>% number(accuracy = 0.1)
+	)
+
+bb_with_all_loci <-
+	bb_locus_tbl %>%
+	mutate(all_loci_sampled = if_else(n_loci == 79, "true", "false")) %>%
+	tabyl_fmt(all_loci_sampled) %>%
+	split(.$all_loci_sampled)
 ```
@@
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
-
-Taxon sampling for the backbone phylogeny (plastome genes) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.
+Taxon sampling for the backbone phylogeny (plastome loci) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.
+
+The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
+`r sp_mult_loci$true$n` species (`r sp_mult_loci$true$percent`) were sampled for more than one locus.
+Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
+The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species).
+The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
+Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species).
+`r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.
 
@@
 ```{r dna-stats}
-# Define missing base codes in ape::base.freq
-missing_base_codes <- c("n", "-", "?")
-
-# Calculate frequence of missing bases in each dataset
-plastome_base_freq <- base.freq(plastome_alignment, all = TRUE)
-plastome_missing_freq <- sum(plastome_base_freq[missing_base_codes])
-sanger_base_freq <- base.freq(sanger_alignment, all = TRUE)
-sanger_missing_freq <- sum(sanger_base_freq[missing_base_codes])
+# Calculate frequency of missing bases in each dataset overall
+plastome_missing_freq <- base_freq_missing(plastome_alignment)
+sanger_missing_freq <- base_freq_missing(sanger_alignment)
+
+# Calculate frequency of missing bases by locus for Sanger loci
+# Concatenate Sanger loci again here to be sure order is the same
+sanger_genes_cat <- do.call(
+	ape::cbind.DNAbin,
+	c(sanger_alignment_tbl$align_trimmed, fill.with.gaps = TRUE))
 
 sanger_missing_by_gene <-
+	# Get start and end position of each gene in concatenated alignment
   sanger_alignment_tbl %>%
-	rename(gene = target) %>%
+	mutate(
+		nbp = map_dbl(align_trimmed, ncol),
+		end = cumsum(nbp),
+		start = end - nbp + 1) %>%
+	# Calculate frequency of missing data
+	select(gene = target, start, end) %>%
+	mutate(
+		missing = map2_dbl(
+			start, end,
+			~base_freq_missing(
+				seqs = sanger_genes_cat,
+				start = .x,
+				end = .y))
+	) %>%
+	select(-start, -end) %>%
+	# Reorder and split for printing
 	mutate(
@@
 		gene = str_replace_all(gene, "([^-]+)", "*\\1*"),
-		# Calculate percent missing data per gene
-		missing = map_dbl(align_trimmed, ~sum(base.freq(., all = TRUE)[missing_base_codes])),
+		gene = fct_reorder(gene, missing),
 		missing = percent(missing)) %>%
-	select(-align_trimmed) %>%
-	arrange(missing) %>%
 	split(.$gene)
@@
 The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
-
 Within leptosporangiate ferns, Osmundales is sister to all other species.
@@
 Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
-Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
+Except for eupolypods II, these ages are all older than those reported in the most recent global fern phylogeny [@Testo2016a].
 Estimated stem ages of fern families were generally older than previous studies (`r figure("divtimes")`).
+We did not compare crown ages of families across studies because differences in crown ages in smaller clades are likely affected by species sampling.
 
@@
 However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
-Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a non-monophyletic Gleicheniales [@Qi2018; @Shen2018].
+Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018].
 We recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
@@
 We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (`r figure("divtimes")`).
-This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
+This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
 Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
@@
 Recent studies using transcriptomics are clarifying the backbone of the fern phylogeny using many (25--2400) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
-The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFLAG) project, which seeks to generate genomic data (ca.
+The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFlag) project, which seeks to generate genomic data (ca.
 300 single-to-low copy nuclear gene regions) for all flagellate plants (bryophytes, lycophytes, ferns, and gymnosperms) [@Breinholt2021].
-GoFLAG data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFLAG markers are to be expected in the near future.
+GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.
 The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data (`r figure("genbank")`).
@@
 FTOL is available via the "ftol" R package, as well as a web portal (FIXME: add URL when ready).
+Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibration points) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom DNA sequences [`r eg`\; @Nitta2021.08.26.457744].
 
@@
 During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/joelnitta/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand).
-We selected R because it is widely used by the ecological research community, well established, and freely available.
+We selected R because it is widely used by the ecological research community, well established, and freely available [@Lai2019].
 The "pteridocat" package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
@@
 
-We want to make it clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
-FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphologic, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
+We want to be clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
+FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphological, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
 
@@
 Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript is not meant to be the last.
-Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree regularly (`ie`, semi-annually).
+Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree on a regular, semi-annual basis.
+We envision that FTOL will be integrated with the next iteration of PPG, PPGII, to provide the most recent hypotheses on the monophyly of various fern taxa, which will in turn enable a more natural classification system.
 
@@
 We are aware that our methodology cannot produce a "perfect" tree, nor is that our goal.
-Each round of analysis involves a step to manually inspect the tree and make taxonomic modifications (`ie`, change opinion regarding synonyms) and flag rogue accessions for exclusion.
-However, it is impossible to catch everything since it is possible the subsequent round of GenBank mining will recover additional rogues among the replacements for the excluded sequences.
-Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked, or proper taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
+Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked (`r eg`, placement of species within genera, which we did not have the resources to inspect), or proper taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
 That is why we have made our methodology (code), data, and software (R packages and docker image) completely open and available to the research community.
-It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through GitHub (https://github.com/joelnitta/ftol) or by contacting us by email.
+It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through the GitHub repository (https://github.com/joelnitta/ftol).
 This way, FTOL will continually improve and keep pace with the currently available data and taxonomic hypotheses of ferns.
@@
 Numbers next to each bar show sampled species out of total number of species.
-Taxonomy follows [@PteridophytePhylogenyGroupI2016].
+Taxonomy follows @PteridophytePhylogenyGroupI2016.
 
@@
 Outgroup (seed plants, lycophytes, and bryophytes) not shown.
-Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`. Taxonomy follows [@PteridophytePhylogenyGroupI2016]; informal clade names in lowercase.
+Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`. Taxonomy follows @PteridophytePhylogenyGroupI2016; informal clade names in lowercase.
 
@@
 **`r figure("divtimes")`**.
-Stem age of fern families (mya) estimated by selected studies.
+Stem age of fern families (Ma) estimated by selected studies.
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].

### 2022-02-16 _targets.R

@@ -41,6 +41,12 @@ tar_plan(
 		pteridocat,
 		store = ftol_cache
 	),
+  tar_files_input(
+  	fossil_ferns_path,
+  	fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
+  ),
+	fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
+		janitor::clean_names(),
 	
 	# GenBank ----
 	# Analyze number of fern accessions and species in GenBank by type of DNA 
@@ -58,7 +64,8 @@ tar_plan(
 		pattern = map(gb_query)
 	),
 	# - Load NCBI taxonomy
-	tar_file(taxdump_zip_file, "working/taxdmp_2022-02-01.zip"),
+	tar_file(taxdump_zip_file, fs::path(
+		ftol_cache, "user/data_raw/taxdmp_2022-02-01.zip")),
 	ncbi_names = load_ncbi_names(
 		taxdump_zip_file,
 		unique(gb_taxa$taxid)

### 2022-02-16 ms/manuscript.Rmd
Add SI pdf

@@
 params:
-  doc_type: doc
+  doc_type: manual
 ---
@@
 	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
-
 # Load packages
-library(conflicted)
-library(rmarkdown)
-library(bookdown)
-library(targets)
-library(assertr)
-library(ape)
-library(scales)
-library(tidyverse)
-library(rgnparser)
-library(ggtree)
-library(glue)
-library(Polychrome)
-library(patchwork)
-library(magrittr)
-library(deeptime)
-library(ggrepel)
-library(ggokabeito)
-library(ggtext)
-library(janitor)
-
-# Resolve conflicts
-conflict_prefer("filter", "dplyr")
-conflict_prefer("extract", "magrittr")
-
-# Load functions
+source(here::here("R/ms_packages.R"))
+```
+
+```{r load-functions, eval = params$doc_type == "manual"}
+# Load functions (only if running interactively)
 source(here::here("R/functions.R"))
@@
 	c(
+		accs_exclude,
+		fern_sanger_seqs_raw,
+		raw_fasta_all,
+		sanger_seqs_combined_filtered,
+		sanger_seqs_rogues_removed,
+		sanger_accessions_selection,
+		plastome_seqs_combined_filtered,
+		plastome_metadata_raw_all,
+		plastome_metadata_renamed,
 		sanger_alignment,
@@
 All sequences were downloaded from GenBank.
-GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+GenBank queries were formatted including locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
 Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
@@
 # Results
+
+## GenBank mining
+
+```{r genbank-stats}
+# Helper function to count number of GenBank accessions in a dataset
+count_accs <- function(data) {
+	data %>%
+	pull(accession) %>%
+	n_distinct() %>%
+	number()
+}
+
+# Manual list of accessions to exclude
+n_accs_exclude <-
+accs_exclude %>%
+	pull(accession) %>%
+	n_distinct() %>%
+	number()
+
+# Sanger ----
+
+# Number of GenBank accessions resulting from inital query
+n_accs_query <- count_accs(fern_sanger_seqs_raw)
+
+# Number of GenBank accessions recovered with superCRUNCH
+n_accs_extract <- count_accs(raw_fasta_all)
+
+# Number of GenBank accessions after excluding manual rogues,
+# names that couldn't be resolved
+n_accs_first_filter <- count_accs(sanger_seqs_combined_filtered)
+
+# Number of GenBank accessions after further excluding
+# rogues based on BLAST
+n_accs_second_filter <- count_accs(sanger_seqs_rogues_removed)
+
+# Number of GenBank accessions in final selection
+# first convert to long form
+sanger_accessions_selection_long <-
+ sanger_accessions_selection %>%
+	select(contains("accession")) %>%
+	pivot_longer(values_to = "accession", everything()) %>%
+	filter(!is.na(accession))
+
+# Count number of raw accessions in final selection
+n_accs_raw_final <- count_accs(sanger_accessions_selection_long)
+
+# Count number of accessions in final selection, split by locus
+n_accs_final <- 
+	sanger_accessions_selection_long %>%
+	# Count accessions containing different loci as distinct
+	mutate(
+		name = str_remove_all(name, "accession_"),
+		accession = paste(name, accession, sep = "_")) %>%
+	count_accs()
+
+# Analyze join by method
+join_by_count <-
+sanger_accessions_selection %>% 
+	assert(not_na, join_by, species) %>%
+	assert(is_uniq, species) %>%
+	tabyl(join_by) %>%
+	mutate(join_by = fct_reorder(join_by, n)) %>%
+	mutate(
+			percent = percent(percent, accuracy = 0.1),
+			n = number(n, accuracy = 1)) %>%
+	split(.$join_by)
+
+# Plastome ----
+# Don't include OG in calculations
+n_plast_accs_query <-
+	plastome_metadata_raw_all %>%
+	left_join(
+		select(plastome_metadata_renamed, accession, outgroup),
+		by = "accession") %>%
+	# excluded accessions lack outgroup info, but are all ferns
+	mutate(outgroup = replace_na(FALSE)) %>%
+	assert(not_na, outgroup) %>%
+	filter(outgroup == FALSE) %>%
+	count_accs()
+n_plast_first_filter <-
+	plastome_metadata_renamed %>%
+  assert(not_na, outgroup) %>%
+	filter(outgroup == FALSE) %>%
+	count_accs()
+n_accs_plast_final <- 
+	plastome_seqs_combined_filtered %>%
+	left_join(
+		select(plastome_metadata_renamed, accession, outgroup),
+		by = "accession") %>%
+	assert(not_na, outgroup) %>%
+	filter(outgroup == FALSE) %>%
+	count_accs()
+```
+
+The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions ("accessions" in this case includes some cases where a single accession contains multiple loci).
+Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
+Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
+`r n_accs_first_filter` accessions were retained after excluding these rogues and species names that could not be resolved to an accepted name in the taxonomic database.
+`r n_accs_second_filter` accessions were retained after excluding rogues identified by the all-by-all BLAST search.
+The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately).
+The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`).
+`r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).
+
+The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions.
+`r n_plast_first_filter` accessions were retained after excluding rogues and species names that could not be resolved.
+Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species).
 
@@
 Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
-The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species).
+The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
 The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
@@
 # Calculate frequency of missing bases in each dataset overall
-plastome_missing_freq <- base_freq_missing(plastome_alignment)
-sanger_missing_freq <- base_freq_missing(sanger_alignment)
+plastome_missing_freq <- base_freq_missing(plastome_alignment) %>%
+	percent(accuracy = 0.1)
+sanger_missing_freq <- base_freq_missing(sanger_alignment) %>%
+	percent(accuracy = 0.1)
 
@@
 
-The Sanger DNA alignment was `r number(ncol(sanger_alignment))` bp with `r percent(sanger_missing_freq, accuracy = 0.1)` missing data (missing bases or gaps) overall; rates of missing data by locus ranged from `r sanger_missing_by_gene %>% first() %>% extract2("missing")` (`r sanger_missing_by_gene %>% first() %>% extract2("gene")`) to `r sanger_missing_by_gene %>% last() %>% extract2("missing")` (`r sanger_missing_by_gene %>% last() %>% extract2("gene")`).
-The plastome DNA alignment was `r number(ncol(plastome_alignment))` bp with `r percent(plastome_missing_freq, accuracy = 0.1)` missing data.
+The Sanger DNA alignment was `r number(ncol(sanger_alignment))` bp with `r sanger_missing_freq` missing data (missing bases or gaps) overall; rates of missing data by locus ranged from `r sanger_missing_by_gene %>% first() %>% extract2("missing")` (`r sanger_missing_by_gene %>% first() %>% extract2("gene")`) to `r sanger_missing_by_gene %>% last() %>% extract2("missing")` (`r sanger_missing_by_gene %>% last() %>% extract2("gene")`).
+The plastome DNA alignment was `r number(ncol(plastome_alignment))` bp with `r plastome_missing_freq` missing data.
 

### 2022-02-16 ms/si.Rmd
Add SI pdf

@@
 title: "Supplemental Information"
+mainfont: Roboto-Regular.ttf
+fontsize: 11pt
+mainfontoptions: 
+  - BoldFont=Roboto-Bold.ttf
+  - ItalicFont=Roboto-Italic.ttf
+  - BoldItalicFont=Roboto-BoldItalic.ttf
 output: 
-  bookdown::word_document2:
-    pandoc_args: [ "--csl", "frontiers.csl"]
+  bookdown::pdf_document2:
+    citation_package: default
     toc: no
     number_sections: no
-    reference_docx: Frontiers_Template.docx
+    fig_caption: yes
+    keep_tex: no
+    latex_engine: xelatex
+    pandoc_args: [ "--csl", "frontiers.csl"]
+header-includes: 
+  - \usepackage{float} # keep figures from floating away
+  - \usepackage{booktabs} # need for kable tables
+  - \usepackage{colortbl} # need for kable tables
+  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
+  - \usepackage{fancyhdr} # enable header
+  - \pagestyle{fancy}  # enable header
+  - \fancypagestyle{plain}{\pagestyle{fancy}} # show header on first page
+  - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
+  - \fancyhead[RE,RO]{Nitta \textit{et al}. Fern Tree of Life SI} # add header on right side
+  - \fancyhead[LO,LE]{} # suppress automatic section header on left side
+  - \cfoot{\thepage} # footer
 editor_options: 
@@
 
-```{r setup, include = FALSE}
+```{r setup-si, include = FALSE}
 knitr::opts_chunk$set(
 	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
-
 # Load packages
-library(conflicted)
-library(rmarkdown)
-library(bookdown)
-library(targets)
-library(assertr)
-library(ape)
-library(scales)
-library(tidyverse)
-library(rgnparser)
-library(ggtree)
-library(glue)
-library(Polychrome)
-library(patchwork)
-library(magrittr)
-library(ggrepel)
-library(ggokabeito)
-library(ggforce)
-
-# Resolve conflicts
-conflict_prefer("filter", "dplyr")
-conflict_prefer("extract", "magrittr")
-
-# Load functions
-source(here::here("R/ms_functions.R"))
-
-# Specify percentage format
-percent <- function(...) {scales::percent(...)}
-number <- function(...) {scales::number(big.mark = ",", ...)}
+library(ggupset)
+library(kableExtra)
+source(here::here("R/ms_packages.R"))
 ```
 
-```{r load-targets}
+```{r load-functions-si, eval = params$doc_type == "manual"}
+# Load functions (only if running interactively)
+source(here::here("R/functions.R"))
+```
+
+```{r load-targets-si}
 # Targets from FTOL workflow
@@
 	c(
+		sanger_alignment_tbl,
 		sanger_sampling,
-		monophy_by_clade,
-		plastid_tree_dated),
+		sanger_accessions_selection),
 	store = ftol_cache
@@
 
-```{r age-biplot}
-# Count number of tips in clades by taxonomy for double checking
-n_tips_in_clades <-
-	sanger_sampling %>%
-	filter(outgroup == FALSE) %>%
-	select(species, family, order) %>%
-	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
-	group_by(taxon) %>%
-	summarize(number_tips_by_tax = n())
+```{r load-captions, cache = FALSE}
+# Source captions numbers from MS
+# 
+# So that caption numbers are in order of appearance in MS
 
-# Get ages for large clades
-clade_ages <-
-	sanger_sampling %>%
-	select(family, order) %>%
-	unique() %>%
-	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
-	filter(!is.na(taxon)) %>%
-	unique() %>%
-	left_join(
-		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
-		by = "taxon") %>%
-	mutate(across(c(mrca, number_tips), parse_number)) %>%
-	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
-	filter(monophyly == "Yes") %>%
-	mutate(
-		# Height is distance above root
-		height = map_dbl(mrca, ~phytools::nodeheight(
-			tree = plastid_tree_dated, node = .))
+# Run inline code when purling
+options(knitr.purl.inline = TRUE)
+
+# Make a temporary file to write out just inline R code from the SI Rmd
+temp_file <- tempfile()
+
+# Generate R script including inline R code from the ms Rmd
+knitr::purl(here::here("ms/manuscript.Rmd"), output = temp_file)
+
+# Trim this R script down to only functions that define figure and table captions
+read_lines(temp_file) %>%
+  magrittr::extract(
+    str_detect(., "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(")
   ) %>%
-	mutate(
-		# Age is the total length of the tree minus height
-		age = max(phytools::nodeHeights(plastid_tree_dated)) - height
-	) %>%
-	# Double check that n_tips is as expected
-	# (so the correct node is being used)
-	left_join(n_tips_in_clades, by = "taxon") %>%
-	verify(all(number_tips == number_tips_by_tax)) %>%
-	select(-number_tips_by_tax)
+  unique %>%
+  write_lines(temp_file)
 
-clade_ages %>%
-	filter(level == "family") %>%
-	select(family = taxon, current_best = age) %>%
-	left_join(other_dates, by = "family") %>%
-	select(matches("best|median")) %>%
-	ggplot(aes(x = .panel_x, y = .panel_y)) +
-	geom_point() +
-	geom_autodensity(na.rm = TRUE) +
-	facet_matrix(vars(everything()), layer.diag = 2, grid.y.diag = FALSE)
+# Source the code. This will load the caption functions in the right order.
+source(temp_file, local = TRUE)
+
+# Delete the temporary file
+fs::file_delete(temp_file)
 ```
 
-Figure 1: Biplot comparing crown ages of fern families (millions of years) across different studies.
+```{r join_by, results = "as-is"}
+# Make table showing method of joining accessions across loci
+sanger_accessions_selection %>% 
+	left_join(select(sanger_sampling, species, outgroup), by = "species") %>%
+	# FIXME why is this NA?
+	# filter(is.na(outgroup)) %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup) %>%
+	assert(not_na, join_by, species) %>%
+	assert(is_uniq, species) %>%
+	tabyl(join_by) %>%
+	arrange(n) %>%
+	adorn_totals() %>%
+	mutate(
+		join_by = str_to_sentence(join_by) %>%
+			str_replace_all("Monophy", "Monophyletic"),
+		n = number(n),
+		percent = case_when(
+			join_by != "Total" ~ percent(percent) %>% 
+				str_replace_all("%", "\\\\%"),
+			TRUE ~ ""
+		)
+	) %>%
+	rename(
+		"Method" = join_by,
+		`\\textit{n}` = n,
+		`\\%` = percent
+	) %>%
+  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
+  kable_styling(latex_options = c("striped", "HOLD_position"))
+```
+
+**`r s_table("join_by")`**. Methods used to join accessions across loci for Sanger dataset. Count is number of species.
+
+\clearpage
+
+```{r locus-upset, fig.width = 7, fig.height = 5}
+# Make tibble with one row per locus per species
+sanger_alignment_tbl %>%
+	mutate(taxon = map(align_trimmed, rownames)) %>%
+	select(locus = target, species = taxon) %>%
+	unnest(cols = species) %>%
+	# Exclude outgroups
+	left_join(
+		unique(select(sanger_sampling, species, outgroup)), by = "species") %>%
+	filter(outgroup == FALSE) %>%
+	select(-outgroup) %>%
+	group_by(species) %>%
+	summarize(locus = list(locus)) %>%
+	ggplot(aes(x = locus)) +
+	geom_bar() +
+	labs(x = "Locus combination", y = "Count (num. species)") +
+	scale_x_upset(n_intersections = 20)
+```
+
+**`r s_figure("locus-upset")`**.
+Frequency of locus combinations by species, Sanger dataset.
+Only top 20 combinations shown.

### 2022-02-16 R/functions.R

@@ -707,16 +707,16 @@ get_stem_family_age <- function(
 
 # Formatting ----
 
-# Specify percentage format
-percent <- function(...) {scales::percent(...)}
-
-# Specify number format
-number <- function(...) {scales::number(big.mark = ",", ...)}
-
 # Abbreviations
 ie <- "*i.e.*"
 eg <- "*e.g.*"
 ca <- "*ca.*"
+
+# Specify custom percentage format
+percent <- function(...) {scales::percent(...)}
+
+# Specify custom number format
+number <- function(...) {scales::number(big.mark = ",", ...)}
 
 #' Convert a data.frame of counts to percentages, with automatic formatting
 #'
@@ -834,8 +834,8 @@ result_file <- function (result_num, extension) {
 # (except I never use the caption in the function, and instead replace with 'blank')
 figure_full <- captioner::captioner(prefix = "Figure ", auto_space = FALSE)
 table_full <- captioner::captioner(prefix = "Table ", auto_space = FALSE)
-s_figure_full <- captioner::captioner(prefix = "Figure S ", auto_space = FALSE)
-s_table_full <- captioner::captioner(prefix = "Table S ", auto_space = FALSE)
+s_figure_full <- captioner::captioner(prefix = "Figure S", auto_space = FALSE)
+s_table_full <- captioner::captioner(prefix = "Table S", auto_space = FALSE)
 
 # - Make a short function that prints only the object type and number, e.g., "Fig. 1"
 figure <- pryr::partial(figure_full, display = "cite", caption = "blank")
@@ -952,7 +952,7 @@ get_tips_in_ape_plot_order <- function(tree) {
 base_freq_missing <- function(
 	seqs, missing_base_codes = c("n", "-", "?"), 
 	start = 1, end = ncol(seqs),
-	freq = TRUE) {
+	freq = FALSE) {
 	assertthat::assert_that(is.matrix(seqs))
 	assertthat::assert_that(end > start)
 	base_freq <- ape::base.freq(seqs[, start:end], all = TRUE, freq = freq)

### 2022-02-16 R/ms_packages.R

@@ -1,0 +1,24 @@
+# Load packages
+library(conflicted)
+library(rmarkdown)
+library(bookdown)
+library(targets)
+library(assertr)
+library(ape)
+library(scales)
+library(tidyverse)
+library(rgnparser)
+library(ggtree)
+library(glue)
+library(Polychrome)
+library(patchwork)
+library(magrittr)
+library(deeptime)
+library(ggrepel)
+library(ggokabeito)
+library(ggtext)
+library(janitor)
+
+# Resolve conflicts
+conflict_prefer("filter", "dplyr")
+conflict_prefer("extract", "magrittr")

### 2022-02-16 _targets.R

@@ -115,5 +115,10 @@ tar_plan(
 		ms,
 		"ms/manuscript.Rmd",
 		output_dir = "results"
+	),
+	tar_render(
+		si,
+		"ms/si.Rmd",
+		output_dir = "results"
 	)
 )

### 2022-02-17 ms/manuscript.Rmd
Add SI fig with Du et al 2021 dates

@@
 		fossil_calibration_tips,
+		ts_fossil_calibration_tips,
 		treepl_cv_results),
@@
 
-We dated the ML tree using treePL [@Smith2012].
+We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
 We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa.
@@
 
-We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (`r figure("divtimes")`).
+We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
@@
 Moreover, our set of calibration points more than doubles the number used by previous studies.
-It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).
-
-<<FIXME: add discussion of specific groups once we have final ages>>
+It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+
+<<FIXME: add discussion of some specific groups once we have final ages>>
 
@@
 ```{r plot-prep}
+## ---- plot-prep-main
+
 # Make vector of outgroup taxa
@@
 	assert(is_uniq, code)
+
+# Make tibble for annotating major clades
+clade_tibble <- 
+	# Use assess_monophy to get MRCA of each major clade
+	# They should all by monophyletic anyways
+	select(sanger_sampling, species, major_clade) %>%
+	assess_monophy(
+		taxon_sampling = .,
+		tree = plastid_tree_dated,
+		tax_levels = "major_clade"
+	) %>%
+	get_result_monophy(1) %>%
+	# Exclude lycophytes
+	filter(!taxon %in% c(
+		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
+	verify(all(monophyly == "Yes")) %>%
+	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
+	left_join(clade_cols_tib, by = "major_clade")
 ```
@@
 # Tree plot ----
-# Make tibble for annotating major clades
-clade_tibble <- 
-	# Use assess_monophy to get MRCA of each major clade
-	# They should all by monophyletic anyways
-	select(sanger_sampling, species, major_clade) %>%
-	assess_monophy(
-		taxon_sampling = .,
-		tree = plastid_tree_dated,
-		tax_levels = "major_clade"
-	) %>%
-	get_result_monophy(1) %>%
-	filter(!taxon %in% c(
-		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
-	verify(all(monophyly == "Yes")) %>%
-	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
-	left_join(clade_cols_tib, by = "major_clade")
 
@@
 ```{r div-times-make}
-# Convert family ages to long format
-family_ages_median <-
-family_stem_ages %>%
-	select(family, current_best = age) %>%
-	left_join(other_dates, by = "family") %>%
+# Convert family ages to almost-long format
+
+# Helper function for subsetting dates to get into
+# almost-long format
+subset_ages <- function(other_dates, author_sel) {
+	other_dates %>%
+		select(family, contains(glue("{author_sel}_"))) %>%
+		rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
+		mutate(author = author_sel)
+}
+
+# Combine dates in almost-long format:
+# one row per family/study combination,
+# cols for median, high, and low date
+all_ages <-
+bind_rows(
+	subset_ages(other_dates, "ts"),
+	subset_ages(other_dates, "rothfels"),
+	select(other_dates, family, median = schuettpelz_best) %>%
+		mutate(author = "schuettpelz"),
+	transmute(
+		family_stem_ages,
+		family,
+		author = "current",
+		median = age
+	)
+) %>%
 	left_join(
-		unique(select(sanger_sampling, family, major_clade)),
-		by = "family"
-	) %>%
-	left_join(clade_cols_tib, by = "major_clade"
-	) %>%
-	mutate(family = fct_reorder(family, current_best)) %>%
-	pivot_longer(values_to = "age", -c(family, color, code, major_clade)) %>%
-	separate(name, into = c("author", "age_type"), sep = "_") %>%
-	assert(not_na, author, age_type) %>%
-	filter(age_type %in% c("best", "median")) %>%
-	assert(not_na, major_clade, color)
-
-# Convert family errors to almost long format
-family_ages_errors <-
-	family_stem_ages %>%
-	select(family, current_best = age) %>%
-	left_join(other_dates, by = "family") %>%
-	mutate(family = fct_reorder(family, current_best)) %>%
-	select(family, matches("ts|rothfels")) %>%
-	select(-matches("median"))
-
-family_ages_errors <-
-	bind_rows(
-		transmute(family_ages_errors, family,
-					 high = ts_high, low = ts_low, author = "ts"),
-		transmute(family_ages_errors, family,
-					 high = rothfels_high, low = rothfels_low, author = "rothfels")
-	)
+		unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
+	left_join(clade_tibble, by = "major_clade") %>%
+	# (some median dates are NA if that family wasn't included in a study)
+	assert(not_na, family, major_clade, code, author) %>%
+	# Set family as a factor ordered by current age estimate
+	left_join(family_stem_ages, by = "family") %>%
+	mutate(family = fct_reorder(family, age)) %>%
+	select(-age)
 
@@
 family_labs_tbl <-
-family_ages_median %>%
+all_ages %>%
 	select(family, code) %>%
@@
 	mutate(label = glue("{family} ({code})"))
+
 family_labs <- family_labs_tbl$label %>%
@@
 div_times_plot <-
-ggplot(mapping = aes(y = family)) +
-	geom_errorbarh(
-		data = family_ages_errors,
-		aes(xmax = high, xmin = low, color = author),
-		position = position_dodge(width = 0.5)
-	) +
-	geom_point(
-		data = filter(family_ages_median, age_type %in% c("best", "median")),
-		aes(x = age, color = author, shape = author),
+ggplot(all_ages) +
+	geom_pointrange(
+		aes(
+			x = median, xmin = low, xmax = high,
+			y = family,
+			color = author, shape = author),
 		position = position_dodge(width = 0.5)) +
@@
 		labels = ages_sampling$label) +
-	scale_y_discrete(labels = family_labs) +
+	scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
 	scale_x_reverse("Age (millions of years)") +
@@
 		height = unit(1, "lines"),
+		# FIXME: change this to true when issue gets resolved
+		# https://github.com/willgearty/deeptime/issues/41
+		expand = FALSE,
 		size = 3) +

### 2022-02-17 ms/si.Rmd
Add SI fig with Du et al 2021 dates

@@
 		sanger_sampling,
-		sanger_accessions_selection),
+		sanger_accessions_selection,
+		plastome_tree,
+		plastid_tree_dated),
 	store = ftol_cache
@@
 		ref_files,
-		other_dates
+		other_dates,
+		du_dates_all,
+		family_stem_ages
 	),
@@
 
+```{r plot-prep-load}
+# Load the plot prep code from the main MS
+knitr::read_chunk(
+	here::here("ms/manuscript.Rmd"),
+	labels = "plot-prep-main",
+	from = "og_taxa <-",
+	to = "left_join\\(clade_cols_tib"
+	)
+```
+
+<!-- Run the plot prep code -->
+```{r, plot-prep-main}
+```
+
 ```{r locus-upset, fig.width = 7, fig.height = 5}
@@
 Only top 20 combinations shown.
+
+\clearpage
+
+```{r du-ages, fig.width = 7, fig.height = 5}
+# Helper function for subsetting dates to get into
+# almost-long format
+subset_ages <- function(other_dates, author_sel) {
+	other_dates %>%
+		select(family, contains(glue("{author_sel}_"))) %>%
+		rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
+		mutate(author = author_sel)
+}
+
+# Filter Du ages to just stem families
+du_dates_stem_fam <-
+	du_dates_all %>%
+	filter(
+		str_detect(affinities, "eae$"),
+		affinities_group == "stem",
+		str_detect(author, "^r")) %>%
+	select(family = affinities, author, median, low, high)
+
+# Make tibble of Du
+du_schemes <- du_dates_stem_fam %>%
+	select(scheme = author) %>%
+	unique() %>%
+	mutate(
+		code = str_split(scheme, "_") %>%
+			map_chr(last) %>%
+			str_to_upper()) %>%
+	arrange(code)
+
+# Combine dates in almost-long format:
+# one row per family/author combination,
+# cols for median, high, and low date
+all_ages_with_du <-
+	bind_rows(
+		du_dates_stem_fam,
+		subset_ages(other_dates, "ts"),
+		subset_ages(other_dates, "rothfels"),
+		select(other_dates, family, median = schuettpelz_best) %>%
+			mutate(author = "schuettpelz"),
+		transmute(
+			family_stem_ages,
+			family,
+			author = "current",
+			median = age
+		)
+	) %>%
+	filter(family %in% du_dates_stem_fam$family) %>%
+	left_join(
+		unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
+	left_join(clade_tibble, by = "major_clade") %>%
+	# (some median dates are NA if that family wasn't included in a author)
+	assert(not_na, family, major_clade, code, author) %>%
+	# Set family as a factor ordered by current age estimate
+	left_join(family_stem_ages, by = "family") %>%
+	mutate(family = fct_reorder(family, age)) %>%
+	select(-age)
+
+# Define shapes and colors for points
+du_ages_colors <- c(
+	current = palette_okabe_ito(order = "vermillion"),
+	rothfels = palette_okabe_ito(order = "bluishgreen"),
+	schuettpelz = palette_okabe_ito(order = "blue"),
+	ts = palette_okabe_ito(order = "orange"),
+	rep(
+		palette_okabe_ito(order = "gray"),
+		length(du_schemes$scheme)) %>%
+		set_names(du_schemes$scheme)
+)
+
+du_ages_shapes <- c(
+	current = 16, # circle
+	rothfels = 16,
+	schuettpelz = 16,
+	ts = 16,
+	seq_along(du_schemes$scheme) %>%
+		subtract(1) %>%
+		set_names(du_schemes$scheme)
+)
+
+# Cross doesn't work because we can't distinguish error bars
+du_ages_shapes[du_ages_shapes == 3] <- 6
+
+# Format names of papers for printing
+du_ages_sampling <-
+	tribble(
+		~source, ~year, ~label,
+		"schuettpelz", 2009, "Schuettpelz & Pryer 2009",
+		"rothfels", 2015, "Rothfels et al. 2015",
+		"ts", 2016, "Testo and Sundue 2016",
+		"current", 2021, "This study",
+	) %>%
+	bind_rows(
+		du_schemes %>%
+			mutate(
+				year = 2020 + 0.1*seq_along(du_schemes$scheme),
+				label = paste("Du et al. 2021 scheme", code)
+			) %>%
+			transmute(source = scheme, year, label)
+	) %>%
+	mutate(source = fct_reorder(source, year))
+
+# Format y-axis text: show major clade in parentheses
+family_labs_tbl <-
+	all_ages_with_du %>%
+	select(family, code) %>%
+	unique() %>%
+	mutate(label = glue("{family} ({code})"))
+
+family_labs <- family_labs_tbl$label %>%
+	set_names(., family_labs_tbl$family)
+
+ggplot(all_ages_with_du) +
+	geom_pointrange(
+		aes(
+			x = median, xmin = low, xmax = high,
+			y = family,
+			color = author, shape = author),
+		position = position_dodge(width = 0.5)) +
+	scale_shape_manual(
+		values = du_ages_shapes,
+		limits = du_ages_sampling$source, 
+		labels = du_ages_sampling$label) +
+	scale_color_manual(
+		values = du_ages_colors,
+		limits = du_ages_sampling$source, 
+		labels = du_ages_sampling$label) +
+	scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
+	scale_x_reverse("Age (millions of years)") +
+	coord_geo(
+		dat = list("periods", "eras"),
+		pos = list("b", "b"),
+		abbrv = list(TRUE, FALSE),
+		xlim = c(250, 0),
+		fill = "transparent",
+		color = "grey20",
+		lab_color = "grey30",
+		height = unit(1, "lines"),
+		# FIXME: change this to true when issue gets resolved
+		# https://github.com/willgearty/deeptime/issues/41
+		expand = FALSE,
+		size = 3) +
+	theme_bw() +
+	theme(
+		axis.title.y = element_blank(),
+		legend.title = element_blank()
+	)
+```
+
+**`r s_figure("du-ages")`**.
+Stem age of fern families (Ma) estimated by selected studies, only including families that were sampled across all studies.
+For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
+For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
+@Du2021 used six different dating schemes; for details, see that paper.
+Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
+Period name abbreviations as follows: FIXME.

### 2022-02-17 R/functions.R

@@ -252,6 +252,48 @@ parse_ts_dates <- function(testo_sundue_2016_si_path) {
 		)
 }
 
+#' Parse divergene dates in the Du et al. 2021 SI file
+#' 
+#' Du X, Lu J, Zhang L, et al (2021) Simultaneous diversification of 
+#' Polypodiales and angiosperms in the Mesozoic. Cladistics 37:518–539. 
+#' https://doi.org/10/gpb22j
+#'
+#' @param du_2021_si_path Path to Du et al. 2021 SI file
+#'
+#' @return Tibble
+#' 
+parse_du_dates <- function(du_2021_si_path) {
+	# Extract dates in original format
+	du_dates_raw <-
+		docxtractr::read_docx(du_2021_si_path) %>%
+		docxtractr::docx_extract_all_tbls(
+			guess_header = TRUE, preserve = FALSE, trim = TRUE) %>%
+		# Table of interest is the third one in the SI
+		magrittr::extract2(3)
+	
+	# Convert to longish format
+	du_dates_raw %>%
+		janitor::clean_names() %>%
+		pivot_longer(names_to = "author", values_to = "age", -lineage) %>%
+		mutate(age = str_remove_all(age, "\\(|,|\\)|ca\\.") %>%
+					 	str_squish()) %>%
+		separate(age,
+						 c("median", "low", "high"),
+						 sep = " ", fill = "right", convert = TRUE) %>%
+		mutate(
+			median = na_if(median, "\\") %>%
+				parse_number(),
+			affinities_group = str_match(
+				lineage,
+				"stem|crown") %>%
+				magrittr::extract(,1),
+			affinities = str_remove_all(lineage, "The|stem|crown|of") %>%
+				str_squish()
+		) %>%
+		assert(not_na, affinities_group) %>%
+		select(author, lineage, affinities, affinities_group, median, low, high)
+}
+
 # Coverage ----
 
 #' Make a tibble of accepted species in pteridocat taxonomic database
@@ -846,8 +888,8 @@ s_table <- pryr::partial(s_table_full, display = "cite", caption = "blank")
 # - Make a short function that prints only the number (e.g., "1")
 figure_num <- function(name) {figure(name) %>% str_remove("Figure ")}
 table_num <- function(name) {table(name) %>% str_remove("Table ")}
-s_figure_num <- function(name) {s_figure(name) %>% str_remove("Figure S ")}
[…]

### 2022-02-17 _targets.R

@@ -41,10 +41,10 @@ tar_plan(
 		pteridocat,
 		store = ftol_cache
 	),
-  tar_files_input(
-  	fossil_ferns_path,
-  	fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
-  ),
+	tar_files_input(
+		fossil_ferns_path,
+		fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
+	),
 	fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
 		janitor::clean_names(),
 	
@@ -107,6 +107,11 @@ tar_plan(
 	# - Stem age of fern families
 	family_stem_ages = get_stem_family_age(
 		sanger_sampling, plastid_tree_dated, ppgi_taxonomy),
+	# - Read in Du 2021 dates
+	tar_files_input(
+		du_2021_si_path,
+		"_targets/user/data_raw/cla12457-sup-0003-tables1-s5.docx"),
+	du_dates_all = parse_du_dates(du_2021_si_path),
 	
 	# Render manuscript ----
 	# track files used for rendering MS

### 2022-02-17 ms/manuscript.Rmd
Add PDF versions of ms and SI
For some weird reason need template .docx in two places :/

@@
 title: "An open and continuously updated fern tree of life (FTOL)"
+mainfont: Roboto-Regular.ttf
+fontsize: 11pt
+mainfontoptions: 
+  - BoldFont=Roboto-Bold.ttf
+  - ItalicFont=Roboto-Italic.ttf
+  - BoldItalicFont=Roboto-BoldItalic.ttf
 output: 
+  bookdown::pdf_document2:
+    citation_package: default
+    toc: no
+    number_sections: yes
+    fig_caption: yes
+    keep_tex: no
+    latex_engine: xelatex
+    pandoc_args: [ "--csl", "frontiers.csl"]
   bookdown::word_document2:
@@
     number_sections: no
-    reference_docx: Frontiers_Template.docx
+    reference_docx: Frontiers_Template_SI.docx
+header-includes: 
+  - \usepackage{float} # keep figures from floating away
+  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
+  - \usepackage{fancyhdr} # enable header
+  - \cfoot{\thepage} # footer
 editor_options: 
@@
 We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
-We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa.
+We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`).
 In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
@@
 
-# References
+# Acknowledgments {-}
+
+Members of the Iwasaki lab provided helpful comments that improved the manuscript.
+This study was supported by JSPS KAKENHI Grant Number 16H06279.
+The authors thank Michael Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.
+
+# References {-}
 <!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
@@
 
-# Figures
+`r pagebreak_pdf(params)`
+
+# Figures {-}
 
@@
 
+`r pagebreak_pdf(params)`
+
 ```{r ftol-make}
@@
 Taxonomy follows @PteridophytePhylogenyGroupI2016.
+
+`r pagebreak_pdf(params)`
 
@@
 
+`r pagebreak_pdf(params)`
+
 ```{r div-times-make}

### 2022-02-17 ms/si.Rmd
Add PDF versions of ms and SI
For some weird reason need template .docx in two places :/

@@
 ---
-title: "Supplemental Information"
+title: "*Supplementary Material*"
 mainfont: Roboto-Regular.ttf
@@
     pandoc_args: [ "--csl", "frontiers.csl"]
+  bookdown::word_document2:
+    pandoc_args: [ "--csl", "frontiers.csl"]
+    toc: no
+    number_sections: no
+    reference_docx: Frontiers_Template_SI.docx
 header-includes: 
@@
   - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
-  - \fancyhead[RE,RO]{Nitta \textit{et al}. Fern Tree of Life SI} # add header on right side
+  - \fancyhead[RE,RO]{Supplementary Material} # add header on right side
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
@@
 library(ggupset)
-library(kableExtra)
+library(flextable)
 source(here::here("R/ms_packages.R"))
@@
 		plastome_tree,
-		plastid_tree_dated),
+		plastid_tree_dated,
+		fossil_calibration_tips),
 	store = ftol_cache
@@
 		du_dates_all,
-		family_stem_ages
+		family_stem_ages,
+		fossil_ferns_raw
 	),
@@
 
-```{r join_by, results = "as-is"}
+```{r fossils}
+# Filter raw fossil data to just fossils used, write out as CSV
+fossils_table <-
+fossil_ferns_raw %>%
+	inner_join(
+		select(fossil_calibration_tips, fossil_taxon), by = "fossil_taxon") %>%
+	verify(nrow(.) == nrow(fossil_calibration_tips)) %>%
+	rowwise() %>%
+	mutate(
+		note = jntools::paste3(notes_21, notes_22sep = ". ") %>%
+			str_replace_all("\\.\\.", "\\.") %>%
+			na_if(".") %>%
+			str_replace_all(" \\.$", "\\.")
+	) %>%
+	ungroup() %>%
+	select(
+		n_fos, minimum_age, node_calibrated, fossil_taxon, 
+		lower = lower_limit_of_oldest_stratigraphic_age,
+		upper = upper_limit_of_oldest_stratigraphic_age,
+		stratigraphic_age, 
+		country, type_locality, formation,
+		organs, specimen_holotype, collection,
+		reference, note
+	)
+
+write_csv(fossils_table, result_file(s_table("fossils"), "csv"))
+```
+
+**`r s_table("fossils")`**.
+Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv")`).
+"n_fos" is the fossil ID number used in our fossil database (https://github.com/pteridophy/fossils).
+"minimum_age" (Ma) is the age used for calibration; "lower" is the lower limit of the oldest stratigraphic age for the fossil; "upper" is the upper limit of the oldest stratigraphic age for the fossil.
+
+\newpage
+
+```{r join-by, results = "markup"}
 # Make table showing method of joining accessions across loci
@@
 		percent = case_when(
-			join_by != "Total" ~ percent(percent) %>% 
-				str_replace_all("%", "\\\\%"),
+			join_by != "Total" ~ percent(percent),
 			TRUE ~ ""
@@
 		"Method" = join_by,
-		`\\textit{n}` = n,
-		`\\%` = percent
-	) %>%
-  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
-  kable_styling(latex_options = c("striped", "HOLD_position"))
+		`%` = percent) %>%
+	flextable() %>%
+	italic(1,2, part = "header") %>%
+	autofit()
 ```
@@
 
-\clearpage
+\newpage
 
@@
 Period name abbreviations as follows: FIXME.
+
+\newpage

### 2022-02-17 R/functions.R

@@ -893,13 +893,23 @@ s_table_num <- function(name) {s_table(name) %>% str_remove("Table ")}
 
 # Pagebreaks ----
 
+#' Pagebreak for MS Word (doc) only
+#' 
+#' Meant for use within Rmd when rendering.
+#' Breaks a page depending if the output format is MS Word
+#' 
+#' @param rmd_params Parameters set in YAML header or rmarkdown::render(). 
+#' Must include `doc_type` (either 'doc' for MS Word or 'pdf' for PDF output)
+#' 
+pagebreak_doc <- function(rmd_params = params) {ifelse(rmd_params$doc_type == "doc", return("\\newpage"), return(""))}
+
 #' Pagebreak for PDF only
 #' 
 #' Meant for use within Rmd when rendering. Requires parameter variable
 #' `doc_type` to be defined (either 'doc' for MS Word or 'pdf' for PDF output)
 #'
 #' Breaks a page depending if the output format is PDF
-pagebreak_pdf <- function(rmd_params = params) {ifelse(rmd_params$doc_type == "pdf", return("\\clearpage"), return(""))}
+pagebreak_pdf <- function(rmd_params = params) {ifelse(rmd_params$doc_type == "pdf", return("\\newpage"), return(""))}
 
 # Phylogeny ----
 

### 2022-02-17 _targets.R

@@ -117,13 +117,32 @@ tar_plan(
 	# track files used for rendering MS
 	tar_file(ref_files, list.files("ms", "references", full.names = TRUE)),
 	tar_render(
-		ms,
+		ms_doc,
 		"ms/manuscript.Rmd",
-		output_dir = "results"
+		output_dir = "results",
+		output_format = "bookdown::word_document2",
+		params = list(doc_type = "doc")
 	),
 	tar_render(
-		si,
+		ms_pdf,
+		"ms/manuscript.Rmd",
+		output_dir = "results",
+		output_format = "bookdown::pdf_document2",
+		params = list(doc_type = "pdf")
+	),
+	tar_render(
+		si_doc,
 		"ms/si.Rmd",
-		output_dir = "results"
+		knit_root_dir = "ms",
+		output_dir = "results",
+		output_format = "bookdown::word_document2",
+		params = list(doc_type = "doc")
+	),
+	tar_render(
+		si_pdf,
+		"ms/si.Rmd",
+		output_dir = "results",
+		output_format = "bookdown::pdf_document2",
+		params = list(doc_type = "pdf")
 	)
 )

### 2022-02-17 ms/manuscript.Rmd
Update MS

@@
 Furthermore, multiple genera included in PPGI were known to be non-monophyletic and included on a tentative basis [@PteridophytePhylogenyGroupI2016].
-Pteridocat includes newly published genera and some genera that were not recognized by PPGI.
-Differences between pteridocat and PPGI are summarized in Table S1.
+Pteridocat includes newly published genera and some genera that were not recognized by PPGI as well as nothogenera, which were not included in PPGI [@Liu2020a].
+Differences between pteridocat and PPGI are summarized in `r s_table("genera_diff")`.
 

### 2022-02-17 ms/si.Rmd
Update MS

@@
 		plastid_tree_dated,
-		fossil_calibration_tips),
+		fossil_calibration_tips,
+		ppgi_taxonomy),
 	store = ftol_cache
@@
 	c(
+		accepted_species,
 		gb_species_by_year,
@@
 
+```{r genera-diff}
+
+# FIXME: need to update ppgi csv file
+# Add one missing nothogenus
+ppgi_taxonomy_mod <-
+	bind_rows(
+		ppgi_taxonomy,
+		tibble(
+			genus = "Phlebosia",
+			notes = "Nothogenus"
+		)
+	)
+
+# Verify that all accepted genera are accounted for
+accepted_species %>%
+	count(genus, name = "n_species") %>%
+	anti_join(select(ppgi_taxonomy_mod, genus, notes)) %>%
+	verify(nrow(.) == 0, success_fun = success_logical)
+
+# Make table of Pteridocat genera not in PPGI
+pteridocat_not_in_ppgi <-
+accepted_species %>%
+	count(genus, name = "n_species") %>%
+	left_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
+	assert(not_na, notes) %>%
+	assert(is_uniq, genus) %>%
+	filter(notes != "original PPGI 2016") %>%
+	transmute(
+		diff = "Pteridocat not in PPGI",
+		genus,
+		n_species,
+		note = notes
+	)
+
+# Make table of PPGI genera not in Pteridocat
+ppgi_not_in_pteridocat <-
+ppgi_taxonomy_mod %>%
+	filter(notes == "original PPGI 2016",
+				 class == "Polypodiopsida") %>%
+	anti_join(accepted_species, by = "genus") %>%
+	count(genus) %>%
+	transmute(
+		diff = "PPGI not in Pteridocat",
+		genus,
+		# FIXME: need to add n species from PPGI and note
+		n_species = NA,
+		note = NA
+	)
+
+genera_diff_table <- 
+	bind_rows(pteridocat_not_in_ppgi, ppgi_not_in_pteridocat)
+
+write_csv(genera_diff_table, result_file(s_table("genera_diff"), "csv"))
+```
+
+**`r s_table("genera_diff")`**.
+Differences in genera included as accepted in Pteridocat taxonomic database (https://github.com/pteridophy/pteridocat) vs.
+@PteridophytePhylogenyGroupI2016 (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
+"diff" is type of difference (Pteridocat genus not in PPGI or PPGI genus not in Pteridocat).
+"n_species" is the number of accepted species in each genus in the system that includes the genus.
+References in "notes" column provided at end of Supplementary Materials.
+
 ```{r fossils}
@@
 **`r s_table("fossils")`**.
-Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv")`).
+Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv") %>% fs::path_file()`).
 "n_fos" is the fossil ID number used in our fossil database (https://github.com/pteridophy/fossils).
@@
 
-**`r s_table("join_by")`**. Methods used to join accessions across loci for Sanger dataset. Count is number of species.
+**`r s_table("join_by")`**.
+Methods used to join accessions across loci for Sanger dataset.
+Count is number of species.
 

### 2022-02-17 R/functions.R

@@ -313,13 +313,20 @@ get_accepted_species <- function(pteridocat, ppgi_taxonomy) {
 		# then don't need is.na()
 		filter(taxonRank == "species" | is.na(taxonRank)) %>%
 		select(scientificName) %>%
+		mutate(scientificName = str_squish(scientificName)) %>%
 		mutate(
 			rgnparser::gn_parse_tidy(scientificName) %>% 
 				select(species = canonicalsimple),
 			species = str_replace_all(species, " ", "_")
 		) %>%
 		# Map on higher level taxonomy
-		mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
+		mutate(
+			genus = case_when(
+				str_detect(scientificName, "^x |^× ") ~ 
+					str_split(scientificName, " ") %>% map_chr(2),
+			TRUE ~ str_split(scientificName, " ") %>% map_chr(1)
+			)
+    ) %>%
 		left_join(
 			select(ppgi_taxonomy, genus, family, 
 						 suborder, order, class), by = "genus"

### 2022-02-18 ms/si.Rmd
Add SI references

--- 2022-02-17 ms/si.Rmd
@@
   - \cfoot{\thepage} # footer
+nocite: |
+  @*
 editor_options: 
@@
 bibliography: 
-- `r here::here("ms/references.yaml")`
-- `r here::here("ms/references_other.yaml")`
-- `r here::here("ms/references_other.bib")`
+- `r here::here("ms/si_references.yaml")`
+- `r here::here("ms/si_references_other.yaml")`
 ---
@@
 		family_stem_ages,
-		fossil_ferns_raw
+		fossil_ferns_raw,
+		pteridocat_ppgi_diff_refs_out,
+		pteridocat_ppgi_diff_notes
 	),
@@
 
-```{r fossils}
+```{r genera-diff-make}
+# FIXME: need to update ppgi csv file
+# Add one missing nothogenus
+ppgi_taxonomy_mod <-
+	bind_rows(
+		ppgi_taxonomy,
+		tibble(
+			genus = "Phlebosia",
+			notes = "Nothogenus"
+		)
+	)
+
+# Verify that all accepted genera are accounted for
+accepted_species %>%
+	count(genus, name = "n_species") %>%
+	anti_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
+	verify(nrow(.) == 0, success_fun = success_logical)
+
+# Make table of Pteridocat genera not in PPGI
+pteridocat_not_in_ppgi <-
+	accepted_species %>%
+	count(genus, name = "n_species") %>%
+	left_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
+	assert(not_na, notes) %>%
+	assert(is_uniq, genus) %>%
+	filter(notes != "original PPGI 2016") %>%
+	transmute(
+		diff = "Pteridocat not in PPGI",
+		genus,
+		n_species
+	) %>%
+	# Add notes
+	left_join(
+		select(pteridocat_ppgi_diff_notes, -ref, -n_species),
+		by = c("diff", "genus")
+	) %>%
+	assert(not_na, note)
+
+# Make table of PPGI genera not in Pteridocat
+ppgi_not_in_pteridocat <-
+	ppgi_taxonomy_mod %>%
+	filter(notes == "original PPGI 2016",
+				 class == "Polypodiopsida") %>%
+	anti_join(accepted_species, by = "genus") %>%
+	count(genus) %>%
+	transmute(
+		diff = "PPGI not in Pteridocat",
+		genus
+	) %>%
+	# Add notes
+	left_join(
+		select(pteridocat_ppgi_diff_notes, -ref), by = c("diff", "genus")
+	) %>%
+	assert(not_na, note)
+
+genera_diff_table <- 
+	bind_rows(pteridocat_not_in_ppgi, ppgi_not_in_pteridocat)
+```
+
+```{r genera-diff-save, eval = params$doc_type == "doc"}
+# Save if rendering to doc
+write_csv(genera_diff_table, result_file(s_table("genera_diff"), "csv"))
+```
+
+**`r s_table("genera_diff")`**.
+Differences in genera included as accepted in Pteridocat taxonomic database (https://github.com/pteridophy/pteridocat) vs.
+PPGI [@PteridophytePhylogenyGroupI2016] (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
+"diff" is type of difference (Pteridocat genus not in PPGI or PPGI genus not in Pteridocat).
+"n_species" is the number of accepted species in each genus (Pteridocat) or estimated number of species (PPGI).
+References in "notes" column provided at end of Supplementary Materials.
+
+```{r fossils-make}
 # Filter raw fossil data to just fossils used, write out as CSV
@@
 	)
+```
 
+```{r fossils-save, eval = params$doc_type == "doc"}
+# Save if rendering to doc
 write_csv(fossils_table, result_file(s_table("fossils"), "csv"))

### 2022-02-18 R/functions.R

@@ -795,7 +795,8 @@ extract_citations <- function(rmd_file) {
 		stringr::str_split(" |;") %>% 
 		unlist %>% 
 		magrittr::extract(., stringr::str_detect(., "@")) %>% 
-		stringr::str_remove_all("\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}|\\\\") %>% 
+		stringr::str_remove_all("^[^@]*") %>%
+		stringr::str_remove_all('\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}|\\\\|\\"') %>% 
 		magrittr::extract(., stringr::str_detect(., "^@|^-@")) %>% 
 		stringr::str_remove_all("^@|^-@") %>% 
 		unique %>% 

### 2022-02-18 R/process_refs.R

@@ -10,6 +10,15 @@ filter_refs_yaml(
   "data_raw/main_library.yaml",
   "ms/references.yaml", silent = TRUE)
 
+# Same for SI
+filter_refs_yaml(
+  c(
+    "ms/si.rmd", # SI
+    "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv" # refs in SI tables
+  ),
+  "data_raw/main_library.yaml",
+  "ms/si_references.yaml", silent = TRUE)
+
 # Write out 'aux' file with citation keys to make collection in Zotero
 # so reference data are easier to work with.
 # The collection can be made with the betterbibtex for Zotero plugin:

### 2022-02-18 _targets.R

@@ -20,6 +20,14 @@ tar_plan(
 		"_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
 	),
 	other_dates = parse_ts_dates(testo_sundue_2016_si_path),
+	# - Notes on differences between Pteridocat and PPGI
+	tar_files_input(
+		pteridocat_ppgi_diff_notes_path,
+		"_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv"
+	),
+	pteridocat_ppgi_diff_notes = read_csv(
+		pteridocat_ppgi_diff_notes_path
+	),
 	# - Targets from FTOL workflow
 	plastome_metadata_renamed = tar_read(
 		plastome_metadata_renamed,

### 2022-02-18 ms/manuscript.Rmd
Add monophyly table

@@
 
+```{r formatters-load}
+# Load formatting functions defined in functions.R
+knitr::read_chunk(
+	here::here("R/functions.R"),
+	labels = "formatters",
+	from = "formatters start",
+	to = "formatters end"
+)
+```
+
+```{r, formatters}
+```
+
 ```{r load-functions, eval = params$doc_type == "manual"}
@@
 		fossil_ferns_raw,
-		family_stem_ages
+		family_stem_ages,
+		fern_monophy_summ_tbl
 	),
@@
 denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
+
+fern_monophy_summ <-
+fern_monophy_summ_tbl %>%
+	mutate(
+		n = as.integer(n),
+		p = percent(percent)
+	) %>%
+	split(.$tax_level) %>%
+	map(~split(., .$monophyly))
 ```
@@
 The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and Athyriaceae.
+
+All sampled orders, suborders, families, and subfamilies were recovered as monophyletic (or monotypic) with the exception of Polypodioideae, which is known to be paraphyletic relative to Grammitidoideae [@PteridophytePhylogenyGroupI2016].
+`r fern_monophy_summ$genus$No$n` genera (`r fern_monophy_summ$genus$No$p`) were non-monophyletic (`r s_table("monophy")`).
 
@@
 ```{r plot-prep}
-## ---- plot-prep-main
+### plot-prep start
 
@@
 	left_join(clade_cols_tib, by = "major_clade")
+### plot-prep end
 ```

### 2022-02-18 ms/si.Rmd
Add monophyly table

@@
 
+```{r formatters-load-si}
+# Load formatting functions defined in functions.R
+knitr::read_chunk(
+	here::here("R/functions.R"),
+	labels = "formatters-si",
+	from = "formatters start",
+	to = "formatters end"
+)
+```
+
+```{r, formatters-si}
+```
+
 ```{r load-functions-si, eval = params$doc_type == "manual"}
@@
 		fossil_calibration_tips,
-		ppgi_taxonomy),
+		ppgi_taxonomy,
+		monophy_by_clade),
 	store = ftol_cache
@@
 		pteridocat_ppgi_diff_refs_out,
-		pteridocat_ppgi_diff_notes
+		pteridocat_ppgi_diff_notes,
+		fern_monophy_summ_tbl
 	),
@@
 
+```{r monophy, results = "markup"}
+fern_monophy_summ_tbl %>%
+	mutate(
+		tax_level = str_to_sentence(tax_level),
+		percent = percent(percent, accuracy = 0.1) %>%
+			str_replace_all("100\\.0", "100")) %>%
+	select(
+		`Taxonomic level` = tax_level, Monophyly = monophyly,
+		n, `%` = percent) %>%
+	flextable() %>%
+	italic(1,3, part = "header") %>%
+	autofit()
+```
+
+**`r s_table("monophy")`**.
+Monophyly by taxonomic level for ferns in FTOL.
+
+\newpage
+
 ```{r plot-prep-load}
@@
 	labels = "plot-prep-main",
-	from = "og_taxa <-",
-	to = "left_join\\(clade_cols_tib"
+	from = "plot-prep start",
+	to = "plot-prep end"
 )

### 2022-02-18 R/functions.R

@@ -26,7 +26,7 @@ fetch_metadata <- function(
 	
 	# If no results, return empty tibble
 	if (initial_genbank_results$count < 1) return(tibble(taxid = NA))
-
+	
 	# Download IDs with maximum set to 1 more than the total number of hits.
 	genbank_results <- rentrez::entrez_search(
 		db = "nucleotide",
@@ -39,7 +39,7 @@ fetch_metadata <- function(
 	# - make "safe" version of rentrez::entrez_summary to catch errors
 	safe_entrez_summary <- purrr::safely(rentrez::entrez_summary)
 	entrez_summary_gb <- function(id, col_select) {
-
+		
 		# Download data
 		res <- safe_entrez_summary(db = "nucleotide", id = id)
 		
@@ -324,9 +324,9 @@ get_accepted_species <- function(pteridocat, ppgi_taxonomy) {
 			genus = case_when(
 				str_detect(scientificName, "^x |^× ") ~ 
 					str_split(scientificName, " ") %>% map_chr(2),
-			TRUE ~ str_split(scientificName, " ") %>% map_chr(1)
+				TRUE ~ str_split(scientificName, " ") %>% map_chr(1)
 			)
-    ) %>%
+		) %>%
 		left_join(
 			select(ppgi_taxonomy, genus, family, 
 						 suborder, order, class), by = "genus"
@@ -754,6 +754,88 @@ get_stem_family_age <- function(
 		select(family, age)
 }
 
+#' Summarize (non-)monophyletic status of ferns in FTOL
+#'
+#' @param monophy_by_clade Tibble with monophyly status by clade
+#' @param ppgi_taxonomy Pteridophyte phylogeny group I taxonomy (genus level
+#' and above)
+#' @param sanger_sampling Tibble including sampling statistics in Sanger 
+#' dataset (one row per species, with columns for higher-level taxonomy
+#' and outgroup status)
+#' @param check Logical; check that Polypodioideae is the only taxonomic
+#' level above genus that is non-monophyletic
+#'
+#' @return Tibble
+#' 
+summarize_fern_monophyly <- function(
+	monophy_by_clade, ppgi_taxonomy,
+	sanger_sampling, check = TRUE) {
+	
+	ppgi_tax_levels <-
+		ppgi_taxonomy %>%
+		select(class:genus) %>%
+		pivot_longer(
+			names_to = "tax_level",
[…]

### 2022-02-18 _targets.R

@@ -49,6 +49,10 @@ tar_plan(
 		pteridocat,
 		store = ftol_cache
 	),
+	monophy_by_clade = tar_read(
+		monophy_by_clade,
+		store = ftol_cache
+	),
 	tar_files_input(
 		fossil_ferns_path,
 		fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
@@ -120,6 +124,9 @@ tar_plan(
 		du_2021_si_path,
 		"_targets/user/data_raw/cla12457-sup-0003-tables1-s5.docx"),
 	du_dates_all = parse_du_dates(du_2021_si_path),
+	# - Summarize fern monophyly
+	fern_monophy_summ_tbl = summarize_fern_monophyly(
+		monophy_by_clade, ppgi_taxonomy, sanger_sampling),
 	
 	# Render manuscript ----
 	# track files used for rendering MS

### 2022-02-18 R/functions.R
Add func to check order of fig and table refs in MS

@@ -1106,3 +1106,47 @@ base_freq_missing <- function(
 # Generate custom error message for assertr
 # https://github.com/ropensci/assertr/issues/70
 err_msg <- function(msg) stop(msg, call. = FALSE)
+
+#' Check order of cited figures and tables in an Rmd file
+#'
+#' @param type Type of citation to check:
+#' - "all": everything
+#' - "fig": all figures
+#' - "s_fig": SI figures
+#' - "tab": all tables
+#' - "s_tab": SI tables
+#' @param rmd_file Path to the Rmd file
+#' @return Character vector: the functions citing the tables and references,
+#' in the order they appear in the Rmd file.
+check_ft_order <- function(type = "all", rmd_file = "ms/manuscript.Rmd") {
+	
+	# Run inline code when purling
+	options(knitr.purl.inline = TRUE)
+	
+	# Make a temporary file to write out just inline R code from the SI Rmd
+	temp_file <- tempfile()
+	
+	# Generate R script including inline R code from the ms Rmd
+	suppressMessages(knitr::purl(rmd_file, output = temp_file))
+	
+	# Trim this R script down to only functions that define figure and table captions
+	refs <- read_lines(temp_file) %>%
+		magrittr::extract(
+			str_detect(., "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(")
+		) %>%
+		unique
+	
+	# Delete the temporary file
+	fs::file_delete(temp_file)
+	
+	switch(
+		type,
+		"all" = refs,
+		"fig" = refs[str_detect(refs, "fig")],
+		"s_fig" = refs[str_detect(refs, "s_fig")],
+		"tab" = refs[str_detect(refs, "tab")],
+		"s_tab" = refs[str_detect(refs, "s_tab")],
+		stop("Must choose 'all', 'fig', 's_fig', 'tab', or 's_tab' for type")
+	)
+	
+}

### 2022-02-18 ms/manuscript.Rmd
Add nonmono gen table

@@
 		family_stem_ages,
-		fern_monophy_summ_tbl
+		fern_monophy_summ_tbl,
+		fern_nonmono_gen
 	),
@@
 
+# Format monophyly status as list for MS
 fern_monophy_summ <-
@@
 	map(~split(., .$monophyly))
+
+top_nonmono_subfam <-
+fern_nonmono_gen %>%
+	group_by(family, subfamily) %>%
+	summarize(n = n(), .groups = "drop") %>%
+	slice_max(n = 3, order_by = n) %>%
+	assert(not_na, family, subfamily) %>%
+	mutate(subfamily = fct_reorder(subfamily, n)) %>%
+	mutate(n = jntools::english2(n)) %>%
+	arrange(desc(subfamily))
+
+# Double check that non-monophyletic genera outside of
+# top three non-monophyletic subfamilies all have less
+# than four genera each
+fern_nonmono_gen %>%
+	anti_join(
+		top_nonmono_subfam,
+		by = c("family", "subfamily")) %>%
+	count(family, subfamily) %>%
+	verify(
+		all(n < 4), 
+		error_fun = error_msg(
+			"Other subfamilies besides top three don't have less than four genera each"),
+		success_fun = success_logical)
 ```
@@
 `r fern_monophy_summ$genus$No$n` genera (`r fern_monophy_summ$genus$No$p`) were non-monophyletic (`r s_table("monophy")`).
+The subfamilies with the most non-monophyletic genera were `r top_nonmono_subfam$subfamily[[1]]` (`r top_nonmono_subfam$n[[1]]`), `r top_nonmono_subfam$subfamily[[2]]` (`r top_nonmono_subfam$n[[2]]`), and `r top_nonmono_subfam$subfamily[[3]]` (`r top_nonmono_subfam$n[[3]]`); other (sub)families each had four or fewer non-monophyletic genera (`r s_table("nonmono_gen")`).
 

### 2022-02-18 ms/si.Rmd
Add nonmono gen table

@@
 		pteridocat_ppgi_diff_notes,
-		fern_monophy_summ_tbl
+		fern_monophy_summ_tbl,
+		fern_nonmono_gen
 	),
@@
 ```{r fossils-save, eval = params$doc_type == "doc"}
-# Save if rendering to doc
 write_csv(fossils_table, result_file(s_table("fossils"), "csv"))
@@
 
+```{r nonmono-gen, eval = params$doc_type == "doc"}
+fern_nonmono_gen %>%
+	arrange(family, subfamily, genus) %>%
+	write_csv(result_file(s_table("nonmono_gen"), "csv"))
+```
+
+**`r s_table("nonmono_gen")`**.
+Non-monophyletic fern genera in FTOL (`r result_file(s_table("nonmono_gen"), "csv") %>% fs::path_file()`).
+"number_tips" is the number of tips (species) in the genus.
+"delta_tips" is the number of tips that do not belong to the genus but are included in the clade descending from the most common recent ancestor of the genus.
+"number_intruders" is the number of intruding taxa into the genus.
+"number_outliers" is the number of tips (species) belonging to the genus that appear in other clades.
+Status of intruders and outliers was determined using the default cutoff level (0.5) in the R package "MonoPhy" [@Schwery2016].
+
+\newpage
+
 ```{r plot-prep-load}
@@
 <!-- Run the plot prep code -->
-```{r, plot-prep-main}
+```{r plot-prep-main}
 ```

### 2022-02-18 R/functions.R

@@ -754,22 +754,20 @@ get_stem_family_age <- function(
 		select(family, age)
 }
 
-#' Summarize (non-)monophyletic status of ferns in FTOL
+#' Filter monophyly table to ferns, add taxonomic level
 #'
-#' @param monophy_by_clade Tibble with monophyly status by clade
+#' @param monophy_by_clade Monophyly status for various taxonomic levels,
+#' including ferns and outgroups
 #' @param ppgi_taxonomy Pteridophyte phylogeny group I taxonomy (genus level
 #' and above)
-#' @param sanger_sampling Tibble including sampling statistics in Sanger 
-#' dataset (one row per species, with columns for higher-level taxonomy
-#' and outgroup status)
-#' @param check Logical; check that Polypodioideae is the only taxonomic
-#' level above genus that is non-monophyletic
+#' @param sanger_sampling Tibble with one row per species in the tree, including
+#' columns "genus", "family", etc. 
 #'
 #' @return Tibble
 #' 
-summarize_fern_monophyly <- function(
+add_tax_levels_to_monophyly <- function(
 	monophy_by_clade, ppgi_taxonomy,
-	sanger_sampling, check = TRUE) {
+	sanger_sampling) {
 	
 	ppgi_tax_levels <-
 		ppgi_taxonomy %>%
@@ -801,6 +799,22 @@ summarize_fern_monophyly <- function(
 		mutate(outgroup = replace_na(outgroup, FALSE)) %>%
 		filter(outgroup != TRUE) %>%
 		assert(not_na, tax_level)
+}
+
+#' Summarize (non-)monophyletic status of ferns in FTOL
+#'
+#' @param monophy_by_clade_tax_level Tibble with monophyly status by clade,
+#' including taxonomic level
+#' @param ppgi_taxonomy Pteridophyte phylogeny group I taxonomy (genus level
+#' and above)
+#' @param check Logical; check that Polypodioideae is the only taxonomic
+#' level above genus that is non-monophyletic
+#'
+#' @return Tibble
+#' 
+summarize_fern_monophyly <- function(
+	monophy_by_clade_tax_level, ppgi_taxonomy,
+	check = TRUE) {
 	
 	if (isTRUE(check)) {
 	monophy_by_clade_tax_level %>%
@@ -835,6 +849,26 @@ summarize_fern_monophyly <- function(
 		arrange(tax_level, monophyly)
 }
 
[…]

### 2022-02-18 R/process_refs.R

@@ -1,6 +1,6 @@
 # Filter a list of references in YAML format to those cited in an RMD file
 library(tidyverse)
-source("R/ms_functions.R")
+source(here::here("R/functions.R"))
 
 # Filter YAML references from Zotero library.
 # data_raw/main_library.yaml has been exported from Zotero

### 2022-02-18 _targets.R

@@ -124,9 +124,19 @@ tar_plan(
 		du_2021_si_path,
 		"_targets/user/data_raw/cla12457-sup-0003-tables1-s5.docx"),
 	du_dates_all = parse_du_dates(du_2021_si_path),
-	# - Summarize fern monophyly
+	# Summarize fern monophyly
+	# - filter to ferns, add taxonomic levels
+	fern_monophy_by_clade_tax_level = add_tax_levels_to_monophyly(
+		monophy_by_clade, ppgi_taxonomy, sanger_sampling
+	),
+	# - calculate summary table
 	fern_monophy_summ_tbl = summarize_fern_monophyly(
-		monophy_by_clade, ppgi_taxonomy, sanger_sampling),
+		fern_monophy_by_clade_tax_level, ppgi_taxonomy
+	),
+	# - make table of non-monophyletic genera
+	fern_nonmono_gen = make_nonmonophy_gen_tab(
+		fern_monophy_by_clade_tax_level, ppgi_taxonomy
+	),
 	
 	# Render manuscript ----
 	# track files used for rendering MS

### 2022-02-19 ms/si.Rmd
Output SI2

@@
 - `r here::here("ms/si_references_other.yaml")`
+- `r here::here("ms/si_references_other.yaml")`
+- `r here::here("ms/si_fossils.bib")`
 ---

### 2022-02-19 ms/si2.Rmd
Output SI2

---
title: "*Supplementary Material 2*"
subtitle: "Fossil Calibration List"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "frontiers.csl"]
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template_SI.docx
header-includes: 
  - \usepackage{float} # keep figures from floating away
  - \usepackage{booktabs} # need for kable tables
  - \usepackage{colortbl} # need for kable tables
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
  - \usepackage{fancyhdr} # enable header
  - \pagestyle{fancy}  # enable header
  - \fancypagestyle{plain}{\pagestyle{fancy}} # show header on first page
  - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
  - \fancyhead[RE,RO]{Supplementary Material} # add header on right side
  - \fancyhead[LO,LE]{} # suppress automatic section header on left side
  - \cfoot{\thepage} # footer
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

---
bibliography: 
- `r here::here("ms/si_fossils.bib")`
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
library(glue)
library(gluedown)
library(targets)
library(tarchetypes)
library(tidyverse)
library(assertr)
```

```{r prep-fossil-data}
# Set path to fossil data
fossils_data_path <- here::here("_targets/user/data_raw/Fossils_Ferns_mod.xlsx")

# Extract fossil categories (first row of fossil data)
fossils_categories <-
  readxl::read_excel(fossils_data_path, n_max = 1) %>%
	pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
	mutate(
		category = str_replace_all(category, "\\.\\.\\.[0-9]+", "skip") %>%
			na_if("skip")) %>%
	fill(category) %>%
	mutate(category = str_to_title(category)) %>%
	filter(category != "Pteridophy")

# Get raw column names
fossils_raw_names <-
  readxl::read_excel(fossils_data_path, skip = 1)

# Get cleaned column names
fossils_clean_names <- fossils_raw_names %>% janitor::clean_names()

# Load fossils in wide format
fossils <-
fossils_clean_names %>%
	# Filter to only those actually used in the study
	filter(used == "X") %>%
	select(-used) %>%
	select(fossil_taxon, everything()) %>%
	# Format reference key for citing with pandoc
	mutate(
		reference = as.character(glue("@{reference}"))
	)

# Make a tibble of column namees for formatting
name_tbl <- tibble(
	raw_names = colnames(fossils_raw_names),
	var = colnames(fossils_clean_names)
) %>%
	right_join(
		tibble(var = colnames(fossils)), by = "var"
	) %>%
	mutate(var_order = 1:nrow(.))

# Convert to long format
fossils_long <- fossils %>%
	mutate(across(everything(), as.character)) %>%
	pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
	left_join(name_tbl, by = "var") %>%
	mutate(
		text = as.character(glue::glue("{raw_names}: {value}"))
	) %>%
	left_join(fossils_categories, by = "raw_names") %>%
	arrange(fossil_taxon, category, var_order)

#' Print a single category of fossil data
#'
#' @param fossil_data Fossil data in long form, filtered to a single
#' fossil
#' @param cat_select Category of data to print
#'
#' @return Character vector

print_fossil_cat <- function(fossil_data, cat_select) {
	data <-
    fossil_data %>% filter(category == cat_select)

	c(
		md_heading(cat_select, 3),
		md_bullet(data$text),
		md_blank()
	)
}

#' Print data for a single fossil
#'
#' @param fossil_data Fossil data in long form
#' @param fossil_select Selected fossil to print
#'
#' @return Character vector
#'
#' @examples
#' print_fossil(fossils_long, "Acrostichum_intertrappeum") %>%
#' 	glue::as_glue()
print_fossil <- function(fossil_data, fossil_select) {
	# Filter data to only selected fossil
	data <-
    fossils_long %>% filter(fossil_taxon == fossil_select)

	# Print name of fossil as header in italics
	header <-
	fossil_select %>% 
		str_replace_all("_", " ") %>%
		md_italic() %>%
		md_heading(2)

	categories <- unique(data$category)

	body <- map(categories, ~print_fossil_cat(data, .)) %>%
		unlist()

	c(header, body, md_blank())
}
```

```{r print-fossils, results = "asis"}
unique(fossils_long$fossil_taxon) %>%
  map(~print_fossil(fossils_long, .)) %>%
	unlist() %>%
	glue::as_glue()
```

### 2022-02-19 _targets.R

@@ -169,5 +169,21 @@ tar_plan(
 		output_dir = "results",
 		output_format = "bookdown::pdf_document2",
 		params = list(doc_type = "pdf")
+	),
+	tar_render(
+		si_2_doc,
+		"ms/si2.Rmd",
+		knit_root_dir = "ms",
+		output_dir = "results",
+		output_format = "bookdown::word_document2",
+		params = list(doc_type = "doc")
+	),
+	tar_render(
+		si_2_pdf,
+		"ms/si2.Rmd",
+		knit_root_dir = "ms",
+		output_dir = "results",
+		output_format = "bookdown::pdf_document2",
+		params = list(doc_type = "pdf")
 	)
 )

### 2022-02-19 ms/si2.Rmd
Update SI2
use modified CSV

@@
 
-```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
+```{r setup-rmd, include = FALSE}
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, results = 'hide')
 library(glue)
 library(gluedown)
-library(targets)
-library(tarchetypes)
 library(tidyverse)
@@
 
-```{r prep-fossil-data}
-# Set path to fossil data
-fossils_data_path <- here::here("_targets/user/data_raw/Fossils_Ferns_mod.xlsx")
-
-# Extract fossil categories (first row of fossil data)
-fossils_categories <-
-  readxl::read_excel(fossils_data_path, n_max = 1) %>%
-	pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
-	mutate(
-		category = str_replace_all(category, "\\.\\.\\.[0-9]+", "skip") %>%
-			na_if("skip")) %>%
-	fill(category) %>%
-	mutate(category = str_to_title(category)) %>%
-	filter(category != "Pteridophy")
-
-# Get raw column names
-fossils_raw_names <-
-  readxl::read_excel(fossils_data_path, skip = 1)
-
-# Get cleaned column names
-fossils_clean_names <- fossils_raw_names %>% janitor::clean_names()
-
-# Load fossils in wide format
-fossils <-
-fossils_clean_names %>%
-	# Filter to only those actually used in the study
-	filter(used == "X") %>%
-	select(-used) %>%
-	select(fossil_taxon, everything()) %>%
-	# Format reference key for citing with pandoc
-	mutate(
-		reference = as.character(glue("@{reference}"))
-	)
-
-# Make a tibble of column namees for formatting
-name_tbl <- tibble(
-	raw_names = colnames(fossils_raw_names),
-	var = colnames(fossils_clean_names)
-) %>%
-	right_join(
-		tibble(var = colnames(fossils)), by = "var"
-	) %>%
-	mutate(var_order = 1:nrow(.))
-
-# Convert to long format
-fossils_long <- fossils %>%
-	mutate(across(everything(), as.character)) %>%
-	pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
-	left_join(name_tbl, by = "var") %>%
-	mutate(
-		text = as.character(glue::glue("{raw_names}: {value}"))
-	) %>%
-	left_join(fossils_categories, by = "raw_names") %>%
-	arrange(fossil_taxon, category, var_order)
-
+```{r custom-funcs}
 #' Print a single category of fossil data
@@
 	header <-
-	fossil_select %>% 
+	data %>% 
+		filter(var == "fossil_taxon_header") %>%
+		pull(value) %>%
 		str_replace_all("_", " ") %>%
 		md_italic() %>%
-		md_heading(2)
+		md_heading(1)
 
-	categories <- unique(data$category)
+	categories <- data %>%
+		filter(!is.na(category)) %>%
+		pull(category) %>%
+		unique()
 
@@
 
-	c(header, body, md_blank())
+	c(header, body, md_rule())
 }
@@
 
-```{r print-fossils, results = "asis"}
+```{r prep-fossil-data}
+# Set path to fossil data
+fossils_data_path <- here::here("_targets/user/data_raw/Fossils_Ferns.csv")
+
+# Extract fossil categories (first row of fossil data)
+fossils_categories <-
+  read_csv(fossils_data_path, n_max = 1) %>%
+	pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
+	mutate(
+		category = str_replace_all(category, "\\.\\.\\.[0-9]+", "skip") %>%
+			na_if("skip")) %>%
+	fill(category) %>%
+	mutate(category = str_to_title(category)) %>%
+	filter(category != "Pteridophy")
+
+# Get raw column names
+fossils_raw_names <-
+  read_csv(fossils_data_path, skip = 1)
+
+# Load data with cleaned column names
+fossils_clean_names <- fossils_raw_names %>% janitor::clean_names()
+
+# Tidy fossils in wide format
+fossils <-
+fossils_clean_names %>%
+	# Filter to only those actually used in the study
+	filter(used == "X") %>%
+	select(-used) %>%
+	select(fossil_taxon, everything())
+
+# Make a tibble of column namees for formatting
+name_tbl <- tibble(
+	raw_names = colnames(fossils_raw_names),
+	var = colnames(fossils_clean_names)
+) %>%
+	right_join(
+		tibble(var = colnames(fossils)), by = "var"
+	) %>%
+	mutate(var_order = 1:nrow(.))
+
+# Convert to long format
+fossils_long <- fossils %>%
+	mutate(
+		across(everything(), as.character),
+		fossil_taxon_header = glue("†{fossil_taxon}")) %>%
+	pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
+	left_join(name_tbl, by = "var") %>%
+	mutate(
+		text = as.character(glue::glue("{raw_names}: {value}"))
+	) %>%
+	left_join(fossils_categories, by = "raw_names") %>%
+	arrange(fossil_taxon, category, var_order)
+```
+
+These data are exported from the FernCal v1.0 database (https://github.com/pteridophy/FossilFerns), which includes `r nrow(fossils_clean_names)` records of fossils to use as calibration points for molecular dating of ferns.
+
+"NFos" is the unique ID of the record in FernCal.
+
+Taxonomy of extant taxa follows pteridocat v1.0 database (https://github.com/pteridophy/pteridocat).
+
+For a summary of these data in CSV format, see Table FIXME <add table number>.
+
+\newpage
+
+```{r print-fossils, include = TRUE, results = "asis", eval = TRUE}
 unique(fossils_long$fossil_taxon) %>%
@@
 ```
+
+# References

### 2022-02-19 R/functions.R
Finally figure out how to change RStudio tabs to spaces

@@ -9,106 +9,106 @@
 #' @return Tibble
 #' 
 fetch_metadata <- function(
-	query = NULL,
-	col_select = c("gi", "caption", "taxid", "title", "slen", "subtype", "subname")) {
-	
-	assertthat::assert_that(assertthat::is.string(query))
-	
-	assertthat::assert_that(is.character(col_select))
-	
-	# Do an initial search without downloading any IDs to see how many hits
-	# we get.
-	initial_genbank_results <- rentrez::entrez_search(
-		db = "nucleotide",
-		term = query,
-		use_history = FALSE
-	)
-	
-	# If no results, return empty tibble
-	if (initial_genbank_results$count < 1) return(tibble(taxid = NA))
-	
-	# Download IDs with maximum set to 1 more than the total number of hits.
-	genbank_results <- rentrez::entrez_search(
-		db = "nucleotide",
-		term = query,
-		use_history = FALSE,
-		retmax = initial_genbank_results$count + 1
-	)
-	
-	# Define internal function to download genbank data into tibble
-	# - make "safe" version of rentrez::entrez_summary to catch errors
-	safe_entrez_summary <- purrr::safely(rentrez::entrez_summary)
-	entrez_summary_gb <- function(id, col_select) {
-		
-		# Download data
-		res <- safe_entrez_summary(db = "nucleotide", id = id)
-		
-		# Early exit if error with entrez
-		if (!is.null(res$error)) {
-			warning("No esummary records found in file, returning empty tibble")
-			return(tibble(taxid = NA))
-		}
-		
-		res$result %>%
-			# Extract selected columns from result
-			purrr::map_dfr(magrittr::extract, col_select) %>%
-			# Make sure taxid column is character
-			mutate(taxid = as.character(taxid)) %>%
-			assert(not_na, taxid)
-	}
-	
-	# Extract list of IDs from search results
-	genbank_ids <- genbank_results$ids
-	
-	# Fetch metadata for each ID and extract selected columns
-	if (length(genbank_ids) == 1) {
-		rentrez_results <- rentrez::entrez_summary(db = "nucleotide", id = genbank_ids) %>%
-			magrittr::extract(col_select) %>%
-			tibble::as_tibble() %>%
-			mutate(taxid = as.character(taxid)) %>%
-			assert(not_na, taxid)
-	} else {
-		# Split input vector into chunks
[…]

### 2022-02-19 _targets.R

@@ -11,179 +11,179 @@ source("R/functions.R")
 ftol_cache <- here::here("working/_targets")
 
 tar_plan(
-	
-	# Load data ----
-	# - Dates from Testo and Sundue 2016 SI
-	# (includes Rothfels 2015, Schuettpelz & Pryer 2009)
-	tar_files_input(
-		testo_sundue_2016_si_path,
-		"_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
-	),
-	other_dates = parse_ts_dates(testo_sundue_2016_si_path),
-	# - Notes on differences between Pteridocat and PPGI
-	tar_files_input(
-		pteridocat_ppgi_diff_notes_path,
-		"_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv"
-	),
-	pteridocat_ppgi_diff_notes = read_csv(
-		pteridocat_ppgi_diff_notes_path
-	),
-	# - Targets from FTOL workflow
-	plastome_metadata_renamed = tar_read(
-		plastome_metadata_renamed,
-		store = ftol_cache
-	),
-	plastome_tree = tar_read(
-		plastome_tree,
-		store = ftol_cache
-	),
-	plastid_tree_dated = tar_read(
-		plastid_tree_dated,
-		store = ftol_cache
-	),
-	ppgi_taxonomy = tar_read(
-		ppgi_taxonomy,
-		store = ftol_cache
-	),
-	pteridocat = tar_read(
-		pteridocat,
-		store = ftol_cache
-	),
-	monophy_by_clade = tar_read(
-		monophy_by_clade,
-		store = ftol_cache
-	),
-	tar_files_input(
-		fossil_ferns_path,
-		fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
-	),
-	fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
-		janitor::clean_names(),
-	
-	# GenBank ----
-	# Analyze number of fern accessions and species in GenBank by type of DNA 
-	# (plastid, mitochondrial, or nuclear)
-	#
-	# - Make dataframe of query terms by year
-	gb_query = make_gb_query(),
-	# - Download taxids one year at a time
-	tar_target(
-		gb_taxa,
-		fetch_taxa_by_year(
-			query = gb_query$query,
-			year = gb_query$year,
-			type = gb_query$type),
-		pattern = map(gb_query)
[…]

### 2022-02-20 ms/manuscript.Rmd
Finally figure out how to change RStudio tabs to spaces
Update MS

@@
 We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
-We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`).
+We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
 In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).

### 2022-02-20 ms/si2.Rmd
Finally figure out how to change RStudio tabs to spaces
Update MS

@@
 
+```{r load-captions-si2, cache = FALSE}
+# Source captions numbers from MS
+# 
+# So that caption numbers are in order of appearance in MS
+
+# Run inline code when purling
+options(knitr.purl.inline = TRUE)
+
+# Make a temporary file to write out just inline R code from the SI Rmd
+temp_file <- tempfile()
+
+# Generate R script including inline R code from the ms Rmd
+knitr::purl(here::here("ms/manuscript.Rmd"), output = temp_file)
+
+# Trim this R script down to only functions that define figure and table captions
+read_lines(temp_file) %>%
+  magrittr::extract(
+    str_detect(., "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(")
+  ) %>%
+  unique %>%
+  write_lines(temp_file)
+
+# Source the code. This will load the caption functions in the right order.
+source(temp_file, local = TRUE)
+
+# Delete the temporary file
+fs::file_delete(temp_file)
+```
+
 ```{r prep-fossil-data}
@@
 
-For a summary of these data in CSV format, see Table FIXME <add table number>.
+For a summary of these data in CSV format, see `r s_table("fossils")`.
 

### 2022-02-21 ms/manuscript.Rmd
Revise intro and methods per ES comments

--- 2022-02-18 ms/manuscript.Rmd
@@
 ---
-title: "An open and continuously updated fern tree of life (FTOL)"
+title: "An Open and Continuously Updated Fern Tree of Life (FTOL)"
 mainfont: Roboto-Regular.ttf
@@
 
-^2^Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., USA
+^2^Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, District of Columbia, United States
 
@@
 
+Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after seed plants.
+Ferns have received intense focus from molecular systematists, resulting in the publication of molecular phylogenies for every major clade and DNA sequences for more than one third of all fern species.
+Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these quickly become outdated.
+
+Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data available via GenBank.
+Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with commonly sequenced plastid regions (many taxa, few loci) to obtain a global, species-level fern phylogeny with high resolution along the backbone and maximal sampling across the tips.
+We use an expert-curated reference taxonomy to resolve synonyms in general compliance with Pteridophyte Phylogeny Group I (PPG I), the community-driven taxonomic system widely accepted by many pteridologists.
+
+The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of more than 20% relative to the previous largest fern phylogeny.
+Using an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points, we find estimated ages for most families and deeper clades to be significantly older than earlier studies.
+
+FTOL and its accompanying datasets including the fossil list and taxonomic database will be updated on a regular basis and made available via a web portal and R packages, enabling immediate access to the most up-to-date, comprehensively-sampled fern phylogeny.
+FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
+We anticipate FTOL will be particularly relevant for macroecological studies at regional to global scales and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
+
+# Introduction
+
 Ferns (ca.
-12,000 spp.) are the second most diverse clade of vascular plants after seed plants and have received intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for more than one third of all fern species deposited in GenBank.
-Global trees including all ferns have been published periodically, but as molecular systematics research continues at a rapid pace, such trees quickly become outdated.
-
-Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data on GenBank.
-Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with Sanger-sequenced loci (many taxa, few loci) to obtain a global fern phylogeny with high resolution along the backbone and maximum sampling across the tips.
-We use an expert-curated reference taxonomy to resolve synonyms in compliance with Pteridophyte Phylogeny Group I (PPGI), the community-driven taxonomic system widely accepted by many pteridologists.
-
-The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of more than 20% relative to the previous largest fern phylogeny.
-We also used an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points to date the tree, and estimated ages for most families and deeper clades to be significantly older compared to previous studies.
-
-FTOL and its accompanying datasets including the fossil list and taxonomic database will be updated on a semi-annual basis and be made available via a web portal and R packages, enabling researchers to have immediate access to the most recent, comprehensively sampled fern phylogeny.
-FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales from smaller clades to the entire tree.
-We anticipate FTOL will be particularly relevant for macroecological studies at the regional to global scale and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
-
-# Introduction
-
-Ferns (ca.
-12,000 species) are the second most diverse clade of vascular plants after seed plants (ca.
-300,000 species), and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
-A well-sampled molecular phylogeny is key to any investigation of fern evolutionary biology.
-Fortunately, ferns have received relatively intense focus from molecular systematists, resulting in the publication of molecular trees for nearly all major clades and DNA sequences for more than one third all currently recognized fern species deposited in GenBank.
+`r number(12000)` species) are the second most diverse clade of vascular plants after seed plants (ca.
+`r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
+Key to any investigation of evolutionary biology in this group is a well-sampled phylogeny.
+Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data for more than one third of all currently recognized fern species.
 Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.
 
-Past efforts to construct such a global phylogeny have steadily increased in number of species as sequences in GenBank have accumulated [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, accumulation of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny of ferns [@Testo2016a\; `r figure("genbank")`].
-There is a need therefore, not only for a global fern phylogeny, but one that is self-updating to keep pace with the continued accumulation of molecular data.
-This will eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
-
-Multiple efforts exist to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], a tree for all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
-While these may be appropriate in some cases, they may not produce an optimal fern phylogeny due to the need to take "one-size-fits-all" approaches to accommodate such a wide phylogenetic breadth. 
-By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can be used "as-is" by biologists studying these organisms.
-Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be possible with a "tree of all life" or "tree of all plants".
-
-Recently, a genus-level taxonomy of ferns and lycophytes was established following a open, community-driven model [@PteridophytePhylogenyGroupI2016\; hereafter "PPGI"].
-Thanks to this community-first approach, PPGI has been widely accepted and used (cited 881 times on Google Scholar as of writing).
-However, there were clearly problematic (non-monophyletic) genera at the time PPGI was published, and many new taxonomic changes have been (and will continue to be) made since [e.g., @Almeida2017; @Shang2018; @Zhang2020].
-It is anticipated that "PPGII" will be an online, open resource that can be updated as necessary (Schuettpelz, pers. comm.).
-We envision an open, continuously updated global fern phylogeny that is directly integrated with PPGII so that taxonomic decisions reflecting phylogeny can be made based on community consensus using the most recently available data.
-
-Here, we leverage our specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
-We plan to run the pipeline on a semi-annual basis and make the results freely available online through a web portal and R package.
+Past efforts to construct such a global phylogeny have steadily expanded their sampling as sequences have been added to GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
+Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny of ferns [@Testo2016a\; `r figure("genbank")`].
+There is a need therefore, not only for a global fern phylogeny, but one that is continuously updated to keep pace with the rapid accumulation of molecular data.
+This would eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
+
+Multiple frameworks have been put forth to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
+While these may be appropriate in some cases, they are not expected to produce an optimal fern phylogeny due to the need to take a "one-size-fits-all" approach to accommodate such a wide phylogenetic breadth.
+By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can then be used "as-is" by biologists studying these organisms.
+Furthermore, there is much to be gained from integrating a carefully designed global fern phylogeny with the fern systematics community that would not be as easily accomplished with a "tree of all life" or "tree of all plants".
+
+Recently, a genus-level taxonomy of ferns and lycophytes was established using an inclusive, community-driven approach  [@PteridophytePhylogenyGroupI2016\; hereafter "PPG I"].
+PPG I has been widely accepted and used, but there were problematic (non-monophyletic) genera included at the time of publication, and many taxonomic changes have been (and will continue to be) proposed since [`r eg`, @Almeida2017; @Shang2018; @Zhang2020].
+The next iteration of PPG (`r ie`, PPG II) will ideally be an online, open resource that can be updated as necessary.
+We envision an open, continuously updated global fern phylogeny that could be directly integrated with PPG II such that taxonomic decisions can be made based on community consensus and the most recently available data.
+
+Here, we leverage specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
+We plan to run the pipeline on a regular basis and make the results freely available online through a web portal and R package.
 This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
-We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
+We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
 
@@
 The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
-Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome loci").
-Here, the locus *trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*; the locus *rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
-The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes), then combined with the two spacers (79 loci total).
-The plastome loci include all seven Sanger loci.
+Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
+Here, the *trnL*--*trnF* locus includes the *trnL* intron, the 3' *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
+The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
 
@@
 All sequences were downloaded from GenBank.
-GenBank queries were formatted including locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+GenBank queries were formatted to include locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
 Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
 There is no formal definition of genomic vs.
-Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.
+Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.
 
@@
 
-One goal of this project is to generate a phylogeny that is consistent with PPGI.
-Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPGI.
+This project aimed to generate a phylogeny that is consistent with PPG I, while accounting for taxonomic changes that have been made since the publication of PPG I.
+Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPG I.
 Furthermore, the NCBI taxonomy is not curated specifically for ferns and includes many fern synonyms.
-Therefore, we standardized all species names in the GenBank sequences against a reference taxonomy.
-We designed the reference taxonomy for this project based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPGI (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
-
-We used the taxastand R package [@Nitta2021] to resolve species names, which can account for differences in formatting of taxonomic authors (`r eg`, parenthetical basionym author present or absent) as well as perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
-We manually inspected any fuzzily matched or non-matching names.
-This revealed some species names in GenBank that were lacking in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
-Therefore, we updated and edited the World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
-We refer to the resulting taxonomic database as "pteridocat", and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
-
-There are differences between pteridocat and PPGI, mostly at the genus level.
-In the time since PPGI was published, several new genera have been published.
-Furthermore, multiple genera included in PPGI were known to be non-monophyletic and included on a tentative basis [@PteridophytePhylogenyGroupI2016].
-Pteridocat includes newly published genera and some genera that were not recognized by PPGI as well as nothogenera, which were not included in PPGI [@Liu2020a].
-Differences between pteridocat and PPGI are summarized in `r s_table("genera_diff")`.
+Therefore, we standardized all species names in the GenBank sequences against a newly generated reference taxonomy.
+Our reference taxonomy is based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPG I (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
+
+To resolve species names, we used the taxastand R package [@Nitta2021], which can account for differences in formatting of taxonomic authors (`r eg`, parenthetical basionym author present or absent) and perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
+
+We manually inspected any fuzzy matched or non-matching names.
+This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
+We thus updated and edited the initial World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
+We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
+
+There are differences between pteridocat and PPG I, mostly at the genus level.
+In the time since PPG I was introduced, several new genera have been published.
+Furthermore, multiple genera included in PPG I were known to be non-monophyletic and provisionally circumscribed [@PteridophytePhylogenyGroupI2016].
+Pteridocat includes newly published genera, some genera that were not recognized by PPG I, as well as nothogenera, which were not included in PPG I [@Liu2020a].
+Differences between pteridocat and PPG I are summarized in `r s_table("genera_diff")`.
 
@@
 
-Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminations (hereafter collectively "rogues").
+Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminations (hereafter collectively referred to as "rogues").
 We removed putative rogues from Sanger sequences as follows.
@@
 Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
-We compared the families (following PPGI) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contamination and excluded from further analysis.
+We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contaminant and excluded from further analysis.
 For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
 Species belonging to monotypic families were not considered for this filtering.
-While this method cannot account for rogues at finer taxonomic levels, it is an efficient method to remove obviously incorrect sequences.
+While this method cannot account for rogues at finer taxonomic levels, it is an efficient approach for removing obviously erroneous sequences.
 We inspected all accessions flagged this way as rogues before excluding them.
-In the case that that mismatch of family was due to incorrect taxonomy (`r eg`, species name on GenBank matches the correct family but name used in the taxonomic database does not), we updated the taxonomic database accordingly.
+In cases where the family mismatch was due to incorrect taxonomy (`r eg`, the species name in GenBank matches the correct family but the name used in the taxonomic database does not), we updated the taxonomic database accordingly.
 
@@
 
-A typical step in multilocus workflows is to concatenate loci across samples.
-For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
+A typical step in multilocus phylogenetic workflows is to concatenate loci across samples.
+For phylogenetic studies aiming to include one tip per species, this is often done by concatenating the longest sequence per locus within each species.
 However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
@@
 First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
-We then concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
+Then, we concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
 Accessions not meeting any of these conditions were not concatenated.
-All calculations of sequence length excluded missing bases.
-
-We then selected the final set of concatenated loci for each species based on presence of *rbcL* and sequence length.
+All calculations of sequence length excluded missing bases ("?", "N", or "-").
+
+We selected the final set of concatenated loci for each species based on presence of *rbcL* and sequence length.
 We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
@@
 1. If accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length.
-2. Otherwise, if accessions include *rbcL*, select the accession with the longest *rbcL* sequence.
+2. Otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence.
 3. Otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
@@
 
-For non-spacer regions, we aligned each locus separately with MAFFT with automatic adjustment for sequence direction and other settings on default.
-Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
+For non-spacer regions (genes), we aligned each locus separately in MAFFT with automatic adjustment for sequence direction and other settings on default.
+Spacer regions are difficult to align across higher taxonomic levels within ferns (`r eg`, family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer scales (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
 Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "\-\-merge" option.
@@
 For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "\-\-merge".
-Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data.
+Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
 We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
 
-Next, we removed poorly aligned regions from the alignments using trimAl.
-We removed regions with >1% or >5% of sequences having gaps from coding regions or spacers, respectively.
+We removed poorly aligned regions, including those with >1% or >5% of sequences having gaps, from the alignments using trimAl [@Capella-Gutierrez2009].
 
@@
 
-Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
+We used the backbone (plastome) phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
 We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
@@
 However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants.
-
-Based on the results of this inspection, we updated the list of rogue accessions to exclude, updated the taxonomic database, then re-ran all analyses up to this step.
-When re-running the analysis, it is possible that there were newly added GenBank accessions in place of the excluded rogues that themselves are rogues.
-Therefore, we repeated this process until the monophyly of all higher-level (`r eg` genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
+While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants FIXME <<add citation>>.
+
+Based on the results of this inspection, we updated the list of rogue accessions to be excluded, updated the taxonomic database, and then re-ran all analyses up to this step.
+During each iteration of the analysis, it is possible that the GenBank accessions added in place of the excluded rogues themselves include rogues.
+Therefore, we repeated this process until the monophyly of all higher-level (`r eg`, genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
 
@@
 We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
-We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`).
-In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
-We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil species or genus) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
-Taxon concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be calibrated for a given fossil will also change.
-We also applied two calibration points outside of ferns including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma [@Testo2016a].
-All calibration points other than the root are minimum ages.
-
-We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); a value of `r smooth_best` was selected with the smallest chi-squared value for final analysis.
+We selected `r n_calibrations_ferns` fern fossils to use as constraints from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
+In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
+We assigned fossils to lineages only after consulting the original publication (as opposed to simply relying on the fossil name) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
+Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained for a given fossil could change.
+We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a.
+All constraints other than the root are minimum ages.
+
+We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude).
+A value of `r smooth_best` was selected, based on the smallest chi-squared value, for the final analysis.
 To construct confidence intervals for the estimated divergence times, we constructed 100 bootstrap trees (trees with identical topology but different branchlengths) using IQ-TREE and dated each using treePL with the same settings used for the ML tree [@Maurin2020].
@@
 The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
-Code used to generate FTOL is available at https://github.com/joelnitta/ftol.
+Code used to generate FTOL and compile this manuscript is available at https://github.com/pteridophylo/ftol_flow and https://github.com/pteridophylo/ftol_ms, respectively.
 A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.
@@
 `r n_accs_second_filter` accessions were retained after excluding rogues identified by the all-by-all BLAST search.
-The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately).
+The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`).
 The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`).
@@
 `r n_plast_first_filter` accessions were retained after excluding rogues and species names that could not be resolved.
-Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species).
+Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species; `r s_table("accs")`).
 
@@
 Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree on a regular, semi-annual basis.
-We envision that FTOL will be integrated with the next iteration of PPG, PPGII, to provide the most recent hypotheses on the monophyly of various fern taxa, which will in turn enable a more natural classification system.
+We envision that FTOL will be integrated with the next iteration of PPG, PPG II, to provide the most recent hypotheses on the monophyly of various fern taxa, which will in turn enable a more natural classification system.
 

### 2022-02-21 ms/si.Rmd
Finally figure out how to change RStudio tabs to spaces
Update MS
Add accessions table

@@
 ---
-title: "*Supplementary Material*"
+title: "*Supplementary Material 1*"
+subtitle: "Tables and Figures"
 mainfont: Roboto-Regular.ttf
@@
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
+  - \setlength{\headheight}{13.6pt}
   - \cfoot{\thepage} # footer
-nocite: |
-  @*
 editor_options: 
@@
     ppgi_taxonomy,
-		monophy_by_clade),
+    monophy_by_clade,
+    acc_table_long),
   store = ftol_cache
@@
 "minimum_age" (Ma) is the age used for calibration; "lower" is the lower limit of the oldest stratigraphic age for the fossil; "upper" is the upper limit of the oldest stratigraphic age for the fossil.
+For detailed notes and full references for each fossil, see Appendix S2.
+
+```{r accs-save, eval = params$doc_type == "doc"}
+write_csv(acc_table_long, result_file(s_table("accs"), "csv"))
+```
+
+**`r s_table("accs")`**.
+GenBank accessions used in this study (`r result_file(s_table("accs"), "csv") %>% fs::path_file()`).
+"species" is the species name as it appears in the phylogenetic tree.
+"seq_len" is the total number of bases excluding missing bases ("?", "N", or "-") and only includes bases belonging to loci used in this study (not the entire accession).
+"sci_name" is the scientific name in the Pteridocat taxonomic database.
+"ncbi_name" is the scientific name used in the NCBI taxonomic database.
+"ncbi_taxid" is the NCBI taxonomy database unique identifier.
+"outgroup_status" is "in" for ingroup taxa (ferns) and "out" for outgroup taxa.
 

### 2022-02-21 R/functions.R

@@ -1184,3 +1184,10 @@ check_ft_order <- function(type = "all", rmd_file = "ms/manuscript.Rmd") {
   )
   
 }
+
+# Count the number of non-missing bases in a DNA sequence
+#' @param seq List of class DNAbin of length one.
+count_non_missing <- function(seq) {
+  bases <- ape::base.freq(seq, all = TRUE, freq = TRUE)
+  sum(bases[!names(bases) %in% c("n", "N", "-", "?")])
+}

### 2022-02-21 ms/manuscript.Rmd
Fix expansion in family stem ages plot
https://github.com/willgearty/deeptime/issues/41

@@
   scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
-  scale_x_reverse("Age (millions of years)") +
+  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
   coord_geo(
@@
     # https://github.com/willgearty/deeptime/issues/41
-    expand = FALSE,
+    expand = TRUE,
     size = 3) +

### 2022-02-21 ms/si.Rmd
Fix expansion in family stem ages plot
https://github.com/willgearty/deeptime/issues/41

@@
     height = unit(1, "lines"),
-    # FIXME: change this to true when issue gets resolved
-    # https://github.com/willgearty/deeptime/issues/41
-    expand = FALSE,
+    expand = TRUE,
     size = 3) +

### 2022-02-21 ms/manuscript.Rmd
Update MS

--- 2022-02-21 ms/manuscript.Rmd
@@
 Ferns have received intense focus from molecular systematists, resulting in the publication of molecular phylogenies for every major clade and DNA sequences for more than one third of all fern species.
-Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these quickly become outdated.
+Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these become quickly outdated.
 
@@
 The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
+Here, "locus" refers to an individual gene, intergenic spacer, or a unit comprised of these in the case of *trnL*--*trnF*.
 
@@
 However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants FIXME <<add citation>>.
+While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants [@Turland2018].
 

### 2022-02-21 R/functions.R
Load angiosperm trees for LTT plots

@@ -294,6 +294,72 @@ parse_du_dates <- function(du_2021_si_path) {
     select(author, lineage, affinities, affinities_group, median, low, high)
 }
 
+#' Load dated angiosperm trees from Barahona et al 2020
+#' 
+#' Ramírez-Barahona S, Sauquet H, Magallón S (2020) The delayed and 
+#' geographically heterogeneous diversification of flowering plant families. 
+#' Nature Ecology & Evolution 1–7. https://doi.org/10/gg45rb
+#'
+#' @param zip_path Path to zip file of supplemental data from 
+#' Barahona et al 2020, downloaded from
+#' https://zenodo.org/record/4721917/files/spiritu-santi/angiosperm-time-tree-2.0-v2.0.zip
+#' 
+#' @return List of trees named after analysis scheme
+#' 
+load_barahona_trees <- function(zip_path) {
+  
+  # Make temp dir for unzipping
+  temp_dir <- fs::path(tempdir(), "angiosperm-time-tree")
+  
+  if (fs::dir_exists(temp_dir)) fs::dir_delete(temp_dir)
+  
+  fs::dir_create(temp_dir)
+  
+  # Files we want are in a zip file within the zip file
+  # Unzip Data10_mcctrees.zip first
+  
+  unzip(
+    zipfile = zip_path,
+    files = "spiritu-santi-angiosperm-time-tree-2.0-7303f80/data/Supplementary/Data10_mcctrees.zip",
+    junkpaths = TRUE,
+    exdir = temp_dir
+  )
+  
+  # Tree files of interest
+  
+  tree_files <- c(
+    "CC_complete_MCCv_2.tre",
+    "CC_conservative_MCCv_2.tre",
+    "RC_complete_MCCv_2.tre",
+    "RC_conservative_MCCv_2.tre",
+    "UC_complete_MCCv_2.tre",
+    "UC_conservative_MCCv_2.tre"
+  )
+  
+  tree_names <- str_remove_all(tree_files, "_MCCv_2\\.tre")
+  
+  # Next unzip the tree files within Data10_mcctrees.zip
+  
+  unzip(
+    zipfile = fs::path(temp_dir, "Data10_mcctrees.zip"),
+    files = tree_files,
+    junkpaths = TRUE,
+    exdir = temp_dir
+  )
+  
+  # Read in as list of trees, named by analysis scheme
+  res <-
+    tree_files %>%
[…]

### 2022-02-21 _targets.R

@@ -28,6 +28,19 @@ tar_plan(
   pteridocat_ppgi_diff_notes = read_csv(
     pteridocat_ppgi_diff_notes_path
   ),
+  # - Fossil ferns data
+  tar_files_input(
+    fossil_ferns_path,
+    fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
+  ),
+  fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
+    janitor::clean_names(),
+  # - Seed plant trees from Barahona et al 2020
+  tar_files_input(
+    spermato_trees_path,
+    "_targets/user/data_raw/angiosperm-time-tree-2.0-v2.0.zip"
+  ),
+  spermato_trees = load_barahona_trees(spermato_trees_path),
   # - Targets from FTOL workflow
   plastome_metadata_renamed = tar_read(
     plastome_metadata_renamed,
@@ -53,12 +66,6 @@ tar_plan(
     monophy_by_clade,
     store = ftol_cache
   ),
-  tar_files_input(
-    fossil_ferns_path,
-    fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
-  ),
-  fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
-    janitor::clean_names(),
   
   # GenBank ----
   # Analyze number of fern accessions and species in GenBank by type of DNA 

### 2022-02-21 ms/si.Rmd
Update SI2 (fossils)
with latest data from ferncal

https://github.com/spiritu-santi/FernsCals/commit/651321d043d58a3a347925a48db590a5b18d6258

@@
   verify(nrow(.) == nrow(fossil_calibration_tips)) %>%
-  rowwise() %>%
-  mutate(
-    note = jntools::paste3(notes_21, notes_22sep = ". ") %>%
-      str_replace_all("\\.\\.", "\\.") %>%
-      na_if(".") %>%
-      str_replace_all(" \\.$", "\\.")
-  ) %>%
-  ungroup() %>%
   select(
@@
     organs, specimen_holotype, collection,
-    reference, note
+    reference
   )

### 2022-02-21 ms/si2.Rmd
Update SI2 (fossils)
with latest data from ferncal

https://github.com/spiritu-santi/FernsCals/commit/651321d043d58a3a347925a48db590a5b18d6258

@@
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
+  - \setlength{\headheight}{13.6pt}
   - \cfoot{\thepage} # footer
@@
 
-```{r custom-funcs}
-#' Print a single category of fossil data
-#'
-#' @param fossil_data Fossil data in long form, filtered to a single
-#' fossil
-#' @param cat_select Category of data to print
-#'
-#' @return Character vector
-
-print_fossil_cat <- function(fossil_data, cat_select) {
-  data <-
-    fossil_data %>% filter(category == cat_select)
-
-  c(
-    md_heading(cat_select, 3),
-    md_bullet(data$text),
-    md_blank()
+```{r load-targets-si-2}
+# Targets from MS workflow
+tar_load(
+  fossil_ferns_path,
+  store = here::here("_targets")
 )
-}
-
-#' Print data for a single fossil
-#'
-#' @param fossil_data Fossil data in long form
-#' @param fossil_select Selected fossil to print
-#'
-#' @return Character vector
-#'
-#' @examples
-#' print_fossil(fossils_long, "Acrostichum_intertrappeum") %>%
-#' 	glue::as_glue()
-print_fossil <- function(fossil_data, fossil_select) {
-  # Filter data to only selected fossil
-  data <-
-    fossils_long %>% filter(fossil_taxon == fossil_select)
-
-  # Print name of fossil as header in italics
-  header <-
-    data %>% 
-    filter(var == "fossil_taxon_header") %>%
-    pull(value) %>%
-    str_replace_all("_", " ") %>%
-    md_italic() %>%
-    md_heading(1)
-
-  categories <- data %>%
-    filter(!is.na(category)) %>%
-    pull(category) %>%
-    unique()
-
-  body <- map(categories, ~print_fossil_cat(data, .)) %>%
-    unlist()
-
-  c(header, body, md_rule())
-}
 ```
@@
 ```{r prep-fossil-data}
-# Set path to fossil data
-fossils_data_path <- here::here("_targets/user/data_raw/Fossils_Ferns.csv")
-
 # Extract fossil categories (first row of fossil data)
 fossils_categories <-
-  read_csv(fossils_data_path, n_max = 1) %>%
+  read_csv(fossil_ferns_path, n_max = 1) %>%
   pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
@@
 fossils_raw_names <-
-  read_csv(fossils_data_path, skip = 1)
+  read_csv(fossil_ferns_path, skip = 1)
 
@@
 
-```{r print-fossils, include = TRUE, results = "asis", eval = TRUE}
-unique(fossils_long$fossil_taxon) %>%
-  map(~print_fossil(fossils_long, .)) %>%
+```{r print-fossils, include = TRUE, results = "asis"}
+taxa <- unique(fossils_long$fossil_taxon)
+
+map(taxa, ~print_fossil(fossils_long, .)) %>%
   unlist() %>%

### 2022-02-21 R/functions.R

@@ -966,6 +966,60 @@ tabyl_fmt <- function(...) {
     )
 }
 
+#' Print a single category of fossil data
+#'
+#' @param fossil_data Fossil data in long form, filtered to a single
+#' fossil
+#' @param cat_select Category of data to print
+#'
+#' @return Character vector
+
+print_fossil_cat <- function(fossil_data, cat_select) {
+  data <-
+    fossil_data %>% filter(category == cat_select)
+  
+  c(
+    md_heading(cat_select, 3),
+    md_bullet(data$text),
+    md_blank()
+  )
+}
+
+#' Print data for a single fossil
+#'
+#' @param fossil_data Fossil data in long form
+#' @param fossil_select Selected fossil to print
+#'
+#' @return Character vector
+#'
+#' @examples
+#' print_fossil(fossils_long, "Acrostichum intertrappeum") %>%
+#' 	glue::as_glue()
+print_fossil <- function(fossil_data, fossil_select) {
+  # Filter data to only selected fossil
+  data <-
+    fossil_data %>% filter(fossil_taxon == fossil_select)
+  
+  # Print name of fossil as header in italics
+  header <-
+    data %>% 
+    filter(var == "fossil_taxon_header") %>%
+    pull(value) %>%
+    str_replace_all("_", " ") %>%
+    md_italic() %>%
+    md_heading(1)
+  
+  categories <- data %>%
+    filter(!is.na(category)) %>%
+    pull(category) %>%
+    unique()
+  
+  body <- map(categories, ~print_fossil_cat(data, .)) %>%
+    unlist()
+  
+  c(header, body, md_rule())
+}
+
 # References ----
 
 #' Extract citations (formatted like `[@key]`) from an Rmd file

### 2022-02-21 _targets.R

@@ -31,7 +31,7 @@ tar_plan(
   # - Fossil ferns data
   tar_files_input(
     fossil_ferns_path,
-    fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
+    fs::path("_targets/user/data_raw/Fossils_Ferns.csv")
   ),
   fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
     janitor::clean_names(),

### 2022-02-22 ms/manuscript.Rmd
Update MS

@@
 
-# Number of GenBank accessions in final selection
+# Number of GenBank accessions in final selection, excluding plastome
 # first convert to long form
@@
   sanger_accessions_selection %>%
+  # Remove outgroup and plastome species
+  anti_join(
+    unique(select(plastome_seqs_combined_filtered, species)),
+    by = "species") %>%
+  # Remove plastome species
   select(contains("accession")) %>%
@@
 
-# Count number of raw accessions in final selection
+# Count number of raw accessions in final Sanger selection
 n_accs_raw_final <- count_accs(sanger_accessions_selection_long)
 
-# Count number of accessions in final selection, split by locus
+# Count number of accessions in final Sanger selection, split by locus
 n_accs_final <- 
@@
 
+n_species_sanger <- 
+  sanger_accessions_selection %>% 
+  # Remove outgroup and plastome species
+  anti_join(
+    unique(select(plastome_seqs_combined_filtered, species)),
+    by = "species") %>%
+  pull(species) %>%
+  n_distinct() %>%
+  number()
+
 # Analyze join by method
@@
   sanger_accessions_selection %>% 
+  # Remove outgroup and plastome species
+  anti_join(
+    unique(select(plastome_seqs_combined_filtered, species)),
+    by = "species") %>%
   assert(not_na, join_by, species) %>%
@@
 
-The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions ("accessions" in this case includes some cases where a single accession contains multiple loci).
+The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions (note that in some cases a single accession may contain multiple loci).
 Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
 Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
-`r n_accs_first_filter` accessions were retained after excluding these rogues and species names that could not be resolved to an accepted name in the taxonomic database.
-`r n_accs_second_filter` accessions were retained after excluding rogues identified by the all-by-all BLAST search.
-The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`).
-The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`).
-`r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).
+After excluding these and species names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
+After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
+The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species (not including species in the plastome dataset).
+The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`); `r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).
 

### 2022-02-22 ms/si.Rmd
Update MS

@@
     monophy_by_clade,
-    acc_table_long),
+    acc_table_long,
+    plastome_seqs_combined_filtered),
   store = ftol_cache
@@
 sanger_accessions_selection %>% 
+    # Remove outgroup and plastome species
+  anti_join(
+    unique(select(plastome_seqs_combined_filtered, species)),
+    by = "species") %>%
   left_join(select(sanger_sampling, species, outgroup), by = "species") %>%

### 2022-02-22 R/functions.R

@@ -1048,6 +1048,7 @@ extract_citations <- function(rmd_file) {
 #' @param yaml_in String or list; if string, the path to the YAML file to filter.
 #' If list, should be result of reading in a YAML file with yaml::read_yaml()
 #' @param yaml_out Path to write filtered YAML reference file
+#' @param silent Logical; should a warning be issued for missing references?
 #'
 #' @return NULL; externally, the filtered YAML will be written to `yaml_out`
 #' 

### 2022-02-23 R/functions.R
Use tar_file_read

@@ -229,13 +229,13 @@ count_ncbi_species_by_year <- function(gb_taxa, ncbi_names, year_range) {
 #' the evolution of ferns. Molecular Phylogenetics and Evolution 105:200–211. 
 #' https://doi.org/10.1016/j.ympev.2016.09.003
 #'
-#' @param testo_sundue_2016_si_path Path to Testo and Sundue 2016 SI file
+#' @param file Path to Testo and Sundue 2016 SI file
 #'
 #' @return Tibble
 #' 
-parse_ts_dates <- function(testo_sundue_2016_si_path) {
+parse_ts_dates <- function(file) {
   readxl::read_excel(
-    testo_sundue_2016_si_path,
+    file,
     # Divergence dates are in the 5th sheet
     sheet = 5,
     skip = 1
@@ -358,6 +358,16 @@ load_barahona_trees <- function(zip_path) {
   
   return(res)
   
+}
+
+#' Load fossil fern data
+#'
+#' @param file Path to fossil fern data (CSV file)
+#'
+#' @return Tibble
+load_fossil_data <- function(file) {
+  read_csv(file, skip = 1) %>%
+    janitor::clean_names()
 }
 
 # Coverage ----
@@ -1311,4 +1321,45 @@ check_ft_order <- function(type = "all", rmd_file = "ms/manuscript.Rmd") {
 count_non_missing <- function(seq) {
   bases <- ape::base.freq(seq, all = TRUE, freq = TRUE)
   sum(bases[!names(bases) %in% c("n", "N", "-", "?")])
+}
+
+#' Extract the best-scoring model from an IQTREE report
+#'
+#' @param iqtree_log Raw IQTREE report file (ending in .iqtree) read into R
+#'
+#' @return Best-scoring model **by BIC**
+#' 
+extract_iqtree_mod <- function(iqtree_log) {
+  iqtree_log[
+    str_detect(iqtree_log, "Best-fit model according to BIC")] %>%
+    str_match("BIC: ([^$]+)$") %>%
+    magrittr::extract(,2)
+}
+
+#' Extract number of iterations from an IQTREE log
+#'
+#' @param iqtree_log Raw IQTREE log file (ending in .log) read into R
+#'
+#' @return Number of iterations used by IQTREE search
+#' 
+extract_iqtree_iter <- function(iqtree_log) {
+  iqtree_log[
+    str_detect(iqtree_log, "TREE SEARCH COMPLETED AFTER")] %>%
[…]

### 2022-02-23 _targets.R

@@ -15,32 +15,40 @@ tar_plan(
   # Load data ----
   # - Dates from Testo and Sundue 2016 SI
   # (includes Rothfels 2015, Schuettpelz & Pryer 2009)
-  tar_files_input(
-    testo_sundue_2016_si_path,
-    "_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx"
+  tar_file_read(
+    other_dates,
+    "_targets/user/data_raw/1-s2.0-S1055790316302287-mmc2.xlsx",
+    parse_ts_dates(file = !!.x)
   ),
-  other_dates = parse_ts_dates(testo_sundue_2016_si_path),
   # - Notes on differences between Pteridocat and PPGI
-  tar_files_input(
-    pteridocat_ppgi_diff_notes_path,
-    "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv"
-  ),
-  pteridocat_ppgi_diff_notes = read_csv(
-    pteridocat_ppgi_diff_notes_path
+  tar_file_read(
+    pteridocat_ppgi_diff_notes,
+    "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv",
+    readr::read_csv(file = !!.x)
   ),
   # - Fossil ferns data
-  tar_files_input(
-    fossil_ferns_path,
-    fs::path("_targets/user/data_raw/Fossils_Ferns.csv")
+  tar_file_read(
+    fossil_ferns_raw,
+    "_targets/user/data_raw/Fossils_Ferns.csv",
+    load_fossil_data(file = !!.x)
   ),
-  fossil_ferns_raw = read_csv(fossil_ferns_path, skip = 1) %>%
-    janitor::clean_names(),
   # - Seed plant trees from Barahona et al 2020
-  tar_files_input(
-    spermato_trees_path,
-    "_targets/user/data_raw/angiosperm-time-tree-2.0-v2.0.zip"
+  tar_file_read(
+    spermato_trees,
+    "_targets/user/data_raw/angiosperm-time-tree-2.0-v2.0.zip",
+    load_barahona_trees(zip_path = !!.x)
   ),
-  spermato_trees = load_barahona_trees(spermato_trees_path),
+  # Iqtree logs
+  tar_file_read(
+    plastome_iqtree_report,
+    fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.iqtree"), #nolint
+    readr::read_lines(file = !!.x)
+  ),
[…]

### 2022-02-23 ms/manuscript.Rmd
Update MS

@@
     fern_monophy_summ_tbl,
-    fern_nonmono_gen
+    fern_nonmono_gen,
+    plastome_iqtree_report,
+    plastome_iqtree_log
   ),
@@
 We thus updated and edited the initial World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
-We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/joelnitta/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
+We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
 
@@
 The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
-Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].
+Node support was assessed with up to `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013; @Hoang2018].
 
@@
 Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained for a given fossil could change.
-We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a.
+We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
 All constraints other than the root are minimum ages.
@@
 The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
-Code used to generate FTOL and compile this manuscript is available at https://github.com/pteridophylo/ftol_flow and https://github.com/pteridophylo/ftol_ms, respectively.
+Code used to generate FTOL and compile this manuscript is available at https://github.com/fernphy/ftol and https://github.com/fernphy/ftol_ms, respectively.
 A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.
@@
 # Count number of raw accessions in final Sanger selection
+# (excluding plastome)
 n_accs_raw_final <- count_accs(sanger_accessions_selection_long)
@@
 # Count number of accessions in final Sanger selection, split by locus
+# (excluding plastome)
 n_accs_final <- 
@@
 
+# Count number of species in final Sanger selection
+# (excluding plastome)
 n_species_sanger <- 
@@
 # Analyze join by method
+# (excluding plastome)
 join_by_count <-
@@
 n_plast_accs_query <-
+  # initial search includes some accessions that were later excluded
   plastome_metadata_raw_all %>%
@@
 After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
-The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species (not including species in the plastome dataset).
+The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) after excluding species in the plastome dataset included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species.
 The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`); `r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).
 
-The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions.
-`r n_plast_first_filter` accessions were retained after excluding rogues and species names that could not be resolved.
-Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species; `r s_table("accs")`).
+The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions, of which `r n_plast_first_filter` accessions were retained after excluding rogues and species with names that could not be resolved.
+Selection of the longest set of concatenated sequences per species yielded `r n_accs_plast_final` accessions representing unique species (`r s_table("accs")`).
 
@@
 
-Taxon sampling for the current version of FTOL includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat).
+Taxon sampling for FTOL v1.0.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat v1.0.0).
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
@@
 The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
-`r sp_mult_loci$true$n` species (`r sp_mult_loci$true$percent`) were sampled for more than one locus.
-Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
+A majority of species (`r sp_mult_loci$true$n`; `r sp_mult_loci$true$percent`) were sampled for more than one locus.
+Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors are standard deviations unless otherwise mentioned).
 The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
 The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
-Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species).
-`r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.
+Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species); `r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.
 
@@
 
-## Topology
-
-```{r topology-stats}
+## Phylogeny
+
+```{r phylogeny-stats}
+# Bootstrap values for selected clades
 ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
@@
     n = as.integer(n),
-    p = percent(percent)
+    p = percent(percent, accuracy = 1)
   ) %>%
@@
 
+The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_report)`.
+The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)`.
+
 Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
@@
 Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales. 
-
 The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
 There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).
-
-Within Polypodiineae, Hypodematiaceae is sister to all remaining species, which diverge successfully in the order Didymochlaenaceae, Dryopteridaceae, Lomariopsidaceae, Nephrolepidaceae, Tectariaceae, Oleandraceae, then finally Polypodiaceae sister to Davalliaceae.
-
-Within Aspleniineae, Cystopteridaceae is sister to the remaining species, which are split into two clades. 
-One clade consists of Rhacidosoraceae, Diplaziopsidaceae, Desmophlebiaceae, Hemidictyaceae and Aspleniaceae.
-The other clade includes Thelypteridaceae, Woodsiaceae, a clade comprised of Onocleaceae and Blechnaceae, and Athyriaceae.
 
@@
 The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
-We recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
-This contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have recovered Equisetidae as sister to all other ferns; some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae with low support [@Pryer2001; @Qiu2006; @Qiu2007].
-Aside from alternative resolutions of Equisetidae dependent on nuclear envelope or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
+Here, we recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with many other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
+However, this contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have all recovered Equisetidae as sister to all other ferns.
+Some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae, albeit with generally low support [@Pryer2001; @Qiu2006; @Qiu2007].
+Aside from alternative resolutions of Equisetidae dependent on genomic compartment or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
 The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is lacking.
@@
 The monophyly of Polypodiidae (leptosporangiate ferns) is well supported across many studies and not in doubt, as is the status of Osmundales as the first lineage to diverge from the remainder of leptosporangiates [@Pryer2001; @Schneider2004d; @Schuettpelz2007; @Kuo2011; @Testo2016a].
-However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining non-Osmundales leptosporangiates [@Pryer2004a; @Lehtonen2017].
+However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining (non-Osmundales) leptosporangiates [@Pryer2004a; @Lehtonen2017].
 Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018].
-We recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
+Here, we recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
 However, current plastome sampling only includes `r plastome_gleich_genera_n` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
@@
 
+Relationships of families within Aspleniineae (eupolypods II) have been difficult resolve due to the ancient, rapid radiation of this clade [@Rothfels2012].
+Our analysis robustly resolves the relationships between all families in Aspleniineae and is in agreement with a recent plastome analysis with similar sampling [@Du2021].
+Notably, previous phylogenomic studies that had different or less well-supported topologies did not sample all eupolypod II families [Desmophlebiaceae and Hemidictyaceae absent\; @Wei2017a; @Qi2018; @Shen2018].
+The family sister to the remainder of Aspleniineae has been resolved as either Aspleniaceae [`r eg`, @Schneider2004d; @Testo2016a; @Shen2018] or Cystopteridaceae [`r eg`, @Kuo2011; @Wei2017a; @Qi2018].
+Here, we recovered Cystopteridaceae as sister to the remainder of eupolypod II families, which are split into two clades.
+One clade consists of Rhachidosoraceae, Diplaziopsidaceae, Aspleniaceae, Desmophlebiaceae, and Hemidictyaceae [RHADD clade of @Du2021; Clade E of @Sundue2013].
+The other clade includes Thelypteridaceae, Woodsiaceae, Athyriaceae, Onocleaceae, and Blechnaceae [WOBAT clade of @Du2021; Clade B of @Sundue2013].
+Each of these two clades is supported by morphological synapomorphies [@Sundue2013; @Du2021].
+
 Taken together, our results for nodes of contention in the fern phylogeny generally agree with other plastid phylogenomic analyses and are well within the realm of plausible hypotheses generated to date.
@@
 
-We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
-This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
-Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
-However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
+We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+This may be due to our use of a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
+Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
+However, we found inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
 Moreover, our set of calibration points more than doubles the number used by previous studies.
-It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 
@@
 
-Currently, the vast majority of publicly available DNA sequence data for ferns is Sanger-sequenced plastid gene regions (`r figure("genbank")`).
 Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
-However, the plastid tree does not necessarily agree with trees inferred from other data sources.
-Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing, and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
+However, a tree derived from plastid data may not reflect those inferred from other data sources.
+Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing [`r eg`, @Sessa2012b; @Zhou2017a; @Wei2021a] and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
 Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
-Therefore, we have no expectation that there is a single fern tree of life; rather, multiple trees exist that each capture an element of biological reality, and FTOL is merely one such possible tree.
-
-Recent studies using transcriptomics are clarifying the backbone of the fern phylogeny using many (25--2400) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
+Therefore, a major future research goal for fern molecular systematics should be to combine nuclear and plastid datasets to infer species trees with comprehensive sampling.
+
+Recent transcriptomic studies are gradually clarifying the backbone of the fern phylogeny using many (25--`r number(2400)`) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
 The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFlag) project, which seeks to generate genomic data (ca.
-300 single-to-low copy nuclear gene regions) for all flagellate plants (bryophytes, lycophytes, ferns, and gymnosperms) [@Breinholt2021].
+300 single-to-low copy nuclear gene regions) for all flagellate plants [bryophytes, lycophytes, ferns, and gymnosperms\; @Breinholt2021].
 GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.
-The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets still is far less that available for plastid data (`r figure("genbank")`).
-Furthermore, many subclades of ferns are under active investigation using plastid markers using both Sanger and next-gen sequencing, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
-We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for several more years.
+
+However, the rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets is still far less than that available for plastid data (`r figure("genbank")`).
+Furthermore, many subclades of ferns are under active investigation using both Sanger and next-gen sequencing of plastid markers, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
+We therefore expect that the methodology outlined here will continue to be useful to generate a maximally sampled plastid fern tree of life for many years to come.
 
@@
 We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
-FTOL is available via the "ftol" R package, as well as a web portal (FIXME: add URL when ready).
-Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibration points) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom DNA sequences [`r eg`\; @Nitta2021.08.26.457744].
-
-A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data from joining [@Page2008].
-During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/joelnitta/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand).
-We selected R because it is widely used by the ecological research community, well established, and freely available [@Lai2019].
+FTOL is available via the "ftolr" R package, as well as a web portal (https://fernphy.github.io).
+Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibration points) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2021.08.26.457744].
+
+A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data merging [@Page2008].
+During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/fernphy/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand).
+We selected R because it is widely used by the biological research community, well established, and freely available [@Lai2019].
 The "pteridocat" package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
@@
 We want to be clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
-FTOL cannot substitute for systematic careful studies at finer taxonomic scales that include sampling multiple individuals per species and/or other sources (`r eg`, morphological, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
+FTOL cannot substitute for careful systematic studies at finer taxonomic scales that include sampling of multiple individuals per species and/or other sources (`r eg`, morphological, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
 
@@
 Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript is not meant to be the last.
-Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree on a regular, semi-annual basis.
-We envision that FTOL will be integrated with the next iteration of PPG, PPG II, to provide the most recent hypotheses on the monophyly of various fern taxa, which will in turn enable a more natural classification system.
+Rather, we plan to re-run these analyses as additional data become available on GenBank, and release updated versions of the tree on a regular basis.
+We envision that FTOL will be integrated with the next iteration of the Pteridophyte Phylogeny Group classification, PPG II, to provide the most recent hypotheses on the monophyly of various fern taxa, which will in turn enable a more natural classification system.
 
@@
 We are aware that our methodology cannot produce a "perfect" tree, nor is that our goal.
-Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked (`r eg`, placement of species within genera, which we did not have the resources to inspect), or proper taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
+Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked (`r eg`, placement of species within genera, which we did not have the resources to inspect) or because an updated taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
 That is why we have made our methodology (code), data, and software (R packages and docker image) completely open and available to the research community.
-It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through the GitHub repository (https://github.com/joelnitta/ftol).
+It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through the GitHub repository (https://github.com/fernphy/ftol).
 This way, FTOL will continually improve and keep pace with the currently available data and taxonomic hypotheses of ferns.
@@
 
-Members of the Iwasaki lab provided helpful comments that improved the manuscript.
-This study was supported by JSPS KAKENHI Grant Number 16H06279.
+Members of the Iwasaki lab provided comments that improved the manuscript.
+Alexander White provided helpful comments on an early version of the analysis.
+This study was supported by JSPS KAKENHI Grant Number 16H06279 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
 The authors thank Michael Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.

### 2022-02-23 ms/si.Rmd
Update MS

@@
   doc_type: manual
+nocite: '@*'
 ---
@@
 - `r here::here("ms/si_references_other.yaml")`
-- `r here::here("ms/si_references_other.yaml")`
-- `r here::here("ms/si_fossils.bib")`
 ---
@@
     by = "species") %>%
-  left_join(select(sanger_sampling, species, outgroup), by = "species") %>%
-  # FIXME why is this NA?
-  # filter(is.na(outgroup)) %>%
-  filter(outgroup == FALSE) %>%
-  select(-outgroup) %>%
   assert(not_na, join_by, species) %>%

### 2022-02-23 ms/manuscript.Rmd
Update MS

@@
     fern_sanger_seqs_raw,
+    fossil_calibration_tips,
+    monophy_by_clade,
+    plastid_calibration_dates,
+    plastid_tree_dated,
+    plastome_alignment,
+    plastome_alignment_tbl,
+    plastome_metadata_raw_all,
+    plastome_metadata_renamed,
+    plastome_metadata_renamed,
+    plastome_seqs_combined_filtered,
+    plastome_tree,
+    ppgi_taxonomy,
+    pteridocat,
     raw_fasta_all,
+    sanger_accessions_selection,
+    sanger_alignment,
+    sanger_alignment_tbl,
     sanger_seqs_combined_filtered,
     sanger_seqs_rogues_removed,
-    sanger_accessions_selection,
-    plastome_seqs_combined_filtered,
-    plastome_metadata_raw_all,
-    plastome_metadata_renamed,
-    sanger_alignment,
-    plastome_alignment,
-    plastome_metadata_renamed,
-    ppgi_taxonomy,
-    plastid_calibration_dates,
-    pteridocat,
-    sanger_alignment_tbl,
-    plastome_alignment_tbl,
-    monophy_by_clade,
-    plastome_tree,
-    plastid_tree_dated,
-    fossil_calibration_tips,
-    ts_fossil_calibration_tips,
-    treepl_cv_results),
+    treepl_cv_results,
+    ts_fossil_calibration_tips
+  ),
   store = ftol_cache
@@
     fern_nonmono_gen,
-    plastome_iqtree_report,
     plastome_iqtree_log
@@
 We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
-All constraints other than the root are minimum ages.
+All constraints other than the root are minimum ages (`r s_figure("fossil_map")`).
 
@@
 
-The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_report)`.
+The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_log)`.
 The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)`.

### 2022-02-23 ms/si.Rmd
Update MS

@@
     acc_table_long,
-    plastome_seqs_combined_filtered),
+    plastome_seqs_combined_filtered,
+    sanger_tree_rooted,
+    fossil_calibrations_for_treepl),
   store = ftol_cache
@@
 
+```{r fossil-map}
+# Make tibble with node number and label for ggtree
+node_labs_tbl <-
+  fossil_calibrations_for_treepl %>%
+  transmute(
+    fossil_label = str_remove_all(min, "min = "),
+    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_tree_rooted, c(.x, .y))),
+    calibration_type = case_when(
+      !is.na(min) & is.na(max) ~ "minimum",
+      !is.na(min) & !is.na(max) ~ "fixed",
+      TRUE ~ NA_character_
+    )
+  ) %>%
+  assert(not_na, calibration_type) %>%
+  right_join(
+    tibble(
+      node = Ntip(sanger_tree_rooted) + seq_along(sanger_tree_rooted$node.label)
+    ), by = "node"
+  ) %>%
+  mutate(
+    fossil_label = replace_na(fossil_label, ""),
+    calibration_type = replace_na(calibration_type, "none")) %>%
+  arrange(node)
+
+# Make plot
+fossil_map_plot <- 
+  ggtree(
+    sanger_tree_rooted,
+    branch.length = "none", size = 0.1) %<+% node_labs_tbl +
+  # Add node points
+  geom_nodepoint(
+    aes(color = calibration_type), size = 1) +
+  scale_color_manual(
+    values = c(
+      minimum = palette_okabe_ito("vermillion"), 
+      fixed = palette_okabe_ito("skyblue"),
+      none = "transparent")) +
+  # Add node labels
+  geom_text(aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5) +
+  theme(legend.position = "bottom")
+
+# Save plot
+ggsave(
+  plot = fossil_map_plot,
+  file = (result_file(s_figure("fossil_map"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("fossil_map")`**.
+Maximum likelihood fern phylongey with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
+Number for each constraint is age in Ma.
+
+\newpage
+
 ```{r locus-upset, fig.width = 7, fig.height = 5}
@@
 Period name abbreviations as follows: FIXME.
-
-\newpage

### 2022-02-23 R/functions.R

@@ -1323,16 +1323,17 @@ count_non_missing <- function(seq) {
   sum(bases[!names(bases) %in% c("n", "N", "-", "?")])
 }
 
-#' Extract the best-scoring model from an IQTREE report
+#' Extract the best-scoring model from an IQTREE log
 #'
-#' @param iqtree_log Raw IQTREE report file (ending in .iqtree) read into R
+#' @param iqtree_log Raw IQTREE log file (ending in .log) read into R
 #'
-#' @return Best-scoring model **by BIC**
+#' @return Best-scoring model
 #' 
 extract_iqtree_mod <- function(iqtree_log) {
   iqtree_log[
-    str_detect(iqtree_log, "Best-fit model according to BIC")] %>%
-    str_match("BIC: ([^$]+)$") %>%
+    # Formatted for IQTREE2
+    str_detect(iqtree_log, "Best-fit model: ")] %>%
+    str_match("([^ ]+) chosen") %>%
     magrittr::extract(,2)
 }
 

### 2022-02-23 _targets.R

@@ -40,11 +40,6 @@ tar_plan(
   ),
   # Iqtree logs
   tar_file_read(
-    plastome_iqtree_report,
-    fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.iqtree"), #nolint
-    readr::read_lines(file = !!.x)
-  ),
-  tar_file_read(
     plastome_iqtree_log,
     fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.log"), #nolint
     readr::read_lines(file = !!.x)

### 2022-02-24 R/functions.R
Update func to read in fossils
latest fossils CSV doesn't have extra header row

@@ -366,7 +366,7 @@ load_barahona_trees <- function(zip_path) {
 #'
 #' @return Tibble
 load_fossil_data <- function(file) {
-  read_csv(file, skip = 1) %>%
+  read_csv(file) %>%
     janitor::clean_names()
 }
 

### 2022-02-24 ms/manuscript.Rmd
Update MS

@@
 We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
-We selected `r n_calibrations_ferns` fern fossils to use as constraints from a curated list of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
+We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
+The full database is available at https://fernphy.github.com/FernCal.
 In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).

### 2022-02-24 ms/si.Rmd
Update MS

--- 2022-02-23 ms/si.Rmd
@@
 
+```{r fossil-map}
+# Make tibble with node number and label for ggtree
+node_labs_tbl <-
+  fossil_calibrations_for_treepl %>%
+  transmute(
+    fossil_label = str_remove_all(min, "min = "),
+    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_tree_rooted, c(.x, .y))),
+    calibration_type = case_when(
+      !is.na(min) & is.na(max) ~ "minimum",
+      !is.na(min) & !is.na(max) ~ "fixed",
+      TRUE ~ NA_character_
+    )
+  ) %>%
+  assert(not_na, calibration_type)
+
+# Make plot
+fossil_map_plot <- 
+  ggtree(
+    sanger_tree_rooted,
+    branch.length = "none", size = 0.1) %<+% node_labs_tbl +
+  # Add node points
+  geom_nodepoint(
+    aes(color = calibration_type), size = 1) +
+  scale_color_manual(
+    "Calibration type",
+    values = c(
+      minimum = palette_okabe_ito("vermillion"), 
+      fixed = palette_okabe_ito("skyblue")),
+    na.value = "transparent"
+  ) +
+  # Add node labels
+  geom_text(aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5) +
+  theme(legend.position = "bottom")
+
+# Save plot
+ggsave(
+  plot = fossil_map_plot,
+  file = (result_file(s_figure("fossil_map"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("fossil_map")`**.
+Maximum likelihood fern phylongey with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
+Number for each constraint is age in Ma.
+
+\newpage
+
 ```{r locus-upset, fig.width = 7, fig.height = 5}

### 2022-02-24 R/functions.R

@@ -742,19 +742,19 @@ assess_monophy <- function(
 #' to define the clade
 #' @param sanger_sampling Tibble with one row per species in the tree, including
 #' columns "genus", "family", etc.
-#' @param plastid_tree_dated Dated phylogeny
+#' @param sanger_tree_dated Dated phylogeny
 #'
 #' @return Character vector: age of the clade defined by the two genera
 #' 
-get_crown_age_from_genus_pair <- function(tip_pair, sanger_sampling, plastid_tree_dated) {
-  tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
+get_crown_age_from_genus_pair <- function(tip_pair, sanger_sampling, sanger_tree_dated) {
+  tree_height <- max(phytools::nodeHeights(sanger_tree_dated))
   sanger_sampling %>%
     filter(genus %in% tip_pair) %>%
     group_by(genus) %>%
     slice(1) %>%
     pull(species) %>%
-    ape::getMRCA(plastid_tree_dated, .) %>%
-    phytools::nodeheight(plastid_tree_dated, .) %>%
+    ape::getMRCA(sanger_tree_dated, .) %>%
+    phytools::nodeheight(sanger_tree_dated, .) %>%
     magrittr::subtract(tree_height, .) %>%
     number(accuracy = 0.1)
 }
@@ -766,23 +766,23 @@ get_crown_age_from_genus_pair <- function(tip_pair, sanger_sampling, plastid_tre
 #'
 #' @param sanger_sampling Tibble with one row per species in the tree, including
 #' columns "genus", "family", etc.
-#' @param plastid_tree_dated Dated phylogeny 
+#' @param sanger_tree_dated Dated phylogeny 
 #' @param ppgi_taxonomy PPGI taxonomic system for ferns and lycophytes
 #'
 #' @return Tibble with columns "family" and "age" (stem age of the family)
 get_stem_family_age <- function(
   sanger_sampling, 
-  plastid_tree_dated, ppgi_taxonomy) {
+  sanger_tree_dated, ppgi_taxonomy) {
   
   # Get overall tree height
-  tree_height <- max(phytools::nodeHeights(plastid_tree_dated))
+  tree_height <- max(phytools::nodeHeights(sanger_tree_dated))
   
   # Get ages for families
   family_monophy <- 
     select(sanger_sampling, species, family) %>%
     assess_monophy(
       taxon_sampling = .,
-      tree = plastid_tree_dated,
[…]

### 2022-02-24 _targets.R
Make FTOL cache a file target

@@ -6,10 +6,6 @@ library(glue)
 
 source("R/functions.R")
 
-# Specify path to FTOL cache
-# FIXME: change to ftol_cache when ready
-ftol_cache <- here::here("working/_targets")
-
 tar_plan(
   
   # Load data ----
@@ -38,6 +34,12 @@ tar_plan(
     "_targets/user/data_raw/angiosperm-time-tree-2.0-v2.0.zip",
     load_barahona_trees(zip_path = !!.x)
   ),
+  # Specify path to FTOL cache
+  # FIXME: change to ftol_cache when ready
+  tar_file(
+    ftol_cache,
+    "working/_targets"
+  ),
   # Iqtree logs
   tar_file_read(
     plastome_iqtree_log,
@@ -53,8 +55,8 @@ tar_plan(
     plastome_tree,
     store = ftol_cache
   ),
-  plastid_tree_dated = tar_read(
-    plastid_tree_dated,
+  sanger_tree_dated = tar_read(
+    sanger_tree_dated,
     store = ftol_cache
   ),
   ppgi_taxonomy = tar_read(
@@ -114,7 +116,7 @@ tar_plan(
   # - Make Sanger sampling table
   sanger_sampling = make_sanger_sampling_tbl(
     plastome_metadata_renamed,
-    sanger_tree = plastid_tree_dated,
+    sanger_tree = sanger_tree_dated,
     ppgi_taxonomy = ppgi_taxonomy),
   # - Crown age of various clades
   crown_ages = list(
@@ -124,10 +126,10 @@ tar_plan(
     eupoly_i = c("Hypodematium", "Polypodium"),
     eupoly_ii = c("Cystopteris", "Asplenium")) %>%
     map(~get_crown_age_from_genus_pair(
-      ., sanger_sampling, plastid_tree_dated)),
+      ., sanger_sampling, sanger_tree_dated)),
   # - Stem age of fern families
   family_stem_ages = get_stem_family_age(
-    sanger_sampling, plastid_tree_dated, ppgi_taxonomy),
+    sanger_sampling, sanger_tree_dated, ppgi_taxonomy),
   # - Read in Du 2021 dates
   tar_file_read(
     du_dates_all,

### 2022-02-24 ms/manuscript.Rmd
Add SI LTT plot

@@
 Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
-Except for eupolypods II, these ages are all older than those reported in the most recent global fern phylogeny [@Testo2016a].
-Estimated stem ages of fern families were generally older than previous studies (`r figure("divtimes")`).
+Except for crown ferns, these ages are all significantly older than those reported in the most recent global fern phylogeny [@Testo2016a].
+Estimated stem ages of fern families were also older than previous studies (`r figure("divtimes")`).
 We did not compare crown ages of families across studies because differences in crown ages in smaller clades are likely affected by species sampling.
@@
 We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
-This may be due to our use of a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
+This is almost certainly due to our use of a completely revised and greatly expanded set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
 Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
@@
 It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+
+Notably, the scenario of fern diversification suggested by our results strongly conflicts with the hypothesis that Polypodiales diversified in the shadow of angiosperms [@Schneider2004d].
+Rather, we estimate that polypod diversification occurred coincidentally or even preceding the diversification of angiosperms (`r s_figure("ltt")`).
 

### 2022-02-24 ms/si.Rmd
Add SI LTT plot

--- 2022-02-23 ms/si.Rmd
@@
 
+```{r load-functions-si, eval = params$doc_type == "manual"}
+# Load functions (only if running interactively)
+source("_targets.R")
+```
+
 ```{r formatters-load-si}
@@
 ```{r, formatters-si}
-```
-
-```{r load-functions-si, eval = params$doc_type == "manual"}
-# Load functions (only if running interactively)
-source(here::here("R/functions.R"))
 ```
@@
     acc_table_long,
-    plastome_seqs_combined_filtered),
+    plastome_seqs_combined_filtered,
+    sanger_tree_rooted,
+    fossil_calibrations_for_treepl,
+    sanger_tree_dated),
   store = ftol_cache
@@
     fern_monophy_summ_tbl,
-    fern_nonmono_gen
+    fern_nonmono_gen,
+    spermato_trees
   ),
@@
 
-```{r locus-upset, fig.width = 7, fig.height = 5}
+```{r fossil-map-make}
+# Make tibble with node number and label for ggtree
+node_labs_tbl <-
+  fossil_calibrations_for_treepl %>%
+  transmute(
+    fossil_label = str_remove_all(min, "min = "),
+    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_tree_rooted, c(.x, .y))),
+    calibration_type = case_when(
+      !is.na(min) & is.na(max) ~ "minimum",
+      !is.na(min) & !is.na(max) ~ "fixed",
+      TRUE ~ NA_character_
+    )
+  ) %>%
+  assert(not_na, calibration_type)
+
+# Make plot
+fossil_map_plot <- 
+  ggtree(
+    sanger_tree_rooted,
+    branch.length = "none", size = 0.1, color = "grey30") %<+% node_labs_tbl +
+  # Add node points
+  geom_nodepoint(
+    aes(color = calibration_type), size = 0.6) +
+  scale_color_manual(
+    "Calibration type",
+    values = c(
+      minimum = palette_okabe_ito("vermillion"), 
+      fixed = palette_okabe_ito("skyblue")),
+    na.value = "transparent"
+  ) +
+  # Add node labels
+  geom_text(aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  theme(legend.position = "bottom")
+```
+
+```{r fossil-map-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = fossil_map_plot,
+  file = (result_file(s_figure("fossil_map"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("fossil_map")`**.
+Maximum likelihood fern phylongey with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
+Number for each constraint is age in Ma.
+
+\newpage
+
+```{r locus-upset, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
 # Make tibble with one row per locus per species
@@
 
-```{r du-ages, fig.width = 7, fig.height = 5}
+```{r du-ages, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
 # Helper function for subsetting dates to get into
@@
 \newpage
+
+```{r ltt, fig.width = 7, fig.height = 6, eval = params$doc_type != "manual"}
+# Make lineages through time (LTT) plot comparing Angiosperms and polypods
+
+## Angiosperms ##
+# FIXME: if we are really going to use this, move it to _targets.R
+# Li 2019 data
+li_species_tree <- ape::read.tree(here::here("_targets/user/data_raw/2881_dating_Species_extracted.tree"))
+li_og <- read_lines(here::here("_targets/user/data_raw/2881_dating_Species_extracted_og.txt"))
+li_angio_tree <- ape::drop.tip(li_species_tree, li_og)
+
+li_tt_data <- make_ltt_tib(li_angio_tree, "Li_2019")
+
+# Define outgroup taxa
+og <- c(
+  "Pinales_Pinaceae_Pinus_spp", "Gnetales_Welwitschiaceae_Welwitschia_spp",
+  "Cycadales_Cycadaceae_Cycas_spp", "Gnetales_Gnetaceae_Gnetum_spp", 
+  "Pinales_Cupressaceae_Metasequoia_glyptostroboides", 
+  "Ginkgoales_Ginkgoaceae_Ginkgo_spp")
+
+# Drop outgroups
+angio_trees <-
+  map(spermato_trees, ~ape::drop.tip(., og))
+
+# Get data for LTT plot
+angio_tt_data <-
+  map2_df(angio_trees, names(angio_trees), ~make_ltt_tib(.x, .y))
+
+## Polypods ##
+
+# Get data for LTT plot
+poly_tt_data <-
+  sanger_sampling %>%
+  filter(order == "Polypodiales") %>%
+  pull(species) %>% 
+  ape::keep.tip(sanger_tree_dated, .) %>%
+  make_ltt_tib(., "Polypodiales")
+
+## Combine and plot ##
+tt_data <-
+  bind_rows(angio_tt_data, poly_tt_data, li_tt_data)
+
+ltt_dataset_cols <- c(
+CC_complete = palette_okabe_ito("orange"),
+CC_conservative = palette_okabe_ito("skyblue"),
+RC_complete = palette_okabe_ito("bluishgreen"),
+RC_conservative = palette_okabe_ito("blue"),
+UC_complete = palette_okabe_ito("vermillion"),
+UC_conservative = palette_okabe_ito("reddishpurple"),
+Li_2019 = palette_okabe_ito("yellow"),
+Polypodiales = palette_okabe_ito("black")
+)
+
+raw_count <- ggplot(tt_data, aes(x = time, y = N, color = dataset)) +
+  geom_step() +
+  scale_y_continuous("Number of lineages") +
+  scale_x_reverse("Age (millions of years)") +
+  coord_geo(
+    dat = list("periods", "eras"),
+    pos = list("b", "b"),
+    abbrv = list(TRUE, FALSE),
+    xlim = c(310, 0),
+    fill = "transparent",
+    color = "grey20",
+    lab_color = "grey30",
+    height = unit(1, "lines"),
+    expand = FALSE,
+    size = 3) +
+  scale_color_manual("Dataset", values = ltt_dataset_cols) +
+  theme_bw()
+
+ggplot(tt_data, aes(x = time, y = log_n_percent, color = dataset)) +
+  geom_step() +
+  scale_y_continuous("Log number of lineages (as percent)", labels = scales::percent) +
+  scale_x_reverse("Age (millions of years)") +
+  coord_geo(
+    dat = list("periods", "eras"),
+    pos = list("b", "b"),
+    abbrv = list(TRUE, FALSE),
+    xlim = c(310, 0),
+    fill = "transparent",
+    color = "grey20",
+    lab_color = "grey30",
+    height = unit(1, "lines"),
+    expand = FALSE,
+    size = 3) +
+  scale_color_manual("Dataset", values = ltt_dataset_cols) +
+  theme_bw() +
+  theme(legend.position = "bottom")
+```
+
+**`r s_figure("ltt")`**.
+Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramirez-Barahona2020].
+Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020.
+Log number of lineages shown as percent of total to enable comparison across trees.

### 2022-02-24 R/functions.R

@@ -1199,6 +1199,26 @@ node_height_from_tips <- function(phy, tips) {
     phytools::nodeheight(phy, .)
 }
 
+#' Make a dataframe for a lineages-through-time (LTT) plot
+#' 
+#' Time is in units (millions of years) before present
+#'
+#' @param phy Phylogeny
+#' @param dataset Name of dataset to add (optional)
+#'
+make_ltt_tib <- function(phy, dataset = NULL) {
+  res <- ape:::ltt.plot.coords(phy, backward = TRUE, type = "S") %>%
+    as_tibble() %>%
+    mutate(time = time*-1) %>%
+    mutate(
+      log_n = log(N),
+      log_n_percent = log_n/max(log_n),
+      n_percent = N/max(N)
+    )
+  if (!is.null(dataset)) res <- mutate(res, dataset = dataset)
+  return(res)
+}
+
 # Etc -----
 
 #' Function to get support value in BS%

### 2022-02-25 ms/si.Rmd
Read fossils from ftol cache

@@
     fossil_calibrations_for_treepl,
-    sanger_tree_dated),
+    sanger_tree_dated,
+    fossil_ferns_all),
   store = ftol_cache
@@
     family_stem_ages,
-    fossil_ferns_raw,
     pteridocat_ppgi_diff_refs_out,
@@
 fossils_table <-
-  fossil_ferns_raw %>%
+  fossil_ferns_all %>%
   inner_join(
-    select(fossil_calibration_tips, fossil_taxon), by = "fossil_taxon") %>%
+    unique(select(fossil_calibration_tips, n_fos)), by = "n_fos") %>%
   verify(nrow(.) == nrow(fossil_calibration_tips)) %>%

### 2022-02-25 _targets.R

@@ -22,12 +22,6 @@ tar_plan(
     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv",
     readr::read_csv(file = !!.x)
   ),
-  # - Fossil ferns data
-  tar_file_read(
-    fossil_ferns_raw,
-    "_targets/user/data_raw/Fossils_Ferns.csv",
-    load_fossil_data(file = !!.x)
-  ),
   # - Seed plant trees from Barahona et al 2020
   tar_file_read(
     spermato_trees,

### 2022-02-25 build.R
Add build script

@@ -1,0 +1,2 @@
+#!/usr/bin/env Rscript
+targets::tar_make()

### 2022-02-26 ms/manuscript.Rmd
WIP: add bs values

@@
     sanger_alignment_tbl,
+    sanger_con_tree_rooted,
     sanger_seqs_combined_filtered,
@@
 
+# Mean bootstrap for consensus tree
+tibble(
+  bs = sanger_con_tree_rooted$node.label)
+
 # Format monophyly status as list for MS
@@
 
+Bootstrap support was generally across the tree (mean BS) and particularly high at deeper nodes.
+Relationships within some more recently diverged clades were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).
+
 ## Divergence times
@@
 Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
-Except for crown ferns, these ages are all significantly older than those reported in the most recent global fern phylogeny [@Testo2016a].
-Estimated stem ages of fern families were also older than previous studies (`r figure("divtimes")`).
+Our estimates for leptosporangiates and Polypodiales are both `r ca` 50 ma older than the most recent global fern phylogeny [@Testo2016a], while other ages are similar.
+Estimated stem ages of fern families were mostly older than previous studies (`r figure("divtimes")`).
 We did not compare crown ages of families across studies because differences in crown ages in smaller clades are likely affected by species sampling.

### 2022-02-26 ms/si.Rmd
WIP: add bs values

@@
     sanger_tree_dated,
+    sanger_con_tree_rooted,
     fossil_ferns_all),
@@
 
+```{r bs-plot}
+# Parse bootstrap values from IQTREE
+phy <- sanger_con_tree_rooted
+
+bs_values <-
+tibble(
+  label_orig = phy$node.label
+) %>%
+  mutate(label_new = case_when(
+    label_orig == "Root" ~ NaN,
+    TRUE ~ parse_number(label_orig) * 0.01
+  )) %>%
+  pull(label_new)
+
+# Make tibble for printing BS values
+bs_tibble <- tibble(
+  # hard-code node ID: internal nodes start after tip nodes,
+  # and phy$node.label is in the same order as internal nodes
+  node = 1:Nnode(phy) + Ntip(phy), 
+  # Don't print BS = 100%
+  bootstrap = bs_values)
+
+# Make plot
+# optional: round version
+bs_plot_round <- 
+  ggtree(
+    phy,
+    layout = "circular",
+    branch.length = "none", 
+    size = 0.1, color = "grey30") %<+% bs_tibble +
+  # Add node points
+  geom_nodepoint(
+    aes(color = bootstrap), size = 0.2) +
+  scale_color_viridis_c(
+    "BS",
+    option = "B", na.value = "transparent",
+    breaks = 0.25*0:4, labels = percent(0.25*0:4))
+
+# Make plot
+bs_plot <- 
+  ggtree(
+    phy,
+    size = 0.1, color = "grey30") %<+% bs_tibble +
+  # Add node points
+  geom_nodepoint(
+    aes(color = bootstrap), size = 0.1) +
+  scale_color_viridis_c(
+    "BS",
+    option = "B", na.value = "transparent",
+    breaks = 0.25*0:4, labels = percent(0.25*0:4)) +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  theme(legend.position = "bottom")
+```
+
+```{r bs-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = bs_plot,
+  file = (result_file(s_figure("bs"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("bs")`**.
+Consensus fern phylogeny with node labels bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
+
+\newpage
+
 ```{r du-ages, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}

### 2022-02-26 ms/manuscript.Rmd
WIP: add bs values

@@
     sanger_seqs_rogues_removed,
-    treepl_cv_results,
     ts_fossil_calibration_tips
@@
     fern_nonmono_gen,
-    plastome_iqtree_log
+    plastome_iqtree_log,
+    treepl_cv_results
   ),

### 2022-02-26 _targets.R

@@ -40,6 +40,11 @@ tar_plan(
     fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.log"), #nolint
     readr::read_lines(file = !!.x)
   ),
+  tar_file_read(
+    treepl_cv_results,
+    fs::path(ftol_cache, "user/intermediates/treepl/treepl_cv_out.txt"),
+    readr::read_lines(file = !!.x)
+  ),
   # - Targets from FTOL workflow
   plastome_metadata_renamed = tar_read(
     plastome_metadata_renamed,

### 2022-02-26 ms/manuscript.Rmd
Add bootstrap tree

@@
     plastid_calibration_dates,
-    plastid_tree_dated,
+    sanger_tree_dated,
     plastome_alignment,
@@
     plastome_seqs_combined_filtered,
-    plastome_tree,
+    plastome_tree_rooted,
     ppgi_taxonomy,
@@
     accepted_species,
-    sanger_sampling,
-    coverage_by_rank,
-    total_coverage,
     bb_coverage_by_rank,
     bb_total_coverage,
-    gb_species_by_year,
-    ref_files,
-    other_dates,
+    coverage_by_rank,
     crown_ages,
-    fossil_ferns_raw,
     family_stem_ages,
@@
     fern_nonmono_gen,
+    fossil_ferns_raw,
+    gb_species_by_year,
+    other_dates,
     plastome_iqtree_log,
+    ref_files,
+    sanger_iqtree_log,
+    sanger_sampling,
+    total_coverage,
     treepl_cv_results
@@
 # Bootstrap values for selected clades
-ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
-hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
-cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
-denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
-
-# Mean bootstrap for consensus tree
-tibble(
-  bs = sanger_con_tree_rooted$node.label)
+ferns_bs <- get_bs(plastome_tree_rooted, c("Equisetum_arvense", "Danaea_nodosa"))
+hym_gle_bs <- get_bs(plastome_tree_rooted, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
+cyath_salv_bs <- get_bs(plastome_tree_rooted, c("Plagiogyria_adnata", "Azolla_filiculoides"))
+denn_poly_bs <- get_bs(plastome_tree_rooted, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))
+
+# Tibble of bootstrap for consensus trees
+bs_tib <-
+tibble(bs = sanger_con_tree_rooted$node.label, dataset = "sanger") %>%
+bind_rows(tibble(bs = plastome_tree_rooted$node.label, dataset = "plastome")) %>%
+  mutate(
+    bs = na_if(bs, "Root") %>% 
+      parse_number
+    )
+
+# Percent of plastome phylogeny with 100% BS support
+plastome_bs100 <-
+bs_tib %>%
+  filter(dataset == "plastome") %>%
+  filter(!is.na(bs)) %>%
+  mutate(is_100 = if_else(bs == 100, "true", "false")) %>%
+  tabyl(is_100) %>%
+  mutate(percent = percent(percent, accuracy = 0.1)) %>%
+  split(.$is_100)
+
+# Mean BS values for printing
+bs <-
+bs_tib %>%
+  group_by(dataset) %>%
+  summarize(
+    mean = mean(bs, na.rm = TRUE),
+    sd = sd(bs, na.rm = TRUE)
+  ) %>%
+  mutate(across(c(mean, sd), ~number(., accuracy = 0.1))) %>%
+  split(.$dataset)
 
@@
     success_fun = success_logical)
-```
-
-The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_log)`.
-The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)`.
+
+# Check that model for plastome and sanger was same
+assertthat::assert_that(
+  extract_iqtree_mod(sanger_iqtree_log) == 
+    extract_iqtree_mod(plastome_iqtree_log),
+  msg = "Same model not selected for plastome and sanger trees"
+)
+```
+
+The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger phylogenies.
+The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` and `r extract_iqtree_iter(sanger_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)` and `r extract_iqtree_corr(sanger_iqtree_log)` for the plastome and Sanger phylogenies, respectively.
 
 Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
-The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (`r figure("backbone")`).
+The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (Figures `r figure_num("backbone")`, `r s_figure_num("backbone_full")`).
 The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
@@
 
-Bootstrap support was generally across the tree (mean BS) and particularly high at deeper nodes.
-Relationships within some more recently diverged clades were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).
+Bootstrap support was generally across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`; `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
+Relationships within some genera were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).
 
@@
   pull(species)
-
-plastome_tree_rooted <- phytools::reroot(
-  plastome_tree,
-  getMRCA(
-    plastome_tree,
-    c("Physcomitrium_patens", "Marchantia_polymorpha", "Anthoceros_angustus")
-  )) %>%
-  ladderize()
 
@@
     taxon_sampling = .,
-    tree = plastid_tree_dated,
+    tree = sanger_tree_dated,
     tax_levels = "major_clade"
@@
 a <- ggtree(
-  plastid_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
+  sanger_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
   ftol_geom_cladelab(
@@
 # Convert BS support values to numeric
-phy$node.label <- parse_number(phy$node.label)
+# phy$node.label <- parse_number(phy$node.label)
 # Scale to total height of 100 for easy plotting
@@
 # Make tibble for printing BS values
-bs_tibble <- tibble(
-  # hard-code node ID: internal nodes start after tip nodes,
-  # and phy$node.label is in the same order as internal nodes
-  node = 1:Nnode(phy) + Ntip(phy), 
-  # Don't print BS = 100%
-  bootstrap = ifelse(phy$node.label == 100, "", phy$node.label))
+bs_tibble <- make_bs_tbl(phy) %>%
+  # only print values < 100
+  mutate(bs = case_when(
+    bs == 100 ~ "",
+    TRUE ~ as.character(bs)
+  ))
 
@@
   # now we can show BS values using geom_text()
-  geom_text(aes(label = bootstrap), hjust = -.25, size = 2.5) +
+  geom_text(aes(label = bs), hjust = -.25, size = 2.5) +
   # add tip labels

### 2022-02-26 ms/si.Rmd
Add bootstrap tree

--- 2022-02-25 ms/si.Rmd
@@
   c(
-    sanger_alignment_tbl,
+    acc_table_long,
+    fossil_calibration_tips,
+    fossil_calibrations_for_treepl,
+    fossil_ferns_all,
+    monophy_by_clade,
+    plastome_seqs_combined_filtered,
+    plastome_tree,
+    plastome_tree_rooted,
+    ppgi_taxonomy,
+    sanger_accessions_selection,
+    sanger_con_tree_rooted,
     sanger_sampling,
-    sanger_accessions_selection,
-    plastome_tree,
-    plastid_tree_dated,
-    fossil_calibration_tips,
-    ppgi_taxonomy,
-    monophy_by_clade,
-    acc_table_long,
-    plastome_seqs_combined_filtered,
+    sanger_tree_dated,
     sanger_tree_rooted,
-    fossil_calibrations_for_treepl,
-    sanger_tree_dated,
-    fossil_ferns_all),
+    sanger_alignment_tbl
+  ),
   store = ftol_cache
@@
   c(
-    accepted_species,
-    gb_species_by_year,
-    ref_files,
-    other_dates,
     du_dates_all,
     family_stem_ages,
-    pteridocat_ppgi_diff_refs_out,
-    pteridocat_ppgi_diff_notes,
     fern_monophy_summ_tbl,
     fern_nonmono_gen,
-    spermato_trees
+    gb_species_by_year,
+    other_dates,
+    pteridocat_ppgi_diff_notes,
+    pteridocat_ppgi_diff_refs_out,
+    ref_files,
+    spermato_trees,
+    accepted_species
   ),
@@
 
+```{r backbone-full-plot}
+phy <- plastome_tree_rooted
+
+# Make tibble for printing BS values
+bs_tibble <- make_bs_tbl(phy) %>%
+  # only print values < 100
+  mutate(bs = case_when(
+    bs == 100 ~ "",
+    TRUE ~ as.character(bs)
+  ))
+
+# Make plot
+bb_full_plot <- 
+  ggtree(phy, size = 0.2, color = "grey10") %<+% 
+  bs_tibble +
+  # now we can show BS values using geom_text()
+  geom_text(
+    aes(label = bs), hjust = -.25, size = 1.0,
+    color = "grey50") +
+  # add tip labels
+  geom_tiplab(aes(label = label), size = 1.0)
+```
+
+```{r backbone-full-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = bb_full_plot,
+  file = (result_file(s_figure("backbone_full"), "pdf")),
+  height = 20, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("bs")`**.
+Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
+Values at nodes are bootstrap support (%); only values less than 100% are shown.
+
+\newpage
+
+```{r bs-plot}
+phy <- sanger_con_tree_rooted
+
+# Make tibble for printing BS values
+bs_tibble <- make_bs_tbl(phy)
+
+# Make plot
+# optional: round version
+bs_plot_round <- 
+  ggtree(
+    phy,
+    layout = "circular",
+    branch.length = "none", 
+    size = 0.1, color = "grey30") %<+% bs_tibble +
+  # Add node points
+  geom_nodepoint(
+    aes(color = bs), size = 0.2) +
+  scale_color_viridis_c(
+    "BS",
+    option = "B", na.value = "transparent",
+    breaks = 0.25*0:4, labels = percent(0.25*0:4))
+
+# Make plot
+bs_plot <- 
+  ggtree(
+    phy,
+    size = 0.1, color = "grey30") %<+% bs_tibble +
+  # Add node points
+  geom_nodepoint(
+    aes(color = bs), size = 0.1) +
+  scale_color_viridis_c(
+    "BS",
+    option = "B", na.value = "transparent",
+    breaks = 0.25*0:4, labels = percent(0.25*0:4)) +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  theme(legend.position = "bottom")
+```
+
+```{r bs-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = bs_plot,
+  file = (result_file(s_figure("bs"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("bs")`**.
+Consensus phylogeny (Sanger dataset) with node labels showing bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
+The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
+
+\newpage
+
 ```{r du-ages, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}

### 2022-02-26 R/functions.R

@@ -1219,6 +1219,26 @@ make_ltt_tib <- function(phy, dataset = NULL) {
   return(res)
 }
 
+#' Make a tibble of bootstrap values and their node numbers in a phylogeny
+#'
+#' @param phy Phylogeny with bootstrap values as a number or "Root"
+#'
+#' @return Tibble of bootstrap values
+make_bs_tbl <- function(phy) {
+  bs_values = tibble(bs = phy$node.label) %>%
+    mutate(
+      bs = na_if(bs, "Root") %>% 
+        parse_number) %>%
+    pull(bs)
+  
+  tibble(
+    # hard-code node ID: internal nodes start after tip nodes,
+    # and phy$node.label is in the same order as internal nodes
+    node = 1:Nnode(phy) + Ntip(phy), 
+    bs = bs_values)
+}
+
+
 # Etc -----
 
 #' Function to get support value in BS%
@@ -1381,6 +1401,7 @@ extract_iqtree_iter <- function(iqtree_log) {
 extract_iqtree_corr <- function(iqtree_log) {
   iqtree_log[
     str_detect(iqtree_log, "Bootstrap correlation coefficient of split occurrence frequencies: ")] %>%
+    last() %>%
     str_match("Bootstrap correlation coefficient of split occurrence frequencies: ([^$]+)$") %>%
     magrittr::extract(,2)
 }

### 2022-02-26 _targets.R

@@ -41,6 +41,12 @@ tar_plan(
     readr::read_lines(file = !!.x)
   ),
   tar_file_read(
+    sanger_iqtree_log,
+    fs::path(ftol_cache, "user/intermediates/iqtree/sanger/sanger_alignment.phy.2022-02-23.log"), #nolint
+    readr::read_lines(file = !!.x)
+  ),
+  # treePL results
+  tar_file_read(
     treepl_cv_results,
     fs::path(ftol_cache, "user/intermediates/treepl/treepl_cv_out.txt"),
     readr::read_lines(file = !!.x)

### 2022-03-09 ms/manuscript.Rmd
Fix order of major clades in Fig 2

@@
 backbone_taxonomy <-
-  tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
+  plastome_tree_rooted %>%
+  ladderize() %>%
+  get_tips_in_ape_plot_order() %>%
+  tibble(species = .) %>%
   # Add higher taxonomy
@@
 
+# Make tibble of backbone tips with higher-level taxonomy
+sanger_clade_exemp_taxa <-
+  sanger_tree_dated %>%
+  ladderize() %>%
+  get_tips_in_ape_plot_order() %>%
+  tibble(species = .) %>%
+  # Add higher taxonomy
+  left_join(sanger_sampling, by = "species") %>%
+  # Add column for phylogenetic order
+  mutate(phy_order = seq_len(nrow(.))) %>%
+  # Outgroup lacks family data, so call "og"
+  mutate(
+    family = case_when(
+      outgroup == TRUE ~ "og",
+      TRUE ~ family),
+    major_clade = case_when(
+      outgroup == TRUE ~ "og",
+      TRUE ~ major_clade)
+  ) %>%
+# Make exemplar set of species in backbone representing plot groups
+  # Slice to one exemplar species per major clade
+  group_by(major_clade) %>%
+  slice(1) %>%
+  ungroup() %>%
+  # Slicing arranges alphabetically, so put back in (unladderized) phylo order
+  arrange(phy_order) %>%
+  # Drop outgroup
+  filter(family != "og")
+
 # Format y-axis text: show major clade abbreviations in bold
@@
   filter(rank == "major_clade") %>%
-  left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
+  left_join(sanger_clade_exemp_taxa, by = c(name = "major_clade")) %>%
   arrange(phy_order) %>%

### 2022-03-09 R/functions.R

@@ -827,7 +827,7 @@ get_stem_family_age <- function(
     ) %>%
     assert(is_uniq, family) %>%
     assert(not_na, everything()) %>%
-    select(family, age)
+    select(family, stem_node, age)
 }
 
 #' Filter monophyly table to ferns, add taxonomic level

### 2022-03-09 _targets.R
Use tar_file_read() instead of tar_read()

@@ -6,6 +6,10 @@ library(glue)
 
 source("R/functions.R")
 
+# Specify path to FTOL cache
+# FIXME: change to ftol_cache when ready
+ftol_cache <- "working/_targets"
+
 tar_plan(
   
   # Load data ----
@@ -28,13 +32,7 @@ tar_plan(
     "_targets/user/data_raw/angiosperm-time-tree-2.0-v2.0.zip",
     load_barahona_trees(zip_path = !!.x)
   ),
-  # Specify path to FTOL cache
-  # FIXME: change to ftol_cache when ready
-  tar_file(
-    ftol_cache,
-    "working/_targets"
-  ),
-  # Iqtree logs
+  # - Targets from FTOL workflow
   tar_file_read(
     plastome_iqtree_log,
     fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.log"), #nolint
@@ -45,36 +43,40 @@ tar_plan(
     fs::path(ftol_cache, "user/intermediates/iqtree/sanger/sanger_alignment.phy.2022-02-23.log"), #nolint
     readr::read_lines(file = !!.x)
   ),
-  # treePL results
   tar_file_read(
     treepl_cv_results,
     fs::path(ftol_cache, "user/intermediates/treepl/treepl_cv_out.txt"),
     readr::read_lines(file = !!.x)
   ),
-  # - Targets from FTOL workflow
-  plastome_metadata_renamed = tar_read(
+  tar_file_read(
     plastome_metadata_renamed,
-    store = ftol_cache
+    fs::path(ftol_cache, "objects/plastome_metadata_renamed"),
+    readRDS(file = !!.x)
   ),
-  plastome_tree = tar_read(
+  tar_file_read(
     plastome_tree,
-    store = ftol_cache
+    fs::path(ftol_cache, "objects/plastome_tree"),
+    readRDS(file = !!.x)
   ),
-  sanger_tree_dated = tar_read(
+  tar_file_read(
     sanger_tree_dated,
-    store = ftol_cache
+    fs::path(ftol_cache, "objects/sanger_tree_dated"),
+    readRDS(file = !!.x)
   ),
-  ppgi_taxonomy = tar_read(
+  tar_file_read(
     ppgi_taxonomy,
-    store = ftol_cache
+    fs::path(ftol_cache, "objects/ppgi_taxonomy"),
+    readRDS(file = !!.x)
   ),
[…]

### 2022-03-23 ms/manuscript.Rmd
Update MS
fix target names

@@
     accs_exclude,
+    con_fossil_calibration_tips,
     fern_sanger_seqs_raw,
-    fossil_calibration_tips,
-    monophy_by_clade,
     plastid_calibration_dates,
-    sanger_tree_dated,
     plastome_alignment,
@@
     sanger_alignment_tbl,
+    sanger_con_tree_dated,
     sanger_con_tree_rooted,
@@
 n_calibrations_ferns <-
-  fossil_calibration_tips %>%
+  con_fossil_calibration_tips %>%
   assert(is_uniq, node_calibrated) %>%
@@
 n_calibrations_total <-
-  fossil_calibration_tips %>%
+  con_fossil_calibration_tips %>%
   assert(is_uniq, node_calibrated) %>%
@@
   mutate(subfamily = fct_reorder(subfamily, n)) %>%
-  mutate(n = jntools::english2(n)) %>%
+  mutate(n = map_chr(n, jntools::english2)) %>%
   arrange(desc(subfamily))
@@
 assertthat::assert_that(
-  extract_iqtree_mod(sanger_iqtree_log) == 
-    extract_iqtree_mod(plastome_iqtree_log),
+  unique(extract_iqtree_mod(sanger_iqtree_log)) == 
+    unique(extract_iqtree_mod(plastome_iqtree_log)),
   msg = "Same model not selected for plastome and sanger trees"
@@
     taxon_sampling = .,
-    tree = sanger_tree_dated,
+    tree = sanger_con_tree_dated,
     tax_levels = "major_clade"
@@
 sanger_clade_exemp_taxa <-
-  sanger_tree_dated %>%
+  sanger_con_tree_dated %>%
   ladderize() %>%
@@
 a <- ggtree(
-  sanger_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
+  sanger_con_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
   ftol_geom_cladelab(

### 2022-03-23 ms/si.Rmd
Update MS
fix target names

@@
     acc_table_long,
-    fossil_calibration_tips,
-    fossil_calibrations_for_treepl,
+    con_fossil_calibration_tips,
+    con_fossil_calibrations_for_treepl,
     fossil_ferns_all,
-    monophy_by_clade,
     plastome_seqs_combined_filtered,
@@
     sanger_accessions_selection,
+    sanger_sampling,
+    sanger_con_tree_dated,
     sanger_con_tree_rooted,
-    sanger_sampling,
-    sanger_tree_dated,
-    sanger_tree_rooted,
     sanger_alignment_tbl
@@
   inner_join(
-    unique(select(fossil_calibration_tips, n_fos)), by = "n_fos") %>%
-  verify(nrow(.) == nrow(fossil_calibration_tips)) %>%
+    unique(select(con_fossil_calibration_tips, n_fos)), by = "n_fos") %>%
+  verify(nrow(.) == nrow(con_fossil_calibration_tips)) %>%
   select(
@@
 node_labs_tbl <-
-  fossil_calibrations_for_treepl %>%
+  con_fossil_calibrations_for_treepl %>%
   transmute(
     fossil_label = str_remove_all(min, "min = "),
-    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_tree_rooted, c(.x, .y))),
+    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_con_tree_rooted, c(.x, .y))),
     calibration_type = case_when(
@@
   ggtree(
-    sanger_tree_rooted,
+    sanger_con_tree_rooted,
     branch.length = "none", size = 0.1, color = "grey30") %<+% node_labs_tbl +
@@
   pull(species) %>% 
-  ape::keep.tip(sanger_tree_dated, .) %>%
+  ape::keep.tip(sanger_con_tree_dated, .) %>%
   make_ltt_tib(., "Polypodiales")

### 2022-03-23 ms/si2.Rmd
Update MS
fix target names

@@
 # Targets from MS workflow
+
+# Specify path to FTOL cache
+# FIXME: change to ftol_cache when ready
+ftol_cache <- here::here("working/_targets")
+
 tar_load(
-  fossil_ferns_path,
-  store = here::here("_targets")
+  con_fossil_calibration_tips,
+  store = ftol_cache
 )
+
+fossil_ferns_path <- fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
 ```
@@
 ```{r prep-fossil-data}
-# Extract fossil categories (first row of fossil data)
+# Read all fossils (including those not used in this study)
+# Get raw column names
+fossils_raw_names <- read_csv(fossil_ferns_path)
+# Load data with cleaned column names
+fossils_clean_names <- janitor::clean_names(fossils_raw_names)
+
+# Make tibble of fossil data categories
+raw_col_names <- colnames(fossils_raw_names)
 fossils_categories <-
-  read_csv(fossil_ferns_path, n_max = 1) %>%
-  pivot_longer(names_to = "category", values_to = "raw_names", everything()) %>%
+  tibble(raw_name = raw_col_names) %>%
   mutate(
-    category = str_replace_all(category, "\\.\\.\\.[0-9]+", "skip") %>%
-      na_if("skip")) %>%
-  fill(category) %>%
-  mutate(category = str_to_title(category)) %>%
-  filter(category != "Pteridophy")
+    category = case_when(
+      raw_name %in% raw_col_names[1:4] ~ "Calibration",
+      raw_name %in% raw_col_names[5:12] ~ "Fossil Identity",
+      raw_name %in% raw_col_names[13:18] ~ "Fossil Age",
+      raw_name %in% raw_col_names[19:21] ~ "Fossil Locality",
+      raw_name %in% raw_col_names[22:24] ~ "Fossil Specimen",
+      TRUE ~ NA_character_
+    )
+  )
 
-# Get raw column names
-fossils_raw_names <-
-  read_csv(fossil_ferns_path, skip = 1)
-
-# Load data with cleaned column names
-fossils_clean_names <- fossils_raw_names %>% janitor::clean_names()
+# Get list of fossils used
+n_fos_used <-
+con_fossil_calibration_tips %>%
+  filter(!is.na(n_fos)) %>%
+  assert(is_uniq, n_fos) %>%
+  select(n_fos)
 
@@
   # Filter to only those actually used in the study
-  filter(used == "X") %>%
-  select(-used) %>%
-  select(fossil_taxon, everything())
+  inner_join(
+    n_fos_used, by = "n_fos"
+  ) %>%
+  verify(all(sort(.$n_fos) == sort(n_fos_used$n_fos)))
 
-# Make a tibble of column namees for formatting
+# Make a tibble of column names for formatting
 name_tbl <- tibble(
-  raw_names = colnames(fossils_raw_names),
+  raw_name = colnames(fossils_raw_names),
   var = colnames(fossils_clean_names)
@@
 # Convert to long format
-fossils_long <- fossils %>%
+fossils_long <- 
+  fossils %>%
   mutate(
@@
   mutate(
-    text = as.character(glue::glue("{raw_names}: {value}"))
+    text = as.character(glue::glue("{raw_name}: {value}"))
   ) %>%
-  left_join(fossils_categories, by = "raw_names") %>%
+  left_join(fossils_categories, by = "raw_name") %>%
   arrange(fossil_taxon, category, var_order)

### 2022-03-23 _targets.R

@@ -10,6 +10,88 @@ source("R/functions.R")
 # FIXME: change to ftol_cache when ready
 ftol_cache <- "working/_targets"
 
+# Define Rmd targets outside of main workflow
+tmp <- capture.output({
+  ms_doc_tar <- tar_render(
+    ms_doc,
+    "ms/manuscript.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::word_document2",
+    params = list(doc_type = "doc")
+  )
+  ms_pdf_tar <- tar_render(
+    ms_pdf,
+    "ms/manuscript.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::pdf_document2",
+    params = list(doc_type = "pdf")
+  )
+  si_doc_tar <- tar_render(
+    si_doc,
+    "ms/si.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::word_document2",
+    params = list(doc_type = "doc")
+  )
+  si_pdf_tar <- tar_render(
+    si_pdf,
+    "ms/si.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::pdf_document2",
+    params = list(doc_type = "pdf")
+  )
+  si_doc_2_tar <- tar_render(
+    si_doc_2,
+    "ms/si2.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::word_document2",
+    params = list(doc_type = "doc")
+  )
+  si_pdf_2_tar <- tar_render(
+    si_pdf_2,
+    "ms/si2.Rmd",
+    output_dir = "results",
+    output_format = "bookdown::pdf_document2",
+    params = list(doc_type = "pdf")
+  )
+  },
+  type = "message"
+)
+
+
+# tar_render(
+#   si_doc,
+#   "ms/si.Rmd",
+#   knit_root_dir = "ms",
+#   output_dir = "results",
+#   output_format = "bookdown::word_document2",
+#   params = list(doc_type = "doc")
+# ),
+# tar_render(
+#   si_pdf,
+#   "ms/si.Rmd",
+#   output_dir = "results",
+#   output_format = "bookdown::pdf_document2",
+#   params = list(doc_type = "pdf")
+# ),
+# tar_render(
+#   si_2_doc,
+#   "ms/si2.Rmd",
+#   knit_root_dir = "ms",
+#   output_dir = "results",
+#   output_format = "bookdown::word_document2",
[…]

### 2022-03-23 ms/si.Rmd
Use rmdref to format citations in SI tables

@@
   doc_type: manual
-nocite: '@*'
 ---
@@
     pteridocat_ppgi_diff_notes,
-    pteridocat_ppgi_diff_refs_out,
+    pteridocat_ppgi_diff_notes_file,
     ref_files,
@@
 
+```{r load-pandoc, cache = FALSE}
+# Specify bibliography and CSL files for pandoc
+bib <- c(
+  here::here("ms/si_references.yaml"),
+  here::here("ms/si_references_other.yaml"),
+  here::here("ms/si_fossils.bib"))
+csl <- here::here("ms/frontiers.csl")
+```
+
+```{r extract-table-citations}
+# Make list of references cited in pteridocat vs. ppgi differences table
+# to include in bibliography
+table_citations <- extract_citations(pteridocat_ppgi_diff_notes_file)$key %>%
+  paste0("@", .) %>%
+  paste0(collapse = ", ")
+```
+
+---
+nocite: |
+  `r table_citations`
+---
+
 ```{r genera-diff-make}
@@
   left_join(
-    select(pteridocat_ppgi_diff_notes, -ref, -n_species),
+    select(pteridocat_ppgi_diff_notes, -n_species),
     by = c("diff", "genus")
@@
   left_join(
-    select(pteridocat_ppgi_diff_notes, -ref), by = c("diff", "genus")
+    pteridocat_ppgi_diff_notes, by = c("diff", "genus")
   ) %>%
@@
 genera_diff_table <- 
-  bind_rows(pteridocat_not_in_ppgi, ppgi_not_in_pteridocat)
+  bind_rows(pteridocat_not_in_ppgi, ppgi_not_in_pteridocat) %>%
+  # Format references
+  mutate(note = cite_pd(
+    note, bib = bib, csl = csl, context = pteridocat_ppgi_diff_notes_file)) %>%
+  # Check that all notes have references properly formatted
+  # pandoc leaves a '?' if not
+  assert(function(x) str_detect(x, "\\?", negate = TRUE), note)
 ```
@@
 References in "notes" column provided at end of Supplementary Materials.
+"World Ferns" refers to @Hassler2022, which was modified to create Pteridocat.
 
 ```{r fossils-make}
-# Filter raw fossil data to just fossils used, write out as CSV
-fossils_table <-
+# Filter raw fossil data to just fossils used
+fossils_table_raw <-
   fossil_ferns_all %>%
@@
     organs, specimen_holotype, collection,
-    reference
+    reference_raw = reference,
+    notes_on_affinities
   )
+
+# Write out list of cited refs as "context" file
+temp_context <- fs::path_abs(tempfile(fileext = ".txt"))
+
+fossils_table_raw %>%
+  select(reference_raw, notes_on_affinities) %>%
+  pivot_longer(everything()) %>%
+  filter(!is.na(value)) %>%
+  pull(value) %>%
+  write_lines(temp_context)
+
+# Format references with pandoc
+fossils_table <-
+fossils_table_raw %>%
+  mutate(reference = cite_pd(
+    reference_raw, bib = bib, csl = csl, context = temp_context)) %>%
+  select(-reference_raw, -notes_on_affinities)
+
+fs::file_delete(temp_context)
 ```
@@
 Log number of lineages shown as percent of total to enable comparison across trees.
+
+## References

### 2022-03-23 R/functions.R

@@ -1030,74 +1030,6 @@ print_fossil <- function(fossil_data, fossil_select) {
   c(header, body, md_rule())
 }
 
-# References ----
-
-#' Extract citations (formatted like `[@key]`) from an Rmd file
-#'
-#' @param rmd_file Character vector; path to Rmd file
-#'
-#' @return Data frame with one column 'key'
-#' 
-extract_citations <- function(rmd_file) {
-  read_lines(rmd_file) %>%
-    stringr::str_split(" |;") %>% 
-    unlist %>% 
-    magrittr::extract(., stringr::str_detect(., "@")) %>% 
-    stringr::str_remove_all("^[^@]*") %>%
-    stringr::str_remove_all('\\[|\\]|\\)|\\(|\\.$|,|\\{|\\}|\\\\|\\"') %>% 
-    magrittr::extract(., stringr::str_detect(., "^@|^-@")) %>% 
-    stringr::str_remove_all("^@|^-@") %>% 
-    unique %>% 
-    sort %>%
-    tibble(key = .)
-}
-
-#' Filter a list of references in YAML format to those occurring in an Rmd file
-#'
-#' @param rmd_file Character vector; Path to Rmd file(s)
-#' @param yaml_in String or list; if string, the path to the YAML file to filter.
-#' If list, should be result of reading in a YAML file with yaml::read_yaml()
-#' @param yaml_out Path to write filtered YAML reference file
-#' @param silent Logical; should a warning be issued for missing references?
-#'
-#' @return NULL; externally, the filtered YAML will be written to `yaml_out`
-#' 
-filter_refs_yaml <- function(rmd_file, yaml_in = "ms/main_library.yaml", yaml_out = "ms/references.yaml", silent = FALSE) {
-  
-  # Parse RMD file and extract citation keys
-  citations <- purrr::map_df(rmd_file, extract_citations)
-  
-  # Read in YAML including all references exported from Zotero
-  if (inherits(yaml_in, "character")) {
-    ref_yaml <- yaml::read_yaml(yaml_in)
-  } else if (inherits(yaml_in, "list")) {
-    ref_yaml <- yaml_in
-  } else {
-    stop("`yaml_in` must be a path to a YAML file (string) or a list read in with yaml::read_yaml()")
-  }
-  
-  # Extract all citation keys from full YAML
-  cite_keys_all <- purrr::map_chr(ref_yaml$references, "id") %>%
-    tibble::tibble(
-      key = .,
-      order = 1:length(.)
[…]

### 2022-03-23 R/ms_packages.R

@@ -18,6 +18,7 @@ library(ggrepel)
 library(ggokabeito)
 library(ggtext)
 library(janitor)
+library(rmdref)
 
 # Resolve conflicts
 conflict_prefer("filter", "dplyr")

### 2022-03-23 R/process_refs.R

@@ -1,23 +1,23 @@
 # Filter a list of references in YAML format to those cited in an RMD file
 library(tidyverse)
-source(here::here("R/functions.R"))
+library(rmdref)
 
 # Filter YAML references from Zotero library.
 # data_raw/main_library.yaml has been exported from Zotero
 # like this: file -> "export library" -> "Better CSL YAML"
 filter_refs_yaml(
-  "ms/manuscript.Rmd",
-  "data_raw/main_library.yaml",
-  "ms/references.yaml", silent = TRUE)
+  rmd_file = "ms/manuscript.Rmd",
+  yaml_in = "data_raw/main_library.yaml",
+  yaml_out = "ms/references.yaml", silent = TRUE)
 
 # Same for SI
 filter_refs_yaml(
-  c(
+  rmd_file = c(
     "ms/si.rmd", # SI
     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv" # refs in SI tables
   ),
-  "data_raw/main_library.yaml",
-  "ms/si_references.yaml", silent = TRUE)
+  yaml_in = "data_raw/main_library.yaml",
+  yaml_out = "ms/si_references.yaml", silent = TRUE)
 
 # Write out 'aux' file with citation keys to make collection in Zotero
 # so reference data are easier to work with.

### 2022-03-25 ms/manuscript.Rmd
Update MS

@@
 Recently the whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others.
+
+Although some relationships within Polypodiineae (eupolypods I) had previously been resolved differently between various studies using Sanger sequencing, such as Nephrolepidaceae sister to Lomariopsidaceae [@Schuettpelz2007; @Zhang2015] vs.
+Nephrolepidaceae sister to Tectariaceae, Oleandraceae, Davalliaceae, and Polypodiaceae [@Kuo2011; @Lehtonen2011; @Liu2013; @Testo2016a], our study is in agreement with both nuclear [@Qi2018; @Shen2018] and plastid phylogenomic analyses [@Du2021] that support the latter.
+Similarly, although Didymochlaenaceae had previously been identified as either sister to the remainder of Polypodiineae [@Kuo2011; @Zhang2015; @Testo2016a] or nested with Hypodematiaceae [@Schuettpelz2007; @Lehtonen2011] by studies using Sanger sequencing, our study as well as nuclear [@Qi2018] and plastid phylogenomic analyses [@Du2021] indicate that Hypodematiaceae is sister to the remainder of the clade.
 

### 2022-03-25 ms/si.Rmd
Update MS

@@
   - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
-  - \fancyhead[RE,RO]{Supplementary Material} # add header on right side
+  - \fancyhead[RE,RO]{Supplementary Material 1} # add header on right side
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
@@
     con_fossil_calibrations_for_treepl,
+    fossil_ferns_raw,
     fossil_ferns_all,
@@
     sanger_con_tree_rooted,
-    sanger_alignment_tbl
+    sanger_alignment_tbl,
+    ts_fossil_calibration_tips
   ),
@@
     reference_raw, bib = bib, csl = csl, context = temp_context)) %>%
-  select(-reference_raw, -notes_on_affinities)
+  select(-reference_raw, -notes_on_affinities) %>%
+  # Make sure all references in the table have entries in the bibliography
+  assert(function(x) str_detect(x, "\\?", negate = TRUE), reference)
 

### 2022-03-25 ms/si2.Rmd
Update MS

@@
   - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
-  - \fancyhead[RE,RO]{Supplementary Material} # add header on right side
+  - \fancyhead[RE,RO]{Supplementary Material 2} # add header on right side
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
@@
 
+```{r load-functions-si-2, eval = params$doc_type == "manual"}
+# Load functions (only if running interactively)
+source("_targets.R")
+```
+
 ```{r load-targets-si-2}
@@
 tar_load(
+  c(
     con_fossil_calibration_tips,
-  store = ftol_cache
+    fossil_ferns_raw,
+    fossil_ferns_all),
+  store = ftol_cache,
 )
 
-fossil_ferns_path <- fs::path(ftol_cache, "user/data_raw/Fossils_Ferns.csv")
+# Targets from MS workflow
+tar_load(
+  fossil_notes,
+  store = here::here("_targets")
+)
 ```
@@
 
-```{r prep-fossil-data}
-# Read all fossils (including those not used in this study)
-# Get raw column names
-fossils_raw_names <- read_csv(fossil_ferns_path)
-# Load data with cleaned column names
-fossils_clean_names <- janitor::clean_names(fossils_raw_names)
+```{r tidy-fossil-data}
+# fossil_ferns_raw is read in from Fossils_Ferns.csv directly
+# fossil_ferns_all is same, but with clean names by janitor::clean_names()
 
 # Make tibble of fossil data categories
-raw_col_names <- colnames(fossils_raw_names)
+raw_col_names <- colnames(fossil_ferns_raw)
 fossils_categories <-
@@
 fossils <-
-  fossils_clean_names %>%
+  fossil_ferns_all %>%
   # Filter to only those actually used in the study
@@
   ) %>%
-  verify(all(sort(.$n_fos) == sort(n_fos_used$n_fos)))
+  verify(all(sort(.$n_fos) == sort(n_fos_used$n_fos))) %>%
+  # Replace with formatted data
+  select(
+    -node_calibrated, -full_taxon_name, 
+    -affinities, -notes_on_affinities, -notes_on_age) %>%
+  left_join(fossil_notes, by = "n_fos") %>%
+  # Put column names back in proper order
+  select(all_of(colnames(fossil_ferns_all)))
+```
 
+```{r write-new-formatted, eval = FALSE}
+# If needed, write out notes for further editing (if need to add new rows)
+fossil_ferns_all %>%
+  inner_join(
+    n_fos_used, by = "n_fos"
+  ) %>%
+  anti_join(fossil_notes, by = "n_fos") %>%
+  select(n_fos,
+         node_calibrated, full_taxon_name, 
+         affinities, notes_on_affinities, notes_on_age) %>%
+  bind_rows(fossil_notes) %>%
+  assert(is_uniq, n_fos) %>%
+  arrange(n_fos) %>%
+  write_csv(here::here("_targets/user/data_raw/fossils_used_formatted.csv"))
+```
+
+```{r format-for-printing}
 # Make a tibble of column names for formatting
 name_tbl <- tibble(
-  raw_name = colnames(fossils_raw_names),
-  var = colnames(fossils_clean_names)
+  raw_name = colnames(fossil_ferns_raw),
+  var = colnames(fossil_ferns_all)
 ) %>%
@@
     across(everything(), as.character),
-    fossil_taxon_header = glue("†{fossil_taxon}")) %>%
+    fossil_taxon_header = fossil_taxon) %>%
   pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
@@
   left_join(fossils_categories, by = "raw_name") %>%
-  arrange(fossil_taxon, category, var_order)
+  arrange(fossil_taxon, category, var_order) %>%
+  # Drop NA entries
+  filter(!is.na(value))
 ```
 
-These data are exported from the FernCal v1.0 database (https://github.com/pteridophy/FossilFerns), which includes `r nrow(fossils_clean_names)` records of fossils to use as calibration points for molecular dating of ferns.
+These data are exported from the FernCal v1.0 database (https://github.com/fernphy/ferncal), which includes `r nrow(fossil_ferns_raw)` records of fossils to use as calibration points for molecular dating of ferns.
 

### 2022-03-25 _targets.R

@@ -58,40 +58,6 @@ tmp <- capture.output({
   type = "message"
 )
 
-
-# tar_render(
-#   si_doc,
-#   "ms/si.Rmd",
-#   knit_root_dir = "ms",
-#   output_dir = "results",
-#   output_format = "bookdown::word_document2",
-#   params = list(doc_type = "doc")
-# ),
-# tar_render(
-#   si_pdf,
-#   "ms/si.Rmd",
-#   output_dir = "results",
-#   output_format = "bookdown::pdf_document2",
-#   params = list(doc_type = "pdf")
-# ),
-# tar_render(
-#   si_2_doc,
-#   "ms/si2.Rmd",
-#   knit_root_dir = "ms",
-#   output_dir = "results",
-#   output_format = "bookdown::word_document2",
-#   params = list(doc_type = "doc")
-# ),
-# tar_render(
-#   si_2_pdf,
-#   "ms/si2.Rmd",
-#   knit_root_dir = "ms",
-#   output_dir = "results",
-#   output_format = "bookdown::pdf_document2",
-#   params = list(doc_type = "pdf")
-# )
-
-
 tar_plan(
   
   # Load data ----
@@ -106,6 +72,12 @@ tar_plan(
   tar_file_read(
     pteridocat_ppgi_diff_notes,
     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv",
+    readr::read_csv(file = !!.x)
+  ),
+  # - Formatted fossil notes
+  tar_file_read(
+    fossil_notes,
+    "_targets/user/data_raw/fossils_used_formatted.csv",
     readr::read_csv(file = !!.x)
   ),
   # - Seed plant trees from Barahona et al 2020

### 2022-03-25 ms/manuscript.Rmd
Update MS
Add TS fossil date plots

@@
     fossil_ferns_raw,
+    ftol_ts_comp_mod,
     gb_species_by_year,
@@
 
+```{r calibration-stats}
+ftol_ts_comp_stats <-
+ftol_ts_comp_mod %>%
+  broom::glance()
+```
+
 We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 This is almost certainly due to our use of a completely revised and greatly expanded set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
+Our set of fossil calibration points did not add any extremely old fossils (> 200 Ma) that would be expected to strongly push back ages across the tree [save for stem Marattiaceae, which is well-known for its extensive fossil record\; @Rothwell2018]; rather, the vast majority of newly added calibration points are younger than 150 Ma (`r s_figure("comp-fossils")`).
+
 Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
@@
 It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
-
-Notably, the scenario of fern diversification suggested by our results strongly conflicts with the hypothesis that Polypodiales diversified in the shadow of angiosperms [@Schneider2004d].
+To test this hypothesis, we conducted additional analysis using the fossil constraints of @Testo2016a but otherwise the exact same methods (FTOL analyzed with treePL).
+The resulting stem family ages show a much closer agreement with @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is responsible for the older ages observed in the current study.
+
+Notably, the scenario of fern diversification suggested by our results conflicts with the hypothesis that Polypodiales diversified "in the shadow of angiosperms" [@Schneider2004d].
 Rather, we estimate that polypod diversification occurred coincidentally or even preceding the diversification of angiosperms (`r s_figure("ltt")`).
-
-<<FIXME: add discussion of some specific groups once we have final ages>>
+Our estimated age of (`r crown_ages$polypodiales` ma) for crown Polypodiales is considerably older than other recent studies [@Testo2016a; @Du2021] and the fossil record, which only dates back to the Early Cretaceous [@Schneider2001; @Chen1997; @Deng2002; @Schneider2016c; @Regalado2018a].
+Molecular ages that are significantly older than the fossil record should be treated with caution; yet, our study is in agreement with others in suggesting a "long fuse" between initial appearance of Polypodiales their subsequent diversification and widespread preservation in the fossil record [@Testo2016a; @Du2021].
+
+Due to the large size of our dataset, carrying out more detailed molecular dating analyses (`r eg`, Bayesian analysis) is computationally difficult with currently available methods [`r eg`, BEAST\; @Drummond2007; @Bouckaert2014].
+Here, we have prioritized computational speed and simplicity, since we anticipate re-running the pipeline on a regular basis.
+We therefore consider our dated tree as a starting point, and not the final word, for a re-evaluation of divergence times in ferns.
+Future studies should focus on utilizing our greatly expanded fern fossil dataset to conduct more thorough molecular dating analyses, possibly including alternative schemes for the age of the root and testing the effects of parameters used for setting priors.
 
@@
   left_join(clade_cols_tib, by = "major_clade")
+
+### Prep for stem family age plots ###
+
+# Helper function for subsetting dates to get into
+# almost-long format
+subset_ages <- function(other_dates, author_sel) {
+  other_dates %>%
+    select(family, contains(glue("{author_sel}_"))) %>%
+    rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
+    mutate(author = author_sel)
+}
+
+# Define shapes and colors for points for stem family age plots
+ages_shapes <- c(
+  current = 16, # circle
+  rothfels = 17, # triangle
+  schuettpelz = 15, # square
+  ts = 18 # diamond
+)
+
+ages_colors <- c(
+  current = palette_okabe_ito(order = "vermillion"),
+  rothfels = palette_okabe_ito(order = "bluishgreen"),
+  schuettpelz = palette_okabe_ito(order = "blue"),
+  ts = palette_okabe_ito(order = "orange")
+)
+
+# Format names of papers for printing for stem family age plots
+ages_sampling <-
+  tribble(
+    ~source, ~year, ~label,
+    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
+    "rothfels", 2015, "Rothfels et al. 2015",
+    "ts", 2016, "Testo and Sundue 2016",
+    "current", 2021, "This study",
+  ) %>%
+  mutate(source = fct_reorder(source, year))
+
+# Provide names of periods in captions
+# Make sure to read and check that they are all there!
+period_names <-
+  periods %>%
+  # Filter to range including Ordovician to Neogene
+  filter(min_age < 475) %>%
+  filter(min_age > 1) %>%
+  arrange(desc(min_age)) %>%
+  mutate(label = glue("{abbr} ({name})")) %>%
+  pull(label) %>%
+  paste(collapse = ", ")
+
 ### plot-prep end
@@
 ```{r div-times-make}
-# Convert family ages to almost-long format
-
-# Helper function for subsetting dates to get into
-# almost-long format
-subset_ages <- function(other_dates, author_sel) {
-  other_dates %>%
-    select(family, contains(glue("{author_sel}_"))) %>%
-    rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
-    mutate(author = author_sel)
-}
-
-# Combine dates in almost-long format:
+# Combine family stem ages into almost-long format:
 # one row per family/study combination,
 # cols for median, high, and low date
-all_ages <-
+all_family_stem_ages <-
   bind_rows(
@@
 
-# Define shapes and colors for points
-ages_shapes <- c(
-  current = 16, # circle
-  rothfels = 17, # triangle
-  schuettpelz = 15, # square
-  ts = 18 # diamond
-)
-
-ages_colors <- c(
-  current = palette_okabe_ito(order = "vermillion"),
-  rothfels = palette_okabe_ito(order = "bluishgreen"),
-  schuettpelz = palette_okabe_ito(order = "blue"),
-  ts = palette_okabe_ito(order = "orange")
-)
-
-# Format names of papers for printing
-ages_sampling <-
-  tribble(
-    ~source, ~year, ~label,
-    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
-    "rothfels", 2015, "Rothfels et al. 2015",
-    "ts", 2016, "Testo and Sundue 2016",
-    "current", 2021, "This study",
-  ) %>%
-  mutate(source = fct_reorder(source, year))
-
 # Format y-axis text: show major clade in parentheses
 family_labs_tbl <-
-  all_ages %>%
+  all_family_stem_ages %>%
   select(family, code) %>%
@@
 
-# Provide names of periods in caption
-# Make sure to read and check that they are all there!
-period_names <-
-  periods %>%
-  # Filter to range including Ordovician to Neogene
-  filter(min_age < 475) %>%
-  filter(min_age > 1) %>%
-  arrange(desc(min_age)) %>%
-  mutate(label = glue("{abbr} ({name})")) %>%
-  pull(label) %>%
-  paste(collapse = ", ")
-
+# Make plot
 div_times_plot <-
-  ggplot(all_ages) +
+  ggplot(all_family_stem_ages) +
   geom_pointrange(

### 2022-03-25 ms/si.Rmd
Update MS
Add TS fossil date plots

@@
 
-```{r, formatters-si}
+```{r run-formatters-si, eval = params$doc_type == "manual"}
+source_chunk("formatters-si")
+```
+
+```{r formatters-si, eval = params$doc_type != "manual"}
 ```
@@
     spermato_trees,
-    accepted_species
+    ts_family_stem_ages,
+    accepted_species,
+    ftol_ts_comp_mod
   ),
@@
   paste0(collapse = ", ")
+```
+
+```{r plot-prep-load}
+# Load the plot prep code from the main MS
+knitr::read_chunk(
+  here::here("ms/manuscript.Rmd"),
+  labels = "plot-prep-main",
+  from = "plot-prep start",
+  to = "plot-prep end"
+)
+```
+
+```{r run-plot-prep, eval = params$doc_type == "manual"}
+source_chunk("plot-prep-main")
+```
+
+```{r plot-prep-main, eval = params$doc_type != "manual"}
 ```
@@
 \newpage
-
-```{r plot-prep-load}
-# Load the plot prep code from the main MS
-knitr::read_chunk(
-  here::here("ms/manuscript.Rmd"),
-  labels = "plot-prep-main",
-  from = "plot-prep start",
-  to = "plot-prep end"
-)
-```
-
-<!-- Run the plot prep code -->
-```{r plot-prep-main}
-```
 
@@
 
+```{r comp-fossils, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
+ts_fossil_calibration_tips %>%
+  select(minimum_age, node_calibrated) %>%
+  mutate(dataset = "ts") %>%
+  bind_rows(
+    con_fossil_calibration_tips %>%
+      select(minimum_age, node_calibrated) %>%
+      mutate(dataset = "current")
+  ) %>%
+  mutate(node_calibrated = fct_reorder(node_calibrated, minimum_age)) %>%
+  ggplot(aes(x = minimum_age, y = node_calibrated, fill = dataset)) +
+  geom_col(position = position_dodge()) +
+  scale_x_continuous("Age (millions of years)", expand = expansion(mult = c(0,0.1))) +
+  theme_bw() +
+  scale_fill_manual(
+    values = c(current = "grey30", ts = "grey80"),
+    labels = c(current = "This Study", ts = "Testo and Sundue 2016")
+  ) +
+  theme(
+    axis.title.y = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom"
+  )
+```
+
+**`r s_figure("comp-fossils")`**.
+Age of nodes constrained by fossils in this study vs. @Testo2016a. Wide bars indicate the node was only constrained in one study; narrow bars indicate the the node was constrained in both studies.
+
+\newpage
+
+```{r comp-stem-ages, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
+# Combine family stem ages into almost-long format:
+# one row per family/study combination,
+# cols for median, high, and low date
+all_family_stem_ages <-
+  bind_rows(
+    subset_ages(other_dates, "ts"),
+    transmute(
+      ts_family_stem_ages,
+      family,
+      author = "current",
+      median = age
+    )
+  ) %>%
+  left_join(
+    unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
+  left_join(clade_tibble, by = "major_clade") %>%
+  # (some median dates are NA if that family wasn't included in a study)
+  assert(not_na, family, major_clade, code, author) %>%
+  # Set family as a factor ordered by current age estimate
+  left_join(ts_family_stem_ages, by = "family") %>%
+  mutate(family = fct_reorder(family, age)) %>%
+  select(-age)
+
+# Format y-axis text: show major clade in parentheses
+family_labs_tbl <-
+  all_family_stem_ages %>%
+  select(family, code) %>%
+  unique() %>%
+  mutate(label = glue("{family} ({code})"))
+
+family_labs <- family_labs_tbl$label %>%
+  set_names(., family_labs_tbl$family)
+
+# Make plot
+ggplot(all_family_stem_ages) +
+  geom_pointrange(
+    aes(
+      x = median, xmin = low, xmax = high,
+      y = family,
+      color = author, shape = author),
+    position = position_dodge(width = 0.5)) +
+  scale_shape_manual(
+    values = ages_shapes,
+    limits = ages_sampling$source, 
+    labels = ages_sampling$label) +
+  scale_color_manual(
+    values = ages_colors,
+    limits = ages_sampling$source, 
+    labels = ages_sampling$label) +
+  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
+  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
+  coord_geo(
+    dat = list("periods", "eras"),
+    pos = list("b", "b"),
+    abbrv = list(TRUE, FALSE),
+    xlim = c(500, 0),
+    fill = "transparent",
+    color = "grey20",
+    lab_color = "grey30",
+    height = unit(1, "lines"),
+    expand = TRUE,
+    size = 3) +
+  guides(shape = guide_legend(nrow = 2)) +
+  theme_bw() +
+  theme(
+    axis.title.y = element_blank(),
+    legend.title = element_blank(),
+    legend.position = "bottom"
+  )
+```
+
+**`r s_figure("comp-stem-ages")`**.
+Comparison of stem age of fern families (Ma) estimated by @Testo2016a and those estimated using methodology (treePL) and phylogeny (FTOL) of the current study with the fossil constraints of @Testo2016a.
+For @Testo2016a, error bars indicate lower and upper 95% highest posterior density levels and point indicates median
+For this study, point indicates best (most likely) estimate.
+Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
+Period name abbreviations as in `r figure("divtimes")`.
+
+\newpage
+
+```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}
+# For checking that model data is same as plot data
+mod_fitted_data <-
+ftol_ts_comp_mod %>%
+  broom::augment() %>%
+  arrange(ftol)
+
+mod_fitted_data_ts <-
+ftol_ts_comp_mod %>%
+  broom::augment() %>%
+  arrange(ts)
+
+# Make dataset for plotting
+comp_plot_data <-
+all_family_stem_ages %>%
+  filter(author == "ts") %>%
+  select(family, ts = median, ts_high = high, ts_low = low) %>%
+  inner_join(
+    filter(all_family_stem_ages, author == "current") %>%
+      select(family, ftol = median),
+    by = "family"
+  ) %>%
+  # Double check that model data is same as plot data
+  arrange(ftol) %>%
+  verify(all(.$ftol == mod_fitted_data$ftol)) %>%
+  arrange(ts) %>%
+  verify(all(.$ts == mod_fitted_data_ts$ts))
+
+# Extract R2 and P-value for linear mod
+stats <- ftol_ts_comp_mod %>%
+  broom::glance() %>%
+  mutate(label = glue::glue(
+    "*R*<sup>2</sup> = {number(r.squared, accuracy = 0.01)}\n
+    {pval} = {sci_number(p.value)}"
+  )) %>%
+  pull(label)
+
+# Extract fitted line for linear mod
+fitted <-
+broom::tidy(ftol_ts_comp_mod) %>%
+  mutate(term = janitor::make_clean_names(term)) %>%
+  split(.$term)
+
+ggplot(comp_plot_data, aes(y = ts, x = ftol)) +
+  geom_errorbar(aes(ymax = ts_high, ymin = ts_low)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1, color = "grey60", lty = 2) +
+  geom_abline(
+    intercept = fitted$intercept$estimate,
+    slope = fitted$ts$estimate, color = "grey20"
+  ) +
+  theme_bw() +
+  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
+  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
+  labs(x = "This study", y = "Testo and Sundue (2016)") +
+  annotate(
+    "richtext", x = 0, y = Inf, 
+    label.color = NA,
+    label = stats, hjust = 0, vjust = 1)
+```
+
+**`r s_figure("comp-stem-points")`**.
+Comparison of stem age of fern families estimated by @Testo2016a and those estimated using methodology (treePL) and phylogeny (FTOL) of the current study with the fossil constraints of @Testo2016a.
+All units in millions of years (Ma).
+For @Testo2016a, error bars indicate lower and upper 95% highest posterior density levels and point indicates median.
+For this study, point indicates best (most likely) estimate.
+Dashed indicates 1:1 relationship.
+Solid line indicates linear model fit.
+
+\newpage
+
 ```{r ltt, fig.width = 7, fig.height = 6, eval = params$doc_type != "manual"}
@@
 
+\newpage
+
 ## References

### 2022-03-25 R/functions.R

@@ -946,12 +946,41 @@ make_nonmonophy_gen_tab <- function(
     select(family, subfamily, everything())
 }
 
+#' Fit a model comparing stem family ages estimated from Testo and Sundue (2016)
+#' fossils using FTOL vs. the original Testo and Sundue (2016) ages
+#'
+#' @param other_dates Tibble with fern family stem ages read in from
+#' Testo and Sundue (2016)
+#' @param ts_family_stem_ages Tibble with fern family stem ages estimated 
+#' from Testo and Sundue (2016) fossils using FTOL
+#'
+#' @return Model
+#' 
+fit_stem_comp_mod <- function(other_dates, ts_family_stem_ages) {
+  
+  # Filter to Testo and Sundue 2016 dates
+  ts_dates <- 
+    other_dates %>%
+    select(family, ts = ts_median)
+  
+  # Make dataset comparing dates
+  date_comp <-
+    select(ts_family_stem_ages, family, ftol = age) %>%
+    # Only include families in common between both datasets
+    inner_join(ts_dates, by = "family")
+  
+  # Construct linear model
+  lm(ftol ~ ts, data = date_comp)
+}
+
 # Formatting ----
 
 # Abbreviations
 ie <- "*i.e.*"
 eg <- "*e.g.*"
 ca <- "*ca.*"
+r2 <- "*R*^2^"
+pval <- "*P*"
 
 ### formatters start
 # Specify custom percentage format
@@ -959,6 +988,10 @@ percent <- function(...) {scales::percent(...)}
 
 # Specify custom number format
 number <- function(...) {scales::number(big.mark = ",", ...)}
+
+# Specify custom number format for scientific notation
+sci_number <- scales::label_scientific()
+
 ### formatters end
 
 #' Convert a data.frame of counts to percentages, with automatic formatting
@@ -1030,7 +1063,7 @@ print_fossil <- function(fossil_data, fossil_select) {
   c(header, body, md_rule())
 }
 
-# Figure output ----
+# Figures ----
 
 #' Generate a path to save a results file
 #' 
@@ -1336,4 +1369,34 @@ extract_iqtree_corr <- function(iqtree_log) {
     last() %>%
     str_match("Bootstrap correlation coefficient of split occurrence frequencies: ([^$]+)$") %>%
     magrittr::extract(,2)
+}
+
+#' Source a chunk of code read in with knitr::read_chunk()
+#' 
+#' This only works if knitr::read_chunk() has already been called to load
[…]

### 2022-03-25 _targets.R

@@ -132,6 +132,11 @@ tar_plan(
     fs::path(ftol_cache, "objects/con_monophy_by_clade"),
     readRDS(file = !!.x)
   ),
+  tar_file_read(
+    ts_sanger_tree_dated,
+    fs::path(ftol_cache, "objects/ts_sanger_tree_dated"),
+    readRDS(file = !!.x)
+  ),
   
   # GenBank ----
   # Analyze number of fern accessions and species in GenBank by type of DNA 
@@ -191,6 +196,10 @@ tar_plan(
   # - Stem age of fern families
   family_stem_ages = get_stem_family_age(
     sanger_sampling, sanger_tree_dated, ppgi_taxonomy),
+  ts_family_stem_ages = get_stem_family_age(
+    sanger_sampling, ts_sanger_tree_dated, ppgi_taxonomy),
+  # - Model comparing T&S 2016 to FTOL estimated with T&S 2016 fossils
+  ftol_ts_comp_mod = fit_stem_comp_mod(other_dates, ts_family_stem_ages),
   # - Read in Du 2021 dates
   tar_file_read(
     du_dates_all,

### 2022-03-29 ms/manuscript.Rmd
Bump version of ftol_cache

@@
     ref_files,
-    sanger_iqtree_log,
+    sanger_iqtree_log_2000,
     sanger_sampling,
@@
 
-The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of more than 20% relative to the previous largest fern phylogeny.
+The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of `r ca` 40% relative to the previous largest fern phylogeny.
 Using an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points, we find estimated ages for most families and deeper clades to be significantly older than earlier studies.
@@
 
-Here, we leverage specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL).
+Here, we leverage specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL; `r figure("workflow")`).
 We plan to run the pipeline on a regular basis and make the results freely available online through a web portal and R package.
@@
 This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
-We thus updated and edited the initial World Ferns database using the dwctaxon R package (FIXME add citation), which is designed to work with taxonomic data in the Darwin Core format.
-We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL.
+We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
+We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@joel_nitta_2022_6388787].
 
@@
 The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
-Node support was assessed with up to `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013; @Hoang2018].
+Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013; @Hoang2018], which were then used to construct an extended, majority-rule consensus tree (the "backbone phylogeny").
 
@@
 
-We used the backbone (plastome) phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
+We used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
 We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
@@
 
-We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
+We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset, with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
+IQ-TREE was initially run using 1,000 iterations (default), but at the end of this run the bootstrap correlation coefficient of split occurrence frequencies was 0.96, which is below the threshold for convergence (0.99).
+We then continued the search for another 1,000 iterations (2,000 total), but the correlation coefficient fluctuated between 0.96 and 0.98 without overall improvement and again failed to converge.
+This suggests that the search was stuck in local optima.
+As our alignment contains many species with very similar sequences, it is likely that the search algorithm is unable to optimize many highly similar topologies that only vary in positions of closely related species at the tips.
+Considering that additional iterations were unlikely to converge, we therefore conducted 10 independent runs of IQTREE with 1,000 iterations each, and selected the run with the best (highest) combined log-likelihood of the ML and consensus trees [@Zhou2018].
+
 We did not conduct a more exhaustive analysis testing various partitions of the data or other phylogenetic inference methods or models as our goal is to produce a single species-level phylogeny that is a reasonable hypothesis of fern evolution, not to interrogate the outcomes of multiple, more or less equally applicable methods.
@@
 
-We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
+Dating was conducted separately on the ML tree and the consensus tree.
+We rooted the tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
+
 We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
-The full database is available at https://fernphy.github.com/FernCal.
+The full database is available at https://fernphy.github.com/ferncal.
 In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
@@
 A value of `r smooth_best` was selected, based on the smallest chi-squared value, for the final analysis.
-To construct confidence intervals for the estimated divergence times, we constructed 100 bootstrap trees (trees with identical topology but different branchlengths) using IQ-TREE and dated each using treePL with the same settings used for the ML tree [@Maurin2020].
 
@@
 assertthat::assert_that(
-  unique(extract_iqtree_mod(sanger_iqtree_log)) == 
+  unique(extract_iqtree_mod(sanger_iqtree_log_2000)) == 
     unique(extract_iqtree_mod(plastome_iqtree_log)),
@@
 The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger phylogenies.
-The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` and `r extract_iqtree_iter(sanger_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)` and `r extract_iqtree_corr(sanger_iqtree_log)` for the plastome and Sanger phylogenies, respectively.
+Two of the 10 runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, which was selected as the final tree, did not converge (correlation coefficient 0.977 after 1,000 iterations).
+The consensus tree had higher log-likelihood (-1183776.726) than the ML tree (-1183911.595).
 
@@
 
+The consensus tree had higher log-likelihood, so we only present and divergence times estimated from the consensus tree.
 We estimate the crown age of ferns to be `r crown_ages$ferns` ma.
@@
 It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
-To test this hypothesis, we conducted additional analysis using the fossil constraints of @Testo2016a but otherwise the exact same methods (FTOL analyzed with treePL).
-The resulting stem family ages show a much closer agreement with @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is responsible for the older ages observed in the current study.
+To test this hypothesis, we conducted an additional analysis using the fossil constraints of @Testo2016a but otherwise the same methods (FTOL analyzed with treePL).
+The resulting stem family ages show a much closer agreement with those of @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is likely responsible for the older ages observed in the current study.
 
@@
 
+```{r workflow-save, eval = params$doc_type == "doc" }
+# Copy workflow image if rendering to doc
+fs::file_copy(
+  here::here("images/workflow.pdf"),
+  result_file(figure("workflow"), "pdf"),
+  overwrite = TRUE)
+```
+
+```{r workflow-print, fig.width = 7, fig.height = 5, eval = params$doc_type == "pdf", results = "show"}
+# Print plot if rendering to PDF
+knitr::include_graphics(here::here("images/workflow.pdf"))
+```
+
+**`r figure("workflow")`**.
+Summary of workflow to construct the Fern Tree of Life (FTOL).
+The workflow is automated except for steps in boxes with red outlines and a hand symbol.
+Sequences originating from whole plastomes are in blue; sequences typically obtained by Sanger sequencing are in orange.
+For details of each step, see Materials and Methods.
+
+`r pagebreak_pdf(params)`
+
 ```{r ftol-make}
@@
   ftol_geom_cladelab(
-    clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li", "De")], 17) +
+    clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li")], 17) +
+  ftol_geom_cladelab("De", 36) +
   ftol_geom_cladelab("Pt", 26) +
@@
     "Polypodiales" = c("Polypodiaceae", "Saccolomataceae"),
-    "Polypodiidae\n(leptosporangiates)" = c("Polypodiaceae", "Osmundaceae")
+    "Polypodiidae\n(leptosporan-\ngiates)" = c("Polypodiaceae", "Osmundaceae")
   ) %>%
@@
     data = other_clade_labels %>%
-      filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
+      filter(clade_label == "Polypodiidae\n(leptosporan-\ngiates)"),
     aes(x = x, y = y, label = clade_label),
@@
     hjust = 0,
-    xlim = c(NA, 64),
+    xlim = c(NA, 52),
     ylim = c(32, NA),

### 2022-03-29 ms/si.Rmd
Bump version of ftol_cache

@@
 # FIXME: change to ftol_cache when ready
-ftol_cache <- here::here("working/_targets")
+ftol_cache <- here::here("ftol_cache")
 
@@
     con_fossil_calibrations_for_treepl,
-    fossil_ferns_raw,
     fossil_ferns_all,
@@
 ```{r fossils-make}
+# Write out list of cited refs as "context" file
+temp_context <- fs::path_abs(tempfile(fileext = ".txt"))
+
+fossil_ferns_all %>%
   # Filter raw fossil data to just fossils used
-fossils_table_raw <-
-  fossil_ferns_all %>%
   inner_join(
@@
   verify(nrow(.) == nrow(con_fossil_calibration_tips)) %>%
+  select(where(is.character)) %>%
+  pivot_longer(everything()) %>%
+  filter(!is.na(value)) %>%
+  pull(value) %>%
+  write_lines(temp_context)
+
+# Format references with pandoc
+fossils_table <-
+  fossil_ferns_all %>%
+  # Filter raw fossil data to just fossils used
+  inner_join(
+    unique(select(con_fossil_calibration_tips, n_fos)), by = "n_fos") %>%
+  verify(nrow(.) == nrow(con_fossil_calibration_tips)) %>%
+  # Select columns
   select(
@@
     organs, specimen_holotype, collection,
-    reference_raw = reference,
-    notes_on_affinities
-  )
-
-# Write out list of cited refs as "context" file
-temp_context <- fs::path_abs(tempfile(fileext = ".txt"))
-
-fossils_table_raw %>%
-  select(reference_raw, notes_on_affinities) %>%
-  pivot_longer(everything()) %>%
-  filter(!is.na(value)) %>%
-  pull(value) %>%
-  write_lines(temp_context)
-
-# Format references with pandoc
-fossils_table <-
-fossils_table_raw %>%
+    reference_raw = reference
+  ) %>%
+  # Format references
   mutate(reference = cite_pd(
     reference_raw, bib = bib, csl = csl, context = temp_context)) %>%
-  select(-reference_raw, -notes_on_affinities) %>%
+  select(-reference_raw) %>%
   # Make sure all references in the table have entries in the bibliography
@@
 
-**`r s_figure("bs")`**.
+**`r s_figure("backbone_full")`**.
 Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
@@
 
-\newpage
-
 ```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}

### 2022-03-29 ms/si2.Rmd
Bump version of ftol_cache

@@
 # Specify path to FTOL cache
-# FIXME: change to ftol_cache when ready
-ftol_cache <- here::here("working/_targets")
+ftol_cache <- here::here("ftol_cache")
 
@@
     con_fossil_calibration_tips,
-    fossil_ferns_raw,
-    fossil_ferns_all),
+    fossil_ferns_raw),
   store = ftol_cache,
@@
 tar_load(
+  c(
     fossil_notes,
+    fossil_meta),
   store = here::here("_targets")
@@
 ```{r tidy-fossil-data}
-# fossil_ferns_raw is read in from Fossils_Ferns.csv directly
-# fossil_ferns_all is same, but with clean names by janitor::clean_names()
-
-# Make tibble of fossil data categories
-raw_col_names <- colnames(fossil_ferns_raw)
-fossils_categories <-
-  tibble(raw_name = raw_col_names) %>%
+# Add category to metadata
+fossil_meta <-
+  fossil_meta %>%
   mutate(
     category = case_when(
-      raw_name %in% raw_col_names[1:4] ~ "Calibration",
-      raw_name %in% raw_col_names[5:12] ~ "Fossil Identity",
-      raw_name %in% raw_col_names[13:18] ~ "Fossil Age",
-      raw_name %in% raw_col_names[19:21] ~ "Fossil Locality",
-      raw_name %in% raw_col_names[22:24] ~ "Fossil Specimen",
+      variable %in% colnames(fossil_ferns_raw)[1:4] ~ "Calibration",
+      variable %in% colnames(fossil_ferns_raw)[5:12] ~ "Fossil Identity",
+      variable %in% colnames(fossil_ferns_raw)[13:18] ~ "Fossil Age",
+      variable %in% colnames(fossil_ferns_raw)[19:21] ~ "Fossil Locality",
+      variable %in% colnames(fossil_ferns_raw)[22:24] ~ "Fossil Specimen",
       TRUE ~ NA_character_
-    )
+    ),
+    var_order = 1:nrow(.)
   )
@@
 # Tidy fossils in wide format
-fossils <-
-  fossil_ferns_all %>%
+fossils_long <-
+  fossil_ferns_raw %>%
   # Filter to only those actually used in the study
@@
   verify(all(sort(.$n_fos) == sort(n_fos_used$n_fos))) %>%
-  # Replace with formatted data
-  select(
-    -node_calibrated, -full_taxon_name, 
-    -affinities, -notes_on_affinities, -notes_on_age) %>%
-  left_join(fossil_notes, by = "n_fos") %>%
-  # Put column names back in proper order
-  select(all_of(colnames(fossil_ferns_all)))
-```
-
-```{r write-new-formatted, eval = FALSE}
-# If needed, write out notes for further editing (if need to add new rows)
-fossil_ferns_all %>%
-  inner_join(
-    n_fos_used, by = "n_fos"
-  ) %>%
-  anti_join(fossil_notes, by = "n_fos") %>%
-  select(n_fos,
-         node_calibrated, full_taxon_name, 
-         affinities, notes_on_affinities, notes_on_age) %>%
-  bind_rows(fossil_notes) %>%
-  assert(is_uniq, n_fos) %>%
-  arrange(n_fos) %>%
-  write_csv(here::here("_targets/user/data_raw/fossils_used_formatted.csv"))
-```
-
-```{r format-for-printing}
-# Make a tibble of column names for formatting
-name_tbl <- tibble(
-  raw_name = colnames(fossil_ferns_raw),
-  var = colnames(fossil_ferns_all)
-) %>%
-  right_join(
-    tibble(var = colnames(fossils)), by = "var"
-  ) %>%
-  mutate(var_order = 1:nrow(.))
-
-# Convert to long format
-fossils_long <- 
-  fossils %>%
   mutate(
     across(everything(), as.character),
+    # Add fossil taxon header
     fossil_taxon_header = fossil_taxon) %>%
-  pivot_longer(names_to = "var", values_to = "value", -fossil_taxon) %>%
-  left_join(name_tbl, by = "var") %>%
+  pivot_longer(names_to = "variable", values_to = "value", -fossil_taxon) %>%
+  left_join(
+    select(fossil_meta, -description), by = "variable") %>%
   mutate(
-    text = as.character(glue::glue("{raw_name}: {value}"))
+    text = as.character(glue::glue("{formatted_name}: {value}"))
   ) %>%
-  left_join(fossils_categories, by = "raw_name") %>%
   arrange(fossil_taxon, category, var_order) %>%

### 2022-03-29 R/functions.R

@@ -1046,7 +1046,7 @@ print_fossil <- function(fossil_data, fossil_select) {
   # Print name of fossil as header in italics
   header <-
     data %>% 
-    filter(var == "fossil_taxon_header") %>%
+    filter(variable == "fossil_taxon_header") %>%
     pull(value) %>%
     str_replace_all("_", " ") %>%
     md_italic() %>%

### 2022-03-29 _targets.R

@@ -7,8 +7,7 @@ library(glue)
 source("R/functions.R")
 
 # Specify path to FTOL cache
-# FIXME: change to ftol_cache when ready
-ftol_cache <- "working/_targets"
+ftol_cache <- here::here("ftol_cache")
 
 # Define Rmd targets outside of main workflow
 tmp <- capture.output({
@@ -74,10 +73,10 @@ tar_plan(
     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv",
     readr::read_csv(file = !!.x)
   ),
-  # - Formatted fossil notes
+  # - Fossil ferns metadata
   tar_file_read(
-    fossil_notes,
-    "_targets/user/data_raw/fossils_used_formatted.csv",
+    fossil_meta,
+    "_targets/user/data_raw/fern_fossils_metadata.csv",
     readr::read_csv(file = !!.x)
   ),
   # - Seed plant trees from Barahona et al 2020
@@ -91,6 +90,12 @@ tar_plan(
     plastome_iqtree_log,
     fs::path(ftol_cache, "user/intermediates/iqtree/plastome/plastome_alignment.phy.log"), #nolint
     readr::read_lines(file = !!.x)
+  ),
+  # original log file from first, single Sanger run
+  tar_file_read(
+    sanger_iqtree_log_2000, # 2000 iterations were run total
+    fs::path(ftol_cache, "user/intermediates/iqtree/sanger/sanger_alignment.phy.2022-02-23.log"), #nolint
+    readr::read_lines(file = !!.x),
   ),
   tar_file_read(
     sanger_iqtree_log,

### 2022-04-01 ms/manuscript.Rmd
Update MS

@@
     number_sections: no
-    reference_docx: Frontiers_Template_SI.docx
+    reference_docx: Frontiers_Template.docx
 header-includes: 
@@
     fern_sanger_seqs_raw,
+    fossil_ferns_raw,
     plastid_calibration_dates,
@@
     fern_nonmono_gen,
-    fossil_ferns_raw,
     ftol_ts_comp_mod,
@@
 
-Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after seed plants.
-Ferns have received intense focus from molecular systematists, resulting in the publication of molecular phylogenies for every major clade and DNA sequences for more than one third of all fern species.
+Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after angiosperms.
+They have been the subject of numerous molecular phylogenetic studies, resulting in the publication of trees for every major clade and DNA sequences from nearly half of all species.
 Global fern phylogenies have been published periodically, but as molecular systematics research continues at a rapid pace, these become quickly outdated.
 
-Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data available via GenBank.
+Here, we develop a mostly automated, reproducible, open pipeline to generate a continuously updated fern tree of life (FTOL) from DNA sequence data available in GenBank.
 Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with commonly sequenced plastid regions (many taxa, few loci) to obtain a global, species-level fern phylogeny with high resolution along the backbone and maximal sampling across the tips.
-We use an expert-curated reference taxonomy to resolve synonyms in general compliance with Pteridophyte Phylogeny Group I (PPG I), the community-driven taxonomic system widely accepted by many pteridologists.
-
-The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of `r ca` 40% relative to the previous largest fern phylogeny.
-Using an updated and expanded list of `r n_calibrations_ferns` fern fossil calibration points, we find estimated ages for most families and deeper clades to be significantly older than earlier studies.
-
-FTOL and its accompanying datasets including the fossil list and taxonomic database will be updated on a regular basis and made available via a web portal and R packages, enabling immediate access to the most up-to-date, comprehensively-sampled fern phylogeny.
+We use an curated reference taxonomy to resolve synonyms in general compliance with the community-driven Pteridophyte Phylogeny Group I classification.
+
+The current FTOL includes `r comma(n_sanger_ferns)` species, an increase of `r ca` 40% relative to the most recently published global fern phylogeny.
+Using an updated and expanded list of `r n_calibrations_ferns` fern fossil constraints, we find estimated ages for most families and deeper clades to be considerably older than earlier studies.
+
+FTOL and its accompanying datasets, including the fossil list and taxonomic database, will be updated on a regular basis and are available via a web portal (https://fernphy.github.io) and R packages, enabling immediate access to the most up-to-date, comprehensively sampled fern phylogeny.
 FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
@@
 Ferns (ca.
-`r number(12000)` species) are the second most diverse clade of vascular plants after seed plants (ca.
+`r number(12000)` species) are the second most diverse clade of vascular plants after angiosperms (ca.
 `r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
-Key to any investigation of evolutionary biology in this group is a well-sampled phylogeny.
-Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data for more than one third of all currently recognized fern species.
-Thus, there is both sufficient sampling and a pressing need for a globally sampled fern phylogeny.
-
-Past efforts to construct such a global phylogeny have steadily expanded their sampling as sequences have been added to GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global phylogeny of ferns [@Testo2016a\; `r figure("genbank")`].
-There is a need therefore, not only for a global fern phylogeny, but one that is continuously updated to keep pace with the rapid accumulation of molecular data.
-This would eliminate the need for researchers to "rebuild the wheel" each time they want to use a globally sampled fern phylogeny.
-
-Multiple frameworks have been put forth to automatically or semi-automatically generate trees for any arbitrary part of the tree of life [@Antonelli2016], all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015] which would subsume a global fern phylogeny.
-While these may be appropriate in some cases, they are not expected to produce an optimal fern phylogeny due to the need to take a "one-size-fits-all" approach to accommodate such a wide phylogenetic breadth.
+Key to any investigation of evolutionary history in this group is a well-sampled phylogeny.
+Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data from nearly half of all currently recognized fern species.
+Thus, there is both a pressing need and sufficient sampling for a globally sampled fern phylogeny.
+
+Past efforts to construct such a global phylogeny have steadily expanded their sampling, at first by mostly generating new sequences, then later by mining GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
+Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
+There is a need therefore, not only for a revised global fern phylogeny, but also one that is continuously updated to keep pace with the rapid accumulation of molecular data going forward.
+Such an effort would eliminate the need for researchers to "rebuild the wheel" each time the need for a globally sampled fern phylogeny arises.
+
+Multiple frameworks have been put forth to automatically or semi-automatically generate trees for any particular part of the tree of life [@Antonelli2016], all plants [@Eiserhardt2018], or even the entire tree of life at once [@Hinchliff2015], which would of course subsume a global fern phylogeny.
+While such approaches are well-suited to some studies, they cannot be expected to produce an optimal fern phylogeny due to the use of "one-size-fits-all" methods to accommodate such a wide phylogenetic breadth.
 By focusing methods and datasets specifically on ferns, it should be possible to generate a higher quality end-product (tree) that can then be used "as-is" by biologists studying these organisms.
@@
 PPG I has been widely accepted and used, but there were problematic (non-monophyletic) genera included at the time of publication, and many taxonomic changes have been (and will continue to be) proposed since [`r eg`, @Almeida2017; @Shang2018; @Zhang2020].
-The next iteration of PPG (`r ie`, PPG II) will ideally be an online, open resource that can be updated as necessary.
+The next iteration of the PPG classification (`r ie`, PPG II) will ideally be an online, open resource that can be updated as necessary.
 We envision an open, continuously updated global fern phylogeny that could be directly integrated with PPG II such that taxonomic decisions can be made based on community consensus and the most recently available data.
 
-Here, we leverage specialized taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL; `r figure("workflow")`).
-We plan to run the pipeline on a regular basis and make the results freely available online through a web portal and R package.
-This will enable anybody interested in the biology of ferns to have access to the most recent hypothesis of fern phylogeny.
-We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern phylogeny; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
+Here, we leverage taxonomic knowledge to design a custom, fully reproducible, mostly automated pipeline to generate a maximally sampled global fern tree of life (FTOL; `r figure("workflow")`).
+We plan to run the pipeline on a regular basis and make the results freely available online through a web portal (https://fernphy.github.io) and R package [@ftol2022c].
+This will enable anybody interested in the biology of ferns to have access to the most current hypothesis of fern phylogeny, as well as an associated time-tree dated using a curated list of fern fossils.
+We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern relationships; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
 
@@
 The plastid genome has been the most widely sequenced genomic compartment in ferns by far (`r figure("genbank")`), so we used plastid loci to build our tree.
-Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
+Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used in molecular analyses of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
 Here, the *trnL*--*trnF* locus includes the *trnL* intron, the 3' *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
@@
 All sequences were downloaded from GenBank.
-GenBank queries were formatted to include locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
+GenBank queries were formatted to include the locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify a date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query results in the same set of sequences regardless of when the query is made (with the possible exception of accessions that were embargoed at the time of querying).
 Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
 There is no formal definition of genomic vs.
-Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.
+Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term.
 
@@
 
-This project aimed to generate a phylogeny that is consistent with PPG I, while accounting for taxonomic changes that have been made since the publication of PPG I.
+This project aimed to generate a phylogeny that is consistent with PPG I, while accounting for taxonomic changes that have been made since its publication.
 Species names in GenBank, which use the NCBI taxonomy [@Federhen2012; @Schoch2020], do not necessarily conform to PPG I.
@@
 
-We manually inspected any fuzzy matched or non-matching names.
+We manually inspected any fuzzily matched or non-matching names.
 This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
 We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
-We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@joel_nitta_2022_6388787].
+We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@ftol2022a].
 
@@
 
-Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminations (hereafter collectively referred to as "rogues").
+Another potential issue with GenBank sequences is the presence of misidentified sequences, poor quality sequences, and contaminants (hereafter collectively referred to as "rogues").
 We removed putative rogues from Sanger sequences as follows.
@@
 Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
-We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a contaminant and excluded from further analysis.
-For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family.
+We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, , that query was considered a rogue and excluded from further analysis.
+For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family to avoid false positives.
 Species belonging to monotypic families were not considered for this filtering.
@@
 A typical step in multilocus phylogenetic workflows is to concatenate loci across samples.
-For phylogenetic studies aiming to include one tip per species, this is often done by concatenating the longest sequence per locus within each species.
+For phylogenetic studies aiming to include one tip per species, this is often done by concatenating the longest sequence per locus within each species, regardless of source.
 However, such an approach is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
-Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
+Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may contain multiple loci, but we already split those out using superCRUNCH as described above).
 First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
@@
 We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
-Our procedure for selecting accessions for each species is as follows.
-
-1. If accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length.
-2. Otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence.
-3. Otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
-
-The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
-For these, we selected one specimen per species with the longest combined sequence length across all plastome loci.
+Our procedure for selecting accessions for each species is as follows: 1) if accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length; 2) otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence; 3) otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
+These steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
+For these, we selected the one specimen per species with the longest combined sequence length across all plastome loci.
 
@@
 This retains the indels within each subalignment while aligning across subalignments.
-For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "\-\-merge".
+Sequences from families with fewer than three species each could not be used for subalignments, so these were added as "singletons" during the MAFFT "\-\-merge".
 Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
@@
 We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
-We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect [`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus\; expected monophyly follows @PteridophytePhylogenyGroupI2016].
+We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect (`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus; expected monophyly follows PPG I, 2016).
 However, this was not possible in all cases, particularly in groups that are in need of taxonomic revision and are known to include non-monophyletic genera (`r eg`, cheilanthoid ferns, grammitid ferns, microsoroid ferns).
-While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that would require valid publication according to the International Code of Nomenclature for Algae, Fungi, and Plants [@Turland2018].
+While we consider our phylogeny may serve as a guide for future taxonomic revisions, we did not make any taxonomic changes that were not already validly published according to the International Code of Nomenclature for Algae, Fungi, and Plants [@Turland2018].
+Some unpublished names appearing in GenBank and World Ferns databases were not excluded.
 
@@
 During each iteration of the analysis, it is possible that the GenBank accessions added in place of the excluded rogues themselves include rogues.
-Therefore, we repeated this process until the monophyly of all higher-level (`r eg`, genus rank and higher) taxa that were expected to be monophyletic according to @PteridophytePhylogenyGroupI2016 was either confirmed or could not be achieved due to outstanding taxonomic issues.
+Therefore, we repeated this process until the monophyly of all higher-level (`r eg`, genus rank and higher) taxa that were expected to be monophyletic according to PPG I (2016) was either confirmed or could not be achieved due to outstanding taxonomic issues.
 
@@
 
-We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset, with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
+We generated the final Sanger phylogeny using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset, with the same model selection procedure and bootstrapping as used for the backbone phylogeny.
 IQ-TREE was initially run using 1,000 iterations (default), but at the end of this run the bootstrap correlation coefficient of split occurrence frequencies was 0.96, which is below the threshold for convergence (0.99).
-We then continued the search for another 1,000 iterations (2,000 total), but the correlation coefficient fluctuated between 0.96 and 0.98 without overall improvement and again failed to converge.
-This suggests that the search was stuck in local optima.
+We then continued the search for another 1,000 iterations (2,000 total), but the correlation coefficient fluctuated between 0.96 and 0.98 without overall improvement or convergence.
+This suggests that the search was stuck in a local optimum.
 As our alignment contains many species with very similar sequences, it is likely that the search algorithm is unable to optimize many highly similar topologies that only vary in positions of closely related species at the tips.
@@
 Dating was conducted separately on the ML tree and the consensus tree.
-We rooted the tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
-
-We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa (`r s_table("fossils")`, Appendix S2).
-The full database is available at https://fernphy.github.com/ferncal.
+We rooted the tree with bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
+
+We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa called "ferncal" [@ftol2022b\; `r s_table("fossils")`, Appendix S2].
+The full ferncal database is available at https://fernphy.github.com/ferncal.
 In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
 We assigned fossils to lineages only after consulting the original publication (as opposed to simply relying on the fossil name) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
-Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained for a given fossil could change.
-We also applied two constraints outside of ferns, including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
+Taxonomic concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be constrained with a given fossil could change.
+We applied one fossil constraint outside of ferns (stem euphyllophytes; `r fossils$Euphyllophytes$age` Ma), and fixed the root age of the tree (land plants) at 475 Ma as in @Testo2016a and @Qi2018.
 All constraints other than the root are minimum ages (`r s_figure("fossil_map")`).
@@
 
-The workflow is managed in R [@RCoreTeam2020a] with the targets package [@Landau2021].
-Code used to generate FTOL and compile this manuscript is available at https://github.com/fernphy/ftol and https://github.com/fernphy/ftol_ms, respectively.
-A docker image to run the code is available at https://hub.docker.com/r/joelnitta/ftol.
+The workflow is managed in R v4.1.1 [@RCoreTeam2021] with the "targets" package [@Landau2021].
+Input data are available at https://doi.org/10.6084/m9.figshare.19474316.v1 [@Nitta2022figshare].
+Code used to generate FTOL and compile this manuscript are available at https://github.com/fernphy/ftol and https://github.com/fernphy/ftol_ms, respectively.
+Docker images to run the analysis and compile this manuscript are available at https://hub.docker.com/r/joelnitta/ftol and https://hub.docker.com/r/joelnitta/ftol_ms, respectively.
 
@@
 
-The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions (note that in some cases a single accession may contain multiple loci).
+The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions (note that in some cases a single GenBank accession may contain multiple loci).
 Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
-Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
-After excluding these and species names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
+Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` rogues to be excluded.
+After excluding these and accessions with names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
 After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
@@
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
-Taxon sampling for the backbone phylogeny (plastome loci) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.
+Taxon sampling for the backbone phylogeny (derived from plastome loci) includes `r bb_total_sampling$species$n` species (`r bb_total_sampling$species$coverage_p`), `r bb_total_sampling$genus$coverage` genera (`r bb_total_sampling$genus$coverage_p`), `r bb_total_sampling$family$coverage` families (`r bb_total_sampling$family$coverage_p`), and `r bb_total_sampling$order$coverage` orders of ferns.
 
@@
 A majority of species (`r sp_mult_loci$true$n`; `r sp_mult_loci$true$percent`) were sampled for more than one locus.
-Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors are standard deviations unless otherwise mentioned).
+A mean of `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors are standard deviations unless otherwise mentioned).
 The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
@@
 
-The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger phylogenies.
-Two of the 10 runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, which was selected as the final tree, did not converge (correlation coefficient 0.977 after 1,000 iterations).
-The consensus tree had higher log-likelihood (-1183776.726) than the ML tree (-1183911.595).
-
-Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
+The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger analyses.
+Two of the ten runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, from which the final tree was selected, did not converge (correlation coefficient 0.977 after 1,000 iterations).
+The consensus tree had higher log-likelihood (-1183776.726) than the ML tree (-1183911.595), so we only present the topology and divergence times estimated from the consensus tree.
+
+Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS ≥ `r percent(.98)` in the backbone phylogeny unless otherwise indicated).
 The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (Figures `r figure_num("backbone")`, `r s_figure_num("backbone_full")`).
@@
 
-All sampled orders, suborders, families, and subfamilies were recovered as monophyletic (or monotypic) with the exception of Polypodioideae, which is known to be paraphyletic relative to Grammitidoideae [@PteridophytePhylogenyGroupI2016].
-`r fern_monophy_summ$genus$No$n` genera (`r fern_monophy_summ$genus$No$p`) were non-monophyletic (`r s_table("monophy")`).
+All sampled orders, suborders, families, and subfamilies were recovered as monophyletic (or monotypic) with the exception of Polypodioideae, which is known to be paraphyletic relative to Grammitidoideae (PPG I, 2016).
+`r fern_monophy_summ$genus$No$n %>% english::english() %>% str_to_sentence()` genera (`r fern_monophy_summ$genus$No$p`) were non-monophyletic (`r s_table("monophy")`).
 The subfamilies with the most non-monophyletic genera were `r top_nonmono_subfam$subfamily[[1]]` (`r top_nonmono_subfam$n[[1]]`), `r top_nonmono_subfam$subfamily[[2]]` (`r top_nonmono_subfam$n[[2]]`), and `r top_nonmono_subfam$subfamily[[3]]` (`r top_nonmono_subfam$n[[3]]`); other (sub)families each had four or fewer non-monophyletic genera (`r s_table("nonmono_gen")`).
 
-Bootstrap support was generally across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`; `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
+Bootstrap support was generally moderate to high across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`; `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
 Relationships within some genera were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).
@@
 
-The consensus tree had higher log-likelihood, so we only present and divergence times estimated from the consensus tree.
-We estimate the crown age of ferns to be `r crown_ages$ferns` ma.
-Ages of other major crown groups are leptosporangiates (`r crown_ages$leptos` ma), Polypodiales (`r crown_ages$polypodiales` ma), eupolypods I (`r crown_ages$eupoly_i` ma), and eupolypods II (`r crown_ages$eupoly_ii` ma).
-Our estimates for leptosporangiates and Polypodiales are both `r ca` 50 ma older than the most recent global fern phylogeny [@Testo2016a], while other ages are similar.
+We estimate the crown age of ferns to be `r crown_ages$ferns` million years (Ma) old.
+Ages of other major crown groups are: leptosporangiates (`r crown_ages$leptos` Ma), Polypodiales (`r crown_ages$polypodiales` Ma), eupolypods I (`r crown_ages$eupoly_i` Ma), and eupolypods II (`r crown_ages$eupoly_ii` Ma).
+Our estimates for leptosporangiates and Polypodiales are both `r ca` 50 Ma older than the most recent global fern phylogeny [@Testo2016a], while other ages are similar.
 Estimated stem ages of fern families were mostly older than previous studies (`r figure("divtimes")`).
@@
 
-The phylogenetic position of Equisetidae relative to other ferns has long been a point of contention in molecular systematics of ferns.
+The phylogenetic position of Equisetidae relative to other ferns has long been contentious.
 Here, we recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with many other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
@@
 Aside from alternative resolutions of Equisetidae dependent on genomic compartment or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
-The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is lacking.
+The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is so far lacking.
 
@@
 Here, we recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
-However, current plastome sampling only includes `r plastome_gleich_genera_n` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
+However, current plastome sampling only includes `r plastome_gleich_genera_n %>% jntools::english2()` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
 
@@
 Most previous plastid studies based on Sanger sequencing have recovered (P, (D, e)) [@Kuo2011; @Schuettpelz2006; @Schuettpelz2007; @Testo2016a; but see @Lehtonen2011].
-We recover (D, (P, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] or nuclear data [@Rothfels2015a; @Qi2018; @Shen2018] as well as a plastid supermatrix [@Lehtonen2011].
+We recover (D, (P, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] and nuclear data [@Rothfels2015a; @Qi2018; @Shen2018], as well as a plastid supermatrix [@Lehtonen2011].
 Recently the whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others.
@@
 
-Relationships of families within Aspleniineae (eupolypods II) have been difficult resolve due to the ancient, rapid radiation of this clade [@Rothfels2012].
+Relationships of families within Aspleniineae (eupolypods II) have been difficult to resolve due to the ancient, rapid radiation of this clade [@Rothfels2012].
 Our analysis robustly resolves the relationships between all families in Aspleniineae and is in agreement with a recent plastome analysis with similar sampling [@Du2021].
@@
 
-## Revisiting fossil calibration points in ferns
+## Revisiting the timeline of fern diversification
 
@@
 
-Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
-However, we found inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
-Moreover, our set of calibration points more than doubles the number used by previous studies.
-It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using a similar, more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+Several recent studies exploring divergence times across a global fern phylogeny all used a similar set of 24--26 fossil calibration points [@Schuettpelz2009; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
+However, there were a few inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
+More importantly, our set of calibration points more than doubles the number used by previous studies and it is likely that this expanded dataset is responsible for the older stem age estimates for many families (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 To test this hypothesis, we conducted an additional analysis using the fossil constraints of @Testo2016a but otherwise the same methods (FTOL analyzed with treePL).
-The resulting stem family ages show a much closer agreement with those of @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is likely responsible for the older ages observed in the current study.
-
-Notably, the scenario of fern diversification suggested by our results conflicts with the hypothesis that Polypodiales diversified "in the shadow of angiosperms" [@Schneider2004d].
-Rather, we estimate that polypod diversification occurred coincidentally or even preceding the diversification of angiosperms (`r s_figure("ltt")`).
-Our estimated age of (`r crown_ages$polypodiales` ma) for crown Polypodiales is considerably older than other recent studies [@Testo2016a; @Du2021] and the fossil record, which only dates back to the Early Cretaceous [@Schneider2001; @Chen1997; @Deng2002; @Schneider2016c; @Regalado2018a].
-Molecular ages that are significantly older than the fossil record should be treated with caution; yet, our study is in agreement with others in suggesting a "long fuse" between initial appearance of Polypodiales their subsequent diversification and widespread preservation in the fossil record [@Testo2016a; @Du2021].
+The resulting stem family ages show a much closer agreement with those of @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is the primarily contributor to the older ages observed in the current study.
+
+Notably, the scenario of fern diversification suggested by our results somewhat conflicts with the hypothesis that Polypodiales diversified "in the shadow of angiosperms" [@Schneider2004d; @Schuettpelz2009].
+Rather, we estimate that much of polypod diversification coincides with or even precedes the diversification and rise to ecological dominance of angiosperms during the Late Cretaceous [@Benton2022] (`r s_figure("ltt")`).
+Our estimated age of (`r crown_ages$polypodiales` Ma) for crown Polypodiales is considerably older than other recent studies [@Testo2016a; @Du2021] and the fossil record, which only dates back to the Early Cretaceous [@Schneider2001; @Chen1997; @Deng2002; @Schneider2016c; @Regalado2018a].
+Molecular ages that are significantly older than the fossil record should be treated with caution; yet, our study is in agreement with others in suggesting a "long fuse" between initial appearance of Polypodiales and their subsequent diversification and widespread preservation in the fossil record [@Testo2016a; @Du2021].
 
@@
 We therefore consider our dated tree as a starting point, and not the final word, for a re-evaluation of divergence times in ferns.
-Future studies should focus on utilizing our greatly expanded fern fossil dataset to conduct more thorough molecular dating analyses, possibly including alternative schemes for the age of the root and testing the effects of parameters used for setting priors.
+Future studies should focus on utilizing our greatly expanded fern fossil dataset to conduct more thorough molecular dating analyses, possibly including alternative schemes for the age of the root and testing the effects of different parameters used for setting priors in Bayesian analyses.
 
@@
 Plastid sequences are convenient for phylogenetic analysis because they are essentially a single, uniparentally inherited linkage group, thus free from recombination.
-However, a tree derived from plastid data may not reflect those inferred from other data sources.
+However, a tree derived from plastid data may not necessarily mirror those inferred from other data sources.
 Conflict between plastid and nuclear phylogenies has been frequently observed in narrowly focused (`r eg`, genus level) studies using traditional Sanger sequencing [`r eg`, @Sessa2012b; @Zhou2017a; @Wei2021a] and has recently been demonstrated at deeper levels within Polypodiaceae using phylogenomic approaches [@Wei2022].
-Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
+Such conflict does not necessarily reflect insufficient methodology or sampling, but rather may be due to processes including (but not limited to) introgression, lineage sorting, and hybridization at deep phylogenetic levels.
 Therefore, a major future research goal for fern molecular systematics should be to combine nuclear and plastid datasets to infer species trees with comprehensive sampling.
@@
 300 single-to-low copy nuclear gene regions) for all flagellate plants [bryophytes, lycophytes, ferns, and gymnosperms\; @Breinholt2021].
-GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.
-
-However, the rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets is still far less than that available for plastid data (`r figure("genbank")`).
+GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in the recognition of multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.
+
+The rapid growth of genomic data notwithstanding, species level sampling of such nuclear phylogenomic datasets is still far less than that available from the plastome (`r figure("genbank")`).
 Furthermore, many subclades of ferns are under active investigation using both Sanger and next-gen sequencing of plastid markers, and plastid data for previously unsampled species will likely continue to grow at a rapid pace.
@@
 We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
-FTOL is available via the "ftolr" R package, as well as a web portal (https://fernphy.github.io).
-Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibration points) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2021.08.26.457744].
+FTOL is available via the "ftolr" R package [@ftol2022c], as well as a web portal (https://fernphy.github.io).
+Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibrations) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2021.08.26.457744].
 
 A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data merging [@Page2008].
-During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package (https://github.com/fernphy/pteridocat) and the "taxastand" package (https://github.com/joelnitta/taxastand).
+During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package [@ftol2022a] and the "taxastand" package [@joel_nitta_2021_5726391].
 We selected R because it is widely used by the biological research community, well established, and freely available [@Lai2019].
@@
 
-One feature of FTOL that sets it apart from the vast majority of other fern phylogenetic studies is its iterative nature.
+One feature of FTOL that sets it apart from the vast majority of other phylogenetic studies is its iterative nature.
 Unlike most other published fern phylogenies, the current version of FTOL described in this manuscript is not meant to be the last.
@@
     ~source, ~year, ~label,
-    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
+    "schuettpelz", 2009, "Schuettpelz and Pryer 2009",
     "rothfels", 2015, "Rothfels et al. 2015",
@@
     "Hasebe1995", 1995, 107, "Hasebe et al. 1995",
-    "Schuettpelz2007", 2007, 400, "Schuettpelz et al. 2007",
+    "Schuettpelz2007", 2007, 400, "Schuettpelz and Pryer 2007",
     "Lehtonen2011", 2011, 2957, "Lehtonen 2011",
@@
 ggsave(
-  result_file(figure("genbank"), "pdf"),
+  result_file(figure("genbank"), "tiff"),
   genbank_plot, width = 7, height = 5, units = "in")
@@
 @Schuettpelz2007 did not attempt exhaustive sampling but rather proportional sampling according to lineage size.
-The relatively small increase in number of species in 2021 is likely due to accessions that are still embargoed at the time of writing.
+The relatively small increase in number of species in 2021 may be due to accessions that are still embargoed at the time of writing.
 Taxonomy of GenBank species follows NCBI [@Federhen2012].
@@
 fs::file_copy(
-  here::here("images/workflow.pdf"),
-  result_file(figure("workflow"), "pdf"),
+  here::here("images/workflow.tiff"),
+  result_file(figure("workflow"), "tiff"),
   overwrite = TRUE)
@@
 The workflow is automated except for steps in boxes with red outlines and a hand symbol.
+Numbers of sequences and species are approximate.
 Sequences originating from whole plastomes are in blue; sequences typically obtained by Sanger sequencing are in orange.
@@
 a <- ggtree(
-  sanger_con_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
+  sanger_con_tree_dated, 
+    layout = "fan", right = FALSE, size = 0.1, open.angle = 2) + 
   ftol_geom_cladelab(
@@
   scale_color_manual(values = clade_cols) +
+  geom_treescale(
+    width = 100, # 100 Ma
+    fontsize = 2, 
+    y = 5600, # Just above "top" of tree
+    label = "100 Ma", 
+    offset = 1000, # hack to get default scale width (lacks units) to go away
+    offset.label = -10) +
   theme(legend.position = "none")
@@
 ggsave(
-  result_file(figure("ftol"), "pdf"),
+  result_file(figure("ftol"), "tiff"),
   ftol_plot, width = 7, height = 7, units = "in")
@@
 ggsave(
-  result_file(figure("backbone"), "pdf"),
+  result_file(figure("backbone"), "tiff"),
   backbone_plot, width = 7, height = 9, units = "in")
@@
 Fern tree of life (FTOL) backbone phylogeny.
-One exemplar tip is shown per family.
+One exemplar tip is shown per family (all families were found to be monophyletic; see Results).
 Ultrafast bootstrap support values (%) shown at nodes; unlabeled nodes are 100%.
@@
 ggsave(
-  result_file(figure("divtimes"), "pdf"),
+  result_file(figure("divtimes"), "tiff"),
   div_times_plot, width = 7, height = 8, units = "in")
@@
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
-For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
+For other studies, point indicates best (most likely) estimate [@Schuettpelz2009\; this study].
 Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.

### 2022-04-01 ms/si.Rmd
Update MS

@@
 # Specify path to FTOL cache
-# FIXME: change to ftol_cache when ready
 ftol_cache <- here::here("ftol_cache")
@@
 
-# Make table of Pteridocat genera not in PPGI
+# Make table of pteridocat genera not in PPGI
 pteridocat_not_in_ppgi <-
@@
   transmute(
-    diff = "Pteridocat not in PPGI",
+    diff = "pteridocat not in PPGI",
     genus,
@@
 
-# Make table of PPGI genera not in Pteridocat
+# Make table of PPGI genera not in pteridocat
 ppgi_not_in_pteridocat <-
@@
   transmute(
-    diff = "PPGI not in Pteridocat",
+    diff = "PPGI not in pteridocat",
     genus
@@
 **`r s_table("genera_diff")`**.
-Differences in genera included as accepted in Pteridocat taxonomic database (https://github.com/pteridophy/pteridocat) vs.
+Differences in genera included as accepted in "pteridocat" taxonomic database v0.0.1 (https://github.com/fernphy/pteridocat) vs.
 PPGI [@PteridophytePhylogenyGroupI2016] (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
-"diff" is type of difference (Pteridocat genus not in PPGI or PPGI genus not in Pteridocat).
-"n_species" is the number of accepted species in each genus (Pteridocat) or estimated number of species (PPGI).
+"diff" is type of difference (pteridocat genus not in PPGI or PPGI genus not in pteridocat).
+"n_species" is the number of accepted species in each genus (pteridocat) or estimated number of species (PPGI).
 References in "notes" column provided at end of Supplementary Materials.
-"World Ferns" refers to @Hassler2022, which was modified to create Pteridocat.
+"World Ferns" refers to @Hassler2022, which was modified to create pteridocat.
 
@@
 Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv") %>% fs::path_file()`).
-"n_fos" is the fossil ID number used in our fossil database (https://github.com/pteridophy/fossils).
-"minimum_age" (Ma) is the age used for calibration; "lower" is the lower limit of the oldest stratigraphic age for the fossil; "upper" is the upper limit of the oldest stratigraphic age for the fossil.
+"n_fos" is the fossil ID number used in the "ferncal" database v1.0.0 (https://github.com/fernphy/ferncal).
+"minimum_age" is the age used for calibration (Ma); "lower" is the lower limit of the oldest stratigraphic age for the fossil (Ma); "upper" is the upper limit of the oldest stratigraphic age for the fossil (Ma).
 For detailed notes and full references for each fossil, see Appendix S2.
@@
 "seq_len" is the total number of bases excluding missing bases ("?", "N", or "-") and only includes bases belonging to loci used in this study (not the entire accession).
-"sci_name" is the scientific name in the Pteridocat taxonomic database.
-"ncbi_name" is the scientific name used in the NCBI taxonomic database.
+"sci_name" is the scientific name in the pteridocat taxonomic database.
+"ncbi_name" is the scientific name used in the NCBI taxonomic database version 2022-02-01.
 "ncbi_taxid" is the NCBI taxonomy database unique identifier.
@@
 **`r s_figure("fossil_map")`**.
-Maximum likelihood fern phylongey with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
+Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
 Number for each constraint is age in Ma.
@@
 # Make tibble for printing BS values
-bs_tibble <- make_bs_tbl(phy)
-
-# Make plot
-# optional: round version
-bs_plot_round <- 
-  ggtree(
-    phy,
-    layout = "circular",
-    branch.length = "none", 
-    size = 0.1, color = "grey30") %<+% bs_tibble +
-  # Add node points
-  geom_nodepoint(
-    aes(color = bs), size = 0.2) +
-  scale_color_viridis_c(
-    "BS",
-    option = "B", na.value = "transparent",
-    breaks = 0.25*0:4, labels = percent(0.25*0:4))
+bs_tibble <- make_bs_tbl(phy) %>%
+  mutate(bs = bs*0.01) # so percentages print correctly
 
@@
     option = "B", na.value = "transparent",
-    breaks = 0.25*0:4, labels = percent(0.25*0:4)) +
+    labels = scales::percent
+    ) +
   # Add tip labels
@@
   plot = bs_plot,
-  file = (result_file(s_figure("bs"), "pdf")),
+  file = result_file(s_figure("bs"), "pdf"),
   height = 60, width = 8, units = "in",
@@
 Consensus phylogeny (Sanger dataset) with node labels showing bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
+Darker colors indicate lower support.
 The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
@@
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
-For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
+For other studies, point indicates best (most likely) estimate [@Schuettpelz2009\; this study].
 @Du2021 used six different dating schemes; for details, see that paper.
 Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
-Period name abbreviations as follows: FIXME.
+Period name abbreviations as in `r figure("divtimes")`.
 
@@
     legend.title = element_blank(),
+    axis.text.y = element_text(size = 6),
     legend.position = "bottom"
@@
   scale_shape_manual(
-    values = ages_shapes,
-    limits = ages_sampling$source, 
-    labels = ages_sampling$label) +
+    values = ages_shapes[c("ts", "current")],
+    limits = c("ts", "current"), 
+    labels = ages_sampling %>% 
+      filter(source %in% c("current", "ts")) %>%
+      arrange(source) %>%
+      pull(label)) +
   scale_color_manual(
-    values = ages_colors,
-    limits = ages_sampling$source, 
-    labels = ages_sampling$label) +
+    values = ages_colors[c("ts", "current")],
+    limits = c("ts", "current"), 
+    labels = ages_sampling %>% 
+      filter(source %in% c("current", "ts")) %>%
+      arrange(source) %>%
+      pull(label)) +
   scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
@@
 
+\newpage
+
 ```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}
@@
 Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramirez-Barahona2020].
-Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020.
+Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020; for details, see that paper.
 Log number of lineages shown as percent of total to enable comparison across trees.

### 2022-04-01 ms/si2.Rmd
Update MS

@@
 
-These data are exported from the FernCal v1.0 database (https://github.com/fernphy/ferncal), which includes `r nrow(fossil_ferns_raw)` records of fossils to use as calibration points for molecular dating of ferns.
+These data are exported from the FernCal v1.0.0 database (https://github.com/fernphy/ferncal), which includes records of fossils to use as calibration points for molecular dating of ferns.
 
@@
 
-Taxonomy of extant taxa follows pteridocat v1.0 database (https://github.com/pteridophy/pteridocat).
+Taxonomy of extant taxa follows pteridocat v0.0.1 database (https://github.com/fernphy/pteridocat).
 

### 2022-04-01 R/process_refs.R

@@ -3,11 +3,11 @@ library(tidyverse)
 library(rmdref)
 
 # Filter YAML references from Zotero library.
-# data_raw/main_library.yaml has been exported from Zotero
+# main_library.yaml has been exported from Zotero
 # like this: file -> "export library" -> "Better CSL YAML"
 filter_refs_yaml(
   rmd_file = "ms/manuscript.Rmd",
-  yaml_in = "data_raw/main_library.yaml",
+  yaml_in = "_targets/user/data_raw/main_library.yaml",
   yaml_out = "ms/references.yaml", silent = TRUE)
 
 # Same for SI
@@ -16,7 +16,7 @@ filter_refs_yaml(
     "ms/si.rmd", # SI
     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv" # refs in SI tables
   ),
-  yaml_in = "data_raw/main_library.yaml",
+  yaml_in = "_targets/user/data_raw/main_library.yaml",
   yaml_out = "ms/si_references.yaml", silent = TRUE)
 
 # Write out 'aux' file with citation keys to make collection in Zotero

### 2022-04-01 ms/manuscript.Rmd
Use 'my' for timescale, convert workflow to tiff with pdftools

@@
 ```{r workflow-save, eval = params$doc_type == "doc" }
-# Copy workflow image if rendering to doc
-fs::file_copy(
-  here::here("images/workflow.tiff"),
-  result_file(figure("workflow"), "tiff"),
-  overwrite = TRUE)
+# Convert workflow image to tiff for submission
+pdftools::pdf_convert(
+  pdf = here::here("images/workflow.pdf"),
+  filenames = result_file(figure("workflow"), "tiff"),
+  format = "tiff",
+  dpi = 300)
 ```
@@
     y = 5600, # Just above "top" of tree
-    label = "100 Ma", 
+    label = "100 my", 
     offset = 1000, # hack to get default scale width (lacks units) to go away

### 2022-06-20 install_latex.R
Need install_latex.R for Dockerfile

@@ -1,0 +1,50 @@
+# Install latex packages using tinytex
+
+# This would happen automatically anyways when rendering the Rmd to pdf,
+# but requires downloading packages and may not work during updates of Tex Live.
+# Better to install to the docker image once and keep them there.
+
+tinytex::tlmgr_update()
+
+latex_packages <- c(
+  "amsmath",
+  "auxhook",
+  "bigintcalc",
+  "bitset",
+  "booktabs",
+  "colortbl",
+  "etexcmds",
+  "etoolbox",
+  "euenc",
+  "fancyhdr",
+  "float",
+  "fontspec",
+  "geometry",
+  "gettitlestring",
+  "hycolor",
+  "hyperref",
+  "iftex",
+  "infwarerr",
+  "intcalc",
+  "kvdefinekeys",
+  "kvoptions",
+  "kvsetkeys",
+  "latex-amsmath-dev",
+  "letltxmacro",
+  "ltxcmds",
+  "mdwtools",
+  "multirow",
+  "pdfescape",
+  "pdftexcmds",
+  "refcount",
+  "rerunfilecheck",
+  "stringenc",
+  "tipa",
+  "unicode-math",
+  "uniquecounter",
+  "xcolor",
+  "xunicode",
+  "zapfding"
+)
+
+tinytex::tlmgr_install(latex_packages)

### 2022-06-20 R/dev_packages.R
Add dev_packages.R
for working in vs-code

@@ -1,0 +1,4 @@
+# Packages that don't need to be loaded directly, but
+# should be tracked by renv
+library(languageserver)
+library(httpgd)

### 2022-06-20 ms/manuscript.Rmd
Use read_from_cache()
as work-around to tar_load(), which stopped working with ftol_cache

@@
 
-tar_load(
-  c(
-    accs_exclude,
-    con_fossil_calibration_tips,
-    fern_sanger_seqs_raw,
-    fossil_ferns_raw,
-    plastid_calibration_dates,
-    plastome_alignment,
-    plastome_alignment_tbl,
-    plastome_metadata_raw_all,
-    plastome_metadata_renamed,
-    plastome_metadata_renamed,
-    plastome_seqs_combined_filtered,
-    plastome_tree_rooted,
-    ppgi_taxonomy,
-    pteridocat,
-    raw_fasta_all,
-    sanger_accessions_selection,
-    sanger_alignment,
-    sanger_alignment_tbl,
-    sanger_con_tree_dated,
-    sanger_con_tree_rooted,
-    sanger_seqs_combined_filtered,
-    sanger_seqs_rogues_removed,
-    ts_fossil_calibration_tips
-  ),
-  store = ftol_cache
-)
+accs_exclude <- read_from_cache("accs_exclude", ftol_cache)
+con_fossil_calibration_tips <- read_from_cache("con_fossil_calibration_tips", ftol_cache) # nolint
+fern_sanger_seqs_raw <- read_from_cache("fern_sanger_seqs_raw", ftol_cache)
+fossil_ferns_raw <- read_from_cache("fossil_ferns_raw", ftol_cache)
+plastome_alignment <- read_from_cache("plastome_alignment", ftol_cache)
+plastome_alignment_tbl <- read_from_cache("plastome_alignment_tbl", ftol_cache)
+plastome_metadata_raw_all <- read_from_cache("plastome_metadata_raw_all", ftol_cache) # nolint
+plastome_metadata_renamed <- read_from_cache("plastome_metadata_renamed", ftol_cache) # nolint
+plastome_metadata_renamed <- read_from_cache("plastome_metadata_renamed", ftol_cache) # nolint
+plastome_seqs_combined_filtered <- read_from_cache("plastome_seqs_combined_filtered", ftol_cache) # nolint
+plastome_tree_rooted <- read_from_cache("plastome_tree_rooted", ftol_cache)
+pteridocat <- read_from_cache("pteridocat", ftol_cache)
+raw_fasta_all <- read_from_cache("raw_fasta_all", ftol_cache)
+sanger_accessions_selection <- read_from_cache("sanger_accessions_selection", ftol_cache) # nolint
+sanger_alignment <- read_from_cache("sanger_alignment", ftol_cache)
+sanger_alignment_tbl <- read_from_cache("sanger_alignment_tbl", ftol_cache)
+sanger_con_tree_rooted <- read_from_cache("sanger_con_tree_rooted", ftol_cache)
+sanger_seqs_combined_filtered <- read_from_cache("sanger_seqs_combined_filtered", ftol_cache) # nolint
+sanger_seqs_rogues_removed <- read_from_cache("sanger_seqs_rogues_removed", ftol_cache) # nolint
+ts_fossil_calibration_tips <- read_from_cache("ts_fossil_calibration_tips", ftol_cache) # nolint
 
@@
     total_coverage,
-    treepl_cv_results
+    treepl_cv_results,
+    ppgi_taxonomy,
+    sanger_tree_dated
   ),

### 2022-06-20 ms/si.Rmd
Use read_from_cache()
as work-around to tar_load(), which stopped working with ftol_cache

@@
 
-tar_load(
-  c(
-    acc_table_long,
-    con_fossil_calibration_tips,
-    con_fossil_calibrations_for_treepl,
-    fossil_ferns_all,
-    plastome_seqs_combined_filtered,
-    plastome_tree,
-    plastome_tree_rooted,
-    ppgi_taxonomy,
-    sanger_accessions_selection,
-    sanger_sampling,
-    sanger_con_tree_dated,
-    sanger_con_tree_rooted,
-    sanger_alignment_tbl,
-    ts_fossil_calibration_tips
-  ),
-  store = ftol_cache
-)
+acc_table_long <- read_from_cache("acc_table_long", ftol_cache)
+con_fossil_calibration_tips <- read_from_cache("con_fossil_calibration_tips", ftol_cache) # nolint
+con_fossil_calibrations_for_treepl <- read_from_cache("con_fossil_calibrations_for_treepl", ftol_cache) # nolint
+fossil_ferns_all <- read_from_cache("fossil_ferns_all", ftol_cache)
+plastome_seqs_combined_filtered <- read_from_cache("plastome_seqs_combined_filtered", ftol_cache) # nolint
+plastome_tree <- read_from_cache("plastome_tree", ftol_cache)
+plastome_tree_rooted <- read_from_cache("plastome_tree_rooted", ftol_cache)
+sanger_accessions_selection <- read_from_cache("sanger_accessions_selection", ftol_cache) # nolint
+sanger_con_tree_dated <- read_from_cache("sanger_con_tree_dated", ftol_cache)
+sanger_con_tree_rooted <- read_from_cache("sanger_con_tree_rooted", ftol_cache)
+sanger_alignment_tbl <- read_from_cache("sanger_alignment_tbl", ftol_cache)
+ts_fossil_calibration_tips <- read_from_cache("ts_fossil_calibration_tips", ftol_cache) # nolint
 
@@
     accepted_species,
-    ftol_ts_comp_mod
+    ftol_ts_comp_mod,
+    ppgi_taxonomy,
+    sanger_tree_dated,
+    sanger_sampling
   ),

### 2022-06-20 ms/si2.Rmd
Use read_from_cache()
as work-around to tar_load(), which stopped working with ftol_cache

@@
 ```{r load-targets-si-2}
-# Targets from MS workflow
+# Targets from FTOL workflow
 
@@
 
-tar_load(
-  c(
-    con_fossil_calibration_tips,
-    fossil_ferns_raw),
-  store = ftol_cache,
+con_fossil_calibration_tips <- read_from_cache(
+  "con_fossil_calibration_tips", ftol_cache)
+fossil_ferns_raw <- read_from_cache(
+  "fossil_ferns_raw", ftol_cache
 )

### 2022-06-20 R/functions.R

@@ -1206,6 +1206,35 @@ make_bs_tbl <- function(phy) {
 
 # Etc -----
 
+# Work-around for tar_load()
+# for some reason, doesn't work with tar_load(store = ftol_cache)
+read_from_cache <- function(object, cache) {
+  readRDS(
+    fs::path(cache,
+    glue::glue("objects/{object}"))
+  )
+}
+
+#' Correct PPGI taxonomy
+#'
+#' @param ppgi_taxonomy Dataframe; PPGI taxonomy with incorrect entry for
+#' Dryopolystichum
+#'
+#' @return ppgi_taxonomy with correct data for Dryopolystichum
+fix_ppgi <- function(ppgi_taxonomy) {
+  ppgi_taxonomy %>%
+    mutate(
+      family = case_when(
+        genus == "Dryopolystichum" ~ "Lomariopsidaceae",
+        TRUE ~ family
+      ),
+      notes = case_when(
+        genus == "Dryopolystichum" ~ "Updated classification Chen et al 2017 Phytokeys",
+        TRUE ~ notes
+      )
+    )
+}
+
 #' Function to get support value in BS%
 #'
 #' @param phy an object of class "phylo".

### 2022-06-20 ms/si.Rmd
Don't need to add Phlebosia in si.Rmd anymore

@@
 ```{r genera-diff-make}
-# FIXME: need to update ppgi csv file
-# Add one missing nothogenus
-ppgi_taxonomy_mod <-
-  bind_rows(
-    ppgi_taxonomy,
-    tibble(
-      genus = "Phlebosia",
-      notes = "Nothogenus"
-    )
-  )
-
 # Verify that all accepted genera are accounted for
@@
   count(genus, name = "n_species") %>%
-  anti_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
+  anti_join(select(ppgi_taxonomy, genus, notes), by = "genus") %>%
   verify(nrow(.) == 0, success_fun = success_logical)
@@
   count(genus, name = "n_species") %>%
-  left_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
+  left_join(select(ppgi_taxonomy, genus, notes), by = "genus") %>%
   assert(not_na, notes) %>%
@@
 ppgi_not_in_pteridocat <-
-  ppgi_taxonomy_mod %>%
+  ppgi_taxonomy %>%
   filter(notes == "original PPGI 2016",

### 2022-06-20 ms/si2.Rmd
Add dagger for extinct taxa

@@
   magrittr::extract(
-    str_detect(., "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(")
+    str_detect(
+      .,
+    "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(" # nolint
+    )
   ) %>%
@@
       variable %in% colnames(fossil_ferns_raw)[1:4] ~ "Calibration",
-      variable %in% colnames(fossil_ferns_raw)[5:12] ~ "Fossil Identity",
-      variable %in% colnames(fossil_ferns_raw)[13:18] ~ "Fossil Age",
-      variable %in% colnames(fossil_ferns_raw)[19:21] ~ "Fossil Locality",
-      variable %in% colnames(fossil_ferns_raw)[22:24] ~ "Fossil Specimen",
+      variable %in% colnames(fossil_ferns_raw)[5:13] ~ "Fossil Identity",
+      variable %in% colnames(fossil_ferns_raw)[14:19] ~ "Fossil Age",
+      variable %in% colnames(fossil_ferns_raw)[20:22] ~ "Fossil Locality",
+      variable %in% colnames(fossil_ferns_raw)[23:25] ~ "Fossil Specimen",
       TRUE ~ NA_character_
     ),
-    var_order = 1:nrow(.)
+    var_order = seq_along(variable)
   )
@@
     across(everything(), as.character),
+    extinct = case_when(
+      extinct == "extinct" ~ "^†^",
+      extinct == "extant" ~ "",
+      TRUE ~ NA_character_
+    )) %>%
+  assert(not_na, extinct, fossil_taxon) %>%
   # Add fossil taxon header
-    fossil_taxon_header = fossil_taxon) %>%
+  mutate(fossil_taxon_header = glue::glue("{fossil_taxon}{extinct}")) %>%
   pivot_longer(names_to = "variable", values_to = "value", -fossil_taxon) %>%
@@
   # Drop NA entries
-  filter(!is.na(value))
+  filter(!is.na(value)) %>%
+  filter(variable != "extinct")
 ```
 
-These data are exported from the FernCal v1.0.0 database (https://github.com/fernphy/ferncal), which includes records of fossils to use as calibration points for molecular dating of ferns.
+These data are exported from the FernCal v1.0.1 database (https://github.com/fernphy/ferncal), which includes records of fossils to use as calibration points for molecular dating of ferns.
 

### 2022-06-20 ms/si.Rmd
Fix formatting

@@
 knitr::opts_chunk$set(
-  echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
+  echo = FALSE, warning = FALSE, message = FALSE, results = "hide",
+  cache = FALSE)
 # Load packages

### 2022-06-20 ms/si2.Rmd
Fix formatting

@@
 ```{r setup-rmd, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, results = 'hide')
+knitr::opts_chunk$set(echo = FALSE, include = FALSE,
+  warning = FALSE, message = FALSE, results = "hide")
 library(glue)
@@
 ```{r load-captions-si2, cache = FALSE}
-# Source captions numbers from MS
-# 
+# Source captions numbers from MS.
 # So that caption numbers are in order of appearance in MS

### 2022-06-20 _targets.R
Implement temporary fix for Dryopolystichum
Need to fix in PPGI data in ftol
See https://github.com/fernphy/ftol/issues/16

@@ -58,7 +58,6 @@ tmp <- capture.output({
 )
 
 tar_plan(
-  
   # Load data ----
   # - Dates from Testo and Sundue 2016 SI
   # (includes Rothfels 2015, Schuettpelz & Pryer 2009)
@@ -123,18 +122,16 @@ tar_plan(
     readRDS(file = !!.x)
   ),
   tar_file_read(
-    ppgi_taxonomy,
+    ppgi_taxonomy_raw,
     fs::path(ftol_cache, "objects/ppgi_taxonomy"),
     readRDS(file = !!.x)
   ),
+  # FIXME: Fix ppgi to treat Dryopolystichum as Lomariopsis
+  # can remove this once https://github.com/fernphy/ftol/issues/16 is resolved
+  ppgi_taxonomy = fix_ppgi(ppgi_taxonomy_raw),
   tar_file_read(
     pteridocat,
     fs::path(ftol_cache, "objects/pteridocat"),
-    readRDS(file = !!.x)
-  ),
-  tar_file_read(
-    con_monophy_by_clade,
-    fs::path(ftol_cache, "objects/con_monophy_by_clade"),
     readRDS(file = !!.x)
   ),
   tar_file_read(
@@ -142,7 +139,11 @@ tar_plan(
     fs::path(ftol_cache, "objects/ts_sanger_tree_dated"),
     readRDS(file = !!.x)
   ),
-  
+  tar_file_read(
+    og_to_drop,
+    fs::path(ftol_cache, "objects/og_to_drop"),
+    readRDS(file = !!.x)
+  ),
   # GenBank ----
   # Analyze number of fern accessions and species in GenBank by type of DNA 
   # (plastid, mitochondrial, or nuclear)
@@ -161,7 +162,7 @@ tar_plan(
   # - Load NCBI taxonomy
   tar_file_read(
     ncbi_names,
-    fs::path(ftol_cache, "user/data_raw/taxdmp_2022-02-01.zip"),
+    "_targets/user/data_raw/taxdmp_2022-02-01.zip",
     load_ncbi_names(taxdump_zip_file = !!.x, taxid_keep = unique(gb_taxa$taxid))
   ),
   # - Count number of species and accessions in GenBank
@@ -184,9 +185,10 @@ tar_plan(
     plastome_tree, sanger_sampling, accepted_species, "total_sampling"),
   
   # Analyze monophyly and ages ----
-  # - Make Sanger sampling table
+  # - Make Sanger sampling table, not including Zygnema
[…]

### 2022-06-20 ms/manuscript.Rmd
Use sanger_tree_dated from _targets workflow

@@
     taxon_sampling = .,
-    tree = sanger_con_tree_dated,
+    tree = sanger_tree_dated,
     tax_levels = "major_clade"
@@
 sanger_clade_exemp_taxa <-
-  sanger_con_tree_dated %>%
+  sanger_tree_dated %>%
   ladderize() %>%
@@
 a <- ggtree(
-  sanger_con_tree_dated, 
+  sanger_tree_dated, 
     layout = "fan", right = FALSE, size = 0.1, open.angle = 2) + 

### 2022-06-20 ms/manuscript.Rmd
Update MS according to review

@@
 knitr::opts_chunk$set(
-  echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
+  echo = FALSE, warning = FALSE, message = FALSE,
+  results = 'hide', cache = FALSE)
 # Load packages
@@
 
-<!-- Don't indent -->
+<!-- Do not indent -->
 \setlength{\parindent}{0ex}
@@
 Our tailored sampling strategy combines whole plastomes (few taxa, many loci) with commonly sequenced plastid regions (many taxa, few loci) to obtain a global, species-level fern phylogeny with high resolution along the backbone and maximal sampling across the tips.
-We use an curated reference taxonomy to resolve synonyms in general compliance with the community-driven Pteridophyte Phylogeny Group I classification.
+We use a curated reference taxonomy to resolve synonyms in general compliance with the community-driven Pteridophyte Phylogeny Group I classification.
 
@@
 
-Recently, a genus-level taxonomy of ferns and lycophytes was established using an inclusive, community-driven approach  [@PteridophytePhylogenyGroupI2016\; hereafter "PPG I"].
+Recently, a taxonomy of ferns and lycophytes was established at the genus level and higher using an inclusive, community-driven approach  [@PteridophytePhylogenyGroupI2016\; hereafter "PPG I"].
 PPG I has been widely accepted and used, but there were problematic (non-monophyletic) genera included at the time of publication, and many taxonomic changes have been (and will continue to be) proposed since [`r eg`, @Almeida2017; @Shang2018; @Zhang2020].
@@
 We plan to run the pipeline on a regular basis and make the results freely available online through a web portal (https://fernphy.github.io) and R package [@ftol2022c].
-This will enable anybody interested in the biology of ferns to have access to the most current hypothesis of fern phylogeny, as well as an associated time-tree dated using a curated list of fern fossils.
+This will enable anybody interested in the biology of ferns to have access to the most current hypothesis of fern phylogeny, and an associated time-tree dated using a curated list of fern fossils.
 We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern relationships; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
@@
 Our sampling includes two major categories of sequence data: 1) seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used in molecular analyses of ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci"); and 2) a much larger set of single-copy loci typically obtained from next-generation sequencing of the plastome ("plastome loci").
-Here, the *trnL*--*trnF* locus includes the *trnL* intron, the 3' *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
+Here, the *trnL*--*trnF* locus includes the *trnL* intron, the `r "3'"` *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
 The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
@@
 
-All sequences were downloaded from GenBank.
-GenBank queries were formatted to include the locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify a date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query results in the same set of sequences regardless of when the query is made (with the possible exception of accessions that were embargoed at the time of querying).
+The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server and a local database including only ferns and outgroups (selected seed plants, lycophytes, bryophytes, and algae) created using the restez R package [@Bennett2018a].
+GenBank accession numbers corresponding to sequences potentially matching each locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cuteoff matching the current GenBank release.
 Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
+Sequences were then pulled from the local database in FASTA format by GenBank accession number.
 There is no formal definition of genomic vs.
@@
 To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
-Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
+Briefly, this involves querying candidate FASTA files from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
 Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
@@
 We manually inspected any fuzzily matched or non-matching names.
-This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, as well as some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
+This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, and some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
 We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
@@
 Furthermore, multiple genera included in PPG I were known to be non-monophyletic and provisionally circumscribed [@PteridophytePhylogenyGroupI2016].
-Pteridocat includes newly published genera, some genera that were not recognized by PPG I, as well as nothogenera, which were not included in PPG I [@Liu2020a].
+Pteridocat includes newly published genera, some genera that were not recognized by PPG I, and nothogenera, which were not included in PPG I [@Liu2020a].
 Differences between pteridocat and PPG I are summarized in `r s_table("genera_diff")`.
@@
 
+In addition to this automated approach, we also implemented a manual approach for Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in a previous version of FTOL (v1.0.0) and likely has high rates of misidentification on GenBank.
+For the manual approach, we used a curated list of accessions from a previous phylogenetic analysis of this family [@Patel2019].
+After resolving names in the list of @Patel2019 to pteridocat and correcting some errors (wrong GenBank accession number listed in small number of cases), accessions in this list were used preferentially over those identified by the automated approach.
+
 ## Sequence alignment
@@
 
-Dating was conducted separately on the ML tree and the consensus tree.
-We rooted the tree with bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
+Dating was conducted separately on the ML tree and the consensus tree using penalized likelihood with treePL [@Smith2012].
+We initially rooted the tree on *Zygnema circumcarinatum* Czurda, a member of the group of algae (Zygnematophyceae) thought to be most closely related to land plants [@Donoghue2021].
+Since there is no way to objectively divide branch length between the branch leading to the outgroup and the rest of the tree, we then trimmed *Z.
+circumcarinatum* from the phylogeny; this effectively positioned bryophytes as the outgroup on a branch or branches whose length has been estimated from the data [@Sauquet2013].
 
@@
 The phylogenetic position of Equisetidae relative to other ferns has long been contentious.
-Here, we recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with many other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018].
-However, this contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have all recovered Equisetidae as sister to all other ferns.
+Here, we recovered Equisetidae as sister to Ophioglossidae (Psilotales + Ophioglossales), in agreement with many other plastid phylogenomic analyses [@Grewe2013; @Ruhfel2014; @Gitzendanner2018b; @Lehtonen2019; @Kuo2018; @OneThousandPlantTranscriptomesInitiative2019].
+However, this contradicts nuclear phylogenomic analyses [@Rothfels2015a; @Qi2018; @Shen2018; @OneThousandPlantTranscriptomesInitiative2019], most plastid analyses with smaller numbers (`r ca` 3--17) of genes [@Schneider2004d; @Rai2010; @Kuo2011; @Testo2016a] and an analysis including mitochondrial data [@Knie2015], which have all recovered Equisetidae as sister to all other ferns.
 Some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae, albeit with generally low support [@Pryer2001; @Qiu2006; @Qiu2007].
@@
 However, the subsequent placement of Hymenophyllales differs across studies: some recover Hymenophyllales as the next diverging lineage after Osmundales [@Pryer2001; @Testo2016a; @Schneider2004d; @Schuettpelz2006; @Schuettpelz2007], while others recover a clade comprising Hymenophyllales sister to Gleicheniales, which then together are sister to the remaining (non-Osmundales) leptosporangiates [@Pryer2004a; @Lehtonen2017].
-Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018].
-Here, we recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
-However, current plastome sampling only includes `r plastome_gleich_genera_n %>% jntools::english2()` out of `r ppgi_gleich_genera_n` genera of Gleicheniales; additional taxon sampling may help to resolve this relationship.
+Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018; @Shu2022].
+Here, we recovered a monophyletic Gleicheniales (including Gleicheniaceae, Dipteridaceae, and Matoniaceae) sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
+However, our plastome sampling only includes `r plastome_gleich_genera_n %>% jntools::english2()` out of `r ppgi_gleich_genera_n` genera of Gleicheniales, and lacks Matoniaceae.
+While our study was in revision, 42 additional fern plastomes were published, including seven families (`r ie`, Anemiaceae, Culcitaceae, Dipteridaceae, Loxsomataceae, Matoniaceae, Metaxyaceae, and Thyrsopteridaceae) that are not included in our sampling since they are not in the most recent available GenBank release [release 249\; @Du2022].
+The addition of Matoniaceae results in significantly different relationships between Hymenophyllales and Gleicheniales; @Du2022 recovered Matoniaceae sister to Dipteridaceae, and these together sister to the remaining leptosporangiates.
+Futhermore, @Du2022 did not recover a sister relationship between Hymenophyllaceae and Gleicheniaceae, nor monophyly of Gleicheniales.
+Interestingly, addition of Matoniaceae transcriptome data also supported the sister status of Matoniaceae + Dipteridaceae, as well as the non-monophyly of Gleicheniales in another recent study [@Shu2022].
 
@@
 We recover (D, (P, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] and nuclear data [@Rothfels2015a; @Qi2018; @Shen2018], as well as a plastid supermatrix [@Lehtonen2011].
-Recently the whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others.
+The whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others; a similar study with expanded sampling generally supported ((D, P), e) [@Du2022].
 
 Although some relationships within Polypodiineae (eupolypods I) had previously been resolved differently between various studies using Sanger sequencing, such as Nephrolepidaceae sister to Lomariopsidaceae [@Schuettpelz2007; @Zhang2015] vs.
-Nephrolepidaceae sister to Tectariaceae, Oleandraceae, Davalliaceae, and Polypodiaceae [@Kuo2011; @Lehtonen2011; @Liu2013; @Testo2016a], our study is in agreement with both nuclear [@Qi2018; @Shen2018] and plastid phylogenomic analyses [@Du2021] that support the latter.
-Similarly, although Didymochlaenaceae had previously been identified as either sister to the remainder of Polypodiineae [@Kuo2011; @Zhang2015; @Testo2016a] or nested with Hypodematiaceae [@Schuettpelz2007; @Lehtonen2011] by studies using Sanger sequencing, our study as well as nuclear [@Qi2018] and plastid phylogenomic analyses [@Du2021] indicate that Hypodematiaceae is sister to the remainder of the clade.
+Nephrolepidaceae sister to Tectariaceae, Oleandraceae, Davalliaceae, and Polypodiaceae [@Kuo2011; @Lehtonen2011; @Liu2013; @Testo2016a], our study is in agreement with both nuclear [@Qi2018; @Shen2018] and plastid phylogenomic analyses [@Du2021; @Du2022] that support the latter.
+Similarly, although Didymochlaenaceae had previously been identified as either sister to the remainder of Polypodiineae [@Kuo2011; @Zhang2015; @Testo2016a] or nested with Hypodematiaceae [@Schuettpelz2007; @Lehtonen2011] by studies using Sanger sequencing, our study as well as nuclear [@Qi2018] and plastid phylogenomic analyses [@Du2021; @Du2022] indicate that Hypodematiaceae is sister to the remainder of the clade.
 
@@
 
-We recovered older crown ages for ferns and large clades therein, as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+We recovered older crown ages for ferns and large clades therein, and older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 This is almost certainly due to our use of a completely revised and greatly expanded set of fossil calibration points relative to previous studies, which not only resulted in a more densely constrained tree but also a higher number of fern families with minimum fossil ages.
@@
 
+## Comparison with other automated phylogeny pipelines
+
+Ferns are not unique in having large amounts of publicly available DNA sequence data, and several pipeline tools exist that can leverage such data to automatically generate phylogenies [@Antonelli2016; @Bennett2018; @Drori2018a; @Pearse2013; @Portik2020; @Smith2019a; @Xu2015].
+In particular, superSMART [@Antonelli2016] and pyPHLAWD [@Smith2019a] are recently developed pipelines that can generate maximally sampled phylogenies for any higher taxon of choice; the latter was used to generate a broadly inclusive seed plant phylogeny [@Smith2018].
+
+However, although we make use of some of the functionality of these tools [`r eg`, restez for generating a local copy of GenBank\; superCRUNCH for identifying orthologous sequences\; @Bennett2018a; @Portik2020], our pipeline is mostly custom-built for ferns.
+We chose this approach because it allowed us to implement various steps specific to ferns that would not be possible using a fully automated, taxon-agnostic pipeline like superSMART or pyPHLAWD, which we believe ultimately results in a higher-quality fern phylogeny.
+For example, we integrate our phylogeny with a custom taxonomy for ferns (pteridocat), tailor alignment strategies to locus and taxon (fern-wide alignment for coding loci, nested alignment within families for non-coding loci), inspect intermediate results and modify input based on expert taxonomic knowledge (curation of sample inclusion and exclusion lists), and employ a custom set of criteria for selecting and concatenating sequences within species.
+While we have not sought to make our workflow available as a general tool due to its high degree of specialization, some of the methods employed here could be used in other studies, such as use of a backbone phylogeny based on next-generation data to constrain a tree built with Sanger data.
+
 ## Accesibility and usage of FTOL
@@
 We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
-FTOL is available via the "ftolr" R package [@ftol2022c], as well as a web portal (https://fernphy.github.io).
+FTOL is available via the "ftolr" R package [@ftol2022c], and a web portal (https://fernphy.github.io).
 Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibrations) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2021.08.26.457744].
@@
 Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked (`r eg`, placement of species within genera, which we did not have the resources to inspect) or because an updated taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
-That is why we have made our methodology (code), data, and software (R packages and docker image) completely open and available to the research community.
+One example of progressive refinement is Thelypteridaceae, which had the highest rate of non-monophyletic genera in the previous version of FTOL (v1.0.0).
+After implementing a manual inclusion list for Thelypteridaceae and consulting with a taxonomic expert on this family (Susan Fawcett), the number of non-monophyletic genera dropped to __ in v1.0.1.
+While manual inclusion lists are not an ideal long-term solution because they cannot grow with GenBank, this case demonstrates that such lists are a reasonable option for taxonomically difficult groups, and more importantly, the improvement that results from input from taxonomic experts.
+Another example of the advantage of a continuously updated tree is the publication of 42 additional fern plastome samples while this paper was in revision [@Du2022].
+While we were unable to integrate these data into the current version of FTOL, they will be automatically added when the next version of GenBank data are released, and we expected the next version of FTOL will reflect this new knowledge.
+
+To facilitate community-driven improvements, we have made our methodology (code), data, and software (R packages and docker image) completely open and available.
 It is our hope that other researchers using FTOL will contribute by making edits and suggestions, preferably through the GitHub repository (https://github.com/fernphy/ftol).
@@
 Alexander White provided helpful comments on an early version of the analysis.
+Susan Fawcett provided helpful comments on taxonomy of Thelypteridaceae.
 This study was supported by JSPS KAKENHI Grant Number 16H06279 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).

### 2022-06-21 ms/manuscript.Rmd
Extract stats for best-scoring tree

@@
 
+```{r phylogeny-stats-log}
+sanger_log_files <- fs::dir_ls(
+ here::here("ftol_cache/user/intermediates/iqtree/"),
+ recurse = TRUE,
+ regexp = "sanger_[0-9]+/sanger_alignment.phy.log"
+ )
+
+rename_sanger <- function(x) {
+  names(x) <- str_match(names(x), "sanger_[0-9]") %>%
+    magrittr::extract(,1)
+  x
+}
+
+sanger_ml_log_rep <- sanger_log_files %>%
+  map(read_lines) %>%
+  unlist()
+
+ml_scores <- sanger_ml_log_rep[
+  str_detect(sanger_ml_log_rep, "BEST SCORE FOUND")] %>%
+  rename_sanger() %>%
+  map_dbl(parse_number)
+
+con_scores <- sanger_ml_log_rep[
+  str_detect(sanger_ml_log_rep, "Log-likelihood of consensus tree")] %>%
+  rename_sanger() %>%
+  map_chr(~str_remove_all(., "Log-likelihood of consensus tree: ")) %>%
+  map_dbl(parse_number)
+combined_scores <- ml_scores + con_scores
+
+assertthat::assert_that(
+  !any(is.na(ml_scores))
+)
+assertthat::assert_that(
+  !any(is.na(con_scores))
+)
+assertthat::assert_that(
+  length(ml_scores) == length(con_scores),
+  msg = "Length of ML and consensus scores don't match"
+)
+
+best_index <- which(combined_scores == max(combined_scores))
+best_ml_score <- ml_scores[best_index]
+best_con_score <- con_scores[best_index]
+
+assertthat::assert_that(
+  best_ml_score < best_con_score,
+  msg = "Best ML score not less than best consensus score"
+)
+
+# check model selected
+sanger_models <- sanger_log_files %>%
+  map(read_lines) %>%
+  map(extract_iqtree_mod) %>%
+  map(unique) %>%
+  rename_sanger()
+
+assertthat::assert_that(
+  length(unique(sanger_models)) == 1,
+  msg = "More than one sanger model selected"
+)
+assertthat::assert_that(
+  extract_iqtree_mod(plastome_iqtree_log) == unique(sanger_models),
+  msg = "Model selected for plastome and sanger don't match"
+)
+
+# check number of runs converged
+no_converge <- sanger_ml_log_rep[
+  str_detect(sanger_ml_log_rep, "bootstrap analysis did not converge")] %>%
+  rename_sanger()
+n_converge <- length(sanger_log_files) - length(no_converge) %>%
+  english::english()
+
+assertthat::assert_that(
+  names(best_index) %in% names(no_converge),
+  msg = "Best scoring tree *did* converge")
+```
+
 ```{r phylogeny-stats}
@@
 The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger analyses.
-Two of the ten runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, from which the final tree was selected, did not converge (correlation coefficient 0.977 after 1,000 iterations).
-The consensus tree had higher log-likelihood (-1183776.726) than the ML tree (-1183911.595), so we only present the topology and divergence times estimated from the consensus tree.
+`r str_to_sentence(n_converge)` of the ten runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, from which the final tree was selected, did not converge (correlation coefficient 0.976 after `r number(1000)` iterations).
+The consensus tree had higher log-likelihood (`r best_con_score`) than the ML tree (`r best_ml_score`), so we only present the topology and divergence times estimated from the consensus tree.
 

### 2022-06-21 ms/manuscript.Rmd
Add histogram of sequence lengths

@@
 There is no formal definition of genomic vs.
-Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term.
+Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term (**`r s_figure("seqlen")`**.).
 

### 2022-06-21 ms/si.Rmd
Add histogram of sequence lengths

@@
 
+```{r seqlen, fig.width = 7, fig.height = 4, eval = params$doc_type != "manual"}
+fern_slen_tib <- read_from_cache(
+  "fern_slen_tib", here::here("_supp_store")) %>%
+  # Add column for plastome or not
+  mutate(
+    plastome = if_else(
+      str_detect(def, "complete genome"),
+      "A. 'complete genome'",
+      "B. other"
+    ),
+    plastome = factor(plastome, levels = c("A. 'complete genome'", "B. other"))
+  )
+
+ggplot(fern_slen_tib, aes(slen)) +
+  geom_vline(xintercept = 7000, color = "red") +
+  geom_histogram(bins = 100) +
+  scale_x_log10() +
+  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
+  labs(x = "Sequence length (bp)", y = "Count (num. accessions)") +
+  facet_wrap(vars(plastome))
+```
+
+**`r s_figure("seqlen")`**.
+Histogram of sequence lengths (bp; log-scale) of fern accessions in GenBank release v249.
+Fern accessions obtained by querying GenBank for “Polypodiopsida[ORGN]” with length 10--200,000 bp.
+**A.** Sequences with the term 'complete genome' in their description (`r eg`, "Angiopteris evecta chloroplast, complete genome").
+**B.** Other sequences (those without 'complete genome' in their description).
+Vertical red line at 7,000 bp indicates empirical length cutoff used to differentiate between between Sanger (≤ 7,000 bp) and plastome (> 7,000 bp) accessions.
+Note that "other" includes some whole plastome accessions that cannot be distinguished based on description alone (e.g., "UNVERIFIED: Nephrolepis biserrata plastid sequence"); these are mostly above 7,000 bp.
+
+\newpage
+
 ```{r fossil-map-make}

### 2022-06-21 ms/si.Rmd
Fix some typos in fossil data

@@
   ) %>%
+  # Fix some spelling errors, missing data
+  mutate(
+    formation = case_when(
+      formation == "Dominicam amber deposti" ~ "Dominican amber deposit",
+      TRUE ~ formation
+    ),
+    type_locality =  case_when(
+      n_fos == 33 ~ "Burdigalian amber bearing strata of Dominican Republic",
+      TRUE ~ type_locality
+    )
+  ) %>%
   # Format references

### 2022-06-21 ms/si2.Rmd
Fix some typos in fossil data

@@
   fossil_ferns_raw %>%
+  # Fix some spelling errors, missing data
+  mutate(
+    formation = case_when(
+      formation == "Dominicam amber deposti" ~ "Dominican amber deposit",
+      TRUE ~ formation
+    ),
+    type_locality =  case_when(
+      n_fos == 33 ~ "Burdigalian amber bearing strata of Dominican Republic",
+      TRUE ~ type_locality
+    )
+  ) %>%
   # Filter to only those actually used in the study

### 2022-06-21 ms/si.Rmd
Add rogues table

@@
 ts_fossil_calibration_tips <- read_from_cache("ts_fossil_calibration_tips", ftol_cache) # nolint
+sanger_seqs_rogues_inspected <- read_from_cache(
+  "sanger_seqs_rogues_inspected", ftol_cache
+)
 
@@
 For detailed notes and full references for each fossil, see Appendix S2.
+
+```{r rogues-make}
+rogues_table <-
+  sanger_seqs_rogues_inspected %>%
+  # FIXME: manually correct one species
+  # should delete this after correcting
+  # https://github.com/fernphy/ftol/issues/16
+  filter(species != "Dryopolystichum_phaeostigma") %>%
+  arrange(species, target) %>%
+  rename(locus = target)
+```
+
+```{r rogues-save, eval = params$doc_type == "doc"}
+# Save if rendering to doc
+write_csv(rogues_table, result_file(s_table("rogues"), "csv"))
+```
+
+**`r s_table("rogues")`**.
+GenBank accessions identified as rogues by all-by-all BLAST.
+"q_family" is the family of the GenBank query (accession).
+"s_family" is the family of the top three best matches (only cases where the top three matches all belong to the same family included).
+All accessions in this table were excluded from further analysis.
 

### 2022-06-21 ms/manuscript.Rmd
Add rogues table
Update MS

--- 2022-06-21 ms/manuscript.Rmd
@@
 
-The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server and a local database including only ferns and outgroups (selected seed plants, lycophytes, bryophytes, and algae) created using the restez R package [@Bennett2018a].
+The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server (https://ftp.ncbi.nlm.nih.gov/genbank/) and a local database including only ferns and outgroups (selected seed plants, lycophytes, bryophytes, and algae) created using the restez R package [@Bennett2018a].
 GenBank accession numbers corresponding to sequences potentially matching each locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cuteoff matching the current GenBank release.
@@
 There is no formal definition of genomic vs.
-Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term.
+Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (≤ `r number(7000)` bp) and plastome (> `r number(7000)` bp) accessions with the "[SLEN]" term (`r s_figure("seqlen")`).
 
@@
 Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
-We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, , that query was considered a rogue and excluded from further analysis.
+We compared the families (following PPG I) of the matching sequences to each query; in the case the top three best matches belonged to a different family than the query, that query was considered a rogue and excluded from further analysis.
 For Cyatheales and Saccolomatineae, which include several closely related small families, we used the order and sub-order level, respectively, instead of family to avoid false positives.
 Species belonging to monotypic families were not considered for this filtering.
-While this method cannot account for rogues at finer taxonomic levels, it is an efficient approach for removing obviously erroneous sequences.
+While this method cannot account for rogues at finer taxonomic levels, it is an efficient approach for removing clearly erroneous sequences.
 We inspected all accessions flagged this way as rogues before excluding them.
@@
 After excluding these and accessions with names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
-After further excluding rogues identified by the all-by-all BLAST search, `r n_accs_second_filter` accessions were retained.
+After further excluding rogues identified by the all-by-all BLAST search `r s_table("rogues")`, `r n_accs_second_filter` accessions were retained.
 The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) after excluding species in the plastome dataset included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species.
@@
 
-Taxon sampling for FTOL v1.0.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in Pteridocat v1.0.0).
+Taxon sampling for FTOL v1.1.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in pteridocat v0.1.0).
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
@@
 `r str_to_sentence(n_converge)` of the ten runs converged (correlation coefficient of split occurrence frequencies >0.99), but the run with the highest log-likelihood, from which the final tree was selected, did not converge (correlation coefficient 0.976 after `r number(1000)` iterations).
-The consensus tree had higher log-likelihood (`r best_con_score`) than the ML tree (`r best_ml_score`), so we only present the topology and divergence times estimated from the consensus tree.
+The consensus tree had higher log-likelihood (`r number(best_con_score)`) than the ML tree (`r number(best_ml_score)`), so we only present the topology and divergence times estimated from the consensus tree.
 
@@
 The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
-Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales. 
+Salviniales was recovered as sister to Cyatheales, which are in turn sister to Polypodiales. 
 The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
@@
 Ages of other major crown groups are: leptosporangiates (`r crown_ages$leptos` Ma), Polypodiales (`r crown_ages$polypodiales` Ma), eupolypods I (`r crown_ages$eupoly_i` Ma), and eupolypods II (`r crown_ages$eupoly_ii` Ma).
-Our estimates for leptosporangiates and Polypodiales are both `r ca` 50 Ma older than the most recent global fern phylogeny [@Testo2016a], while other ages are similar.
+Our estimates for leptosporangiates and Polypodiales are both `r ca` 30--40 Ma older than the most recent global fern phylogeny [@Testo2016a], while other ages are similar (eupolypods I) or slightly younger (eupolypods II).
 Estimated stem ages of fern families were mostly older than previous studies (`r figure("divtimes")`).
-We did not compare crown ages of families across studies because differences in crown ages in smaller clades are likely affected by species sampling.
+We did not compare crown ages of families across studies because differences in crown ages of smaller clades are likely affected by species sampling.
 
@@
 However, there were a few inconsistencies in the application of these fossils due to differences in taxonomic concepts between the original fossil publication and the studies in which they were used.
-More importantly, our set of calibration points more than doubles the number used by previous studies and it is likely that this expanded dataset is responsible for the older stem age estimates for many families (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
+More importantly, our set of calibration points approximately doubles the number used by previous studies and it is likely that this expanded dataset is responsible for the older stem age estimates for many families (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
 To test this hypothesis, we conducted an additional analysis using the fossil constraints of @Testo2016a but otherwise the same methods (FTOL analyzed with treePL).
@@
 For example, we integrate our phylogeny with a custom taxonomy for ferns (pteridocat), tailor alignment strategies to locus and taxon (fern-wide alignment for coding loci, nested alignment within families for non-coding loci), inspect intermediate results and modify input based on expert taxonomic knowledge (curation of sample inclusion and exclusion lists), and employ a custom set of criteria for selecting and concatenating sequences within species.
-While we have not sought to make our workflow available as a general tool due to its high degree of specialization, some of the methods employed here could be used in other studies, such as use of a backbone phylogeny based on next-generation data to constrain a tree built with Sanger data.
+While we have not sought to make our workflow available as a general tool due to its high degree of specialization, some of the methods employed here could be adapted and implemented in other phylogeny pipelines.
 
@@
 Indeed, we anticipate that there will almost always be tips in the tree that need correction, either because they get overlooked (`r eg`, placement of species within genera, which we did not have the resources to inspect) or because an updated taxonomic treatment is not yet available (`r eg`, cheilanthoid ferns).
-One example of progressive refinement is Thelypteridaceae, which had the highest rate of non-monophyletic genera in the previous version of FTOL (v1.0.0).
-After implementing a manual inclusion list for Thelypteridaceae and consulting with a taxonomic expert on this family (Susan Fawcett), the number of non-monophyletic genera dropped to __ in v1.0.1.
+One example of progressive refinement is Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in the previous version of FTOL (v1.0.0).
+After implementing a manual inclusion list for Thelypteridaceae and consulting with a taxonomic expert on this family (Susan Fawcett), the number of non-monophyletic genera dropped to `r top_nonmono_subfam %>% filter(family == "Thelypteridaceae") %>% pull(n)`.
 While manual inclusion lists are not an ideal long-term solution because they cannot grow with GenBank, this case demonstrates that such lists are a reasonable option for taxonomically difficult groups, and more importantly, the improvement that results from input from taxonomic experts.
@@
 **`r figure("genbank")`**.
-Number of fern species in GenBank by year and genomic compartment.
+Number of fern species in GenBank by year and genomic compartment, 1990--2021.
 Points indicate number of species sampled in selected studies of global fern phylogeny [@Hasebe1995; @Schuettpelz2007; @Lehtonen2011; @Testo2016a\; this study].

### 2022-06-22 ms/manuscript.Rmd
Update MS
- add SI fig for t&s calibration points
- fix things according to input from ES, SRB

@@
 
-Ferns (ca.
-`r number(12000)` species) are the second most diverse clade of vascular plants after angiosperms (ca.
-`r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [e.g., @Hennequin2014; @Lehtonen2015], and speciation [e.g., @Kao2020].
+Ferns (`r ca` `r number(12000)` species) are the second most diverse clade of vascular plants after angiosperms (`r ca` `r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [`r eg`, @Hennequin2014; @Lehtonen2015], and speciation [`r eg`, @Kao2020].
 Key to any investigation of evolutionary history in this group is a well-sampled phylogeny.
@@
 Past efforts to construct such a global phylogeny have steadily expanded their sampling, at first by mostly generating new sequences, then later by mining GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, growth of of plastid fern accessions in GenBank show no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
+Indeed, growth of plastid fern accessions in GenBank show no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
 There is a need therefore, not only for a revised global fern phylogeny, but also one that is continuously updated to keep pace with the rapid accumulation of molecular data going forward.
@@
 
-The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server (https://ftp.ncbi.nlm.nih.gov/genbank/) and a local database including only ferns and outgroups (selected seed plants, lycophytes, bryophytes, and algae) created using the restez R package [@Bennett2018a].
-GenBank accession numbers corresponding to sequences potentially matching each locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cuteoff matching the current GenBank release.
-Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
-Sequences were then pulled from the local database in FASTA format by GenBank accession number.
+The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server (https://ftp.ncbi.nlm.nih.gov/genbank/).
+A local database was then created from these data including only ferns and outgroup taxa (select seed plants, lycophytes, bryophytes, and algae) using the restez R package [@Bennett2018a].
+GenBank accession numbers corresponding to sequences potentially matching each target locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cutoff matching the current GenBank release.
+Environmental DNA samples and accessions with names including the terms "aff." or "cf." or hybrid formulas were excluded (no attempt was made to exclude hybrid taxa with standard names, `r ie`, names that are not formulas).
+Sequences were then obtained from the local database in FASTA format by GenBank accession number.
 There is no formal definition of genomic vs.
@@
 
-In addition to this automated approach, we also implemented a manual approach for Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in a previous version of FTOL (v1.0.0) and likely has high rates of misidentification on GenBank.
-For the manual approach, we used a curated list of accessions from a previous phylogenetic analysis of this family [@Patel2019].
-After resolving names in the list of @Patel2019 to pteridocat and correcting some errors (wrong GenBank accession number listed in small number of cases), accessions in this list were used preferentially over those identified by the automated approach.
+We also implemented a manual approach for Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in a previous version of FTOL (v1.0.0) and likely has high rates of misidentification on GenBank.
+This approach used a curated list of accessions from a previous phylogenetic analysis of this family [@Patel2019].
+After resolving the names in @Patel2019 to pteridocat and correcting some errors (incorrect GenBank accession number in a small number of cases), accessions in this list were used preferentially over those identified by the automated approach.
 
@@
   pull(smooth) %>%
-  number(., accuracy = .)
+  formatC(format = "e", digits = 0)
 
@@
   pull(smooth) %>%
-  number(., accuracy = .)
-```
-
-Dating was conducted separately on the ML tree and the consensus tree using penalized likelihood with treePL [@Smith2012].
-We initially rooted the tree on *Zygnema circumcarinatum* Czurda, a member of the group of algae (Zygnematophyceae) thought to be most closely related to land plants [@Donoghue2021].
+  formatC(format = "e", digits = 0)
+```
+
+Dating was conducted separately on the ML tree and the consensus tree using penalized likelihood as implemented in treePL [@Smith2012].
+We initially rooted the tree using *Zygnema circumcarinatum* Czurda, a member of the group of algae (Zygnematophyceae) thought to be most closely related to land plants [@Donoghue2021].
 Since there is no way to objectively divide branch length between the branch leading to the outgroup and the rest of the tree, we then trimmed *Z.
-circumcarinatum* from the phylogeny; this effectively positioned bryophytes as the outgroup on a branch or branches whose length has been estimated from the data [@Sauquet2013].
+circumcarinatum* from the phylogeny; this effectively positioned bryophytes as the outgroup on a branch or branches with length(s) estimated from the data [@Sauquet2013].
 
@@
 
-We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude).
+We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); these values are based on consideration that our tree spans a wide phylogenetic range with high rate heterogeneity, and results of initial analyses in which higher smoothing values sometimes produced spurious date estimates.
 A value of `r smooth_best` was selected, based on the smallest chi-squared value, for the final analysis.
@@
 After excluding these and accessions with names that could not be resolved to an accepted name in the taxonomic database, `r n_accs_first_filter` accessions were retained.
-After further excluding rogues identified by the all-by-all BLAST search `r s_table("rogues")`, `r n_accs_second_filter` accessions were retained.
+After further excluding rogues identified by the all-by-all BLAST search (`r s_table("rogues")`), `r n_accs_second_filter` accessions were retained.
 The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation were not met) after excluding species in the plastome dataset included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately; `r s_table("accs")`) representing `r n_species_sanger` species.
@@
 Another possibility supported by recent transcriptomic studies is Hymenophyllaceae sister to Gleicheniaceae, which are in turn sister to Dipteridaceae, resulting in a paraphyletic Gleicheniales [@Qi2018; @Shen2018; @Shu2022].
-Here, we recovered a monophyletic Gleicheniales (including Gleicheniaceae, Dipteridaceae, and Matoniaceae) sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
+Here, we recovered a monophyletic Gleicheniales sister to Hymenophyllales with moderate support, a relationship that was also observed in some, but not all, analyses of plastome data by @Lehtonen2019 and @Kuo2018.
 However, our plastome sampling only includes `r plastome_gleich_genera_n %>% jntools::english2()` out of `r ppgi_gleich_genera_n` genera of Gleicheniales, and lacks Matoniaceae.
 While our study was in revision, 42 additional fern plastomes were published, including seven families (`r ie`, Anemiaceae, Culcitaceae, Dipteridaceae, Loxsomataceae, Matoniaceae, Metaxyaceae, and Thyrsopteridaceae) that are not included in our sampling since they are not in the most recent available GenBank release [release 249\; @Du2022].
-The addition of Matoniaceae results in significantly different relationships between Hymenophyllales and Gleicheniales; @Du2022 recovered Matoniaceae sister to Dipteridaceae, and these together sister to the remaining leptosporangiates.
-Futhermore, @Du2022 did not recover a sister relationship between Hymenophyllaceae and Gleicheniaceae, nor monophyly of Gleicheniales.
-Interestingly, addition of Matoniaceae transcriptome data also supported the sister status of Matoniaceae + Dipteridaceae, as well as the non-monophyly of Gleicheniales in another recent study [@Shu2022].
+The addition of Matoniaceae results in non-monophyly of Gleicheniales, with Matoniaceae sister to Dipteridaceae, and these together sister to the remaining leptosporangiates [@Du2022].
+Furthermore, @Du2022 did not recover a sister relationship between Hymenophyllaceae and Gleicheniaceae.
+Interestingly, addition of Matoniaceae transcriptome data also supported the non-monophyly of Gleicheniales and the sister status of Matoniaceae + Dipteridaceae in another recent study [@Shu2022].
 
@@
 More importantly, our set of calibration points approximately doubles the number used by previous studies and it is likely that this expanded dataset is responsible for the older stem age estimates for many families (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
-To test this hypothesis, we conducted an additional analysis using the fossil constraints of @Testo2016a but otherwise the same methods (FTOL analyzed with treePL).
+To test this hypothesis, we conducted an additional analysis using the fossil constraints of @Testo2016a but otherwise the same methods (FTOL analyzed with treePL; `r s_figure("ts_fossil_map")`).
 The resulting stem family ages show a much closer agreement with those of @Testo2016a (`r r2` = `r ftol_ts_comp_stats$r.squared %>% number(accuracy = 0.01)`, `r pval` = `r ftol_ts_comp_stats$p.value %>% sci_number()`, linear model; Figures `r s_figure_num("comp-stem-ages")`, `r s_figure_num("comp-stem-points")`), indicating that our expanded set of fossil calibration points, not differences in topology or dating methodology, is the primarily contributor to the older ages observed in the current study.
@@
 Notably, the scenario of fern diversification suggested by our results somewhat conflicts with the hypothesis that Polypodiales diversified "in the shadow of angiosperms" [@Schneider2004d; @Schuettpelz2009].
-Rather, we estimate that much of polypod diversification coincides with or even precedes the diversification and rise to ecological dominance of angiosperms during the Late Cretaceous [@Benton2022] (`r s_figure("ltt")`).
+Rather, we estimate that the origin (`r ie`, stem age) of many polypod families coincides with or even precedes the diversification and rise to ecological dominance of angiosperms during the Late Cretaceous [@Benton2022] (`r s_figure("ltt")`).
 Our estimated age of (`r crown_ages$polypodiales` Ma) for crown Polypodiales is considerably older than other recent studies [@Testo2016a; @Du2021] and the fossil record, which only dates back to the Early Cretaceous [@Schneider2001; @Chen1997; @Deng2002; @Schneider2016c; @Regalado2018a].
@@
 Recent transcriptomic studies are gradually clarifying the backbone of the fern phylogeny using many (25--`r number(2400)`) nuclear genes from representative species spanning the tree [@Rothfels2015a; @Qi2018; @Shen2018].
-The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFlag) project, which seeks to generate genomic data (ca.
-300 single-to-low copy nuclear gene regions) for all flagellate plants [bryophytes, lycophytes, ferns, and gymnosperms\; @Breinholt2021].
+The most comprehensively sampled phylogenomic study targeting ferns is the on-going Genealogy of Flagellate Plants (GoFlag) project, which seeks to generate genomic data (`r ca` `r number(300)` single-to-low copy nuclear gene regions) for all flagellate plants [bryophytes, lycophytes, ferns, and gymnosperms\; @Breinholt2021].
 GoFlag data have recently been used in a phylogenomic analysis and taxonomic revision of Thelypteridaceae resulting in the recognition of multiple new genera [@Fawcett2021; @Fawcett2021a], and additional phylogenomic analyses of other fern groups using GoFlag markers are to be expected in the near future.
@@
 Ferns are not unique in having large amounts of publicly available DNA sequence data, and several pipeline tools exist that can leverage such data to automatically generate phylogenies [@Antonelli2016; @Bennett2018; @Drori2018a; @Pearse2013; @Portik2020; @Smith2019a; @Xu2015].
-In particular, superSMART [@Antonelli2016] and pyPHLAWD [@Smith2019a] are recently developed pipelines that can generate maximally sampled phylogenies for any higher taxon of choice; the latter was used to generate a broadly inclusive seed plant phylogeny [@Smith2018].
-
-However, although we make use of some of the functionality of these tools [`r eg`, restez for generating a local copy of GenBank\; superCRUNCH for identifying orthologous sequences\; @Bennett2018a; @Portik2020], our pipeline is mostly custom-built for ferns.
-We chose this approach because it allowed us to implement various steps specific to ferns that would not be possible using a fully automated, taxon-agnostic pipeline like superSMART or pyPHLAWD, which we believe ultimately results in a higher-quality fern phylogeny.
+In particular, superSMART [@Antonelli2016] and pyPHLAWD [@Smith2019a] are recently developed pipelines that can generate maximally sampled phylogenies for any higher taxon of choice.
+The latter was used to generate a broadly inclusive seed plant phylogeny [@Smith2018].
+
+Although we make use of some of the functionality of these tools [`r eg`, restez for generating a local copy of GenBank\; superCRUNCH for identifying orthologous sequences\; @Bennett2018a; @Portik2020], our pipeline is mostly custom-built for ferns.
+We chose our approach because it allowed us to implement various steps specific to ferns that would not be possible using a fully automated, taxon-agnostic pipeline like superSMART or pyPHLAWD, which we believe ultimately results in a higher-quality fern phylogeny.
 For example, we integrate our phylogeny with a custom taxonomy for ferns (pteridocat), tailor alignment strategies to locus and taxon (fern-wide alignment for coding loci, nested alignment within families for non-coding loci), inspect intermediate results and modify input based on expert taxonomic knowledge (curation of sample inclusion and exclusion lists), and employ a custom set of criteria for selecting and concatenating sequences within species.
@@
 One example of progressive refinement is Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in the previous version of FTOL (v1.0.0).
-After implementing a manual inclusion list for Thelypteridaceae and consulting with a taxonomic expert on this family (Susan Fawcett), the number of non-monophyletic genera dropped to `r top_nonmono_subfam %>% filter(family == "Thelypteridaceae") %>% pull(n)`.
-While manual inclusion lists are not an ideal long-term solution because they cannot grow with GenBank, this case demonstrates that such lists are a reasonable option for taxonomically difficult groups, and more importantly, the improvement that results from input from taxonomic experts.
+After implementing a manual inclusion list for Thelypteridaceae and consulting with a taxonomic expert on this family (S.
+Fawcett), the number of non-monophyletic genera was reduced to `r top_nonmono_subfam %>% filter(family == "Thelypteridaceae") %>% pull(n)`.
+While manual inclusion lists are not an ideal long-term solution because they cannot grow with GenBank, this case demonstrates that such lists are a reasonable option for taxonomically difficult groups, and more importantly, the improvement that can result from input from taxonomic experts.
 Another example of the advantage of a continuously updated tree is the publication of 42 additional fern plastome samples while this paper was in revision [@Du2022].
-While we were unable to integrate these data into the current version of FTOL, they will be automatically added when the next version of GenBank data are released, and we expected the next version of FTOL will reflect this new knowledge.
+While we were unable to integrate these data into the current version of FTOL, they will be automatically added when the next version of GenBank data is released, and we expected the next version of FTOL will reflect this new knowledge.
 
@@
 Members of the Iwasaki lab provided comments that improved the manuscript.
-Alexander White provided helpful comments on an early version of the analysis.
-Susan Fawcett provided helpful comments on taxonomy of Thelypteridaceae.
+A.E. White provided helpful comments on an early version of the analysis.
+S. Fawcett provided helpful comments on taxonomy of Thelypteridaceae.
 This study was supported by JSPS KAKENHI Grant Number 16H06279 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
-The authors thank Michael Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.
+The authors thank M. Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.
 
@@
 Points indicate number of species sampled in selected studies of global fern phylogeny [@Hasebe1995; @Schuettpelz2007; @Lehtonen2011; @Testo2016a\; this study].
+This study plotted as 2021 but includes data until 2022-04-15.
 @Schuettpelz2007 did not attempt exhaustive sampling but rather proportional sampling according to lineage size.
@@
     "Polypodiales" = c("Polypodiaceae", "Saccolomataceae"),
-    "Polypodiidae\n(leptosporan-\ngiates)" = c("Polypodiaceae", "Osmundaceae")
+    "Polypodiidae\n(leptosporangiates)" = c("Polypodiaceae", "Osmundaceae")
   ) %>%
@@
 # bootstrap values onto the ggtree object
+bump <- 5 # for easier fiddling of plot size
 backbone_plot <-
   ggtree(phy) %<+% bs_tibble  +
+  theme(legend.position = "none") +
   # now we can show BS values using geom_text()
@@
     align = "right",
-    extend = 30,
+    extend = 36 + bump,
     gradient.direction = "rt",
@@
   scale_fill_manual(values = clade_cols) +
+  xlim(0, 190 + bump) +
   # label major clades
@@
     ),
-    offset = 28,
+    offset = 34 + bump,
     align = TRUE,
@@
     data = other_clade_labels %>%
-      filter(clade_label == "Polypodiidae\n(leptosporan-\ngiates)"),
+      filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
     aes(x = x, y = y, label = clade_label),
@@
   ) +
-  xlim(0, 180) +
-  coord_cartesian(clip = "off") +
-  theme(legend.position = "none")
+  coord_cartesian(clip = "off")
 ```
@@
 Ultrafast bootstrap support values (%) shown at nodes; unlabeled nodes are 100%.
-Outgroup (seed plants, lycophytes, and bryophytes) not shown.
+Outgroup (seed plants, lycophytes, bryophytes, and algae) not shown.
 Colors of major clades (orders or suborders) correspond to those used in `r figure("ftol")`. Taxonomy follows @PteridophytePhylogenyGroupI2016; informal clade names in lowercase.

### 2022-06-22 ms/si.Rmd
Update MS
- add SI fig for t&s calibration points
- fix things according to input from ES, SRB

@@
 sanger_con_tree_rooted <- read_from_cache("sanger_con_tree_rooted", ftol_cache)
+sanger_con_tree_rooted_pruned <- read_from_cache(
+  "sanger_con_tree_rooted_pruned", ftol_cache)
 sanger_alignment_tbl <- read_from_cache("sanger_alignment_tbl", ftol_cache)
@@
 )
+ts_fossil_calibrations_for_treepl <- read_from_cache(
+  "ts_fossil_calibrations_for_treepl", ftol_cache)
 
@@
     fossil_label = str_remove_all(min, "min = "),
-    node = map2_dbl(taxon_1, taxon_2, ~ape::getMRCA(sanger_con_tree_rooted, c(.x, .y))),
+    node = map2_dbl(
+      taxon_1, taxon_2,
+      ~ape::getMRCA(sanger_con_tree_rooted_pruned, c(.x, .y))),
     calibration_type = case_when(
@@
   ggtree(
-    sanger_con_tree_rooted,
+    sanger_con_tree_rooted_pruned,
     branch.length = "none", size = 0.1, color = "grey30") %<+% node_labs_tbl +
@@
   # Add node labels
-  geom_text(aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+  geom_text(
+    aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
   # Add tip labels
@@
 
-```{r backbone-full-plot}
-phy <- plastome_tree_rooted
-
-# Make tibble for printing BS values
-bs_tibble <- make_bs_tbl(phy) %>%
-  # only print values < 100
-  mutate(bs = case_when(
-    bs == 100 ~ "",
-    TRUE ~ as.character(bs)
-  ))
-
-# Make plot
-bb_full_plot <-
-  ggtree(phy, size = 0.2, color = "grey10") %<+% 
-  bs_tibble +
-  # now we can show BS values using geom_text()
-  geom_text(
-    aes(label = bs), hjust = -.25, size = 1.0,
-    color = "grey50") +
-  # add tip labels
-  geom_tiplab(aes(label = label), size = 1.0)
-```
-
-```{r backbone-full-save, eval = params$doc_type == "doc"}
-ggsave(
-  plot = bb_full_plot,
-  file = (result_file(s_figure("backbone_full"), "pdf")),
-  height = 20, width = 8, units = "in",
-  limitsize = FALSE)
-```
-
-**`r s_figure("backbone_full")`**.
-Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
-Values at nodes are bootstrap support (%); only values less than 100% are shown.
-
-\newpage
-
 ```{r bs-plot}
@@
 The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
+
+\newpage
+
+```{r backbone-full-plot}
+phy <- plastome_tree_rooted
+
+# Make tibble for printing BS values
+bs_tibble <- make_bs_tbl(phy) %>%
+  # only print values < 100
+  mutate(bs = case_when(
+    bs == 100 ~ "",
+    TRUE ~ as.character(bs)
+  ))
+
+# Make plot
+bb_full_plot <-
+  ggtree(phy, size = 0.2, color = "grey10") %<+% 
+  bs_tibble +
+  # now we can show BS values using geom_text()
+  geom_text(
+    aes(label = bs), hjust = -.25, size = 1.0,
+    color = "grey50") +
+  # add tip labels
+  geom_tiplab(aes(label = label), size = 1.0)
+```
+
+```{r backbone-full-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = bb_full_plot,
+  file = (result_file(s_figure("backbone_full"), "pdf")),
+  height = 20, width = 10, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("backbone_full")`**.
+Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
+Values at nodes are bootstrap support (%); only values less than 100% are shown.
 
@@
 
+```{r ts-fossil-map-make}
+# Make tibble with node number and label for ggtree
+ts_node_labs_tbl <-
+  ts_fossil_calibrations_for_treepl %>%
+  transmute(
+    fossil_label = str_remove_all(min, "min = "),
+    node = map2_dbl(
+      taxon_1, taxon_2,
+      ~ape::getMRCA(sanger_con_tree_rooted_pruned, c(.x, .y))),
+    calibration_type = case_when(
+      !is.na(min) & is.na(max) ~ "minimum",
+      !is.na(min) & !is.na(max) ~ "fixed",
+      TRUE ~ NA_character_
+    )
+  ) %>%
+  assert(not_na, calibration_type)
+
+# Make plot
+fossil_map_plot <-
+  ggtree(
+    sanger_con_tree_rooted_pruned,
+    branch.length = "none",
+    size = 0.1, color = "grey30") %<+% ts_node_labs_tbl +
+  # Add node points
+  geom_nodepoint(
+    aes(color = calibration_type), size = 0.6) +
+  scale_color_manual(
+    "Calibration type",
+    values = c(
+      minimum = palette_okabe_ito("vermillion"),
+      fixed = palette_okabe_ito("skyblue")),
+    na.value = "transparent"
+  ) +
+  # Add node labels
+  geom_text(
+    aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+  # Add tip labels
+  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  theme(legend.position = "bottom")
+```
+
+```{r ts-fossil-map-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = fossil_map_plot,
+  file = (result_file(s_figure("ts_fossil_map"), "pdf")),
+  height = 60, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("ts_fossil_map")`**.
+Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints of @Testo2016a.
+Number for each constraint is age in Ma.
+
+\newpage
+
 ```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}

### 2022-06-23 ms/si.Rmd
Update SI figures
- fix spacing in trees
- don't plot Zygnema

@@
 
-```{r seqlen, fig.width = 7, fig.height = 4, eval = params$doc_type != "manual"}
+```{r seqlen-plot, fig.width = 7, fig.height = 4, eval = params$doc_type != "manual"}
 fern_slen_tib <- read_from_cache(
@@
 
-```{r fossil-map-make}
+```{r fossil-map-plot-make}
 # Make tibble with node number and label for ggtree
@@
   # Add tip labels
-  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  geom_tiplab(
+    aes(label = label), size = 0.5, color = "grey30", offset = -0.25) +
   theme(legend.position = "bottom")
@@
 
-```{r fossil-map-save, eval = params$doc_type == "doc"}
+```{r fossil-map-plot-save, eval = params$doc_type == "doc"}
 ggsave(
@@
   file = (result_file(s_figure("fossil_map"), "pdf")),
-  height = 60, width = 8, units = "in",
+  height = 70, width = 8, units = "in",
   limitsize = FALSE)
@@
 Number for each constraint is age in Ma.
-
-\newpage
-
-```{r locus-upset, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
+Outgroup taxon *Zygnema circumcarinatum* (algae) pruned prior to dating analysis and not shown.
+
+\newpage
+
+```{r locus-upset-plot, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
 # Make tibble with one row per locus per species
@@
 
-```{r bs-plot}
-phy <- sanger_con_tree_rooted
+```{r bs-plot-make}
+phy <- sanger_con_tree_rooted_pruned
 
@@
     phy,
-    size = 0.1, color = "grey30") %<+% bs_tibble +
+    size = 0.07, color = "grey30") %<+% bs_tibble +
   # Add node points
   geom_nodepoint(
-    aes(color = bs), size = 0.1) +
+    aes(
+      color = bs), size = 0.1, alpha = 0.9,
+      shape = 16, fill = "transparent") +
   scale_color_viridis_c(
@@
   # Add tip labels
-  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  geom_tiplab(
+    aes(label = label), size = 0.5, color = "grey30", offset = -0.005) +
   theme(legend.position = "bottom")
@@
 
-```{r bs-save, eval = params$doc_type == "doc"}
+```{r bs-plot-save, eval = params$doc_type == "doc"}
 ggsave(
@@
   file = result_file(s_figure("bs"), "pdf"),
-  height = 60, width = 8, units = "in",
+  height = 70, width = 8, units = "in",
   limitsize = FALSE)
@@
 The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
-
-\newpage
-
-```{r backbone-full-plot}
-phy <- plastome_tree_rooted
+Outgroup taxon *Zygnema circumcarinatum* (algae) not shown.
+
+\newpage
+
+```{r backbone-full-plot-make}
+# Don't show Zygnema in plot
+phy <- ape::drop.tip(plastome_tree_rooted, "Zygnema_circumcarinatum")
 
@@
   # add tip labels
-  geom_tiplab(aes(label = label), size = 1.0)
-```
-
-```{r backbone-full-save, eval = params$doc_type == "doc"}
+  geom_tiplab(aes(label = label), size = 1.0, offset = -0.005)
+```
+
+```{r backbone-full-plot-save, eval = params$doc_type == "doc"}
 ggsave(
@@
 Values at nodes are bootstrap support (%); only values less than 100% are shown.
-
-\newpage
-
-```{r du-ages, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
+Outgroup taxon *Zygnema circumcarinatum* (algae) not shown.
+
+\newpage
+
+```{r du-ages-plot, fig.width = 7, fig.height = 5, eval = params$doc_type != "manual"}
 # Helper function for subsetting dates to get into
@@
 
-```{r comp-fossils, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
+```{r comp-fossils-plot, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
 ts_fossil_calibration_tips %>%
@@
 **`r s_figure("comp-fossils")`**.
-Age of nodes constrained by fossils in this study vs. @Testo2016a. Wide bars indicate the node was only constrained in one study; narrow bars indicate the the node was constrained in both studies.
-
-\newpage
-
-```{r comp-stem-ages, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
+Age of nodes constrained by fossils in this study vs. @Testo2016a.
+Wide bars indicate the node was only constrained in one study; narrow bars indicate the the node was constrained in both studies.
+
+\newpage
+
+```{r comp-stem-ages-plot, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
 # Combine family stem ages into almost-long format:
@@
 
-```{r ts-fossil-map-make}
+```{r ts-fossil-map-plot-make}
 # Make tibble with node number and label for ggtree
@@
 # Make plot
-fossil_map_plot <-
+ts_fossil_map_plot <-
   ggtree(
@@
   geom_text(
-    aes(label = fossil_label), hjust = 0, size = 1.0, color = "grey20") +
+    aes(label = fossil_label),
+    hjust = 0, size = 1.0, color = "grey20") +
   # Add tip labels
-  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
+  geom_tiplab(aes(label = label),
+  size = 0.5, color = "grey30", offset = -0.25) +
+  xlim(0, 66.1) +
   theme(legend.position = "bottom")
@@
 
-```{r ts-fossil-map-save, eval = params$doc_type == "doc"}
+```{r ts-fossil-map-plot-save, eval = params$doc_type == "doc"}
 ggsave(
-  plot = fossil_map_plot,
+  plot = ts_fossil_map_plot,
   file = (result_file(s_figure("ts_fossil_map"), "pdf")),
-  height = 60, width = 8, units = "in",
+  height = 70, width = 8, units = "in",
   limitsize = FALSE)
@@
 Number for each constraint is age in Ma.
-
-\newpage
-
-```{r comp-stem-points, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}
+Outgroup taxon *Zygnema circumcarinatum* (algae) pruned prior to dating analysis and not shown.
+
+\newpage
+
+```{r comp-stem-points-plot, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}
 # For checking that model data is same as plot data
@@
 
-```{r ltt, fig.width = 7, fig.height = 6, eval = params$doc_type != "manual"}
+```{r ltt-plot, fig.width = 7, fig.height = 6, eval = params$doc_type != "manual"}
 # Make lineages through time (LTT) plot comparing Angiosperms and polypods

### 2022-06-24 R/process_refs.R
Update references

@@ -29,4 +29,4 @@ extract_citations("ms/manuscript.Rmd") %>%
   # where KEY is the citation key
   mutate(latex = glue::glue("\\citation{[key]}", .open = "[", .close = "]")) %>%
   pull(latex) %>%
-  write_lines("working/ms_keys.aux")
+  write_lines("working/ftol.aux")

### 2022-06-24 R/functions.R
Use seq_len

@@ -1199,10 +1199,9 @@ make_bs_tbl <- function(phy) {
   tibble(
     # hard-code node ID: internal nodes start after tip nodes,
     # and phy$node.label is in the same order as internal nodes
-    node = 1:Nnode(phy) + Ntip(phy), 
+    node = seq_len(Nnode(phy)) + Ntip(phy), 
     bs = bs_values)
 }
-
 
 # Etc -----
 

### 2022-06-24 ms/manuscript.Rmd
Update MS
- Add package versions
- Add grant number

@@
 ts_fossil_calibration_tips <- read_from_cache("ts_fossil_calibration_tips", ftol_cache) # nolint
+renv_pkg_versions <- read_from_cache("renv_pkg_versions", ftol_cache)
+sw_versions <- read_from_cache("sw_versions", ftol_cache)
+ftol_r_ver <- read_from_cache("ftol_r_ver", ftol_cache)
+pkg_versions <- bind_rows(renv_pkg_versions, sw_versions)
 
@@
 
-**Joel H. Nitta^1^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^1,4,5,6,7,8^**
+**Joel H. Nitta^1,9^, Eric Schuettpelz^2^, Santiago Ramírez-Barahona^3^, and Wataru Iwasaki^1,4,5,6,7,8^**
 
@@
 ^8^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan
+
+^9^Current affiliation: Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan
 
@@
 The "plants" division of GenBank release 249 (cutoff date 2022-04-15) was downloaded from the NCBI FTP server (https://ftp.ncbi.nlm.nih.gov/genbank/).
-A local database was then created from these data including only ferns and outgroup taxa (select seed plants, lycophytes, bryophytes, and algae) using the restez R package [@Bennett2018a].
+A local database was then created from these data including only ferns and outgroup taxa (select seed plants, lycophytes, bryophytes, and algae) using the restez R package v`r pkg_ver("restez", pkg_versions)` [@Bennett2018a].
 GenBank accession numbers corresponding to sequences potentially matching each target locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cutoff matching the current GenBank release.
@@
 There is a lack of consensus on sequence annotation in GenBank; the same locus may be annotated using different names, or not annotated at all.
-To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
+To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of superCRUNCH v`r pkg_ver("supercrunch", pkg_versions)` [@Portik2020] to extract target loci from GenBank FASTA files.
 Briefly, this involves querying candidate FASTA files from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
@@
 We then filtered the sequences to a single representative longest sequence per genus.
-Next, we aligned the filtered sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
+Next, we aligned the filtered sequences with MAFFT v`r pkg_ver("mafft", pkg_versions)` [@Katoh2002] and removed poorly aligned regions with trimAl v`r pkg_ver("trimal", pkg_versions)` [@Capella-Gutierrez2009].
 To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
@@
 Furthermore, the NCBI taxonomy is not curated specifically for ferns and includes many fern synonyms.
-Therefore, we standardized all species names in the GenBank sequences against a newly generated reference taxonomy.
-Our reference taxonomy is based on the World Ferns database v.12.8 [@Hassler2022], which conforms to PPG I (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
-
-To resolve species names, we used the taxastand R package [@Nitta2021], which can account for differences in formatting of taxonomic authors (`r eg`, parenthetical basionym author present or absent) and perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
+Therefore, we standardized all species names in the GenBank sequences (NCBI Taxonomy database dump release 2022-05-01, obtained from https://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxdump_archive/) against a newly generated reference taxonomy.
+Our reference taxonomy is based on the World Ferns database v12.8 [@Hassler2022], which conforms to PPG I (with some exceptions explained below) and is available in Darwin Core format [@DarwinCore2009] via the Catalog of Life [@COL2366].
+
+To resolve species names, we used the taxastand R package v`r pkg_ver("taxastand", pkg_versions)` [@joel_nitta_2021_5726391], which can account for differences in formatting of taxonomic authors (`r eg`, parenthetical basionym author present or absent) and perform fuzzy matching, which is needed to account for spelling errors or variations in author names.
 
@@
 This revealed some species names in GenBank that were missing in the World Ferns database, spelling errors, and some names in the database that needed to be treated differently (`r eg`, changes in synonymy).
-We thus updated and edited the initial World Ferns database using the dwctaxon R package [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
+We thus updated and edited the initial World Ferns database using the dwctaxon R package v1.0.0 [@joel_nitta_2022_6388272], which is designed to work with taxonomic data in the Darwin Core format.
 We refer to the resulting taxonomic database as "pteridocat," and have made it freely available online (https://github.com/fernphy/pteridocat) so that other researchers may standardize taxonomic names in their data to match those of FTOL [@ftol2022a].
@@
 Furthermore, multiple genera included in PPG I were known to be non-monophyletic and provisionally circumscribed [@PteridophytePhylogenyGroupI2016].
-Pteridocat includes newly published genera, some genera that were not recognized by PPG I, and nothogenera, which were not included in PPG I [@Liu2020a].
+pteridocat includes newly published genera, some genera that were not recognized by PPG I, and nothogenera, which were not included in PPG I [@Liu2020a].
 Differences between pteridocat and PPG I are summarized in `r s_table("genera_diff")`.
@@
 We removed putative rogues from Sanger sequences as follows.
-First, we constructed a BLAST library including all extracted Sanger sequences.
+First, we constructed a BLAST library including all extracted Sanger sequences using BLAST Suite v`r pkg_ver("blast", pkg_versions)`.
 Next, we conducted one BLAST search for each Sanger sequence against the library (all-by-all BLAST).
@@
 Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions may contain multiple loci, but we already split those out using superCRUNCH as described above).
-First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
+First, we constructed gene trees with FastTree v`r pkg_ver("fasttree", pkg_versions)` [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
 Then, we concatenated loci that met any of the following three conditions: 1) if a species was monophyletic, loci were concatenated by selecting the longest accession per locus within that species; 2) if all accessions originated from the same voucher specimen, loci were concatenated across accessions; 3) if all accessions for a given species originated from only one publication, loci were concatenated across accessions.
@@
 
-We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
+We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE v`r pkg_ver("iqtree2", pkg_versions)` [@Nguyen2015].
 ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
@@
 We inspected the initial Sanger phylogeny to identify any remaining rogues or names that needed updating in the taxonomic database.
-We checked for monophyly at taxonomic levels at or above genus using the R package "MonoPhy" [@Schwery2016].
+We checked for monophyly at taxonomic levels at or above genus using the MonoPhy R package v`r pkg_ver("MonoPhy", pkg_versions)` [@Schwery2016].
 We updated our taxonomic database if the molecular data clearly indicated the current usage of a synonym was incorrect (`r eg`, a taxonomic intruder into an otherwise expected monophyletic genus with a synonym available for that genus; expected monophyly follows PPG I, 2016).
@@
 As our alignment contains many species with very similar sequences, it is likely that the search algorithm is unable to optimize many highly similar topologies that only vary in positions of closely related species at the tips.
-Considering that additional iterations were unlikely to converge, we therefore conducted 10 independent runs of IQTREE with 1,000 iterations each, and selected the run with the best (highest) combined log-likelihood of the ML and consensus trees [@Zhou2018].
+Considering that additional iterations were unlikely to converge, we therefore conducted 10 independent runs of IQ-TREE with 1,000 iterations each, and selected the run with the best (highest) combined log-likelihood of the ML and consensus trees [@Zhou2018].
 
@@
 
-Dating was conducted separately on the ML tree and the consensus tree using penalized likelihood as implemented in treePL [@Smith2012].
+Dating was conducted separately on the ML tree and the consensus tree using penalized likelihood as implemented in the development version of treePL [commit starting with `r pkg_ver("treepl", pkg_versions)`\; @Smith2012].
 We initially rooted the tree using *Zygnema circumcarinatum* Czurda, a member of the group of algae (Zygnematophyceae) thought to be most closely related to land plants [@Donoghue2021].
@@
 
-We selected `r n_calibrations_ferns` fern fossils to use as constraints from a newly curated database of `r n_fern_fossils_total` fern fossil taxa called "ferncal" [@ftol2022b\; `r s_table("fossils")`, Appendix S2].
-The full ferncal database is available at https://fernphy.github.com/ferncal.
+We selected `r n_calibrations_ferns` fern fossils to use as constraints from ferncal v1.0.1, a newly curated database including `r n_fern_fossils_total` fern fossil taxa [@ftol2022b\; `r s_table("fossils")`, Appendix S2].
 In the case of redundant fossils (those assignable to the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil).
@@
 
-We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); these values are based on consideration that our tree spans a wide phylogenetic range with high rate heterogeneity, and results of initial analyses in which higher smoothing values sometimes produced spurious date estimates.
-A value of `r smooth_best` was selected, based on the smallest chi-squared value, for the final analysis.
+We tested rate smoothing parameters in treePL ranging from `r smooth_start` to `r smooth_end` (each varies by one order of magnitude); these values were used because our tree spans a wide phylogenetic range with high rate heterogeneity, and because results of initial analyses with higher smoothing values sometimes produced spurious date estimates.
+A value of `r smooth_best` was ultimately selected, based on the smallest chi-squared value, for the final analysis.
 
@@
 
-The workflow is managed in R v4.1.1 [@RCoreTeam2021] with the "targets" package [@Landau2021].
-Input data are available at https://doi.org/10.6084/m9.figshare.19474316.v1 [@Nitta2022figshare].
+The workflow is managed in R v`r ftol_r_ver` [@RCoreTeam2022] with the targets R package v`r pkg_ver("targets", pkg_versions)` [@Landau2021].
+Input data are available at https://doi.org/10.6084/m9.figshare.19474316 [@Nitta2022figshare].
 Code used to generate FTOL and compile this manuscript are available at https://github.com/fernphy/ftol and https://github.com/fernphy/ftol_ms, respectively.
@@
 
-Ferns are not unique in having large amounts of publicly available DNA sequence data, and several pipeline tools exist that can leverage such data to automatically generate phylogenies [@Antonelli2016; @Bennett2018; @Drori2018a; @Pearse2013; @Portik2020; @Smith2019a; @Xu2015].
-In particular, superSMART [@Antonelli2016] and pyPHLAWD [@Smith2019a] are recently developed pipelines that can generate maximally sampled phylogenies for any higher taxon of choice.
+Ferns are not unique in having large amounts of publicly available DNA sequence data, and several pipeline tools exist that can leverage such data to automatically generate phylogenies [@Antonelli2016; @Bennett2018; @Drori2018a; @Pearse2013; @Portik2020; @Smith2019; @Xu2015].
+In particular, superSMART [@Antonelli2016] and pyPHLAWD [@Smith2019] are recently developed pipelines that can generate maximally sampled phylogenies for any higher taxon of choice.
 The latter was used to generate a broadly inclusive seed plant phylogeny [@Smith2018].
@@
 We have sought to make FTOL easily available to support research on the evolution and ecology of ferns.
-FTOL is available via the "ftolr" R package [@ftol2022c], and a web portal (https://fernphy.github.io).
-Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibrations) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2021.08.26.457744].
+FTOL is available via the ftolr R package [@ftol2022c], and a web portal (https://fernphy.github.io).
+Furthermore, we have made all the underlying data (`r eg`, DNA alignments, fossil calibrations) available so that other researchers can use these to conduct analyses such as further investigations of divergence times or phylogenetic analysis including custom sets of DNA sequences [`r eg`, @Nitta2022a].
 
 A typical step in any analysis that joins data across multiple sources (`r eg`, trait data and a phylogeny) is to resolve taxonomic names so that the usage of synonyms does not prevent data merging [@Page2008].
-During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: the "pteridocat" package [@ftol2022a] and the "taxastand" package [@joel_nitta_2021_5726391].
+During the preparation of FTOL, we developed two additional R packages that enable taxonomic name resolution to join data with FTOL: pteridocat [@ftol2022a] and taxastand [@joel_nitta_2021_5726391].
 We selected R because it is widely used by the biological research community, well established, and freely available [@Lai2019].
-The "pteridocat" package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
-The "taxastand" package includes functions to resolve taxonomic names while taking account variation in taxonomic author format and orthographic variation.
+The pteridocat package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
+The taxastand package includes functions to resolve taxonomic names while taking account variation in taxonomic author format and orthographic variation.
 By using these two packages in combination, it should be straightforward for other researchers to map their own data onto FTOL, thus greatly enabling and enhancing studies including, but not limited to, comparative phylogenetics, biogeography, and community ecology in ferns.
@@
 S. Fawcett provided helpful comments on taxonomy of Thelypteridaceae.
-This study was supported by JSPS KAKENHI Grant Number 16H06279 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
+This study was supported by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
 The authors thank M. Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.

### 2022-06-24 R/functions.R

@@ -1205,6 +1205,15 @@ make_bs_tbl <- function(phy) {
 
 # Etc -----
 
+# Helper function to go from package name to version
+pkg_ver <- function(pkg, renv_pkg_versions) {
+  ver <- renv_pkg_versions %>%
+    filter(package == pkg) %>%
+    pull(version)
+  assertthat::assert_that(assertthat::is.string(ver))
+  ver
+}
+
 # Work-around for tar_load()
 # for some reason, doesn't work with tar_load(store = ftol_cache)
 read_from_cache <- function(object, cache) {

### 2022-06-24 ms/si.Rmd
Fix figure order
Update SI

@@
 **`r s_table("genera_diff")`**.
-Differences in genera included as accepted in "pteridocat" taxonomic database v0.0.1 (https://github.com/fernphy/pteridocat) vs.
+Differences in genera included as accepted in "pteridocat" taxonomic database v1.0.0 (https://github.com/fernphy/pteridocat) vs.
 PPGI [@PteridophytePhylogenyGroupI2016] (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
@@
 Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv") %>% fs::path_file()`).
-"n_fos" is the fossil ID number used in the "ferncal" database v1.0.0 (https://github.com/fernphy/ferncal).
+"n_fos" is the fossil ID number used in the "ferncal" database v1.0.1 (https://github.com/fernphy/ferncal).
 "minimum_age" is the age used for calibration (Ma); "lower" is the lower limit of the oldest stratigraphic age for the fossil (Ma); "upper" is the upper limit of the oldest stratigraphic age for the fossil (Ma).
@@
 "sci_name" is the scientific name in the pteridocat taxonomic database.
-"ncbi_name" is the scientific name used in the NCBI taxonomic database version 2022-02-01.
+"ncbi_name" is the scientific name used in the NCBI taxonomic database version 2022-05-01.
 "ncbi_taxid" is the NCBI taxonomy database unique identifier.
-"outgroup_status" is "in" for ingroup taxa (ferns) and "out" for outgroup taxa.
+"outgroup_status" is "FALSE" for ingroup taxa (ferns) and "TRUE" for outgroup taxa.
 
@@
 
+```{r backbone-full-plot-make}
+# Don't show Zygnema in plot
+phy <- ape::drop.tip(plastome_tree_rooted, "Zygnema_circumcarinatum")
+
+# Make tibble for printing BS values
+bs_tibble <- make_bs_tbl(phy) %>%
+  # only print values < 100
+  mutate(bs = case_when(
+    bs == 100 ~ "",
+    TRUE ~ as.character(bs)
+  ))
+
+# Make plot
+bb_full_plot <-
+  ggtree(phy, size = 0.2, color = "grey10") %<+% 
+  bs_tibble +
+  # now we can show BS values using geom_text()
+  geom_text(
+    aes(label = bs), hjust = -.25, size = 1.0,
+    color = "grey50") +
+  # add tip labels
+  geom_tiplab(aes(label = label), size = 1.0, offset = -0.005)
+```
+
+```{r backbone-full-plot-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = bb_full_plot,
+  file = (result_file(s_figure("backbone_full"), "pdf")),
+  height = 20, width = 10, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("backbone_full")`**.
+Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
+Values at nodes are bootstrap support (%); only values less than 100% are shown.
+Outgroup taxon *Zygnema circumcarinatum* (algae) not shown.
+
+\newpage
+
 ```{r bs-plot-make}
@@
 The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
-Outgroup taxon *Zygnema circumcarinatum* (algae) not shown.
-
-\newpage
-
-```{r backbone-full-plot-make}
-# Don't show Zygnema in plot
-phy <- ape::drop.tip(plastome_tree_rooted, "Zygnema_circumcarinatum")
-
-# Make tibble for printing BS values
-bs_tibble <- make_bs_tbl(phy) %>%
-  # only print values < 100
-  mutate(bs = case_when(
-    bs == 100 ~ "",
-    TRUE ~ as.character(bs)
-  ))
-
-# Make plot
-bb_full_plot <-
-  ggtree(phy, size = 0.2, color = "grey10") %<+% 
-  bs_tibble +
-  # now we can show BS values using geom_text()
-  geom_text(
-    aes(label = bs), hjust = -.25, size = 1.0,
-    color = "grey50") +
-  # add tip labels
-  geom_tiplab(aes(label = label), size = 1.0, offset = -0.005)
-```
-
-```{r backbone-full-plot-save, eval = params$doc_type == "doc"}
-ggsave(
-  plot = bb_full_plot,
-  file = (result_file(s_figure("backbone_full"), "pdf")),
-  height = 20, width = 10, units = "in",
-  limitsize = FALSE)
-```
-
-**`r s_figure("backbone_full")`**.
-Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
-Values at nodes are bootstrap support (%); only values less than 100% are shown.
 Outgroup taxon *Zygnema circumcarinatum* (algae) not shown.
@@
 Wide bars indicate the node was only constrained in one study; narrow bars indicate the the node was constrained in both studies.
+
+\newpage
+
+```{r ts-fossil-map-plot-make}
+# Make tibble with node number and label for ggtree
+ts_node_labs_tbl <-
+  ts_fossil_calibrations_for_treepl %>%
+  transmute(
+    fossil_label = str_remove_all(min, "min = "),
+    node = map2_dbl(
+      taxon_1, taxon_2,
+      ~ape::getMRCA(sanger_con_tree_rooted_pruned, c(.x, .y))),
+    calibration_type = case_when(
+      !is.na(min) & is.na(max) ~ "minimum",
+      !is.na(min) & !is.na(max) ~ "fixed",
+      TRUE ~ NA_character_
+    )
+  ) %>%
+  assert(not_na, calibration_type)
+
+# Make plot
+ts_fossil_map_plot <-
+  ggtree(
+    sanger_con_tree_rooted_pruned,
+    branch.length = "none",
+    size = 0.1, color = "grey30") %<+% ts_node_labs_tbl +
+  # Add node points
+  geom_nodepoint(
+    aes(color = calibration_type), size = 0.6) +
+  scale_color_manual(
+    "Calibration type",
+    values = c(
+      minimum = palette_okabe_ito("vermillion"),
+      fixed = palette_okabe_ito("skyblue")),
+    na.value = "transparent"
+  ) +
+  # Add node labels
+  geom_text(
+    aes(label = fossil_label),
+    hjust = 0, size = 1.0, color = "grey20") +
+  # Add tip labels
+  geom_tiplab(aes(label = label),
+  size = 0.5, color = "grey30", offset = -0.25) +
+  xlim(0, 66.1) +
+  theme(legend.position = "bottom")
+```
+
+```{r ts-fossil-map-plot-save, eval = params$doc_type == "doc"}
+ggsave(
+  plot = ts_fossil_map_plot,
+  file = (result_file(s_figure("ts_fossil_map"), "pdf")),
+  height = 70, width = 8, units = "in",
+  limitsize = FALSE)
+```
+
+**`r s_figure("ts_fossil_map")`**.
+Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints of @Testo2016a.
+Number for each constraint is age in Ma.
+Outgroup taxon *Zygnema circumcarinatum* (algae) pruned prior to dating analysis and not shown.
 
@@
 
-```{r ts-fossil-map-plot-make}
-# Make tibble with node number and label for ggtree
-ts_node_labs_tbl <-
-  ts_fossil_calibrations_for_treepl %>%
-  transmute(
-    fossil_label = str_remove_all(min, "min = "),
-    node = map2_dbl(
-      taxon_1, taxon_2,
-      ~ape::getMRCA(sanger_con_tree_rooted_pruned, c(.x, .y))),
-    calibration_type = case_when(
-      !is.na(min) & is.na(max) ~ "minimum",
-      !is.na(min) & !is.na(max) ~ "fixed",
-      TRUE ~ NA_character_
-    )
-  ) %>%
-  assert(not_na, calibration_type)
-
-# Make plot
-ts_fossil_map_plot <-
-  ggtree(
-    sanger_con_tree_rooted_pruned,
-    branch.length = "none",
-    size = 0.1, color = "grey30") %<+% ts_node_labs_tbl +
-  # Add node points
-  geom_nodepoint(
-    aes(color = calibration_type), size = 0.6) +
-  scale_color_manual(
-    "Calibration type",
-    values = c(
-      minimum = palette_okabe_ito("vermillion"),
-      fixed = palette_okabe_ito("skyblue")),
-    na.value = "transparent"
-  ) +
-  # Add node labels
-  geom_text(
-    aes(label = fossil_label),
-    hjust = 0, size = 1.0, color = "grey20") +
-  # Add tip labels
-  geom_tiplab(aes(label = label),
-  size = 0.5, color = "grey30", offset = -0.25) +
-  xlim(0, 66.1) +
-  theme(legend.position = "bottom")
-```
-
-```{r ts-fossil-map-plot-save, eval = params$doc_type == "doc"}
-ggsave(
-  plot = ts_fossil_map_plot,
-  file = (result_file(s_figure("ts_fossil_map"), "pdf")),
-  height = 70, width = 8, units = "in",
-  limitsize = FALSE)
-```
-
-**`r s_figure("ts_fossil_map")`**.
-Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints of @Testo2016a.
-Number for each constraint is age in Ma.
-Outgroup taxon *Zygnema circumcarinatum* (algae) pruned prior to dating analysis and not shown.
-
-\newpage
-
 ```{r comp-stem-points-plot, fig.width = 6, fig.height = 5, eval = params$doc_type != "manual"}

### 2022-07-21 ms/si2.Rmd
Update MS
Update SI2

@@
 ---
-title: "*Supplementary Material 2*"
+title: "*Appendix A2*"
 subtitle: "Fossil Calibration List"
@@
   - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
-  - \fancyhead[RE,RO]{Supplementary Material 2} # add header on right side
+  - \fancyhead[RE,RO]{Appendix A2} # add header on right side
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
@@
 bibliography: 
-- `r here::here("ms/si_fossils.bib")`
+- `r here::here("ms/ferncal_references.yaml")`
 ---
@@
 tar_load(
-  c(
-    fossil_notes,
-    fossil_meta),
+  c(fossil_meta,
+    ref_files),
   store = here::here("_targets")
@@
 
-Taxonomy of extant taxa follows pteridocat v0.0.1 database (https://github.com/fernphy/pteridocat).
+Taxonomy of extant taxa follows pteridocat v1.0.0 database (https://github.com/fernphy/pteridocat).
 

### 2022-07-21 ms/si.Rmd
Update SI1

@@
 ---
-title: "*Supplementary Material 1*"
+title: "*Appendix A1*"
 subtitle: "Tables and Figures"
@@
   - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
-  - \fancyhead[RE,RO]{Supplementary Material 1} # add header on right side
+  - \fancyhead[RE,RO]{Appendix A1} # add header on right side
   - \fancyhead[LO,LE]{} # suppress automatic section header on left side
@@
 "n_species" is the number of accepted species in each genus (pteridocat) or estimated number of species (PPGI).
-References in "notes" column provided at end of Supplementary Materials.
+References in "note" column provided at end of this appendix.
 "World Ferns" refers to @Hassler2022, which was modified to create pteridocat.
@@
 "minimum_age" is the age used for calibration (Ma); "lower" is the lower limit of the oldest stratigraphic age for the fossil (Ma); "upper" is the upper limit of the oldest stratigraphic age for the fossil (Ma).
-For detailed notes and full references for each fossil, see Appendix S2.
+For detailed notes and full references for each fossil, see Appendix A2.
 
@@
 **`r s_table("rogues")`**.
-GenBank accessions identified as rogues by all-by-all BLAST.
+GenBank accessions identified as rogues by all-by-all BLAST (`r result_file(s_table("rogues"), "csv") %>% fs::path_file()`).
 "q_family" is the family of the GenBank query (accession).
@@
 Methods used to join accessions across loci for Sanger dataset.
+"manual" only used for Thelypteridaceae (see Methods).
 Count is number of species.
@@
 **`r s_figure("du-ages")`**.
-Stem age of fern families (Ma) estimated by selected studies, only including families that were sampled across all studies.
+Stem age of fern families (Ma) estimated by selected studies, only including families with stem ages that were provided across all studies.
 For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
@@
 **`r s_figure("ts_fossil_map")`**.
-Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints of @Testo2016a.
+Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints of @Testo2016a (`r result_file(s_figure("ts_fossil_map"), "pdf") %>% fs::path_file()`).
 Number for each constraint is age in Ma.
@@
 **`r s_figure("ltt")`**.
-Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramirez-Barahona2020].
-Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020; for details, see that paper.
+Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramírez-Barahona2020].
+Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramírez-Barahona2020; for details, see that paper.
 Log number of lineages shown as percent of total to enable comparison across trees.

### 2022-07-21 R/process_refs.R
Extract citations from SI

@@ -30,3 +30,15 @@ extract_citations("ms/manuscript.Rmd") %>%
   mutate(latex = glue::glue("\\citation{[key]}", .open = "[", .close = "]")) %>%
   pull(latex) %>%
   write_lines("working/ftol.aux")
+
+extract_citations(
+   c(
+     "ms/si.rmd", # SI
+     "_targets/user/data_raw/pteridocat_ppgi_diff_notes.csv" # refs in SI tables
+   )
+  ) %>%
+  filter(str_detect(key, "figureHmakeatother", negate = TRUE)) %>%
+  filter(key != "") %>%
+  mutate(latex = glue::glue("\\citation{[key]}", .open = "[", .close = "]")) %>%
+  pull(latex) %>%
+  write_lines("temp/ftol_si1.aux")

### 2022-07-22 ms/manuscript.Rmd
Update MS

@@
 ---
-title: "An Open and Continuously Updated Fern Tree of Life (FTOL)"
+title: "An Open and Continuously Updated Fern Tree of Life"
 mainfont: Roboto-Regular.ttf
@@
 
-Keywords: Fern~1~, Phylogeney~2~, Plastome~3~, PPGI~4~, Pteridophyte~5~, *rbcL*~6~.
+Keywords: fern~1~, phylogeney~2~, plastome~3~, PPGI~4~, pteridophyte~5~, *rbcL*~6~.
 
@@
 
-Ferns, with about `r number(12000)` species, are the second most diverse clade of vascular plants after angiosperms.
+Ferns, with about `r number(12000)` species, are the second most diverse lineage of vascular plants after angiosperms.
 They have been the subject of numerous molecular phylogenetic studies, resulting in the publication of trees for every major clade and DNA sequences from nearly half of all species.
@@
 FTOL and its accompanying datasets, including the fossil list and taxonomic database, will be updated on a regular basis and are available via a web portal (https://fernphy.github.io) and R packages, enabling immediate access to the most up-to-date, comprehensively sampled fern phylogeny.
-FTOL will be useful for anybody studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
+FTOL will be useful for anyone studying this important group of plants over a wide range of taxonomic scales, from smaller clades to the entire tree.
 We anticipate FTOL will be particularly relevant for macroecological studies at regional to global scales and will inform future taxonomic systems with the most recent hypothesis of fern phylogeny.
@@
 
-Ferns (`r ca` `r number(12000)` species) are the second most diverse clade of vascular plants after angiosperms (`r ca` `r number(300000)` species) and are a useful study system for understanding processes of biogeography [@Tryon1986; @Kato1993], community ecology [`r eg`, @Hennequin2014; @Lehtonen2015], and speciation [`r eg`, @Kao2020].
+Ferns (`r ca` `r number(12000)` species) are the second most diverse lineage of vascular plants after angiosperms (`r ca` `r number(300000)` species) and are a useful study system for understanding processes of biogeography [`r eg`, @Tryon1986; @Kato1993], community ecology [`r eg`, @Hennequin2014; @Lehtonen2015], and speciation [`r eg`, @Kao2020].
 Key to any investigation of evolutionary history in this group is a well-sampled phylogeny.
@@
 Past efforts to construct such a global phylogeny have steadily expanded their sampling, at first by mostly generating new sequences, then later by mining GenBank [@Pryer2004a; @Lehtonen2011; @Schuettpelz2007; @Testo2016a].
-Indeed, growth of plastid fern accessions in GenBank show no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
+Indeed, the growth of plastid fern accessions in GenBank shows no sign of slowing since the most recent global fern phylogeny [@Testo2016a\; `r figure("genbank")`].
 There is a need therefore, not only for a revised global fern phylogeny, but also one that is continuously updated to keep pace with the rapid accumulation of molecular data going forward.
-Such an effort would eliminate the need for researchers to "rebuild the wheel" each time the need for a globally sampled fern phylogeny arises.
+Such an effort would eliminate the need for researchers to "reinvent the wheel" each time the need for a globally sampled fern phylogeny arises.
 
@@
 This will enable anybody interested in the biology of ferns to have access to the most current hypothesis of fern phylogeny, and an associated time-tree dated using a curated list of fern fossils.
-We anticipate FTOL will have several impacts for the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern relationships; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
+We anticipate FTOL will have multiple impacts on the field of fern systematics and evolution: 1) it will always provide the most up-to-date snapshot of our collective understanding of fern relationships; 2) it will allow for continuous assessment of taxonomy, and indicate those parts of the tree that are in need of taxonomic revision; and 3) it will be an important source of data for phylogenetic comparative and macroevolutionary studies of ferns.
 
@@
 Here, the *trnL*--*trnF* locus includes the *trnL* intron, the `r "3'"` *trnL* exon, and the *trnL*--*trnF* intergenic spacer; the *rps4*--*trnS* locus includes only the *rps4*--*trnS* intergenic spacer.
-The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*), and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
+The set of plastome loci is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes, including *atpA*, *atpB*, *matK*, *rbcL*, and *rps4*) and then combined with the *trnL*--*trnF* and *rps4*--*trnS* loci (79 loci total).
 Here, "locus" refers to an individual gene, intergenic spacer, or a unit comprised of these in the case of *trnL*--*trnF*.
@@
 A local database was then created from these data including only ferns and outgroup taxa (select seed plants, lycophytes, bryophytes, and algae) using the restez R package v`r pkg_ver("restez", pkg_versions)` [@Bennett2018a].
-GenBank accession numbers corresponding to sequences potentially matching each target locus were obtained by querying the GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cutoff matching the current GenBank release.
+GenBank accession numbers corresponding to sequences potentially matching each target locus were obtained by querying GenBank with terms including the locus name (`r eg`, "rbcL"), "Polypodiopsida[ORGN]", and a date cutoff matching the current GenBank release.
 Environmental DNA samples and accessions with names including the terms "aff." or "cf." or hybrid formulas were excluded (no attempt was made to exclude hybrid taxa with standard names, `r ie`, names that are not formulas).
@@
 To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
-The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from sequences downloaded from GenBank.
+The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from accessions downloaded from GenBank.
 
@@
 We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and maximal sampling of one locus has been shown to improve results in super-matrix phylogenies [@Talavera2021].
-Our procedure for selecting accessions for each species is as follows: 1) if accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length; 2) otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence; 3) otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
+Our procedure for selecting accessions for each species was as follows: 1) if accessions are concatenated including *rbcL* and at least one other locus, select the set of concatenated accessions with the greatest total sequence length; 2) otherwise, if accessions include *rbcL* only, select the accession with the longest *rbcL* sequence; 3) otherwise, select the accession or set of concatenated accessions with the greatest total sequence length.
 These steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
@@
 
-We also implemented a manual approach for Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in a previous version of FTOL (v1.0.0) and likely has high rates of misidentification on GenBank.
+We also implemented a manual approach for Thelypteridaceae, which had the highest number of non-monophyletic genera (16) in a previous version of FTOL (v1.0.0) and likely has high rates of misidentification in GenBank.
 This approach used a curated list of accessions from a previous phylogenetic analysis of this family [@Patel2019].
@@
 For non-spacer regions (genes), we aligned each locus separately in MAFFT with automatic adjustment for sequence direction and other settings on default.
-Spacer regions are difficult to align across higher taxonomic levels within ferns (`r eg`, family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer scales (`r eg`, within genera or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
-Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "\-\-merge" option.
-This retains the indels within each subalignment while aligning across subalignments.
+Spacer regions are difficult to align across higher taxonomic levels within ferns (`r eg`, family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer scales (`r eg`, within genus or family) where slower-evolving genes like *rbcL* may not provide enough resolution.
+Therefore, we first aligned spacer regions for each family with MAFFT separately, then merged these subalignments using the MAFFT "\-\-merge" option.
+This option retains the indels within each subalignment while aligning across subalignments.
 Sequences from families with fewer than three species each could not be used for subalignments, so these were added as "singletons" during the MAFFT "\-\-merge".
-Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned across families, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
+Anemiaceae showed an extremely high number of indels compared to other families and could not be reliably aligned, so we excluded it from the spacer region data (*trnL*--*trnF* and *rps4*--*trnS*).
 We also excluded outgroups from the spacer region data as they cannot be reliably aligned to ferns.
@@
 We used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
-We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.
+We prioritized computational speed for this step, as we needed to repeat it multiple times as described in the next section.
 
@@
 
-Taxon sampling for FTOL v1.1.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in pteridocat v0.1.0).
+Taxon sampling for FTOL v1.1.0 includes `r total_sampling$species$coverage` species (`r total_sampling$species$coverage_p`), `r total_sampling$genus$coverage` genera (`r total_sampling$genus$coverage_p`), `r total_sampling$family$coverage` families, and `r total_sampling$order$coverage` orders of ferns (all coverage values are relative to the number of accepted species in pteridocat v`r pkg_ver("pteridocat", pkg_versions)`).
 Coverage varied by major clade from `r tax_coverage %>% first %>% extract2("coverage")` (`r tax_coverage %>% first %>% extract2("clade")`) to `r tax_coverage %>% last %>% extract2("coverage")` (`r tax_coverage %>% last %>% extract2("clade")`) (`r figure("ftol")`).
@@
 Some earlier plastid analyses based on Sanger data also recovered Equisetidae sister to Marattidae, albeit with generally low support [@Pryer2001; @Qiu2006; @Qiu2007].
-Aside from alternative resolutions of Equisetidae dependent on genomic compartment or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12i346*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
+Aside from alternative resolutions of Equisetidae dependent on genomic compartment or model parameters [@Wickett2014; @Kuo2018], structural support exists for both Equisetidae as sister to Ophioglossidae [`r ca` 550 bp intron in plastid *rps12*\; @Grewe2013] and Equisetidae as sister to all other ferns [`r ca` 70 bp intron in mitochondrial *rpl2*\; @Knie2015].
 The clear contradiction between plastid and nuclear phylogenomic data may indicate ancient hybridization or introgression, but robust support for any particular scenario is so far lacking.
@@
 However, our plastome sampling only includes `r plastome_gleich_genera_n %>% jntools::english2()` out of `r ppgi_gleich_genera_n` genera of Gleicheniales, and lacks Matoniaceae.
-While our study was in revision, 42 additional fern plastomes were published, including seven families (`r ie`, Anemiaceae, Culcitaceae, Dipteridaceae, Loxsomataceae, Matoniaceae, Metaxyaceae, and Thyrsopteridaceae) that are not included in our sampling since they are not in the most recent available GenBank release [release 249\; @Du2022].
+While our study was in revision, 42 additional fern plastomes were published, including seven families (`r ie`, Anemiaceae, Culcitaceae, Dipteridaceae, Loxsomataceae, Matoniaceae, Metaxyaceae, and Thyrsopteridaceae) that are not included in our sampling since they were not in the most recent available GenBank release at the time of our analysis [release 249\; @Du2022].
 The addition of Matoniaceae results in non-monophyly of Gleicheniales, with Matoniaceae sister to Dipteridaceae, and these together sister to the remaining leptosporangiates [@Du2022].
@@
 Within Polypodiales, the relationship between suborders Pteridineae, Dennstaedtiineae and the eupolypods (Polypodiineae and Aspleniineae) has been difficult to resolve (here designated "P", "D", and "e", respectively).
-Most previous plastid studies based on Sanger sequencing have recovered (P, (D, e)) [@Kuo2011; @Schuettpelz2006; @Schuettpelz2007; @Testo2016a; but see @Lehtonen2011].
-We recover (D, (P, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] and nuclear data [@Rothfels2015a; @Qi2018; @Shen2018], as well as a plastid supermatrix [@Lehtonen2011].
+Most previous plastid studies based on Sanger sequencing have recovered (D, (P, e)) [@Kuo2011; @Schuettpelz2006; @Schuettpelz2007; @Testo2016a; but see @Lehtonen2011].
+We recover (P, (D, e)) with moderate support (BS `r denn_poly_bs`); this topology agrees with other phylogenomic studies based on whole plastomes [@Lu2015a; @Lehtonen2019] and nuclear data [@Rothfels2015a; @Qi2018; @Shen2018], as well as a plastid supermatrix [@Lehtonen2011].
 The whole plastome study of @Du2021 recovered a novel topology comprising ((D, P), e) under some analysis settings but (P, (D, e)) under others; a similar study with expanded sampling generally supported ((D, P), e) [@Du2022].
@@
 The pteridocat package includes the pteridocat taxonomic database as a data frame (tibble) in Darwin Core format.
-The taxastand package includes functions to resolve taxonomic names while taking account variation in taxonomic author format and orthographic variation.
+The taxastand package includes functions to resolve taxonomic names while taking into account variation in taxonomic author format and orthographic variation.
 By using these two packages in combination, it should be straightforward for other researchers to map their own data onto FTOL, thus greatly enabling and enhancing studies including, but not limited to, comparative phylogenetics, biogeography, and community ecology in ferns.
@@
 
-We want to be clear that FTOL is no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
+We want to be clear that FTOL is in no way meant to be the "official" fern phylogeny; it is simply one reasonable hypothesis that has been designed to be maximally inclusive at the species level.
 FTOL cannot substitute for careful systematic studies at finer taxonomic scales that include sampling of multiple individuals per species and/or other sources (`r eg`, morphological, nuclear) of data, and such studies continue to be vital to our understanding of fern evolution.
@@
 Another example of the advantage of a continuously updated tree is the publication of 42 additional fern plastome samples while this paper was in revision [@Du2022].
-While we were unable to integrate these data into the current version of FTOL, they will be automatically added when the next version of GenBank data is released, and we expected the next version of FTOL will reflect this new knowledge.
+While we were unable to integrate these data into the current version of FTOL, they will be automatically added when the next version of GenBank data is released, and we expect the next version of FTOL will reflect this new knowledge.
 
@@
 
+# Funding {-}
+This study was supported by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
+
 # Acknowledgments {-}
 
+The authors thank M. Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.
 Members of the Iwasaki lab provided comments that improved the manuscript.
@@
 S. Fawcett provided helpful comments on taxonomy of Thelypteridaceae.
-This study was supported by Japan Society for the Promotion of Science (Kakenhi) Grant numbers 16H06279, 22H04925, and 22K15171 and the Smithsonian National Museum of Natural History Peter Buck Fellowship (JHN).
-The authors thank M. Hassler for maintaining the World Ferns taxonomic database and making it available to use for research.
 
@@
 @Schuettpelz2007 did not attempt exhaustive sampling but rather proportional sampling according to lineage size.
-The relatively small increase in number of species in 2021 may be due to accessions that are still embargoed at the time of writing.
+The relatively small increase in number of species in 2021 may be due to accessions that were still embargoed at the time of writing.
 Taxonomy of GenBank species follows NCBI [@Federhen2012].
@@
     clade_short = map2_chr(major_clade, code, ~str_remove(.x, .y)),
-    label = glue("<b>{code}</b>{clade_short}"))
+    label = glue("**{code}**{clade_short}"))
 
@@
 ggsave(
-  result_file(figure("ftol"), "tiff"),
+  result_file(figure("ftol"), "pdf"),
   ftol_plot, width = 7, height = 7, units = "in")
@@
 Fern Tree of Life (FTOL).
-Tree rooted on bryophytes.
+Tree rooted with algae.
 Inset plot shows coverage by major clade (order or suborder).

### 2022-08-03 ms/manuscript.Rmd
Update ms following 2nd round of proofs

@@
 Fortunately, ferns have received relatively intense focus from molecular systematists, which has resulted in the publication of trees for all major clades and DNA sequence data from nearly half of all currently recognized fern species.
-Thus, there is both a pressing need and sufficient sampling for a globally sampled fern phylogeny.
+Thus, there is both a pressing need and sufficient sampling for a global, species-level fern phylogeny.
 
@@
 
-Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS ≥ `r percent(.98)` in the backbone phylogeny unless otherwise indicated).
+Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS > `r percent(.98)` in the backbone phylogeny unless otherwise indicated).
 The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (Figures `r figure_num("backbone")`, `r s_figure_num("backbone_full")`).
@@
 
-Bootstrap support was generally moderate to high across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`; `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
+Bootstrap support was generally moderate to high across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`, `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
 Relationships within some genera were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).
@@
 
-## Accesibility and usage of FTOL
+## Accesibility and usage
 
@@
 
-## FTOL as a living, community-driven resource
+## A living, community-driven resource
 
